---
title: "cf_metaproteomics_thesis_chapter_II"
author: "Natalia Bastón-Paz"
date: "2026"
output: html_document
---
## Set up
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
require("knitr")
opts_knit$set(root.dir = "C:/RProjects/cf_metaproteomics_thesis_chapter_II")
```

## Clean environments
```{#r Clean environments}
rm(list = ls())
```

## General introduction

Cystic Fibrosis (CF) is a genetic disorder caused by mutations in the CFTR (Cystic Fibrosis Transmembrane Conductance Regulator) gene, leading to defective ion transport and thick mucus in organs such as the lungs and gut. People with CF (pwCF) frequently develop chronic lung infections and gastrointestinal complications.

In recent years, CFTR modulator therapies, particularly the triple combination elexacaftor–tezacaftor–ivacaftor (ETI), have significantly improved clinical outcomes. However, their impact on the gut and lung microbiota, especially at the functional (proteomic) level, remains unclear.

In this script, we analyse fecal metaproteomic data from pwCF before (T0) and after one year of ETI treatment (T12) to explore changes in microbial and human protein profiles.

*Main question:*

- *Has the functional profile of the gut and lung microbiota changed after one year of treatment with ETI?*


## Libraries
```{r Libraries, message=FALSE, warning=FALSE}
library(readr)
library(readxl)
library(tidyverse)
library(factoextra)
library(ggrepel)
library(gplots)
library(shape)
library(openxlsx)
library(ggpubr)
library(vegan)
library(ggplot2)
library(fBasics)
library(purrr)
library(stringr)
library(patchwork)
library(tidyr)
library(dplyr)
```

# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------
# INTESTINAL MICROBIOTA SAMPLES ANALYSIS
# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------

# ------------------------------------------------------------------------------
# PROTEINS
# ------------------------------------------------------------------------------

## Load the data and pre-processing

### Data
```{r Load in data, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
Proteins <- readr::read_tsv("C:/RProjects/cf_metaproteomics_thesis_chapter_II/combined_protein.tsv")
```

### Remove MaxLFQ column
```{r Remove MaxLFQ, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
indices <- grep("MaxLFQ Intensity", colnames(Proteins))
Proteins <- Proteins[, -indices]
```

### Remove rows where all intensities are 0
```{r Remove rows with intensity 0, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
intensity_cols <- grep("Intensity", colnames(Proteins))
intensity_sum <- rowSums(Proteins[, intensity_cols])
Proteins <- Proteins[intensity_sum != 0, ]
```

### Filter by number of peptides and unique spectral counts
```{r Filter number peptides and unique spectral counts, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
###### By number of peptides: use column "Combined Total Peptides"
indices_filtered <- Proteins$`Combined Total Peptides` < 2
Proteins <- subset(Proteins, !indices_filtered)

###### By number of unique spectra (unique peptides): use "Combined Unique Spectral Count"
indices_filtered <- Proteins$`Combined Unique Spectral Count` < 1
Proteins <- subset(Proteins, !indices_filtered)
```

### Save protein file (this is our master file) and copy it
```{r Save proteins file, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
write.xlsx(Proteins, "Proteins.xlsx") # SUPPLEMENTARY TABLE
file.copy("Proteins.xlsx", "Proteins_backup.xlsx", overwrite = TRUE)
```

### Remove non-relevant columns and change names if necessary
```{r Remove non-relevant colums, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
colnames(Proteins)
Proteins <- Proteins[ , -c(2, 4, 6, 7, 8)]
colnames(Proteins)
Proteins <- Proteins[, -(10:228)]

# Rename one sample
colnames(Proteins)[colnames(Proteins) == "24_177_7b Intensity"] <- "24_177_7 Intensity"
```

### Save new protein file
```{r Save new proteins file, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
write.xlsx(Proteins, "Proteins_final.xlsx") # Manipulate this
```

### Filter samples
```{r Filter samples, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
remove_samples <- c(
  "24_177_1 Intensity","24_177_2 Intensity","24_177_3 Intensity","24_177_4 Intensity","24_177_5 Intensity", "24_177_6 Intensity","24_177_9 Intensity","24_177_10 Intensity","24_177_11 Intensity","24_177_12 Intensity",
  "24_177_13 Intensity","24_177_17 Intensity","24_177_18 Intensity","24_177_19 Intensity","24_177_20 Intensity", "24_177_21 Intensity", "24_177_22 Intensity", "24_177_23 Intensity","24_177_25 Intensity","24_177_26 Intensity","24_177_29 Intensity","24_177_30 Intensity",
  "24_177_31 Intensity","24_177_32 Intensity","24_177_33 Intensity", "24_177_35 Intensity", "24_177_36 Intensity","24_177_37 Intensity","24_177_38 Intensity", "24_177_39 Intensity","24_177_40 Intensity","24_177_41 Intensity","24_177_42 Intensity","24_177_45 Intensity",
  "24_177_46 Intensity","24_177_47 Intensity","24_177_48 Intensity","24_177_49 Intensity","24_177_53 Intensity", "24_177_54 Intensity","24_177_55 Intensity","24_177_56 Intensity","24_177_57 Intensity", "24_177_58 Intensity", "24_177_59 Intensity", "24_177_61 Intensity", "24_177_62 Intensity","24_177_65 Intensity","24_177_66 Intensity","24_177_67 Intensity","24_177_68 Intensity",
  "24_177_69 Intensity","24_177_70 Intensity", "24_177_72 Intensity", "24_177_73 Intensity")
Proteins <- Proteins[, setdiff(colnames(Proteins), remove_samples)]
colnames(Proteins)

# Remove rows where all intensities are 0
intensity_cols <- grep("Intensity", colnames(Proteins))
intensity_sum <- rowSums(Proteins[, intensity_cols])
Proteins <- Proteins[intensity_sum != 0, ]
```

### Save refined protein file
```{r Save refined file, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
write.xlsx(Proteins, "Proteins_refined.xlsx")
```

# ------------------------------------------------------------------------------
# ----- Human proteins -----
# ------------------------------------------------------------------------------

### Selection of human proteins from filtered protein table ("Protein")
```{r Select human proteins, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
human <- grep("HUMAN", Proteins$Protein)
Proteins_human <- Proteins[human,]
```

### Save final human protein file
```{r Save final file of human proteins, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
write.xlsx(Proteins_human, "Human_final.xlsx")
```

##### Number of human proteins per sample
```{r Number of human proteins per sample, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
intensity_cols <- Proteins_human[, grepl("Intensity", colnames(Proteins_human))]

num_proteins <- colSums(intensity_cols > 0)
num_proteins
num_proteins <- as.data.frame(num_proteins)
write.xlsx(num_proteins, "num_proteins_human.xlsx") # Open the file and copy (transpose) the values into a table created manually called "Table_prots_and_pepts.xlsx"
```

#### Number of human proteins per group
```{r Number of human proteins per group, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
Sample_data <- read_excel("Samples_CF_paired.xlsx")

Time0 <- Sample_data[Sample_data$Group == "time0", ]
Time0 <- Time0$Name
Time12 <- Sample_data[Sample_data$Group == "time12", ]
Time12 <- Time12$Name
dim(Proteins_human[rowSums(Proteins_human[, Time0] != 0) > 0, ])
dim(Proteins_human[rowSums(Proteins_human[, Time12] != 0) > 0, ])
```

# ------------------------------------------------------------------------------
# ----- Microbial proteins -----
# ------------------------------------------------------------------------------

### Selection of microbial proteins
```{r Select mirobial proteins, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
Proteins_microb <- Proteins[-human,]
```

### Save final microbial protein file
```{r Save microbial proteins file, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
write.xlsx(Proteins_microb, "Micro_final.xlsx")
```

##### Number of microbial proteins per sample
```{r Number of microbial proteins per sample, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
intensity_cols <- Proteins_microb[, grepl("Intensity", colnames(Proteins_microb))]

num_proteins <- colSums(intensity_cols > 0)
num_proteins
num_proteins <- as.data.frame(num_proteins)
write.xlsx(num_proteins, "num_proteins_microb.xlsx") # Open the file and copy (transpose) the values into a table created manually called "Table_prots_and_pepts.xlsx"
```

#### Number of microbial proteins per group
```{r Number of microbial proteins per group, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# We already have the Time0 and Time12 mini tables, so we just need dim as for human proteins
dim(Proteins_microb[rowSums(Proteins_microb[, Time0] != 0) > 0, ])
dim(Proteins_microb[rowSums(Proteins_microb[, Time12] != 0) > 0, ])
```

# ------------------------------------------------------------------------------
# PEPTIDES
# ------------------------------------------------------------------------------
  
## Load the data and pre-processing

### Data
```{r Load in peptides data, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
Peptides <- readr::read_tsv("C:/RProjects/cf_metaproteomics_thesis_chapter_II/combined_peptide.tsv")
```


### Remove MaxLFQ column
```{r Remove column MaxLFQ, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
indices <- grep("MaxLFQ Intensity", colnames(Peptides))
Peptides <- Peptides[, -indices]
```

### Remove rows where all intensities are 0
```{r Remove rows with intensity 0, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
intensity_cols <- grep("Intensity", colnames(Peptides))
intensity_sum <- rowSums(Peptides[, intensity_cols])
Peptides <- Peptides[intensity_sum != 0, ]
```

### Remove rows with NA values
```{r Remove rows with NA values, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Usually they are the same rows across the table, but to be safe we select all intensity columns
intensity_columns <- grep("Intensity", names(Peptides), value = TRUE)
Peptides <- Peptides[complete.cases(Peptides[, intensity_columns]), ]
```


### Save final peptide file --> WARNING: saving as xlsx may crash R and it restarts
```{r Save peptides final file, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
write.xlsx(Peptides, "Peptides.xlsx")
```

### Remove non-relevant columns
```{r Remove non-relevant columns, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
colnames(Peptides)
# Remove Spectral Count and Match Type columns
indices <- grep("Spectral Count", colnames(Peptides))
Peptides <- Peptides[, -indices]
indices <- grep("Match Type", colnames(Peptides))
Peptides <- Peptides[, -indices]
```

### Save final peptide file
```{r Save peptides final file, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
write.csv(Peptides, "Peptides_final.csv")
```

### Filter samples
```{r Filter samples, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
Peptides <- read_csv("C:/RProjects/cf_metaproteomics_thesis_chapter_II/Peptides_final.csv")
# Rename sample
colnames(Peptides)[colnames(Peptides) == "24_177_7b Intensity"] <- "24_177_7 Intensity"

remove_samples <- c(
  "24_177_1 Intensity","24_177_2 Intensity","24_177_3 Intensity","24_177_4 Intensity","24_177_5 Intensity", "24_177_6 Intensity","24_177_9 Intensity","24_177_10 Intensity","24_177_11 Intensity","24_177_12 Intensity",
  "24_177_13 Intensity","24_177_17 Intensity","24_177_18 Intensity","24_177_19 Intensity","24_177_20 Intensity", "24_177_21 Intensity", "24_177_22 Intensity", "24_177_23 Intensity","24_177_25 Intensity","24_177_26 Intensity","24_177_29 Intensity","24_177_30 Intensity",
  "24_177_31 Intensity","24_177_32 Intensity","24_177_33 Intensity", "24_177_35 Intensity", "24_177_36 Intensity","24_177_37 Intensity","24_177_38 Intensity", "24_177_39 Intensity","24_177_40 Intensity","24_177_41 Intensity","24_177_42 Intensity","24_177_45 Intensity",
  "24_177_46 Intensity","24_177_47 Intensity","24_177_48 Intensity","24_177_49 Intensity","24_177_53 Intensity", "24_177_54 Intensity","24_177_55 Intensity","24_177_56 Intensity","24_177_57 Intensity", "24_177_58 Intensity", "24_177_59 Intensity", "24_177_61 Intensity", "24_177_62 Intensity","24_177_65 Intensity","24_177_66 Intensity","24_177_67 Intensity","24_177_68 Intensity",
  "24_177_69 Intensity","24_177_70 Intensity", "24_177_72 Intensity", "24_177_73 Intensity")
Peptides <- Peptides[, setdiff(colnames(Peptides), remove_samples)]

# Remove rows where all intensities are 0
intensity_cols <- grep("Intensity", colnames(Peptides))
intensity_sum <- rowSums(Peptides[, intensity_cols])
Peptides <- Peptides[intensity_sum != 0, ]
```

### Save final peptide file --> It still crashes when saving as xlsx due to one row with many characters --> save as csv
```{r Save peptides refined, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
write.csv(Peptides, "Peptides_refined.csv")
```

# ------------------------------------------------------------------------------
# ----- Human peptides -----
# ------------------------------------------------------------------------------

### Selection of human peptides
```{r Select human peptides, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
human <- grep("HUMAN", Peptides$Protein)
Peptides_human <- Peptides[human,]
```

### Save final human peptide file
```{r Save final human_peptides, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
write.xlsx(Peptides_human, "Human_pept_final.xlsx")
```

#### Number of human peptides per sample
```{r Number of human peptides per sample, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
intensity_cols <- Peptides_human[, grepl("Intensity", colnames(Peptides_human))]

num_peptides <- colSums(intensity_cols > 0)
num_peptides
num_peptides <- as.data.frame(num_peptides)
write.xlsx(num_peptides, "num_peptides_human.xlsx")
```

#### Total intensity of human peptides per sample
```{r Intensity of human peptides, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
intensity_cols <- Peptides_human[, grepl("Intensity", colnames(Peptides_human))]
sum_intensities <- colSums(intensity_cols, na.rm = TRUE)
print(sum_intensities)
sum_intensities <- as.data.frame(sum_intensities)
write.xlsx(sum_intensities, "intensity_peptides_human.xlsx")
```

#### Number of human peptides per group
```{r Number of human peptides per group, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
dim(Peptides_human[rowSums(Peptides_human[, Time0] != 0) > 0, ])
dim(Peptides_human[rowSums(Peptides_human[, Time12] != 0) > 0, ])
```

#### Total intensity of human peptides per group
```{r Intensity of human peptides per group, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
filtered_rows <- Peptides_human[rowSums(Peptides_human[, Time0] != 0) > 0, ]
sum(filtered_rows[, Time0]) # Sum of intensities of all peptides at time 0

filtered_rows <- Peptides_human[rowSums(Peptides_human[, Time12] != 0) > 0, ]
sum(filtered_rows[, Time12]) # Sum of intensities of all peptides at time 12
```

# ------------------------------------------------------------------------------
# ----- Microbial peptides -----
# ------------------------------------------------------------------------------

### Selection of microbial peptides
```{r Select microbial peptides, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
Peptides_microb <- Peptides[-human,]
```

### Save final microbial peptide file
```{r Save file micro_peptides, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
write.xlsx(Peptides_microb, "Micro_pept_final.xlsx")
```

#### Number of microbial peptides per sample
```{r Number of microbial peptides per sample, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
intensity_cols <- Peptides_microb[, grepl("Intensity", colnames(Peptides_microb))]

num_peptides <- colSums(intensity_cols > 0)
num_peptides
num_peptides <- as.data.frame(num_peptides)
write.xlsx(num_peptides, "num_peptides_microb.xlsx")
```

#### Total intensity of microbial peptides per sample
```{r Intensity of microbial peptides, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
intensity_cols <- Peptides_microb[, grepl("Intensity", colnames(Peptides_microb))]
sum_intensities <- colSums(intensity_cols, na.rm = TRUE)
print(sum_intensities)
sum_intensities <- as.data.frame(sum_intensities)
write.xlsx(sum_intensities, "intensity_peptides_microb.xlsx")
```

#### Number of microbial peptides per group
```{r Number of microbial peptides per group, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
dim(Peptides_microb[rowSums(Peptides_microb[, Time0] != 0) > 0, ])
dim(Peptides_microb[rowSums(Peptides_microb[, Time12] != 0) > 0, ])
```

#### Total intensity of microbial peptides per group
```{r Intensity of microbial peptides per group, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
filtered_rows <- Peptides_microb[rowSums(Peptides_microb[, Time0] != 0) > 0, ]
sum(filtered_rows[, Time0])

filtered_rows <- Peptides_microb[rowSums(Peptides_microb[, Time12] != 0) > 0, ]
sum(filtered_rows[, Time12])
```

# ------------------------------------------------------------------------------
# PLOTS
# ------------------------------------------------------------------------------

### Plot number of identifications (proteins/peptides)

#### Prepare table proteins and peptides
```{r Prepare table proteins and peptides, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Table is created outside R
Table_prots_and_pepts <- read_excel("Table_prots_and_pepts.xlsx")

# Reshape table
data_long <- Table_prots_and_pepts %>%
  pivot_longer(cols = -Name, names_to = "SampleID", values_to = "Count") %>%
  mutate(Count = as.numeric(Count))

data_long <- data_long %>%
  mutate(Group = case_when(
    SampleID %in% c("24_177_7 Intensity", "24_177_8 Intensity", "24_177_14 Intensity", "24_177_15 Intensity", "24_177_16 Intensity", "24_177_24 Intensity", "24_177_27 Intensity", "24_177_28 Intensity", "24_177_34 Intensity") ~ "Time0",
    SampleID %in% c("24_177_43 Intensity", "24_177_44 Intensity", "24_177_50 Intensity", "24_177_51 Intensity", "24_177_52 Intensity", "24_177_60 Intensity", "24_177_63 Intensity", "24_177_64 Intensity", "24_177_71 Intensity") ~ "Time12",
    TRUE ~ "Other"
  ))

# Order factor levels
order_groups <- c("Microbial peptides", "Human peptides", "Microbial proteins", "Human proteins")
order_time <- c("Time0", "Time12")
data_long$Name <- factor(data_long$Name, levels = order_groups)
data_long$Group <- factor(data_long$Group, levels = order_time)
```

#### Select the data of interest and apply statistics
```{r Select the data of interest, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Subset human and microbial
Data_all <- subset(data_long,(Name == "Human proteins" | Name == "Human peptides"| Name == "Microbial proteins" | Name == "Microbial peptides"))
Data_all$Name <- factor(Data_all$Name, levels = order_groups)

# Calculate all metrics
summary_data_full <- Data_all %>%
  group_by(Name, Group) %>%
  summarise(
    median_count = median(Count, na.rm = TRUE),
    Q1 = quantile(Count, 0.25, na.rm = TRUE),
    Q3 = quantile(Count, 0.75, na.rm = TRUE),
    mean_count = mean(Count, na.rm = TRUE),
    sd_count = sd(Count, na.rm = TRUE),
    .groups = 'drop'
  )
View(summary_data_full)
write.xlsx(summary_data_full, "summary_data_full.xlsx")

# Normality test
shapiro_results <- Data_all %>%
  group_by(Name, Group) %>%
  summarise(
    p_value = shapiro.test(Count)$p.value
    ) %>%
  ungroup()

shapiro_results <- shapiro_results %>%
  mutate(
  Normality = ifelse(p_value > 0.05, "Normal", "Not Normal")
  )

print(shapiro_results) 

# Determine if both Time0 and Time12 are normal for each Name
normality_summary <- shapiro_results %>%
  group_by(Name) %>%
  summarise(is_normal = all(Normality == "Normal"), .groups = "drop")

# Join normality info to the main data
Data_all_with_norm <- Data_all %>%
  left_join(normality_summary, by = "Name")

# Split by Name and is_normal (1 group per test)
split_data <- Data_all_with_norm %>%
  group_by(Name, is_normal) %>%
  group_split()

# Run t-test or Wilcoxon depending on normality
run_test <- function(df, is_normal) {
  df <- df %>% filter(!is.na(Count), !is.na(Group))
  df$Group <- factor(df$Group)

  if (nlevels(df$Group) != 2) {
    return(tibble(p_value = NA, test_type = NA))
  }

  x <- df$Count[df$Group == "Time0"]
  y <- df$Count[df$Group == "Time12"]

  p_val <- tryCatch({
    if (is_normal) {
      t.test(x, y, paired = TRUE)$p.value
    } else {
      wilcox.test(x, y, paired = TRUE)$p.value
    }
  }, error = function(e) NA)

  test_type <- ifelse(is_normal, "Paired t-test", "Paired Wilcoxon test")

  tibble(p_value = p_val, test_type = test_type)
}

test_results <- Data_all_with_norm %>%
  group_by(Name, is_normal) %>%
  group_split() %>%
  map_dfr(~{
    result <- run_test(.x, unique(.x$is_normal))
    result$Name <- unique(.x$Name)
    result$is_normal <- unique(.x$is_normal)
    result
  })

print(test_results)
write.xlsx(test_results, "Statistical_test_results.xlsx")  # Optional export


# Join with the normality results (shapiro_results)
summary_data_adaptable <- summary_data_full %>%
  left_join(shapiro_results, by = c("Name", "Group"))

# Create the adaptable columns
summary_data_final <- summary_data_adaptable %>%
  mutate(
    # Plot center (Median or Mean)
    center_plot = ifelse(Normality == "Normal", mean_count, median_count),
    
    # Lower error line (Mean - SD or Q1)
    error_min = ifelse(Normality == "Normal", mean_count - sd_count, Q1),
    
    # Upper error line (Mean + SD or Q3)
    error_max = ifelse(Normality == "Normal", mean_count + sd_count, Q3)
  )
View(summary_data_final)
```

#### OPTIONAL: Plot results
```{r Optional Plot results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Prepare test_results for plotting
## Set y.position just above the top of each bar group
y_positions <- summary_data_final %>%
  group_by(Name) %>%
  summarise(y.position = max(error_max) * 1.1)

## Merge with p-values
plot_pvalues <- test_results %>%
  left_join(y_positions, by = "Name") %>%
  mutate(
    group1 = "Time0",
    group2 = "Time12",
    p = round(p_value, 4)
  ) %>%
   dplyr::select(Name, group1, group2, p_value, y.position)

# Plot
p <- ggplot(summary_data_final, aes(x = Name, y = center_plot, fill = Group)) + 
  geom_bar(stat = "identity", 
           position = position_dodge(width = 0.8), 
           width = 0.7,
           color = "black", 
           size = 0.3) + 
  geom_errorbar(aes(ymin = error_min, ymax = error_max), 
                position = position_dodge(width = 0.8), 
                width = 0.2) + 
  scale_fill_manual(values = c("Time0" = "#F4E1A1", "Time12" = "#A1E1E1")) +
  labs(y = "Number of peptides/proteins",
       fill = "Group") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("Prot_pept_sum_feces.svg", p, width = 8, height = 6, units = "in", dpi = 300)
```


### Plot values of intensities (proteins/peptides)

#### Prepare table proteins and peptides
```{r Prepare table proteins and peptides, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
data_long <- Table_prots_and_pepts %>%
  pivot_longer(cols = -Name, names_to = "SampleID", values_to = "Count") %>%
  mutate(Count = as.numeric(Count))

data_long <- data_long %>%
  mutate(Group = case_when(
    SampleID %in% c("24_177_7 Intensity", "24_177_8 Intensity", "24_177_14 Intensity", "24_177_15 Intensity", "24_177_16 Intensity", "24_177_24 Intensity", "24_177_27 Intensity", "24_177_28 Intensity", "24_177_34 Intensity") ~ "Time0",
    SampleID %in% c("24_177_43 Intensity", "24_177_44 Intensity", "24_177_50 Intensity", "24_177_51 Intensity", "24_177_52 Intensity", "24_177_60 Intensity", "24_177_63 Intensity", "24_177_64 Intensity", "24_177_71 Intensity") ~ "Time12",
    TRUE ~ "Other"
  ))

# Order factor levels
order_groups <- c("Microbial peptides intensity", "Human peptides intensity")
order_time <- c("Time0", "Time12")
data_long$Name <- factor(data_long$Name, levels = order_groups)
data_long$Group <- factor(data_long$Group, levels = order_time)
```

#### Select the data of interest and apply statistics
```{r Select the data of interest, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Subset human and microbial
Data_all <- subset(data_long,(Name == "Human peptides intensity"| Name == "Microbial peptides intensity"))
Data_all$Name <- factor(Data_all$Name, levels = order_groups)

# Calculate all metrics
summary_data_full <- Data_all %>%
  group_by(Name, Group) %>%
  summarise(
    median_count = median(Count, na.rm = TRUE),
    Q1 = quantile(Count, 0.25, na.rm = TRUE),
    Q3 = quantile(Count, 0.75, na.rm = TRUE),
    mean_count = mean(Count, na.rm = TRUE),
    sd_count = sd(Count, na.rm = TRUE),
    .groups = 'drop'
  )
View(summary_data_full)
write.xlsx(summary_data_full, "summary_data_full_intensity.xlsx")

# Normality test
shapiro_results <- Data_all %>%
  group_by(Name, Group) %>%
  summarise(
    p_value = shapiro.test(Count)$p.value
    ) %>%
  ungroup()

shapiro_results <- shapiro_results %>%
  mutate(
  Normality = ifelse(p_value > 0.05, "Normal", "Not Normal")
  )

print(shapiro_results)

# Determine if both Time0 and Time12 are normal for each Name
normality_summary <- shapiro_results %>%
  group_by(Name) %>%
  summarise(is_normal = all(Normality == "Normal"), .groups = "drop")

# Join normality info to the main data
Data_all_with_norm <- Data_all %>%
  left_join(normality_summary, by = "Name")

# Split by Name and is_normal (1 group per test)
split_data <- Data_all_with_norm %>%
  group_by(Name, is_normal) %>%
  group_split()

# Run t-test or Wilcoxon depending on normality
run_test <- function(df, is_normal) {
  df <- df %>% filter(!is.na(Count), !is.na(Group))
  df$Group <- factor(df$Group)

  if (nlevels(df$Group) != 2) {
    return(tibble(p_value = NA, test_type = NA))
  }

  x <- df$Count[df$Group == "Time0"]
  y <- df$Count[df$Group == "Time12"]

  p_val <- tryCatch({
    if (is_normal) {
      t.test(x, y, paired = TRUE)$p.value
    } else {
      wilcox.test(x, y, paired = TRUE)$p.value
    }
  }, error = function(e) NA)

  test_type <- ifelse(is_normal, "Paired t-test", "Paired Wilcoxon test")

  tibble(p_value = p_val, test_type = test_type)
}

test_results <- Data_all_with_norm %>%
  group_by(Name, is_normal) %>%
  group_split() %>%
  map_dfr(~{
    result <- run_test(.x, unique(.x$is_normal))
    result$Name <- unique(.x$Name)
    result$is_normal <- unique(.x$is_normal)
    result
  })

print(test_results)
write.xlsx(test_results, "Statistical_test_results_intensity.xlsx")  # Optional export


# Join with the normality results (shapiro_results)
summary_data_adaptable <- summary_data_full %>%
  left_join(shapiro_results, by = c("Name", "Group"))

# Create the adaptable columns
summary_data_final <- summary_data_adaptable %>%
  mutate(
    # Plot center (Median or Mean)
    center_plot = ifelse(Normality == "Normal", mean_count, median_count),
    
    # Lower error line (Mean - SD or Q1)
    error_min = ifelse(Normality == "Normal", mean_count - sd_count, Q1),
    
    # Upper error line (Mean + SD or Q3)
    error_max = ifelse(Normality == "Normal", mean_count + sd_count, Q3)
  )
View(summary_data_final)
```

#### OPTIONAL: Plot results
```{r Optional Plot results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Prepare test_results for plotting
## Set y.position just above the top of each bar group
y_positions <- summary_data_final %>%
  group_by(Name) %>%
  summarise(y.position = max(error_max) * 1.1)

## Merge with p-values
plot_pvalues <- test_results %>%
  left_join(y_positions, by = "Name") %>%
  mutate(
    group1 = "Time0",
    group2 = "Time12",
    p = round(p_value, 4)
  ) %>%
  dplyr::select(Name, group1, group2, p_value, y.position)

# Plot
p2 <- ggplot(summary_data_final, aes(x = Name, y = center_plot, fill = Group)) + 
  geom_bar(stat = "identity", 
           position = position_dodge(width = 0.8), 
           width = 0.7,
           color = "black", 
           size = 0.3) + 
  geom_errorbar(aes(ymin = error_min, ymax = error_max), 
                position = position_dodge(width = 0.8), 
                width = 0.2) + 
  scale_fill_manual(values = c("Time0" = "#F4E1A1", "Time12" = "#A1E1E1")) +
  labs(y = "Peptides intensity",
       fill = "Group") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("Pept_intensity_sum_feces.svg", p2, width = 8, height = 6, units = "in", dpi = 300)
```


### Plot Number of identifications
```{r Number of identifications, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Load the CSV file exported from Table_prots_and_pepts.xlsx (created outside R)
data <- read.csv("C:/RProjects/cf_metaproteomics_thesis_chapter_II/identifications_summary_paired.csv", stringsAsFactors = FALSE)

# Add column for Peptides vs Proteins
data$Type <- ifelse(grepl("peptides", data$Category, ignore.case = TRUE), "Peptides", "Proteins")

# Define color palette
custom_colors <- c("Time 0" = "#F4E1A1", "Time 12" = "#A1E1E1")

# Plot
p_final <- ggplot(data, aes(x = Category, y = Identifications, fill = Time)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", linewidth = 0.25) +
  facet_wrap(~Type, scales = "free") +
  scale_fill_manual(values = custom_colors) +
  labs(
    x = NULL,
    y = "Number of identifications",
    fill = "Time"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

ggsave("Prot_pept_human_micro_feces.svg", p_final, width = 8, height = 6, units = "in", dpi = 300)
```


# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------
# HUMAN PROTEINS
# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------

## Load the data
```{r Load in data, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Load protein data
data <- read_excel("Human_final.xlsx")
colnames(data) <- gsub("[[:space:]]+", "_", colnames(data))

# Load sample data
Sample_data <- read_excel("Samples_CF_paired.xlsx")
Sample_data$Name <- gsub("[[:space:]]+", "_", Sample_data$Name)

# Create sample vectors according to group
time0 <- Sample_data$Name[Sample_data$Group == "time0"]
time12 <- Sample_data$Name[Sample_data$Group == "time12"]

colnames(data)
```

## Presence filter
### Presence filter 10 %
```{r Presence filter 10 perc, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Keep only proteins present in ≥10% of samples in time0 or time12
threshold_Group1 <- ceiling(length(time0) * 0.1) 
threshold_Group2 <- ceiling(length(time12) * 0.1)

colnames(data) <- gsub("[[:space:]]+", "_", colnames(data))
data10 <- data %>%
  filter(rowSums(select(., all_of(time0)) > 0) >= threshold_Group1 | 
           rowSums(select(., all_of(time12)) > 0) >= threshold_Group2)

write.xlsx(data10, "Human_refined_10.xlsx")

Table_prots_human10 <- data10
```

# ------------------------------------------------------------------------------
# ---------- STATISTICS ----------
# ------------------------------------------------------------------------------

### Normality in the 10 % subset
```{r Normality in the 10 percent subset, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Select numeric columns only
numeric_data_10 <- Table_prots_human10 %>% 
  dplyr::select(where(is.numeric))

# Function for the Shapiro test
shapiro_p <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  if (n < 3)    return(NA_real_)           
  if (n > 5000) x <- sample(x, 5000)       
  if (sd(x) == 0) return(NA_real_)         
  shapiro.test(x)$p.value
}

# Apply to all numeric columns
results_10 <- numeric_data_10 %>%
  summarise(across(everything(), shapiro_p))

# Convert to long format and create summary
normality_summary_10 <- tidyr::pivot_longer(results_10, 
                                           cols = everything(),
                                           names_to = "Column",
                                           values_to = "P_Value")

# 5) Non-normal columns (p < 0.05)
non_normal_columns_10 <- normality_summary_10 %>% 
  filter(!is.na(P_Value), P_Value < 0.05)

print(normality_summary_10)
print(non_normal_columns_10)
```

### Relative abundance matrix 10 %
```{r Relative abundance matrix 10 perc, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Normalize intensities by the total intensity of each sample (relative abundance)
colnames(data10)

data_relativ10 <- data10[, 10:27]
data_relativ10 <- sweep(data_relativ10, 2, colSums(data_relativ10, na.rm = TRUE), FUN = "/")

data_matrix10 <- as.matrix(data_relativ10)
```


### Wilcoxon test
#### Wilcoxon test 10 %
```{r Wilcoxon 10 perc, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
res10 <- apply(data_matrix10, 1, function(x) {
  p_val <- wilcox.test(x[time0], x[time12], paired = TRUE)$p.value
  return(p_val)
})
res_adj10 <- p.adjust(res10, method = "fdr")

# Create one column per group with the median abundance per protein
data_relativ10$time0_median <- apply(data_relativ10[, time0], 1, median)
data_relativ10$time12_median <- apply(data_relativ10[, time12], 1, median)

# Create the results table
results10 <- data.frame(
  Name = Table_prots_human10$Entry_Name,
  p_value = res10,
  p_adj =res_adj10,
  time0 = data_relativ10$time0_median,
  time12 = data_relativ10$time12_median)
View(results10)
write.csv(results10, file = "wilcoxon_human10.csv")

# Define a pseudocount to avoid log(0) or division by zero
pseudocount <- 1e-10 
results10$Logfc <- log2((results10[,"time12"] + pseudocount) / (results10[,"time0"] + pseudocount))

# Create a new column indicating in which group the protein intensity is higher
results10$Group <- ifelse(results10$time0 > results10$time12, "time0", "time12")

# Filter by adjusted p-value < 0.05
results_pvalue10 <- results10[results10$p_adj < 0.05,]

# Filter by LogFC < -0.5 or > 0.5
results_pvaluelog10 <- results_pvalue10[results_pvalue10$Logfc > 0.5 | results_pvalue10$Logfc < -0.5, ]

# Inspect the results and save the logFC-filtered table
View(results_pvalue10)
write.csv(results_pvalue10, file = "wilcoxon_human10_sig.csv")
View(results_pvaluelog10)
write.csv(results_pvaluelog10, file = "wilcoxon_human10_sig_log.csv") # This is the results table that will be imported into STRING-db for network analysis
```


# ------------------------------------------------------------------------------
### PCA
# ------------------------------------------------------------------------------
###------------- All ------------
```{r Matrix for PCA all human proteins, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
Table_prots_human <- data
colnames(Table_prots_human)

data_pca <- Table_prots_human[, -c(1:9,28)]

data_pca <- t(data_pca) # transpose
pca1 <- prcomp(data_pca, scale = T) # scale to normalize
summary(pca1)

pca_df <- as.data.frame(pca1$x)
pca_df$Name <- row.names(pca_df)
pca_df$sample_name <- pca_df$Name

Sample_data <- read_excel("Samples_CF_paired.xlsx")
Sample_data$Name <- gsub("[[:space:]]+", "_", Sample_data$Name)

Time0  <- Sample_data$Name[Sample_data$Group == "time0"]
Time12 <- Sample_data$Name[Sample_data$Group == "time12"]

pca_df$Group <- ifelse(pca_df$Name %in% Time0, "Time 0",
                  ifelse(pca_df$Name %in% Time12, "Time 12", NA))

pca_df$Group <- factor(pca_df$Group, levels = c("Time 0", "Time 12"))

# sanity check linking
table(pca_df$Group, useNA = "ifany")

# PERMANOVA (euclidian distances)
distances <- dist(pca_df[, c("PC1", "PC2")])
set.seed(150)
adonis_result <- adonis2(distances ~ Group, data = pca_df)
print(adonis_result)

# --- Save PCA summary (like summary(pca1)) ---
imp <- summary(pca1)$importance
summary_pca1 <- tibble(
  PC = colnames(imp),
  standard_deviation     = as.numeric(imp["Standard deviation", ]),
  proportion_of_variance = as.numeric(imp["Proportion of Variance", ]),
  cumulative_proportion  = as.numeric(imp["Cumulative Proportion", ])
)

write.csv(summary_pca1, file = "PCA_human_summary.csv")

# --- Save PERMANOVA (adonis2) table ---
adonis_results <- as.data.frame(adonis_result) %>%
  tibble::rownames_to_column("term")

write.csv(adonis_results, file = "PCA_human_adonis.csv")
```

#### Variance
```{r Plot PCA variance all human proteins, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Extract axis variance (%) from prcomp()
var_expl <- (pca1$sdev^2) / sum(pca1$sdev^2) * 100
pc1_perc <- round(var_expl[1], 2)
pc2_perc <- round(var_expl[2], 2)

# PERMANOVA stats
F_value  <- adonis_result$F[1]
R2_value <- adonis_result$R2[1]
p_value  <- adonis_result$`Pr(>F)`[1]

F_txt  <- signif(F_value, 3)
R2_txt <- signif(R2_value, 3)
p_txt  <- if (is.na(p_value)) "NA" else if (p_value < 1e-3) "< 0.001" else signif(p_value, 3)
cap_text <- paste0("F = ", F_txt, "   •   R\u00B2 = ", R2_txt, "   •   p = ", p_txt)

# Group labels
pca_df$Group_plot <- factor(pca_df$Group, levels = c("Time 0", "Time 12"))

# Labels
pca_df$label_short <- sub(".*24_177_([0-9]+).*", "\\1", pca_df$Name)

# Colors
col_ellipse <- c("Time 0" = "#F4E1A1", "Time 12" = "#A1E1E1")
col_point   <- c("Time 0" = "#FAF3DC", "Time 12" = "#D6F5F5")

PCA_human_all <- ggplot(pca_df, aes(x = PC1, y = PC2)) +
  stat_ellipse(
    aes(fill = Group_plot, color = Group_plot),
    geom = "polygon", level = 0.95,
    alpha = 0.35, linewidth = 0.6, show.legend = TRUE
  ) +
  geom_point(
    data  = subset(pca_df, Group_plot == "Time 0"),
    shape = 21, size = 3, stroke = 0.3,
    color = "black", fill = col_point["Time 0"]
  ) +
  geom_point(
    data  = subset(pca_df, Group_plot == "Time 12"),
    shape = 21, size = 3, stroke = 0.3,
    color = "black", fill = col_point["Time 12"]
  ) +
  ggrepel::geom_text_repel(
    aes(label = label_short),
    color = "black",
    size = 3,
    box.padding = 0.35,
    point.padding = 0.15,
    max.overlaps = Inf,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = col_ellipse, name = "Time", drop = FALSE) +
  scale_color_manual(values = col_ellipse, guide = "none", drop = FALSE) +
  guides(fill = guide_legend(override.aes = list(alpha = 0.35, linewidth = 0.6))) +
  labs(
    x = paste0("PC1 (", pc1_perc, "%)"),
    y = paste0("PC2 (", pc2_perc, "%)"),
    title = "PCA - All human proteins",
    caption = cap_text
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor      = element_blank(),
    legend.position       = "right",
    plot.title            = element_text(size = 14),
    plot.caption.position = "plot",
    plot.caption          = element_text(hjust = 0, size = 11)
  )

ggsave("PCA_prots_human_all_variance.svg", PCA_human_all, width = 8, height = 6, units = "in", dpi = 300)
```

###------------- 10% ------------
#### 10% human proteins

```{r Matrix for PCA 10 human proteins, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
Table_prots_human10 <- data10
colnames(Table_prots_human10)

data_pca10 <- Table_prots_human10[, -c(1:9,28)]

data_pca10 <- t(data_pca10) # transpose
pca1_10 <- prcomp(data_pca10, scale = T) # scale to normalize
summary(pca1_10)

pca10_df <- as.data.frame(pca1_10$x)
pca10_df$Name <- row.names(pca10_df)
pca10_df$sample_name <- pca10_df$Name

Sample_data <- read_excel("Samples_CF_paired.xlsx")
Sample_data$Name <- gsub("[[:space:]]+", "_", Sample_data$Name)

Time0  <- Sample_data$Name[Sample_data$Group == "time0"]
Time12 <- Sample_data$Name[Sample_data$Group == "time12"]

pca10_df$Group <- ifelse(pca10_df$Name %in% Time0, "Time 0",
                  ifelse(pca10_df$Name %in% Time12, "Time 12", NA))

pca10_df$Group <- factor(pca10_df$Group, levels = c("Time 0", "Time 12"))

# sanity check linking
table(pca10_df$Group, useNA = "ifany")

# PERMANOVA (euclidian distances)
distances10 <- dist(pca10_df[, c("PC1", "PC2")])
set.seed(150)
adonis_result10 <- adonis2(distances10 ~ Group, data = pca10_df)
print(adonis_result10)

# --- Save PCA summary (like summary(pca1_10)) ---
imp <- summary(pca1_10)$importance
summary_pca1_10 <- tibble(
  PC = colnames(imp),
  standard_deviation     = as.numeric(imp["Standard deviation", ]),
  proportion_of_variance = as.numeric(imp["Proportion of Variance", ]),
  cumulative_proportion  = as.numeric(imp["Cumulative Proportion", ])
)

write.csv(summary_pca1_10, file = "PCA_human10_summary.csv")

# --- Save PERMANOVA (adonis2) table ---
adonis_results10 <- as.data.frame(adonis_result10) %>%
  tibble::rownames_to_column("term")

write.csv(adonis_results10, file = "PCA_human10_adonis.csv")
```

#### Variance
```{r Plot PCA variance 10 human proteins, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Extract axis variance (%) from prcomp()
var_expl <- (pca1_10$sdev^2) / sum(pca1_10$sdev^2) * 100
pc1_perc <- round(var_expl[1], 2)
pc2_perc <- round(var_expl[2], 2)

# Extract PERMANOVA stats from adonis2()
F_value  <- adonis_result10$F[1]
R2_value <- adonis_result10$R2[1]
p_value  <- adonis_result10$`Pr(>F)`[1]

F_txt  <- signif(F_value, 3)
R2_txt <- signif(R2_value, 3)
p_txt  <- if (is.na(p_value)) "NA" else if (p_value < 1e-3) "< 0.001" else signif(p_value, 3)
cap_text <- paste0("F = ", F_txt, "   •   R\u00B2 = ", R2_txt, "   •   p = ", p_txt)

# Group labels
pca10_df$Group_plot <- factor(pca10_df$Group, levels = c("Time 0", "Time 12"))

# Labels
pca10_df$label_short <- sub(".*24_177_([0-9]+).*", "\\1", pca10_df$Name)

# Colors
col_ellipse <- c("Time 0" = "#F4E1A1", "Time 12" = "#A1E1E1")
col_point   <- c("Time 0" = "#FAF3DC", "Time 12" = "#D6F5F5")

PCA_human_10_var <- ggplot(pca10_df, aes(x = PC1, y = PC2)) +
  stat_ellipse(
    aes(fill = Group_plot, color = Group_plot),
    geom = "polygon", level = 0.95,
    alpha = 0.35, linewidth = 0.6, show.legend = TRUE
  ) +
  geom_point(
    data  = subset(pca10_df, Group_plot == "Time 0"),
    shape = 21, size = 3, stroke = 0.3,
    color = "black", fill = col_point["Time 0"]
  ) +
  geom_point(
    data  = subset(pca10_df, Group_plot == "Time 12"),
    shape = 21, size = 3, stroke = 0.3,
    color = "black", fill = col_point["Time 12"]
  ) +
  ggrepel::geom_text_repel(
    aes(label = label_short),
    color = "black",
    size = 3,
    box.padding = 0.35,
    point.padding = 0.15,
    max.overlaps = Inf,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = col_ellipse, name = "Time", drop = FALSE) +
  scale_color_manual(values = col_ellipse, guide = "none", drop = FALSE) +
  guides(fill = guide_legend(override.aes = list(alpha = 0.35, linewidth = 0.6))) +
  labs(
    x = paste0("PC1 (", pc1_perc, "%)"),
    y = paste0("PC2 (", pc2_perc, "%)"),
    title = "PCA - 10% human proteins",
    caption = cap_text
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor      = element_blank(),
    legend.position       = "right",
    plot.title            = element_text(size = 14),
    plot.caption.position = "plot",
    plot.caption          = element_text(hjust = 0, size = 11)
  )

ggsave("PCA_prots_human_10_variance.svg", PCA_human_10_var, width = 8, height = 6, units = "in", dpi = 300)
```


# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------
# MICROBIAL PROTEINS
# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------

## Load the data
```{r Load in data, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Load protein data
data <- read_excel("Micro_final.xlsx")
colnames(data) <- gsub("[[:space:]]+", "_", colnames(data))

# Load sample data
Sample_data <- read_excel("Samples_CF_paired.xlsx")
Sample_data$Name <- gsub("[[:space:]]+", "_", Sample_data$Name)

# Create sample vectors according to group
time0 <- Sample_data$Name[Sample_data$Group == "time0"]
time12 <- Sample_data$Name[Sample_data$Group == "time12"]

colnames(data)
Table_prots_micro <- data
```

## Presence filter
### Presence filter 10 %
```{r Presence filter 10 perc, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Keep only proteins present in ≥10% of samples in time0 or time12
threshold_Group1 <- ceiling(length(time0) * 0.1) 
threshold_Group2 <- ceiling(length(time12) * 0.1)

colnames(data) <- gsub("[[:space:]]+", "_", colnames(data))
data10 <- data %>%
  filter(rowSums(select(., all_of(time0)) > 0) >= threshold_Group1 | 
           rowSums(select(., all_of(time12)) > 0) >= threshold_Group2)

write.xlsx(data10, "Micro_refined_10.xlsx")

Table_prots_micro10 <- data10
```

# ------------------------------------------------------------------------------
# ---------- STATISTICS ----------
# ------------------------------------------------------------------------------

### Normality in the 10 % subset
```{r Normality in the 10 percent subset, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Select numeric columns only
numeric_data_10 <- Table_prots_micro10 %>% 
  dplyr::select(where(is.numeric))

# Function for the Shapiro test
shapiro_p <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  if (n < 3)    return(NA_real_)           
  if (n > 5000) x <- sample(x, 5000)       
  if (sd(x) == 0) return(NA_real_)         
  shapiro.test(x)$p.value
}

# Apply to all numeric columns
results_10 <- numeric_data_10 %>%
  summarise(across(everything(), shapiro_p))

# Convert to long format and create summary
normality_summary_10 <- tidyr::pivot_longer(results_10, 
                                           cols = everything(),
                                           names_to = "Column",
                                           values_to = "P_Value")

# 5) Non-normal columns (p < 0.05)
non_normal_columns_10 <- normality_summary_10 %>% 
  filter(!is.na(P_Value), P_Value < 0.05)

print(normality_summary_10)
print(non_normal_columns_10)
```

### Relative abundance matrix 10 %
```{r Relative abundance matrix 10 perc, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Normalize intensities by the total intensity of each sample (relative abundance)
colnames(data10)

data_relativ10 <- data10[, 10:27]
data_relativ10 <- sweep(data_relativ10, 2, colSums(data_relativ10, na.rm = TRUE), FUN = "/")

data_matrix10 <- as.matrix(data_relativ10)
```

### Wilcoxon test
#### Wilcoxon test 10 %
```{r Wilcoxon 10 perc, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
res10 <- apply(data_matrix10, 1, function(x) {
  p_val <- wilcox.test(x[time0], x[time12], paired = TRUE)$p.value
  return(p_val)
})
res_adj10 <- p.adjust(res10, method = "fdr")

# Create one column per group with the median abundance per protein
data_relativ10$time0_median <- apply(data_relativ10[, time0], 1, median)
data_relativ10$time12_median <- apply(data_relativ10[, time12], 1, median)

# Create the results table
results10 <- data.frame(
  Name = Table_prots_micro10$Entry_Name,
  p_value = res10,
  p_adj =res_adj10,
  time0 = data_relativ10$time0_median,
  time12 = data_relativ10$time12_median)
View(results10)
write.csv(results10, file = "wilcoxon_micro10.csv")

# Define a pseudocount to avoid log(0) or division by zero
pseudocount <- 1e-10 
results10$Logfc <- log2((results10[,"time12"] + pseudocount) / (results10[,"time0"] + pseudocount))

# Create a new column indicating in which group the protein intensity is higher
results10$Group <- ifelse(results10$time0 > results10$time12, "time0", "time12")

# Filter by adjusted p-value < 0.05
results_pvalue10 <- results10[results10$p_adj < 0.05,]

# Filter by LogFC < -0.5 or > 0.5
results_pvaluelog10 <- results_pvalue10[results_pvalue10$Logfc > 0.5 | results_pvalue10$Logfc < -0.5, ]

# Inspect the results and save the logFC-filtered table
View(results_pvalue10)
write.csv(results_pvalue10, file = "wilcoxon_micro10_sig.csv")
View(results_pvaluelog10)
write.csv(results_pvaluelog10, file = "wilcoxon_micro10_sig_log.csv")
```

# ------------------------------------------------------------------------------
### PCA
# ------------------------------------------------------------------------------
###------------- All ------------
#### All microbial proteins

```{r Matrix for PCA micro proteins, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
Table_prots_micro <- data
colnames(Table_prots_micro)

data_pca <- Table_prots_micro[, -c(1:9,28)]

data_pca <- t(data_pca) # transpose
pca1 <- prcomp(data_pca, scale = T) # scale to normalize
summary(pca1)

pca_df <- as.data.frame(pca1$x)
pca_df$Name <- row.names(pca_df)
pca_df$sample_name <- pca_df$Name

Sample_data <- read_excel("Samples_CF_paired.xlsx")
Sample_data$Name <- gsub("[[:space:]]+", "_", Sample_data$Name)

Time0  <- Sample_data$Name[Sample_data$Group == "time0"]
Time12 <- Sample_data$Name[Sample_data$Group == "time12"]

pca_df$Group <- ifelse(pca_df$Name %in% Time0, "Time 0",
                  ifelse(pca_df$Name %in% Time12, "Time 12", NA))

pca_df$Group <- factor(pca_df$Group, levels = c("Time 0", "Time 12"))

# sanity check linking
table(pca_df$Group, useNA = "ifany")

# PERMANOVA (euclidian distances)
distances <- dist(pca_df[, c("PC1", "PC2")])
set.seed(150)
adonis_result <- adonis2(distances ~ Group, data = pca_df)
print(adonis_result)

# --- Save PCA summary (like summary(pca1)) ---
imp <- summary(pca1)$importance
summary_pca1 <- tibble(
  PC = colnames(imp),
  standard_deviation     = as.numeric(imp["Standard deviation", ]),
  proportion_of_variance = as.numeric(imp["Proportion of Variance", ]),
  cumulative_proportion  = as.numeric(imp["Cumulative Proportion", ])
)

write.csv(summary_pca1, file = "PCA_micro_summary.csv")

# --- Save PERMANOVA (adonis2) table ---
adonis_results <- as.data.frame(adonis_result) %>%
  tibble::rownames_to_column("term")

write.csv(adonis_results, file = "PCA_micro_adonis.csv")
```

#### Variance
```{r Plot PCA variance all micro proteins, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Extract axis variance (%) from prcomp()
var_expl <- (pca1$sdev^2) / sum(pca1$sdev^2) * 100
pc1_perc <- round(var_expl[1], 2)
pc2_perc <- round(var_expl[2], 2)

# PERMANOVA stats
F_value  <- adonis_result$F[1]
R2_value <- adonis_result$R2[1]
p_value  <- adonis_result$`Pr(>F)`[1]

F_txt  <- signif(F_value, 3)
R2_txt <- signif(R2_value, 3)
p_txt  <- if (is.na(p_value)) "NA" else if (p_value < 1e-3) "< 0.001" else signif(p_value, 3)
cap_text <- paste0("F = ", F_txt, "   •   R\u00B2 = ", R2_txt, "   •   p = ", p_txt)

# Group labels
pca_df$Group_plot <- factor(pca_df$Group, levels = c("Time 0", "Time 12"))

# Labels
pca_df$label_short <- sub(".*24_177_([0-9]+).*", "\\1", pca_df$Name)

# Colors
col_ellipse <- c("Time 0" = "#F4E1A1", "Time 12" = "#A1E1E1")
col_point   <- c("Time 0" = "#FAF3DC", "Time 12" = "#D6F5F5")

PCA_micro_all <- ggplot(pca_df, aes(x = PC1, y = PC2)) +
  stat_ellipse(
    aes(fill = Group_plot, color = Group_plot),
    geom = "polygon", level = 0.95,
    alpha = 0.35, linewidth = 0.6, show.legend = TRUE
  ) +
  geom_point(
    data  = subset(pca_df, Group_plot == "Time 0"),
    shape = 21, size = 3, stroke = 0.3,
    color = "black", fill = col_point["Time 0"]
  ) +
  geom_point(
    data  = subset(pca_df, Group_plot == "Time 12"),
    shape = 21, size = 3, stroke = 0.3,
    color = "black", fill = col_point["Time 12"]
  ) +
  ggrepel::geom_text_repel(
    aes(label = label_short),
    color = "black",
    size = 3,
    box.padding = 0.35,
    point.padding = 0.15,
    max.overlaps = Inf,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = col_ellipse, name = "Time", drop = FALSE) +
  scale_color_manual(values = col_ellipse, guide = "none", drop = FALSE) +
  guides(fill = guide_legend(override.aes = list(alpha = 0.35, linewidth = 0.6))) +
  labs(
    x = paste0("PC1 (", pc1_perc, "%)"),
    y = paste0("PC2 (", pc2_perc, "%)"),
    title = "PCA - All microbial proteins",
    caption = cap_text
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor      = element_blank(),
    legend.position       = "right",
    plot.title            = element_text(size = 14),
    plot.caption.position = "plot",
    plot.caption          = element_text(hjust = 0, size = 11)
  )

ggsave("PCA_prots_micro_all_variance.svg", PCA_micro_all, width = 8, height = 6, units = "in", dpi = 300)
```

###------------- 10% ------------
#### 10% microbial proteins

```{r Matrix for PCA 10 micro proteins, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
Table_prots_micro10 <- data10
colnames(Table_prots_micro10)

data_pca10 <- Table_prots_micro10[, -c(1:9,28)]

data_pca10 <- t(data_pca10) # transpose
pca1_10 <- prcomp(data_pca10, scale = T) # scale to normalize
summary(pca1_10)

pca10_df <- as.data.frame(pca1_10$x)
pca10_df$Name <- row.names(pca10_df)
pca10_df$sample_name <- pca10_df$Name

Sample_data <- read_excel("Samples_CF_paired.xlsx")
Sample_data$Name <- gsub("[[:space:]]+", "_", Sample_data$Name)

Time0  <- Sample_data$Name[Sample_data$Group == "time0"]
Time12 <- Sample_data$Name[Sample_data$Group == "time12"]

pca10_df$Group <- ifelse(pca10_df$Name %in% Time0, "Time 0",
                  ifelse(pca10_df$Name %in% Time12, "Time 12", NA))

pca10_df$Group <- factor(pca10_df$Group, levels = c("Time 0", "Time 12"))

# sanity check linking
table(pca10_df$Group, useNA = "ifany")

# PERMANOVA (euclidian distances)
distances10 <- dist(pca10_df[, c("PC1", "PC2")])
set.seed(150)
adonis_result10 <- adonis2(distances10 ~ Group, data = pca10_df)
print(adonis_result10)

# --- Save PCA summary (like summary(pca1_10)) ---
imp <- summary(pca1_10)$importance
summary_pca1_10 <- tibble(
  PC = colnames(imp),
  standard_deviation     = as.numeric(imp["Standard deviation", ]),
  proportion_of_variance = as.numeric(imp["Proportion of Variance", ]),
  cumulative_proportion  = as.numeric(imp["Cumulative Proportion", ])
)

write.csv(summary_pca1_10, file = "PCA_micro10_summary.csv")

# --- Save PERMANOVA (adonis2) table ---
adonis_results10 <- as.data.frame(adonis_result10) %>%
  tibble::rownames_to_column("term")

write.csv(adonis_results10, file = "PCA_micro10_adonis.csv")
```

#### Variance
```{r Plot PCA variance 10 micro proteins, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Extract axis variance (%) from prcomp()
var_expl <- (pca1_10$sdev^2) / sum(pca1_10$sdev^2) * 100
pc1_perc <- round(var_expl[1], 2)
pc2_perc <- round(var_expl[2], 2)

# Extract PERMANOVA stats from adonis2()
F_value  <- adonis_result10$F[1]
R2_value <- adonis_result10$R2[1]
p_value  <- adonis_result10$`Pr(>F)`[1]

F_txt  <- signif(F_value, 3)
R2_txt <- signif(R2_value, 3)
p_txt  <- if (is.na(p_value)) "NA" else if (p_value < 1e-3) "< 0.001" else signif(p_value, 3)
cap_text <- paste0("F = ", F_txt, "   •   R\u00B2 = ", R2_txt, "   •   p = ", p_txt)

# Group labels
pca10_df$Group_plot <- factor(pca10_df$Group, levels = c("Time 0", "Time 12"))

# Labels
pca10_df$label_short <- sub(".*24_177_([0-9]+).*", "\\1", pca10_df$Name)

# Colors
col_ellipse <- c("Time 0" = "#F4E1A1", "Time 12" = "#A1E1E1")
col_point   <- c("Time 0" = "#FAF3DC", "Time 12" = "#D6F5F5")

PCA_micro_10_var <- ggplot(pca10_df, aes(x = PC1, y = PC2)) +
  stat_ellipse(
    aes(fill = Group_plot, color = Group_plot),
    geom = "polygon", level = 0.95,
    alpha = 0.35, linewidth = 0.6, show.legend = TRUE
  ) +
  geom_point(
    data  = subset(pca10_df, Group_plot == "Time 0"),
    shape = 21, size = 3, stroke = 0.3,
    color = "black", fill = col_point["Time 0"]
  ) +
  geom_point(
    data  = subset(pca10_df, Group_plot == "Time 12"),
    shape = 21, size = 3, stroke = 0.3,
    color = "black", fill = col_point["Time 12"]
  ) +
  ggrepel::geom_text_repel(
    aes(label = label_short),
    color = "black",
    size = 3,
    box.padding = 0.35,
    point.padding = 0.15,
    max.overlaps = Inf,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = col_ellipse, name = "Time", drop = FALSE) +
  scale_color_manual(values = col_ellipse, guide = "none", drop = FALSE) +
  guides(fill = guide_legend(override.aes = list(alpha = 0.35, linewidth = 0.6))) +
  labs(
    x = paste0("PC1 (", pc1_perc, "%)"),
    y = paste0("PC2 (", pc2_perc, "%)"),
    title = "PCA - 10% microbial proteins",
    caption = cap_text
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor      = element_blank(),
    legend.position       = "right",
    plot.title            = element_text(size = 14),
    plot.caption.position = "plot",
    plot.caption          = element_text(hjust = 0, size = 11)
  )

ggsave("PCA_prots_micro_10_variance.svg", PCA_micro_10_var, width = 8, height = 6, units = "in", dpi = 300)
```


# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------
# MICROBIAL FUNCTIONS
# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------

## Load the data and pre-processing

### Samples data
```{r Samples data and vectors, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Load functions data
Sample_data <- read_excel("Samples_CF_paired_intensity.xlsx")
Sample_data$Name <- gsub("[[:space:]]+", "_", Sample_data$Name)

# Create sample vectors according to group
time0 <- Sample_data$Name[Sample_data$Time == "time0"] # Carmen en vez de Time tiene Group (es el archivo Samples_FQ)
time12 <- Sample_data$Name[Sample_data$Time == "time12"] # Carmen en vez de Time tiene Group (es el archivo Samples_FQ)
```

### Functions data
```{r Load in data, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Load functions data
data <- read_delim("functions.tsv", 
                   delim = "\t", escape_double = FALSE, trim_ws = TRUE)
```

### Remove MaxLFQ column
```{r Remove MaxLFQ, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
indices <- grep("MaxLFQ", colnames(data))
data <- data[, -indices]
```

### Remove rows where all intensities are 0
```{r Remove rows with intensity 0, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
intensity_cols <- grep("Intensity", colnames(data))
intensity_sum <- rowSums(data[, intensity_cols])
data <- data[intensity_sum != 0, ]
```

### Load in Micro_final file and filter by Protein
```{r Load in Micro_final file, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
Proteins_filtered <- read_excel("C:/RProjects/cf_metaproteomics_thesis_chapter_II/Micro_final.xlsx")

# Filter by column Protein
filtered_data <- data %>%
  filter(Name %in% Proteins_filtered$Protein)
colnames(filtered_data) <- gsub("[[:space:]]+", "_", colnames(filtered_data))

# Rename sample
colnames(filtered_data)[colnames(filtered_data) == "Intensity_24_177_7b"] <- "Intensity_24_177_7"
```

### Convert colums to UTF-8 and save file
```{r Convert colums to UTF-8 and save file, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
filtered_data <- filtered_data %>%
  mutate(across(where(is.character), ~ iconv(., to = "UTF-8")))
write.xlsx(filtered_data, "Functions_filtered.xlsx")
```

### Remove samples and save new file
```{r Remove samples, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Rename remove_samples from "24_177_X Intensity" -> "Intensity_24_177_X"
remove_samples <- c(
  "Intensity_24_177_1","Intensity_24_177_2","Intensity_24_177_3","Intensity_24_177_4","Intensity_24_177_5",
  "Intensity_24_177_6","Intensity_24_177_9","Intensity_24_177_10","Intensity_24_177_11","Intensity_24_177_12",
  "Intensity_24_177_13","Intensity_24_177_17","Intensity_24_177_18","Intensity_24_177_19","Intensity_24_177_20",
  "Intensity_24_177_21","Intensity_24_177_22","Intensity_24_177_23","Intensity_24_177_25","Intensity_24_177_26",
  "Intensity_24_177_29","Intensity_24_177_30","Intensity_24_177_31","Intensity_24_177_32","Intensity_24_177_33",
  "Intensity_24_177_35","Intensity_24_177_36","Intensity_24_177_37","Intensity_24_177_38","Intensity_24_177_39",
  "Intensity_24_177_40","Intensity_24_177_41","Intensity_24_177_42","Intensity_24_177_45","Intensity_24_177_46",
  "Intensity_24_177_47","Intensity_24_177_48","Intensity_24_177_49","Intensity_24_177_53","Intensity_24_177_54",
  "Intensity_24_177_55","Intensity_24_177_56","Intensity_24_177_57","Intensity_24_177_58","Intensity_24_177_59",
  "Intensity_24_177_61","Intensity_24_177_62","Intensity_24_177_65","Intensity_24_177_66","Intensity_24_177_67",
  "Intensity_24_177_68","Intensity_24_177_69","Intensity_24_177_70","Intensity_24_177_72","Intensity_24_177_73")

filtered_data_clean <- filtered_data[, setdiff(colnames(filtered_data), remove_samples)]

filtered_data_clean <- filtered_data_clean %>%
  mutate(across(where(is.character), ~ iconv(., to = "UTF-8")))
write.xlsx(filtered_data_clean, "Functions_refined.xlsx")
```

### Presence filter
#### Presence filter 10 %
```{r Presence filter 10 perc, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Keep only proteins present in ≥10% of samples in time0 or time12
threshold_Group1 <- ceiling(length(time0) * 0.1)
threshold_Group2 <- ceiling(length(time12) * 0.1)

filtered_data10 <- filtered_data_clean %>%
  filter(rowSums(select(., all_of(time0)) > 0) >= threshold_Group1 | 
           rowSums(select(., all_of(time12)) > 0) >= threshold_Group2)

write.xlsx(filtered_data10, "Functions_refined_10.xlsx")
```

# ------------------------------------------------------------------------------
# ---------- STATISTICS ----------
# ------------------------------------------------------------------------------

### Normality in the 10 % subset
```{r Normality in the 10 percent subset, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Select numeric columns only
colnames(filtered_data10)
numeric_data_10 <- filtered_data10 [, c(7:24)]

# Function for the Shapiro test
shapiro_p <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  if (n < 3)    return(NA_real_)           
  if (n > 5000) x <- sample(x, 5000)       
  if (sd(x) == 0) return(NA_real_)         
  shapiro.test(x)$p.value
}

# Apply to all numeric columns
results_10 <- numeric_data_10 %>%
  summarise(across(everything(), shapiro_p))

# Convert to long format and create summary
normality_summary_10 <- tidyr::pivot_longer(results_10, 
                                           cols = everything(),
                                           names_to = "Column",
                                           values_to = "P_Value")

# 5) Non-normal columns (p < 0.05)
non_normal_columns_10 <- normality_summary_10 %>% 
  filter(!is.na(P_Value), P_Value < 0.05)

print(normality_summary_10)
print(non_normal_columns_10)
```

### Relative abundance matrix 10 %
```{r Relative abundance matrix 10 perc, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Normalize intensities by the total intensity of each sample (relative abundance)
colnames(filtered_data10)

data_relativ10 <- filtered_data10[, c(7:24)]
data_relativ10 <- sweep(data_relativ10, 2, colSums(data_relativ10, na.rm = TRUE), FUN = "/")

data_matrix10 <- as.matrix(data_relativ10)

# Relative abundance matrix (proteins x samples)
write_csv(
  tibble::rownames_to_column(as.data.frame(data_relativ10), var = "ProteinID"),
  file = "microbial_proteins_rel_abundance_10perc.csv")
```

###------------- 10% ------------
### General representation (plot before doing statistics)

```{r General representation, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Group by group and COG and create columns for number of proteins and intensity
data_representation10 <- data_relativ10
data_representation10$Name <- filtered_data10$Name
data_representation10$Taxonomy <- filtered_data10$Taxonomy_name
data_representation10$KEGG_ko <- filtered_data10$KEGG_ko
data_representation10$KEGG_Pathway_Entry <- filtered_data10$KEGG_Pathway_Entry
data_representation10$COG_accession <- filtered_data10$COG_accession
data_representation10$COG_category <- filtered_data10$COG_category
data_representation10$COG_name <- filtered_data10$COG_name

# Remove categories separated by ";" and keep only the first value
data_representation10 <- data_representation10 %>%
  mutate(COG_category = map_chr(str_split(COG_category, ";"), ~ .x[1]))


# Calculate the mean of time0 and time12
data_representation10$time0_mean <- rowMeans(data_relativ10[, time0])
data_representation10$time12_mean <- rowMeans(data_relativ10[, time12])

# Calculate summed values grouped by COG_category
summary_data <- data_representation10 %>%
  group_by(COG_category) %>%
  summarize(
    time0 = sum(time0_mean, na.rm = TRUE),
    time12 = sum(time12_mean, na.rm = TRUE)
    )

# Summed COG-category relative abundance at each timepoint
write_csv(
  summary_data,
  file = "COG_category_summed_rel_abundance_T0_T12_10perc.csv")


data_long <- summary_data %>%
  pivot_longer(cols = c(time0, time12), names_to = "Group", values_to = "Value")

# Change the names in the 'Group' column for plotting
data_long <- data_long %>%
  mutate(Group = case_when(
    Group == "time0" ~ "Time 0",
    Group == "time12" ~ "Time 12"
  ))


# Group and calculate totals per category
data_long <- data_long %>%
  group_by(COG_category) %>%
  mutate(Order = sum(Value)) %>% # Compute the sum per category (or desired metric)
  ungroup()

# Reorder 'COG_category' based on 'Order' in decreasing order
data_long <- data_long %>%
  mutate(COG_category = reorder(COG_category, -Order))

# Select the top 20 categories
top_20 <- data_long %>%
  group_by(COG_category) %>%
  summarise(Order = sum(Value)) %>%
  slice_max(Order, n = 20)

# Filter the original data by these categories
data_top_20 <- data_long %>%
  filter(COG_category %in% top_20$COG_category)

# Create a mapping dictionary
mappings <- list(
  J = "[J]: Translation, ribosomal structure and biogenesis",
  A = "[A]: RNA processing and modification",
  K = "[K]: Transcription",
  L = "[L]: Replication, recombination and repair",
  B = "[B]: Chromatin structure and dynamics",
  D = "[D]: Cell cycle control, cell division, chromosome partitioning",
  Y = "[Y]: Nuclear structure",
  V = "[V]: Defense mechanisms",
  T = "[T]: Signal transduction mechanisms",
  M = "[M]: Cell wall/membrane/envelope biogenesis",
  N = "[N]: Cell motility",
  Z = "[Z]: Cytoskeleton",
  W = "[W]: Extracellular structures",
  U = "[U]: Intracellular trafficking, secretion, and vesicular transport",
  O = "[O]: Posttranslational modification, protein turnover, chaperones",
  C = "[C]: Energy production and conversion",
  G = "[G]: Carbohydrate transport and metabolism",
  E = "[E]: Amino acid transport and metabolism",
  F = "[F]: Nucleotide transport and metabolism",
  H = "[H]: Coenzyme transport and metabolism",
  I = "[I]: Lipid transport and metabolism",
  P = "[P]: Inorganic ion transport and metabolism",
  Q = "[Q]: Secondary metabolites biosynthesis, transport and catabolism",
  R = "[R]: General function prediction only",
  S = "[S]: Function unknown"
)

# Create the new column 'cog_names', handling NA values
data_top_20_2 <- data_top_20 %>%
  rowwise() %>%
  mutate(
    cog_names = ifelse(
      COG_category == "NA", 
      NA_character_,  
      str_split(COG_category, "") %>%    
        unlist() %>%                     
        map_chr(~ mappings[[.x]]) %>%    
        paste(collapse = " / ")          
    )
  ) %>%
  ungroup()

# Reorder 'cog_names' based on 'Order' in decreasing order
data_top_20_2 <- data_top_20_2 %>%
  mutate(cog_names = reorder(cog_names, -Order))

# Plot with restructured data
p20_COG_10 <- ggplot(data_top_20_2, aes(x = cog_names, y = Value, fill = Group)) +
  geom_bar(stat = "identity", 
           position = position_dodge(width = 0.8), 
           color = "black") +   
  labs(
    x = "COG category",
    y = "Relative abundance"
  ) +
  scale_fill_manual(values = c("Time 0" = "#F4E1A1", "Time 12" = "#A1E1E1"), name = "Group") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1, vjust = 1) 
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 25))
ggsave("top20_COG_bars_functmicro10.svg", plot = p20_COG_10, width = 12, height = 8, dpi = 300)

# Data used for the “top 20 COG categories” barplot
# (includes COG_category, Group, Value, Order, cog_names after mapping)
write_csv(data_top_20_2, file = "COG_top20_barplot_data_10perc.csv")
```

#### PANELS PCAs + Top 20
```{r Panels PCA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
panel_prot10_hum_micro <- (PCA_human_10_var | PCA_micro_10_var) /
                (p20_COG_10) + plot_layout(guides = "collect") #& theme(legend.position = "bottom")

panel_prot10_hum_micro <- panel_prot10_hum_micro + plot_annotation(tag_levels = "A")

ggsave("panel_proteins_PCA10_hum_micro_top20_micro10_var.svg", plot = panel_prot10_hum_micro, width = 10, height = 10, bg = "white")
```


### Wilcoxon test
```{r Wilcoxon, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
res10 <- apply(data_matrix10, 1, function(x) {
  p_val <- wilcox.test(x[time0], x[time12], paired = TRUE)$p.value
  return(p_val)
})
res_adj10 <- p.adjust(res10, method = "fdr")

# Create one column per group with the median abundance per protein
data_relativ10$time0_median <- apply(data_relativ10[, time0], 1, median)
data_relativ10$time12_median <- apply(data_relativ10[, time12], 1, median)

# Create the results table
results10 <- data.frame(
  Name = filtered_data10$Name,
  p_value = res10,
  p_adj =res_adj10,
  Taxonomy = filtered_data10$Taxonomy_name,
  COG_category = filtered_data10$COG_category,
  COG_accession = filtered_data10$COG_accession,
  COG_name = filtered_data10$COG_name,
  KEGG_ko = filtered_data10$KEGG_ko,
  KEGG_pathway = filtered_data10$KEGG_Pathway_Entry,
  time0 = data_relativ10$time0_median,
  time12 = data_relativ10$time12_median)
View(results10)
write.csv(results10, file = "wilcoxon_functmicro10_COG.csv")

# Define a pseudocount to avoid log(0) or division by zero
pseudocount <- 1e-10 
results10$Logfc <- log2((results10[,"time12"] + pseudocount) / (results10[,"time0"] + pseudocount))

# Create a new column indicating in which group the protein intensity is higher
results10$Group <- ifelse(results10$time0 > results10$time12, "time0", "time12")

# Filter by adjusted p-value < 0.05
results_pvalue10 <- results10[results10$p_adj < 0.05,]
write.csv(results_pvalue10, file = "wilcoxon_functmicro10_COG_sig.csv")

# Filter by LogFC < -0.5 or > 0.5
results_pvaluelog10 <- results_pvalue10[results_pvalue10$Logfc > 0.5 | results_pvalue10$Logfc < -0.5, ]

# Remove categories separated by ";" and keep only the first value
results_pvaluelog10 <- results_pvaluelog10 %>%
  mutate(COG_category = map_chr(str_split(COG_category, ";"), ~ .x[1]))

# Save the logFC-filtered table
View(results_pvaluelog10)
write.csv(results_pvaluelog10, file = "wilcoxon_functmicro10_COG_sig_log.csv")
```

### Plot p-value log
#### Data preparation for plotting
```{r Data preparation for plotting, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_combined10 <- results_pvaluelog10

summary_proteins10 <- df_combined10 %>%
  pivot_longer(cols = c(time0, time12), 
               names_to = "Group1", 
               values_to = "Intensity")

summary_proteins10 <- summary_proteins10 %>%
  mutate(Intensity = as.numeric(Intensity))
```

#### Dot plot
```{r Dot plot, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Group by COG and obtain the number of proteins and the sum of intensity
summary_proteins_agg10 <- summary_proteins10 %>%
  group_by(COG_category, Group) %>% 
  summarise(
    Sum_Intensity = sum(Intensity, na.rm = TRUE), 
    Num_Proteins = n() 
    ) %>%
  ungroup()

# Create a mapping dictionary
mappings <- list(
  J = "[J]: Translation, ribosomal structure and biogenesis",
  A = "[A]: RNA processing and modification",
  K = "[K]: Transcription",
  L = "[L]: Replication, recombination and repair",
  B = "[B]: Chromatin structure and dynamics",
  D = "[D]: Cell cycle control, cell division, chromosome partitioning",
  Y = "[Y]: Nuclear structure",
  V = "[V]: Defense mechanisms",
  T = "[T]: Signal transduction mechanisms",
  M = "[M]: Cell wall/membrane/envelope biogenesis",
  N = "[N]: Cell motility",
  Z = "[Z]: Cytoskeleton",
  W = "[W]: Extracellular structures",
  U = "[U]: Intracellular trafficking, secretion, and vesicular transport",
  O = "[O]: Posttranslational modification, protein turnover, chaperones",
  C = "[C]: Energy production and conversion",
  G = "[G]: Carbohydrate transport and metabolism",
  E = "[E]: Amino acid transport and metabolism",
  F = "[F]: Nucleotide transport and metabolism",
  H = "[H]: Coenzyme transport and metabolism",
  I = "[I]: Lipid transport and metabolism",
  P = "[P]: Inorganic ion transport and metabolism",
  Q = "[Q]: Secondary metabolites biosynthesis, transport and catabolism",
  R = "[R]: General function prediction only",
  S = "[S]: Function unknown",
  X = "[X]: NAN"
)

# Create the new column 'cog_names', handling NA values
summary_proteins_agg10 <- summary_proteins_agg10 %>%
  rowwise() %>%
  mutate(
    cog_names = ifelse(
      COG_category == "NA", 
      NA_character_,  
      str_split(COG_category, "") %>%                     
        unlist() %>%                                       
        map_chr(~ mappings[[.x]]) %>%                      
        paste(collapse = " / ")                           
    )
  ) %>%
  ungroup()

write.xlsx(summary_proteins_agg10, "Summary_prot_functmicro10_COG_sig_log.xlsx")

# Plot
p_COG_10 <- ggplot(summary_proteins_agg10, 
                aes(x = Group, y = cog_names, 
                    size = Num_Proteins, fill = Sum_Intensity)) + 
  geom_point(alpha = 0.7, 
             shape = 21,       
             color = "black",  
             stroke = 0.3) +   
  scale_size_continuous(range = c(2, 10), name = "Number of Proteins") +
  scale_fill_gradient(low = "#499894", high = "#E15759", name = "Intensity") + 
  labs(
    x = "Group",
    y = "COG Category"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 50))

ggsave("COG_dotplot_prot_functmicro10_COG_sig_log.svg", plot = p_COG_10, width = 6, height = 4, dpi = 300)
```

#### Optional: using KEGG_ko and Taxonomy
```{r KEGG_ko and Taxonomy, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# KEGG_ko
# Group by KEGG_ko and obtain the number of proteins and the sum of intensity
summary_proteins3_10 <- summary_proteins10 %>%
  separate_rows(KEGG_ko, sep = ",")  

summary_proteins_agg3_10 <- summary_proteins3_10 %>%
  group_by(KEGG_ko, Group) %>%
  summarise(
    Sum_Intensity = sum(Intensity, na.rm = TRUE), 
    Num_Proteins = n() 
    ) %>%
  ungroup()
write.xlsx(summary_proteins_agg3_10, "Summary_prot_functmicro10_KEGG_ko_sig_log.xlsx")


p_KEGGko_10 <- ggplot(summary_proteins_agg3_10, 
                   aes(x = Group, y = KEGG_ko, 
                       size = Num_Proteins, fill = Sum_Intensity)) +
  geom_point(alpha = 0.7,
             shape = 21,        
             color = "black",   
             stroke = 0.3) +    
  scale_size_continuous(range = c(2, 10), name = "Number of Proteins") +
  scale_fill_gradient(low = "#499894", high = "#E15759", name = "Intensity") +
  labs(
    x = "Group",
    y = "KEGG_ko"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 40))

ggsave("KEGG_ko_dotplot_prot_functmicro10_KEGG_ko_sig_log.svg", plot = p_KEGGko_10, width = 6, height = 4, dpi = 300)


# Taxonomy
summary_proteins_agg1_10 <- summary_proteins10 %>%
  group_by(Taxonomy, Group) %>%
  summarise(
    Sum_Intensity = sum(Intensity, na.rm = TRUE), 
    Num_Proteins = n() 
    ) %>%
  ungroup()

write.xlsx(summary_proteins_agg1_10, "Summary_prot_functmicro10_Taxonomy_sig_log.xlsx")


p_TAX_10 <- ggplot(summary_proteins_agg1_10, 
                aes(x = Group, y = Taxonomy, 
                    size = Num_Proteins, fill = Sum_Intensity)) +
  geom_point(alpha = 0.7,
             shape = 21,        
             color = "black",   
             stroke = 0.3) +    
  scale_size_continuous(range = c(2, 10), name = "Number of Proteins") +
  scale_fill_gradient(low = "#499894", high = "#E15759", name = "Intensity") +
  labs(
    x = "Group",
    y = "Taxonomy"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
)

ggsave("Taxonomy_dotplot_prot_functmicro10_TAX_sig_log.svg", plot = p_TAX_10, width = 6, height = 4, dpi = 300)
```


# ------------------------------------------------------------------------------
# FUNCTIONAL GROUPING
# ------------------------------------------------------------------------------

Instead of performing the analysis at the individual protein level, we will aggregate the data by functional groups:
- COG_category
- COG_accession
- KEGG_ko
- KEGG_pathway

This allows us to summarize and compare changes in broader functional categories (e.g. metabolic functions, pathways) rather than focusing on single proteins, which can make the biological interpretation more robust and easier to understand.

###------------- 10% ------------

## COG_category
### Collapse protein-level data to COG categories:
```{r Collapse protein-level data to COG categories, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
summary_COGcat10 <- filtered_data10 %>%
  mutate(COG_category = map_chr(str_split(COG_category, ";"), ~ .x[1]))

summary_COGcat10 <- summary_COGcat10 %>%
  group_by(COG_category) %>%
  summarise(across(starts_with("Intensity"), sum, .names = "{col}"))
```

### Relative abundance calculation
```{r Relative abundance calculation, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
colnames(summary_COGcat10)
data_relativ10 <- summary_COGcat10[, c(2:19)]
data_relativ10 <- sweep(data_relativ10, 2, colSums(data_relativ10, na.rm = TRUE), FUN = "/")

data_matrix10 <- as.matrix(data_relativ10)
```

### Wilcoxon test
```{r Wilcoxon COGcat, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
res10 <- apply(data_matrix10, 1, function(x) {
  p_val <- wilcox.test(x[time0], x[time12], paired = TRUE)$p.value
  return(p_val)
})
res_adj10 <- p.adjust(res10, method = "fdr")

# Create one column per group with the median abundance per protein
data_relativ10$time0_median <- apply(data_relativ10[, time0], 1, median)
data_relativ10$time12_median <- apply(data_relativ10[, time12], 1, median)

# Create the results table
results10 <- data.frame(
  Name = summary_COGcat10$COG_category,
  p_value = res10,
  p_adj =res_adj10,
  time0 = data_relativ10$time0_median,
  time12 = data_relativ10$time12_median)
View(results10)
write.csv(results10, file = "wilcoxon_functmicrogroup10_COGcat.csv")

# Define a pseudocount to avoid log(0) or division by zero
pseudocount <- 1e-10 
results10$Logfc <- log2((results10[,"time12"] + pseudocount) / (results10[,"time0"] + pseudocount))

# Create a new column indicating in which group the protein intensity is higher
results10$Group <- ifelse(results10$time0 > results10$time12, "time0", "time12")

# Filter by adjusted p-value < 0.05
results_pvalue10 <- results10[results10$p_adj < 0.05,]
View(results_pvalue10)

# Filter by LogFC < -0.5 or > 0.5
results_pvaluelog10 <- results_pvalue10[results_pvalue10$Logfc > 0.5 | results_pvalue10$Logfc < -0.5, ]

# Save the logFC-filtered table
write.csv(results_pvalue10, file = "wilcoxon_functmicrogroup10_COGcat_sig.csv")
write.csv(results_pvaluelog10, file = "wilcoxon_functmicrogroup10_COGcat_sig_log.csv")
```

### Optional: volcano plot
```{r Volcano plot, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
results10 <- results10 %>%
  mutate(Significance = case_when(
    p_adj < 0.05 & Logfc > 0.5  ~ "Upregulated",
    p_adj < 0.05 & Logfc < -0.5 ~ "Downregulated",
    TRUE ~ "Not significant"
  ))
results10$neg_log10_padj <- -log10(results10$p_adj)

# Volcano plot
vp_COGcat_10 <- ggplot(results10, aes(x = Logfc, y = neg_log10_padj)) +
  geom_point(aes(color = Significance), size = 2, alpha = 0.8) +
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue", "Not significant" = "grey")) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dotted") +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted") +
  labs(x = expression(log[2]~FC), y = expression(-log[10]~FDR), title = "Volcano plot") +
  theme_minimal()
ggsave("COGcat_volcanoplot_functmicrogroup10_COGcat_sig_log.svg", plot = vp_COGcat_10, width = 8, height = 6, dpi = 300)

# With labels
N <- 10  # change this to the number of labels you want

labels_df <- results10 %>%
  arrange(p_adj) %>%
  slice_head(n = N)

vp_COGcat_10_lab <- vp_COGcat_10 +
  geom_text(data = labels_df, aes(label = Name), vjust = -0.6, size = 3, check_overlap = TRUE)

ggsave("COGcat_volcanoplot_labeled_functmicrogroup10_COGcat_sig_log.svg", plot = vp_COGcat_10_lab, width = 8, height = 6, dpi = 300)
```

## COG_accession
### Collapse protein-level data to COG accession:
```{r Collapse protein-level data to COG accession, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
summary_COGacc10 <- filtered_data10 %>%
  mutate(COG_accession = map_chr(str_split(COG_accession, ";"), ~ .x[1]))

summary_COGacc10 <- summary_COGacc10 %>%
  group_by(COG_accession) %>%
  summarise(across(starts_with("Intensity"), sum, .names = "{col}"))
```

### Relative abundance calculation
```{r Relative abundance calculation, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
colnames(summary_COGacc10)
data_relativ10 <- summary_COGacc10[, c(2:19)]
data_relativ10 <- sweep(data_relativ10, 2, colSums(data_relativ10, na.rm = TRUE), FUN = "/")

data_matrix10 <- as.matrix(data_relativ10)
```

### Wilcoxon test
```{r Wilcoxon COGacc, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
res10 <- apply(data_matrix10, 1, function(x) {
  p_val <- wilcox.test(x[time0], x[time12], paired = TRUE)$p.value
  return(p_val)
})
res_adj10 <- p.adjust(res10, method = "fdr")

# Create one column per group with the median abundance per protein
data_relativ10$time0_median <- apply(data_relativ10[, time0], 1, median)
data_relativ10$time12_median <- apply(data_relativ10[, time12], 1, median)

# Create the results table
results10 <- data.frame(
  Name = summary_COGacc10$COG_accession,
  p_value = res10,
  p_adj =res_adj10,
  time0 = data_relativ10$time0_median,
  time12 = data_relativ10$time12_median)
View(results10)
write.csv(results10, file = "wilcoxon_functmicrogroup10_COGacc.csv")

# Define a pseudocount to avoid log(0) or division by zero
pseudocount <- 1e-10 
results10$Logfc <- log2((results10[,"time12"] + pseudocount) / (results10[,"time0"] + pseudocount))

# Create a new column indicating in which group the protein intensity is higher
results10$Group <- ifelse(results10$time0 > results10$time12, "time0", "time12")

# Filter by adjusted p-value < 0.05
results_pvalue10 <- results10[results10$p_adj < 0.05,]
View(results_pvalue10)

# Filter by LogFC < -0.5 or > 0.5
results_pvaluelog10 <- results_pvalue10[results_pvalue10$Logfc > 0.5 | results_pvalue10$Logfc < -0.5, ]

# Save the logFC-filtered table
write.csv(results_pvalue10, file = "wilcoxon_functmicrogroup10_COGacc_sig.csv")
write.csv(results_pvaluelog10, file = "wilcoxon_functmicrogroup10_COGacc_sig_log.csv")
```

### Optional: volcano plot
```{r Volcano plot, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
results10 <- results10 %>%
  mutate(Significance = case_when(
    p_adj < 0.05 & Logfc > 0.5  ~ "Upregulated",
    p_adj < 0.05 & Logfc < -0.5 ~ "Downregulated",
    TRUE ~ "Not significant"
  ))
results10$neg_log10_padj <- -log10(results10$p_adj)

# Volcano plot
vp_COGacc_10 <- ggplot(results10, aes(x = Logfc, y = neg_log10_padj)) +
  geom_point(aes(color = Significance), size = 2, alpha = 0.8) +
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue", "Not significant" = "grey")) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dotted") +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted") +
  labs(x = expression(log[2]~FC), y = expression(-log[10]~FDR), title = "Volcano plot") +
  theme_minimal()
ggsave("COGacc_volcanoplot_functmicrogroup10_COGacc_sig_log.svg", plot = vp_COGacc_10, width = 8, height = 6, dpi = 300)

# With labels
N <- 10  # change this to the number of labels you want

labels_df <- results10 %>%
  arrange(p_adj) %>%
  slice_head(n = N)

vp_COGacc_10_lab <- vp_COGacc_10 +
  geom_text(data = labels_df, aes(label = Name), vjust = -0.6, size = 3, check_overlap = TRUE)

ggsave("COGacc_volcanoplot_labeled_functmicrogroup10_COGacc_sig_log.svg", plot = vp_COGacc_10_lab, width = 8, height = 6, dpi = 300)
```

# KEGG_ko
### Collapse protein-level data to KEGG_ko:
```{r Collapse protein-level data to KEGG_ko, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
summary_KEGGko10 <- filtered_data10 %>%
  separate_rows(KEGG_ko, sep = ",")

summary_KEGGko10 <- summary_KEGGko10 %>%
  group_by(KEGG_ko) %>%
  summarise(across(starts_with("Intensity"), sum, .names = "{col}"))
```

### Relative abundance calculation
```{r Relative abundance calculation, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
colnames(summary_KEGGko10)
data_relativ10 <- summary_KEGGko10[, c(2:19)]
data_relativ10 <- sweep(data_relativ10, 2, colSums(data_relativ10, na.rm = TRUE), FUN = "/")

data_matrix10 <- as.matrix(data_relativ10)
```

### Wilcoxon test
```{r Wilcoxon KEGG_ko, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
res10 <- apply(data_matrix10, 1, function(x) {
  p_val <- wilcox.test(x[time0], x[time12], paired = TRUE)$p.value
  return(p_val)
})
res_adj10 <- p.adjust(res10, method = "fdr")

# Create one column per group with the median abundance per protein
data_relativ10$time0_median <- apply(data_relativ10[, time0], 1, median)
data_relativ10$time12_median <- apply(data_relativ10[, time12], 1, median)

# Create the results table
results10 <- data.frame(
  Name = summary_KEGGko10$KEGG_ko,
  p_value = res10,
  p_adj =res_adj10,
  time0 = data_relativ10$time0_median,
  time12 = data_relativ10$time12_median)
View(results10)
write.csv(results10, file = "wilcoxon_functmicrogroup10_KEGGko.csv")

# Define a pseudocount to avoid log(0) or division by zero
pseudocount <- 1e-10 
results10$Logfc <- log2((results10[,"time12"] + pseudocount) / (results10[,"time0"] + pseudocount))

# Create a new column indicating in which group the protein intensity is higher
results10$Group <- ifelse(results10$time0 > results10$time12, "time0", "time12")

# Filter by adjusted p-value < 0.05
results_pvalue10 <- results10[results10$p_adj < 0.05,]
View(results_pvalue10)

# Filter by LogFC < -0.5 or > 0.5
results_pvaluelog10 <- results_pvalue10[results_pvalue10$Logfc > 0.5 | results_pvalue10$Logfc < -0.5, ]

# Save the logFC-filtered table
write.csv(results_pvalue10, file = "wilcoxon_functmicrogroup10_KEGGko_sig.csv")
write.csv(results_pvaluelog10, file = "wilcoxon_functmicrogroup10_KEGGko_sig_log.csv")
```

### Optional: volcano plot
```{r Volcano plot, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
results10 <- results10 %>%
  mutate(Significance = case_when(
    p_adj < 0.05 & Logfc > 0.5  ~ "Upregulated",
    p_adj < 0.05 & Logfc < -0.5 ~ "Downregulated",
    TRUE ~ "Not significant"
  ))
results10$neg_log10_padj <- -log10(results10$p_adj)

# Volcano plot
vp_KEGGko_10 <- ggplot(results10, aes(x = Logfc, y = neg_log10_padj)) +
  geom_point(aes(color = Significance), size = 2, alpha = 0.8) +
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue", "Not significant" = "grey")) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dotted") +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted") +
  labs(x = expression(log[2]~FC), y = expression(-log[10]~FDR), title = "Volcano plot") +
  theme_minimal()
ggsave("KEGGko_volcanoplot_functmicrogroup10_KEGGko_sig_log.svg", plot = vp_KEGGko_10, width = 8, height = 6, dpi = 300)

# With labels
N <- 10  # change this to the number of labels you want

labels_df <- results10 %>%
  arrange(p_adj) %>%
  slice_head(n = N)

vp_KEGGko_10_lab <- vp_KEGGko_10 +
  geom_text(data = labels_df, aes(label = Name), vjust = -0.6, size = 3, check_overlap = TRUE)

ggsave("KEGGko_volcanoplot_labeled_functmicrogroup10_KEGGko_sig_log.svg", plot = vp_KEGGko_10_lab, width = 8, height = 6, dpi = 300)
```

# KEGG_pathway
### Collapse protein-level data to KEGG_pathway:
```{r Collapse protein-level data to KEGG_pathway, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
summary_KEGG_path10 <- filtered_data10 %>%
  separate_rows(KEGG_Pathway_Entry, sep = ",")

summary_KEGG_path10 <- summary_KEGG_path10 %>%
  filter(!grepl("^map", KEGG_Pathway_Entry, ignore.case = TRUE))

summary_KEGG_path10 <- summary_KEGG_path10 %>%
  group_by(KEGG_Pathway_Entry) %>%
  summarise(across(starts_with("Intensity"), sum, .names = "{col}"))
```

### Relative abundance calculation
```{r Relative abundance calculation, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
colnames(summary_KEGG_path10)
data_relativ10 <- summary_KEGG_path10[, c(2:19)]
data_relativ10 <- sweep(data_relativ10, 2, colSums(data_relativ10, na.rm = TRUE), FUN = "/")

data_matrix10 <- as.matrix(data_relativ10)
```

### Wilcoxon test
```{r Wilcoxon KEGG_pw, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
res10 <- apply(data_matrix10, 1, function(x) {
  p_val <- wilcox.test(x[time0], x[time12], paired = TRUE)$p.value
  return(p_val)
})
res_adj10 <- p.adjust(res10, method = "fdr")

# Create one column per group with the median abundance per protein
data_relativ10$time0_median <- apply(data_relativ10[, time0], 1, median)
data_relativ10$time12_median <- apply(data_relativ10[, time12], 1, median)

# Create the results table
results10 <- data.frame(
  Name = summary_KEGG_path10$KEGG_Pathway_Entry,
  p_value = res10,
  p_adj =res_adj10,
  time0 = data_relativ10$time0_median,
  time12 = data_relativ10$time12_median)
View(results10)
write.csv(results10, file = "wilcoxon_functmicrogroup10_KEGGpw.csv")

# Define a pseudocount to avoid log(0) or division by zero
pseudocount <- 1e-10 
results10$Logfc <- log2((results10[,"time12"] + pseudocount) / (results10[,"time0"] + pseudocount))

# Create a new column indicating in which group the protein intensity is higher
results10$Group <- ifelse(results10$time0 > results10$time12, "time0", "time12")

# Filter by adjusted p-value < 0.05
results_pvalue10 <- results10[results10$p_adj < 0.05,]
View(results_pvalue10)

# Filter by LogFC < -0.5 or > 0.5
results_pvaluelog10 <- results_pvalue10[results_pvalue10$Logfc > 0.5 | results_pvalue10$Logfc < -0.5, ]

# Save the logFC-filtered table
write.csv(results_pvalue10, file = "wilcoxon_functmicrogroup10_KEGGpw_sig.csv")
write.csv(results_pvaluelog10, file = "wilcoxon_functmicrogroup10_KEGGpw_sig_log.csv")
```

### Optional: volcano plot
```{r Volcano plot, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
results10 <- results10 %>%
  mutate(Significance = case_when(
    p_adj < 0.05 & Logfc > 0.5  ~ "Upregulated",
    p_adj < 0.05 & Logfc < -0.5 ~ "Downregulated",
    TRUE ~ "Not significant"
  ))
results10$neg_log10_padj <- -log10(results10$p_adj)

# Volcano plot
vp_KEGGpw_10 <- ggplot(results10, aes(x = Logfc, y = neg_log10_padj)) +
  geom_point(aes(color = Significance), size = 2, alpha = 0.8) +
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue", "Not significant" = "grey")) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dotted") +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted") +
  labs(x = expression(log[2]~FC), y = expression(-log[10]~FDR), title = "Volcano plot") +
  theme_minimal()
ggsave("KEGGpw_volcanoplot_functmicrogroup10_KEGGpw_sig_log.svg", plot = vp_KEGGpw_10, width = 8, height = 6, dpi = 300)

# With labels
N <- 10  # change this to the number of labels you want

labels_df <- results10 %>%
  arrange(p_adj) %>%
  slice_head(n = N)

vp_KEGGpw_10_lab <- vp_KEGGpw_10 +
  geom_text(data = labels_df, aes(label = Name), vjust = -0.6, size = 3, check_overlap = TRUE)

ggsave("KEGGpw_volcanoplot_labeled_functmicrogroup10_KEGGpw_sig_log.svg", plot = vp_KEGGpw_10_lab, width = 8, height = 6, dpi = 300)
```

