---
title: "cf_metagenomics_thesis_chapter_II"
author: "Natalia Bastón-Paz"
date: '2026'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require("knitr")
opts_knit$set(root.dir = "C:/RProjects/cf_metagenomics_thesis_chapter_II")
```

```{#r Clean environments, message=FALSE, warning=FALSE}
rm(list = ls())
```

## General introduction

Cystic Fibrosis (CF) is a genetic disorder that primarily affects the respiratory, digestive, and reproductive systems. It is caused by mutations in the CFTR (Cystic Fibrosis Transmembrane Conductance Regulator) gene, which encodes a protein responsible for regulating chloride and bicarbonate transport across epithelial cell membranes. These mutations result in abnormal ion transport, leading to thick, sticky mucus accumulation in the lungs, gut, pancreas, and other organs. CF is one of the most common autosomal recessive disorders in Caucasian populations, with an incidence of approximately 1 in 2,500–3,500 live births. People with CF (pwCF) typically present with chronic lung infections due to impaired mucocilliary clearance, pancreatic insufficiency, leading to malabsorption and nutritional deficiencies, liver disease, diabetes, and other systemic complications. The CF lung microbiome is characterized by chronic colonization and infection with pathogens such as *Pseudomonas aeruginosa*, and *Staphylococcus aureus*, among others. These infections contribute to progressive lung damage, inflammation, and exacerbations. Significant advances in CF management, including CFTR modulators, improved airway clearance techniques, and antibiotic therapies, have increased life expectancy and quality of life for pwCF. The triple therapy elexacaftor-tezacaftor-ivacaftor (ETI) was approved in the USA in 2019, but its effects over the microbiota remains unclear.

*Questions to be answered:*

-   *Have the gut and lung microbiota composition changed after one year of treatment with ETI?*

## Libraries

```{r Libraries, message=FALSE, warning=FALSE}
library(mia)     
library(miaViz)
library(tidyverse)         
library(castor)
library(readr)
library(ggpubr)
library(utils)
library(ggplot2)          
library(patchwork)
library(plotrix)
library(ComplexUpset)
library(vegan)
library(stats)
library(rstatix)
library(data.table)
library(Maaslin2)
library(MicrobiomeStat)
library(ANCOMBC)
library(ALDEx2)
library(MADAV)
library(SummarizedExperiment)
library(dplyr)
library(tibble)
library(TreeSummarizedExperiment)
library(stringr)
library(here)
library(svglite)
library(aplot)
library(patchwork)
library(tidyr)
library(ggsci)
library(purrr)
library(openxlsx)
library(broom)
library(ggthemes)
library(RColorBrewer)
library(gghalves)
library(scales)
library(ggdist)
library(forcats)
```

## Load the data PED

Optional: If we want to know which samples have 0 counts:

```{r Samples with 0 counts, echo=TRUE, results='hide'}
df <- read.table(here("data", "cf_metaphlan_counts.txt"),
                 header = TRUE, sep = "\t", 
                 row.names = 1, stringsAsFactors = FALSE)

print(names(which(colSums(df) == 100000000)))
```

#### Metadata file

Load in sample metadata file --\> colData_Metaphlan object:

```{r Load in Sample Metadata file, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
Metaphlan_data_cf <- read_delim(here("data", "cf_metadata_profiles_pediatrics.txt"), 
                                "\t", 
                                escape_double = FALSE, 
                                trim_ws = TRUE)
# Fix column names
colnames(Metaphlan_data_cf)[1] <- 'SampleID'
```

Remove the samples that have 0 counts and others of interest:

```{r Remove samples of choice, message=FALSE, warning=FALSE}
samples_to_remove <- c("A10E0","A10E12","A10H0","A10H12","A12E0","A12E12","A12H0","A12H12","A13E0","A13E12","A13H0","A13H12","A16E0","A16E12","A16H0","A16H12","A17E0","A17E12","A17H0","A17H12","A19E0","A19E12","A19H0","A19H12","A1E0","A1E12","A1H0","A1H12","A20E0","A20E12","A20H0","A20H12","A22E0","A22E12","A22H0","A22H12","A24E0","A24E12","A24H0","A24H12","A25E0","A25E12","A25H0","A25H12","A26E0","A26E12","A26H0","A26H12","A27E0","A27E12","A27H0","A27H12","A28E0","A28E12","A28H0","A28H12","A30E0","A30E12","A30H0","A30H12","A3E0","A3E12","A3H0","A3H12","A4E0","A4E12","A4H0","A4H12","A6E0","A6E12","A6H0","A6H12","A7E0","A7E12","A7H0","A7H12","A8E0","A8E12","A8H0","A8H12","A9E0","A9E12","A9H0","A9H12","P11E0","P11E12","P14E0","P14E12","P16E0","P16E12","P17E0","P17E12","P20E0","P20E12","P20H0","P20H12","P21E0","P21E12","P25E0","P25E12","P25H0","P25H12","P26E0","P26E12","P27E0","P27E12","P2E0","P2E12","P3E0","P3E12","P3H0","P3H12","P42E0","P42E12","P42H0","P42H12","P4E0","P4E12","P5E0","P5E12","P5H0","P5H12")
  # Replace with actual SampleIDs
Metaphlan_data_cf <- Metaphlan_data_cf[!Metaphlan_data_cf$SampleID %in% samples_to_remove, ]
```

Convert the column SampleID into row names:

```{r Convert col SampleID into row names, message=FALSE, warning=FALSE}
Metaphlan_data_cf <- Metaphlan_data_cf %>%
  column_to_rownames(var = 'SampleID') %>%                                          
  data.frame()
```

#### Abundance file

Load in abundance file (change file name and FeatureID_column) --> assay_Metaphlan object:

```{r Load in abundance file, message=FALSE, warning=FALSE}
M_assay_cf <- read_delim(here("data", "cf_metaphlan_counts.txt"), "\t", escape_double = FALSE, trim_ws = TRUE)
colnames(M_assay_cf)[1] <- 'FeatureID'
```

Get the taxonomy data from the FeatureID column of the assay_Metaphlan object --> rowData_Metaphlan object:

```{r Get the taxonomy data from FeatureID, message=FALSE, warning=FALSE}
M_rowData_cf <- M_assay_cf[,'FeatureID'] %>% 
  mutate(split = FeatureID) %>%  
  separate(col = split, into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species",'Type'), sep = "\\|.__") %>%
  mutate(Kingdom = str_remove_all(Kingdom,'^k__')) %>%  
  column_to_rownames(var = "FeatureID") %>%        
  data.frame()
```

Filter the assay_Metaphlan object (M_assay_cf) using the rowData_Metaphlan (M_rowData_cf) & colData_Metaphlan (Metaphlan_data_cf) objects and create the TSE:

```{r Filter and create the TSE, message=FALSE, warning=FALSE}
M_assay_cf <- M_assay_cf[,c('FeatureID',rownames(Metaphlan_data_cf))] %>% # The assay data columns are reorder to match the colData
  column_to_rownames(var = 'FeatureID') %>%        
  as.matrix()
M_TSE_cf_unfiltered <- TreeSummarizedExperiment(assays = M_assay_cf, # To create TreeSummarizedExperiment object                                   
                                                    rowData = M_rowData_cf,
                                                     colData = Metaphlan_data_cf)
assayNames(M_TSE_cf_unfiltered) <-'counts'
```

Rename NA's in 'UNCLASSIFIED' row of rowData(TSE_Metaphlan) --> Replaces NA values (and any other values) with "UNCLASSIFIED" in the UNCLASSIFIED row only:

```{r Rename UNCLASSIFIED, message=FALSE, warning=FALSE}
rowData(M_TSE_cf_unfiltered)[rowData(M_TSE_cf_unfiltered)$Kingdom == 'UNCLASSIFIED',1:ncol(rowData(M_TSE_cf_unfiltered))] <- 'UNCLASSIFIED' 
```

Filter to 'Species' level and retain 'UNCLASSIFIED':

```{r Filter to Species and retain UNCLASSIFIED, message=FALSE, warning=FALSE}
M_TSE_cf_unfiltered <- M_TSE_cf_unfiltered[
  (!is.na(rowData(M_TSE_cf_unfiltered)$Species) & is.na(rowData(M_TSE_cf_unfiltered)$Type)) | 
    (rowData(M_TSE_cf_unfiltered)$Kingdom == 'UNCLASSIFIED'),
]
```

Optional filtering steps: Remove all non-bacteria taxa (check first whether Bacteria are in the 'Kingdom' or 'Domain' class). Here we also keep unclassified Bacteria and UNCLASSIFIED

```{r Remove non-bacteria taxa, message=FALSE, warning=FALSE}
M_TSE_cf_bacteria <- M_TSE_cf_unfiltered[rowData(M_TSE_cf_unfiltered)$Kingdom %in% c('UNCLASSIFIED','Bacteria'), ]
```

Remove all singleton features:

```{r Remove singletons, message=FALSE, warning=FALSE}
M_TSE_cf_singletons <- M_TSE_cf_bacteria[apply(assay(M_TSE_cf_bacteria), 1, function(x) { sum(x > 0) > 1 }), ]
```

Remove 'UNCLASSIFIED' feature:

```{r Remove UNCLASSIFIED, message=FALSE, warning=FALSE}
M_TSE_cf_UN <- M_TSE_cf_singletons
M_TSE_cf_no_UN <- M_TSE_cf_singletons[rownames(M_TSE_cf_singletons) != 'UNCLASSIFIED',]
```

*REMEMBER:*

-   *For Relative Abundance use: M_TSE_cf_UN*

-   *For alpha and beta diversity, for differential abundance and for the relative abundance of the differential abundance analysis use: M_TSE_cf_no_UN*

## Create subsets

Subsets by Time:

```{r Subset by Time}
subset_T0 <- M_TSE_cf_UN[, M_TSE_cf_UN$Time %in% 'T0']
subset_T12 <- M_TSE_cf_UN[, M_TSE_cf_UN$Time %in% 'T12']
subset_T0_no_UN <- M_TSE_cf_no_UN[, M_TSE_cf_no_UN$Time %in% 'T0']
subset_T12_no_UN <- M_TSE_cf_no_UN[, M_TSE_cf_no_UN$Time %in% 'T12']
```

Subsets by SampleType:

```{r Subset by SampleType}
subset_feces <- M_TSE_cf_UN[, M_TSE_cf_UN$SampleType %in% 'feces']
subset_sputum <- M_TSE_cf_UN[, M_TSE_cf_UN$SampleType %in% 'sputum']
subset_feces_no_UN <- M_TSE_cf_no_UN[, M_TSE_cf_no_UN$SampleType %in% 'feces']
subset_sputum_no_UN <- M_TSE_cf_no_UN[, M_TSE_cf_no_UN$SampleType %in% 'sputum']
```

# ------------------------------------------------------------------------------
# RELATIVE ABUNDANCE (RA)
# ------------------------------------------------------------------------------

```{r RA of Feces CSV at T0 and T12, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
# Species
rank_level <- "Species"  # "Phylum", "Genus", etc.

RA_agg <- mia::transformAssay(M_TSE_cf_UN, method = "relabundance") %>%
  mia::agglomerateByRank(rank_level, onRankOnly = TRUE, na.rm = TRUE)

ra_mat <- SummarizedExperiment::assay(RA_agg, "relabundance")

ra_wide <- as.data.frame(ra_mat) %>%
  tibble::rownames_to_column(var = "FeatureID")

readr::write_csv(ra_wide, here::here("results", paste0("relabundance_", rank_level, "_wide.csv")))

# Phylum
rank_level <- "Phylum"

RA_agg <- mia::transformAssay(M_TSE_cf_UN, method = "relabundance") %>%
  mia::agglomerateByRank(rank_level, onRankOnly = TRUE, na.rm = TRUE)

ra_mat <- SummarizedExperiment::assay(RA_agg, "relabundance")

ra_wide <- as.data.frame(ra_mat) %>%
  tibble::rownames_to_column(var = "FeatureID")

readr::write_csv(ra_wide, here::here("results", paste0("relabundance_", rank_level, "_wide.csv")))
```

##### Feces at T0 and T12 - PHYLUM

```{r RA of Feces at T0 and T12 phylum, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
RA_bar_cf_fc_phylum <- RA_bar(tse = M_TSE_cf_UN,
                    var = c('Time'),
                    var_levels = c('T0', 'T12'),
                    level = 'Phylum',
                    top_features = 5,
                    feature_order = 'abund',
                    sample_order = 'SampleType',
                    sample_order_levels = NULL,
                    feature_bar = NULL,
                    feature_bar_levels = 'SampleType',
                    group_average = FALSE,
                    plot_title = 'Relative Abundance of Feces (Phylum)',
                    cutoff_features = 'None',
                    show_counts = FALSE)

# Save as SVG
ggsave(filename = here("results", "RA_bar_cf_fc_phylum.svg"), plot = RA_bar_cf_fc_phylum$RA_bar_plot, width = 12, height = 8, dpi = 300, device = "svg")
```

##### Feces at T0 and T12 - SPECIES

```{r RA of Feces at T0 and T12, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
RA_bar_cf_fc_sps <- RA_bar(tse = M_TSE_cf_UN,
                    var = c('Time'),
                    var_levels = c('T0', 'T12'),
                    level = 'Species',
                    top_features = 10,
                    feature_order = 'abund',
                    sample_order = 'SampleType',
                    sample_order_levels = NULL,
                    feature_bar = NULL,
                    feature_bar_levels = 'SampleType',
                    group_average = FALSE,
                    plot_title = 'Relative Abundance of Feces',
                    cutoff_features = 'None',
                    show_counts = FALSE)
# Save as SVG
ggsave(filename = here("results", "RA_bar_cf_fc_sps.svg"), plot = RA_bar_cf_fc_sps$RA_bar_plot, width = 12, height = 8, dpi = 300, device = "svg")
```


##### PANEL PLOTS

```{r Panel RA of Feces at T0 and T12, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
p_RA <- (RA_bar_cf_fc_phylum$RA_bar_plot | RA_bar_cf_fc_sps$RA_bar_plot) +
  patchwork::plot_layout(guides = "collect") &
  ggplot2::theme(legend.position = "bottom")

p_RA <- p_RA + plot_annotation(tag_levels = "A")

ggsave(here("results","RA_feces_ped_horizontal.svg"), p_RA, width = 13, height = 6, bg = "white")

p_RA <- (RA_bar_cf_fc_phylum$RA_bar_plot) / (RA_bar_cf_fc_sps$RA_bar_plot) +
  patchwork::plot_layout(guides = "collect") #&
  #ggplot2::theme(legend.position = "bottom")

p_RA <- p_RA + plot_annotation(tag_levels = "A")

ggsave(here("results","RA_feces_ped_vertical.svg"), p_RA, width = 10, height = 10, bg = "white")
```


##### Stats
###### Phylum

```{r RA stats phylum at T0 and T12 phylum, echo=TRUE, message=FALSE, warning=FALSE}
# 1) Load in the file
df <- read_delim(here("results", "relabundance_Phylum_wide.csv"))

# 2) Long format + extract PatientID and Time (H0 vs H12)
long <- df %>%
  pivot_longer(-FeatureID, names_to = "SampleID", values_to = "abundance") %>%
  mutate(
    Time = if_else(str_detect(SampleID, "H12$"), "H12", "H0"),
    PatientID = str_remove(SampleID, "H(0|12)$"),
    Time = factor(Time, levels = c("H0", "H12"))
  )

# 3) Basic stats by phylum and time
stats_basic <- long %>%
  group_by(FeatureID, Time) %>%
  summarise(
    n = sum(!is.na(abundance)),
    mean = mean(abundance, na.rm = TRUE),
    sd   = sd(abundance, na.rm = TRUE),
    median = median(abundance, na.rm = TRUE),
    iqr = IQR(abundance, na.rm = TRUE),
    .groups = "drop"
  )

# 4) Paired test per phylum: H0 vs H12 (within-patient)
paired_results <- long %>%
  dplyr::select(FeatureID, PatientID, Time, abundance) %>%
  tidyr::pivot_wider(names_from = Time, values_from = abundance) %>%
  dplyr::mutate(
    delta = H12 - H0,
    pair_ok = stats::complete.cases(H0, H12)
  ) %>%
  dplyr::group_by(FeatureID) %>%
  dplyr::summarise(
    n_pairs = sum(pair_ok),

    # Detection prevalence in PAIRED participants only (% with abundance > 0)
    prev_H0  = mean(H0[pair_ok]  > 0, na.rm = TRUE),
    prev_H12 = mean(H12[pair_ok] > 0, na.rm = TRUE),

    # Pooled relative abundance (sum within time; converted to % below)
    sum_H0  = sum(H0[pair_ok],  na.rm = TRUE),
    sum_H12 = sum(H12[pair_ok], na.rm = TRUE),

    # Normality on paired differences
    shapiro_p = if (n_pairs >= 3) stats::shapiro.test(delta[pair_ok])$p.value else NA_real_,

    method = dplyr::case_when(
      n_pairs < 3 ~ NA_character_,
      shapiro_p >= 0.05 ~ "paired_t",
      shapiro_p < 0.05  ~ "wilcoxon_paired"
    ),

    p_value = dplyr::case_when(
      n_pairs < 3 ~ NA_real_,
      shapiro_p >= 0.05 ~ stats::t.test(H12[pair_ok], H0[pair_ok], paired = TRUE)$p.value,
      shapiro_p < 0.05  ~ stats::wilcox.test(H12[pair_ok], H0[pair_ok], paired = TRUE, exact = FALSE)$p.value
    ),

    mean_delta   = mean(delta[pair_ok], na.rm = TRUE),
    median_delta = median(delta[pair_ok], na.rm = TRUE),

    .groups = "drop"
  ) %>%
  dplyr::mutate(
    pooled_prev_H0  = 100 * sum_H0  / sum(sum_H0,  na.rm = TRUE),
    pooled_prev_H12 = 100 * sum_H12 / sum(sum_H12, na.rm = TRUE),
    p_fdr = stats::p.adjust(p_value, method = "fdr")
  ) %>%
  dplyr::select(-sum_H0, -sum_H12) %>%
  dplyr::arrange(p_fdr)

# 5) Save outputs
readr::write_excel_csv(stats_basic, here::here("results", "phyla_basic_stats_by_time.csv"))
readr::write_excel_csv(paired_results, here::here("results", "phyla_paired_tests_H0_vs_H12.csv"))

# View
stats_basic
paired_results
```
###### Species

```{r RA stats species at T0 and T12 phylum, echo=TRUE, message=FALSE, warning=FALSE}
# 1) Load in the file
df <- read_delim(here("results", "relabundance_Species_wide.csv"))

# 2) Long format + extract PatientID and Time (H0 vs H12)
long <- df %>%
  pivot_longer(-FeatureID, names_to = "SampleID", values_to = "abundance") %>%
  mutate(
    Time = if_else(str_detect(SampleID, "H12$"), "H12", "H0"),
    PatientID = str_remove(SampleID, "H(0|12)$"),
    Time = factor(Time, levels = c("H0", "H12"))
  )

# 3) Basic stats by phylum and time
stats_basic <- long %>%
  group_by(FeatureID, Time) %>%
  summarise(
    n = sum(!is.na(abundance)),
    mean = mean(abundance, na.rm = TRUE),
    sd   = sd(abundance, na.rm = TRUE),
    median = median(abundance, na.rm = TRUE),
    iqr = IQR(abundance, na.rm = TRUE),
    .groups = "drop"
  )

# 4) Paired test per phylum: H0 vs H12 (within-patient)
paired_results <- long %>%
  dplyr::select(FeatureID, PatientID, Time, abundance) %>%
  tidyr::pivot_wider(names_from = Time, values_from = abundance) %>%
  dplyr::mutate(
    delta = H12 - H0,
    pair_ok = stats::complete.cases(H0, H12)
  ) %>%
  dplyr::group_by(FeatureID) %>%
  dplyr::summarise(
    n_pairs = sum(pair_ok),

    # Detection prevalence in PAIRED participants only (% with abundance > 0)
    prev_H0  = mean(H0[pair_ok]  > 0, na.rm = TRUE),
    prev_H12 = mean(H12[pair_ok] > 0, na.rm = TRUE),

    # Pooled relative abundance (sum within time; converted to % below)
    sum_H0  = sum(H0[pair_ok],  na.rm = TRUE),
    sum_H12 = sum(H12[pair_ok], na.rm = TRUE),

    # Normality on paired differences
    shapiro_p = if (n_pairs >= 3) stats::shapiro.test(delta[pair_ok])$p.value else NA_real_,

    method = dplyr::case_when(
      n_pairs < 3 ~ NA_character_,
      shapiro_p >= 0.05 ~ "paired_t",
      shapiro_p < 0.05  ~ "wilcoxon_paired"
    ),

    p_value = dplyr::case_when(
      n_pairs < 3 ~ NA_real_,
      shapiro_p >= 0.05 ~ stats::t.test(H12[pair_ok], H0[pair_ok], paired = TRUE)$p.value,
      shapiro_p < 0.05  ~ stats::wilcox.test(H12[pair_ok], H0[pair_ok], paired = TRUE, exact = FALSE)$p.value
    ),

    mean_delta   = mean(delta[pair_ok], na.rm = TRUE),
    median_delta = median(delta[pair_ok], na.rm = TRUE),

    .groups = "drop"
  ) %>%
  dplyr::mutate(
    pooled_prev_H0  = 100 * sum_H0  / sum(sum_H0,  na.rm = TRUE),
    pooled_prev_H12 = 100 * sum_H12 / sum(sum_H12, na.rm = TRUE),
    p_fdr = stats::p.adjust(p_value, method = "fdr")
  ) %>%
  dplyr::select(-sum_H0, -sum_H12) %>%
  dplyr::arrange(p_fdr)

# 5) Save outputs
readr::write_excel_csv(stats_basic, here::here("results", "species_basic_stats_by_time.csv"))
readr::write_excel_csv(paired_results, here::here("results", "species_paired_tests_H0_vs_H12.csv"))

# View
stats_basic
paired_results
```

## Calculate relative abundance

```{r Function to calculate relative abundance, echo=TRUE, message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# RELATIVE ABUNDANCE TABLES BY SAMPLE TYPE AND TAXONOMIC LEVEL
# Phylum in feces, Species in feces
# ------------------------------------------------------------------------------

calc_relative_abundance <- function(tse, rank_level, output_prefix, group_vars = c("Time")) {
  
  # 0. Standardize phylum names before relative abundance calculation
  # This is only applied when calculating relative abundance at Phylum level
  if (rank_level == "Phylum") {
    
    phylum_map <- c(
      "Firmicutes"      = "Bacillota",
      "Bacteroidetes"   = "Bacteroidota",
      "Proteobacteria"  = "Pseudomonadota",
      "Actinobacteria"  = "Actinomycetota",
      "Verrucomicrobia" = "Verrucomicrobiota",
      "Fusobacteria"    = "Fusobacteriota",
      "Spirochaetes"    = "Spirochaetota",
      "Tenericutes"     = "Mycoplasmatota"
    )
    
    phylum <- as.character(SummarizedExperiment::rowData(tse)$Phylum)
    
    phylum[phylum %in% names(phylum_map)] <- phylum_map[
      phylum[phylum %in% names(phylum_map)]
    ]
    
    SummarizedExperiment::rowData(tse)$Phylum <- phylum
  }
  
  # 1. Calculate relative abundance per sample and agglomerate at selected rank
  RA_agg <- mia::transformAssay(tse, method = "relabundance") %>%
    mia::agglomerateByRank(rank_level, onRankOnly = TRUE, na.rm = TRUE)
  
  # 2. Extract relative abundance matrix and convert to percentage
  ra_mat <- SummarizedExperiment::assay(RA_agg, "relabundance") * 100
  
  # 3. Taxonomy table
  tax_table <- as.data.frame(SummarizedExperiment::rowData(RA_agg)) %>%
    tibble::rownames_to_column(var = "FeatureID") %>%
    dplyr::mutate(
      Rank = rank_level,
      Taxon = .data[[rank_level]]
    ) %>%
    dplyr::select(FeatureID, Rank, Taxon)
  
  # 4. Metadata table
  meta_table <- as.data.frame(SummarizedExperiment::colData(RA_agg)) %>%
    tibble::rownames_to_column(var = "SampleID")
  
  # 5. Wide table: taxa as rows, samples as columns
  ra_wide <- as.data.frame(ra_mat) %>%
    tibble::rownames_to_column(var = "FeatureID") %>%
    dplyr::left_join(tax_table, by = "FeatureID") %>%
    dplyr::select(FeatureID, Rank, Taxon, dplyr::everything())
  
  # 6. Long table: one row per taxon and sample
  ra_long <- as.data.frame(ra_mat) %>%
    tibble::rownames_to_column(var = "FeatureID") %>%
    tidyr::pivot_longer(
      cols = -FeatureID,
      names_to = "SampleID",
      values_to = "Relative_abundance_percent"
    ) %>%
    dplyr::left_join(tax_table, by = "FeatureID") %>%
    dplyr::left_join(meta_table, by = "SampleID") %>%
    dplyr::select(
      SampleID,
      SampleType,
      dplyr::all_of(group_vars),
      Rank,
      Taxon,
      Relative_abundance_percent,
      dplyr::everything()
    )
  
  # 7. Summary table by group, usually T0 and T12
  ra_summary <- ra_long %>%
    dplyr::group_by(
      dplyr::across(dplyr::all_of(c("SampleType", group_vars, "Rank", "Taxon")))
    ) %>%
    dplyr::summarise(
      n_samples = dplyr::n(),
      mean_RA_percent = mean(Relative_abundance_percent, na.rm = TRUE),
      sd_RA_percent = sd(Relative_abundance_percent, na.rm = TRUE),
      median_RA_percent = median(Relative_abundance_percent, na.rm = TRUE),
      Q1_RA_percent = quantile(Relative_abundance_percent, 0.25, na.rm = TRUE),
      Q3_RA_percent = quantile(Relative_abundance_percent, 0.75, na.rm = TRUE),
      min_RA_percent = min(Relative_abundance_percent, na.rm = TRUE),
      max_RA_percent = max(Relative_abundance_percent, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::arrange(SampleType, dplyr::across(dplyr::all_of(group_vars)), desc(mean_RA_percent))
  
  # 8. Export CSV files
  readr::write_csv(
    ra_wide,
    here::here("results", paste0(output_prefix, "_wide_by_sample.csv"))
  )
  
  readr::write_csv(
    ra_long,
    here::here("results", paste0(output_prefix, "_long_by_sample.csv"))
  )
  
  readr::write_csv(
    ra_summary,
    here::here("results", paste0(output_prefix, "_summary_by_", paste(group_vars, collapse = "_"), ".csv"))
  )
  
  # 9. Return objects to R environment
  return(list(
    RA_agg = RA_agg,
    wide = ra_wide,
    long = ra_long,
    summary = ra_summary
  ))
}
```

## Run to calculate RA

```{r Run function, echo=TRUE, message=FALSE, warning=FALSE}
# Phylum in feces
RA_phylum_feces <- calc_relative_abundance(
  tse = subset_feces,
  rank_level = "Phylum",
  output_prefix = "pediatric_relabundance_phylum_feces"
)

# Species in feces
RA_species_feces <- calc_relative_abundance(
  tse = subset_feces,
  rank_level = "Species",
  output_prefix = "pediatric_relabundance_species_feces"
)
```

### Check distribution

```{r Check distribution, echo=TRUE, message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# CHECK DISTRIBUTION OF RELATIVE ABUNDANCE VALUES
# Safe version: avoids Shapiro-Wilk errors when all values are identical
# ------------------------------------------------------------------------------

safe_shapiro_p <- function(x) {
  
  x <- x[!is.na(x)]
  
  if (length(x) < 3) {
    return(NA_real_)
  }
  
  if (length(x) > 5000) {
    return(NA_real_)
  }
  
  if (length(unique(x)) < 3) {
    return(NA_real_)
  }
  
  return(shapiro.test(x)$p.value)
}

check_RA_distribution <- function(ra_long) {
  
  distribution_table <- ra_long %>%
    dplyr::group_by(SampleType, Time, Rank, Taxon) %>%
    dplyr::summarise(
      n_samples = sum(!is.na(Relative_abundance_percent)),
      n_nonzero = sum(Relative_abundance_percent > 0, na.rm = TRUE),
      n_unique_values = length(unique(Relative_abundance_percent[!is.na(Relative_abundance_percent)])),
      zero_percent = mean(Relative_abundance_percent == 0, na.rm = TRUE) * 100,
      mean_RA_percent = mean(Relative_abundance_percent, na.rm = TRUE),
      sd_RA_percent = sd(Relative_abundance_percent, na.rm = TRUE),
      median_RA_percent = median(Relative_abundance_percent, na.rm = TRUE),
      Q1_RA_percent = quantile(Relative_abundance_percent, 0.25, na.rm = TRUE),
      Q3_RA_percent = quantile(Relative_abundance_percent, 0.75, na.rm = TRUE),
      min_RA_percent = min(Relative_abundance_percent, na.rm = TRUE),
      max_RA_percent = max(Relative_abundance_percent, na.rm = TRUE),
      shapiro_p = safe_shapiro_p(Relative_abundance_percent),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      distribution = dplyr::case_when(
        n_unique_values < 3 ~ "Not assessed: identical or nearly identical values",
        is.na(shapiro_p) ~ "Not assessed",
        shapiro_p < 0.05 ~ "Non-normal",
        shapiro_p >= 0.05 ~ "Compatible with normality"
      ),
      recommended_summary = dplyr::case_when(
        n_unique_values < 3 ~ "Median [IQR]",
        zero_percent > 20 ~ "Median [IQR]",
        is.na(shapiro_p) ~ "Median [IQR]",
        shapiro_p < 0.05 ~ "Median [IQR]",
        shapiro_p >= 0.05 ~ "Mean +/- SD"
      )
    ) %>%
    dplyr::arrange(SampleType, Time, Rank, desc(mean_RA_percent))
  
  return(distribution_table)
}
```

### Run distribution

```{r Run distribution, echo=TRUE, message=FALSE, warning=FALSE}
dist_phylum_feces <- check_RA_distribution(RA_phylum_feces$long)

dist_species_feces <- check_RA_distribution(RA_species_feces$long)
```

### Export distribution

```{r Export distribution, echo=TRUE, message=FALSE, warning=FALSE}
readr::write_csv(dist_phylum_feces, here::here("results", "pediatric_distribution_phylum_feces.csv"))

readr::write_csv(dist_species_feces, here::here("results", "pediatric_distribution_species_feces.csv"))
```


## Patient by patient FECES

### Taxonomy harmonization
```{r Txonomy harmonization, echo=TRUE, message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# TAXONOMIC NAME STANDARDIZATION
# ------------------------------------------------------------------------------

standardize_taxon_names_in_ra_long <- function(ra_long) {
  
  phylum_map <- c(
    "Firmicutes"      = "Bacillota",
    "Bacteroidetes"   = "Bacteroidota",
    "Proteobacteria"  = "Pseudomonadota",
    "Actinobacteria"  = "Actinomycetota",
    "Verrucomicrobia" = "Verrucomicrobiota",
    "Fusobacteria"    = "Fusobacteriota",
    "Spirochaetes"    = "Spirochaetota",
    "Tenericutes"     = "Mycoplasmatota"
  )
  
  ra_long_std <- ra_long %>%
    dplyr::mutate(
      Taxon = dplyr::case_when(
        Rank == "Phylum" & Taxon %in% names(phylum_map) ~ phylum_map[Taxon],
        TRUE ~ Taxon
      )
    ) %>%
    dplyr::group_by(
      SampleID,
      SampleType,
      Time,
      Rank,
      Taxon,
      dplyr::across(dplyr::any_of(c("PatientID", "AgeGroup", "Sex")))
    ) %>%
    dplyr::summarise(
      Relative_abundance_percent = sum(Relative_abundance_percent, na.rm = TRUE),
      .groups = "drop"
    )
  
  return(ra_long_std)
}
```

### Functions

```{r Functions, echo=TRUE, message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# PAIRED PATIENT-LEVEL RELATIVE ABUNDANCE CHANGES
# For taxa shown in the fecal relative abundance figure
# ------------------------------------------------------------------------------

# Taxa shown in Figure 6A
feces_phyla_fig6 <- c(
  "Bacillota",
  "Actinomycetota",
  "Verrucomicrobiota",
  "Pseudomonadota",
  "Bacteroidota"
)

# Taxa shown in Figure 6B
feces_species_fig6 <- c(
  "Bifidobacterium_longum",
  "Bifidobacterium_pseudocatenulatum",
  "Bifidobacterium_bifidum",
  "Bifidobacterium_adolescentis",
  "Ruminococcus_bromii",
  "Escherichia_coli",
  "Akkermansia_muciniphila",
  "Eubacterium_rectale",
  "Collinsella_aerofaciens"
)

# ------------------------------------------------------------------------------
# Function 1: create patient-level paired table
# ------------------------------------------------------------------------------

get_paired_RA_changes <- function(ra_long,
                                  taxa_keep,
                                  detection_threshold = 0,
                                  change_threshold = 0) {
  
  # Standardize phylum names and merge duplicated taxa after renaming
  ra_long <- standardize_taxon_names_in_ra_long(ra_long)
  
  # Also standardize taxa_keep, in case old phylum names are used
  phylum_map <- c(
    "Firmicutes"      = "Bacillota",
    "Bacteroidetes"   = "Bacteroidota",
    "Proteobacteria"  = "Pseudomonadota",
    "Actinobacteria"  = "Actinomycetota",
    "Verrucomicrobia" = "Verrucomicrobiota",
    "Fusobacteria"    = "Fusobacteriota",
    "Spirochaetes"    = "Spirochaetota",
    "Tenericutes"     = "Mycoplasmatota"
  )
  
  taxa_keep[taxa_keep %in% names(phylum_map)] <- phylum_map[
    taxa_keep[taxa_keep %in% names(phylum_map)]
  ]
  
  df <- ra_long %>%
    dplyr::filter(
      Taxon %in% taxa_keep,
      Time %in% c("T0", "T12"),
      !Taxon %in% c("UNCLASSIFIED", "Unclassified", "Other")
    )
  
  # If PatientID is not present, derive it from SampleID
  if (!"PatientID" %in% colnames(df)) {
    df <- df %>%
      dplyr::mutate(
        PatientID = stringr::str_remove(SampleID, "(H|E)(0|12)$")
      )
  }
  
  paired_df <- df %>%
    dplyr::select(
      PatientID,
      SampleID,
      SampleType,
      Time,
      Rank,
      Taxon,
      Relative_abundance_percent
    ) %>%
    dplyr::group_by(PatientID, SampleType, Rank, Taxon, Time) %>%
    dplyr::summarise(
      SampleID = paste(unique(SampleID), collapse = ";"),
      RA_percent = sum(Relative_abundance_percent, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    tidyr::pivot_wider(
      names_from = Time,
      values_from = c(SampleID, RA_percent),
      names_sep = "_"
    ) %>%
    dplyr::filter(
      !is.na(RA_percent_T0),
      !is.na(RA_percent_T12)
    ) %>%
    dplyr::mutate(
      detected_T0 = RA_percent_T0 > detection_threshold,
      detected_T12 = RA_percent_T12 > detection_threshold,
      delta_T12_minus_T0 = RA_percent_T12 - RA_percent_T0,
      
      detection_status = dplyr::case_when(
        detected_T0 & detected_T12 ~ "detected_T0_and_T12",
        detected_T0 & !detected_T12 ~ "detected_T0_only",
        !detected_T0 & detected_T12 ~ "detected_T12_only",
        !detected_T0 & !detected_T12 ~ "not_detected"
      ),
      
      change_direction = dplyr::case_when(
        delta_T12_minus_T0 > change_threshold ~ "increase",
        delta_T12_minus_T0 < -change_threshold ~ "decrease",
        TRUE ~ "stable"
      )
    ) %>%
    dplyr::arrange(Taxon, PatientID)
  
  return(paired_df)
}

# ------------------------------------------------------------------------------
# Function 2: summarise paired changes by taxon
# ------------------------------------------------------------------------------

summarise_paired_RA_changes <- function(paired_df) {
  
  summary_df <- paired_df %>%
    dplyr::group_by(SampleType, Rank, Taxon) %>%
    dplyr::summarise(
      n_paired = dplyr::n(),
      
      detected_T0_n = sum(detected_T0, na.rm = TRUE),
      detected_T12_n = sum(detected_T12, na.rm = TRUE),
      
      detected_T0_percent = detected_T0_n / n_paired * 100,
      detected_T12_percent = detected_T12_n / n_paired * 100,
      
      detected_T0_only_n = sum(detection_status == "detected_T0_only", na.rm = TRUE),
      detected_T12_only_n = sum(detection_status == "detected_T12_only", na.rm = TRUE),
      detected_both_n = sum(detection_status == "detected_T0_and_T12", na.rm = TRUE),
      not_detected_n = sum(detection_status == "not_detected", na.rm = TRUE),
      
      increased_n = sum(change_direction == "increase", na.rm = TRUE),
      decreased_n = sum(change_direction == "decrease", na.rm = TRUE),
      stable_n = sum(change_direction == "stable", na.rm = TRUE),
      
      median_T0_all = median(RA_percent_T0, na.rm = TRUE),
      Q1_T0_all = quantile(RA_percent_T0, 0.25, na.rm = TRUE),
      Q3_T0_all = quantile(RA_percent_T0, 0.75, na.rm = TRUE),
      
      median_T12_all = median(RA_percent_T12, na.rm = TRUE),
      Q1_T12_all = quantile(RA_percent_T12, 0.25, na.rm = TRUE),
      Q3_T12_all = quantile(RA_percent_T12, 0.75, na.rm = TRUE),
      
      median_delta_all = median(delta_T12_minus_T0, na.rm = TRUE),
      Q1_delta_all = quantile(delta_T12_minus_T0, 0.25, na.rm = TRUE),
      Q3_delta_all = quantile(delta_T12_minus_T0, 0.75, na.rm = TRUE),
      
      median_T0_among_T0_detected = if (sum(detected_T0, na.rm = TRUE) > 0) {
        median(RA_percent_T0[detected_T0], na.rm = TRUE)
      } else {
        NA_real_
      },
      
      median_T12_among_T0_detected = if (sum(detected_T0, na.rm = TRUE) > 0) {
        median(RA_percent_T12[detected_T0], na.rm = TRUE)
      } else {
        NA_real_
      },
      
      median_delta_among_T0_detected = if (sum(detected_T0, na.rm = TRUE) > 0) {
        median(delta_T12_minus_T0[detected_T0], na.rm = TRUE)
      } else {
        NA_real_
      },
      
      patients_detected_T0_only = paste(PatientID[detection_status == "detected_T0_only"], collapse = "; "),
      patients_detected_T12_only = paste(PatientID[detection_status == "detected_T12_only"], collapse = "; "),
      patients_with_decrease = paste(PatientID[change_direction == "decrease"], collapse = "; "),
      patients_with_increase = paste(PatientID[change_direction == "increase"], collapse = "; "),
      
      .groups = "drop"
    ) %>%
    dplyr::arrange(Rank, desc(detected_T0_n), desc(median_T0_all))
  
  return(summary_df)
}

# ------------------------------------------------------------------------------
# Function 3: extract the individual samples with highest RA per taxon and timepoint
# ------------------------------------------------------------------------------

get_top_RA_samples <- function(ra_long,
                               taxa_keep,
                               n_top = 5,
                               detection_threshold = 0) {
  
  # Standardize phylum names and merge duplicated taxa after renaming
  ra_long <- standardize_taxon_names_in_ra_long(ra_long)
  
  # Also standardize taxa_keep
  phylum_map <- c(
    "Firmicutes"      = "Bacillota",
    "Bacteroidetes"   = "Bacteroidota",
    "Proteobacteria"  = "Pseudomonadota",
    "Actinobacteria"  = "Actinomycetota",
    "Verrucomicrobia" = "Verrucomicrobiota",
    "Fusobacteria"    = "Fusobacteriota",
    "Spirochaetes"    = "Spirochaetota",
    "Tenericutes"     = "Mycoplasmatota"
  )
  
  taxa_keep[taxa_keep %in% names(phylum_map)] <- phylum_map[
    taxa_keep[taxa_keep %in% names(phylum_map)]
  ]
  
  df <- ra_long %>%
    dplyr::filter(
      Taxon %in% taxa_keep,
      Time %in% c("T0", "T12"),
      !Taxon %in% c("UNCLASSIFIED", "Unclassified", "Other")
    )
  
  if (!"PatientID" %in% colnames(df)) {
    df <- df %>%
      dplyr::mutate(
        PatientID = stringr::str_remove(SampleID, "(H|E)(0|12)$")
      )
  }
  
  top_df <- df %>%
    dplyr::filter(Relative_abundance_percent > detection_threshold) %>%
    dplyr::group_by(SampleType, Rank, Taxon, Time) %>%
    dplyr::arrange(desc(Relative_abundance_percent), .by_group = TRUE) %>%
    dplyr::mutate(rank_within_taxon_time = dplyr::row_number()) %>%
    dplyr::filter(rank_within_taxon_time <= n_top) %>%
    dplyr::ungroup() %>%
    dplyr::select(
      SampleType,
      Rank,
      Taxon,
      Time,
      rank_within_taxon_time,
      PatientID,
      SampleID,
      Relative_abundance_percent
    )
  
  return(top_df)
}
```

### Run functions

```{r Run functions, echo=TRUE, message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# Apply to fecal phyla
# ------------------------------------------------------------------------------

feces_phylum_patient_changes <- get_paired_RA_changes(
  ra_long = RA_phylum_feces$long,
  taxa_keep = feces_phyla_fig6,
  detection_threshold = 0,
  change_threshold = 0
)

feces_phylum_summary_changes <- summarise_paired_RA_changes(
  feces_phylum_patient_changes
)

feces_phylum_top_samples <- get_top_RA_samples(
  ra_long = RA_phylum_feces$long,
  taxa_keep = feces_phyla_fig6,
  n_top = 5,
  detection_threshold = 0
)

# ------------------------------------------------------------------------------
# Apply to fecal species
# ------------------------------------------------------------------------------

feces_species_patient_changes <- get_paired_RA_changes(
  ra_long = RA_species_feces$long,
  taxa_keep = feces_species_fig6,
  detection_threshold = 0,
  change_threshold = 0
)

feces_species_summary_changes <- summarise_paired_RA_changes(
  feces_species_patient_changes
)

feces_species_top_samples <- get_top_RA_samples(
  ra_long = RA_species_feces$long,
  taxa_keep = feces_species_fig6,
  n_top = 5,
  detection_threshold = 0
)
```

### Export results

```{r Export results, echo=TRUE, message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# EXPORT FECAL PATIENT-LEVEL TABLES
# ------------------------------------------------------------------------------
# Phylum-level fecal tables
readr::write_csv(
  feces_phylum_patient_changes,
  here::here("results", "pediatric_feces_phylum_patient_level_T0_T12_changes.csv")
)

readr::write_csv(
  feces_phylum_summary_changes,
  here::here("results", "pediatric_feces_phylum_summary_T0_T12_changes.csv")
)

readr::write_csv(
  feces_phylum_top_samples,
  here::here("results", "pediatric_feces_phylum_top_samples_by_taxon_time.csv")
)

# Species-level fecal tables
readr::write_csv(
  feces_species_patient_changes,
  here::here("results", "pediatric_feces_species_patient_level_T0_T12_changes.csv")
)

readr::write_csv(
  feces_species_summary_changes,
  here::here("results", "pediatric_feces_species_summary_T0_T12_changes.csv")
)

readr::write_csv(
  feces_species_top_samples,
  here::here("results", "pediatric_feces_species_top_samples_by_taxon_time.csv")
)
```


# ------------------------------------------------------------------------------
# ALPHA DIVERSITY (AD)
# ------------------------------------------------------------------------------

A modification for the AD_plot() function from MADAV package was used in order to enable the paired analysis and export the results.

Indices used for Alpha Diversity:

-   Chao1: estimated species richness

-   Shannon: species richness and evenness

-   Inverse Simpson: dominance (evenness)

### Modified function of AD_plot {MADAV}: ADplot_paired_csv

```{r AD_plot MODIFIED for paired results and CSV, message=FALSE, warning=FALSE}
# Drop-in replacement of AD_plot:
# - Adds `paired` and `export_csv` parameters
# - Exports alpha-diversity values to CSV if requested
# - Uses paired comparisons in geom_pwc() when paired = TRUE
AD_plot_paired_csv <- function (tse,
                     assay_type = "counts",
                     var,
                     var_levels = NULL,
                     level = NULL,
                     index = "inverse_simpson",
                     plot_title = "Alpha-diversity",
                     hide.ns = FALSE,
                     paired = FALSE,            # new argument
                     export_csv = NULL) {       # new argument

  `%>%` <- magrittr::`%>%`
  AD <- tse

  if (is.null(level)) {
    level <- mia::taxonomyRanks(tse)[length(mia::taxonomyRanks(tse))]
    warning(paste("No level was specified, alpha-diversity analysis was performed on",
                  level, "level"))
  }

  if (level %in% colnames(SummarizedExperiment::rowData(AD)) &
      !(level %in% mia::taxonomyRanks(AD))) {
    if ("domain" %in% colnames(SummarizedExperiment::rowData(AD))) {
      colnames(SummarizedExperiment::rowData(AD))[colnames(SummarizedExperiment::rowData(AD)) ==
        "domain"] <- "true_domain_column"
    }
    colnames(SummarizedExperiment::rowData(AD))[colnames(SummarizedExperiment::rowData(AD)) ==
      level] <- "domain"
    rank <- "domain"
  } else if (level %in% mia::taxonomyRanks(AD)) {
    rank <- level
  }

  AD_agg <- mia::agglomerateByRank(AD, rank, onRankOnly = TRUE, na.rm = TRUE)

  if (!is.null(var_levels)) {
    SummarizedExperiment::colData(AD_agg)[, var] <- factor(
      SummarizedExperiment::colData(AD_agg)[, var], levels = var_levels
    )
  } else {
    SummarizedExperiment::colData(AD_agg)[, var] <- factor(
      SummarizedExperiment::colData(AD_agg)[, var]
    )
  }

  set_levels <- levels(SummarizedExperiment::colData(AD_agg)[, var])

  if (index %in% c("coverage", "fisher", "gini_simpson", "inverse_simpson",
                   "log_modulo_skewness", "shannon")) {
    AD_index <- mia::estimateDiversity(AD_agg, index = index, assay.type = assay_type)
  } else if (index %in% c("ace", "chao1", "hill", "observed")) {
    AD_index <- mia::estimateRichness(AD_agg, index = index, assay.type = assay_type)
  }

  # NEW: build a tidy alpha-diversity table and optionally export it
  alpha_df <- SummarizedExperiment::colData(AD_index) %>%
    as.data.frame() %>%
    dplyr::transmute(
      SampleID  = rownames(.),
      PatientID = .data[["PatientID"]],
      Time      = .data[[var]],
      AlphaValue = .data[[index]]
  )
  if (!is.null(export_csv)) {
    readr::write_csv(alpha_df, export_csv)
    message("Alpha-diversity exported to: ", export_csv)
  }

  y_label <- stringr::str_replace_all(index, "_", " ") %>%
    stringr::str_to_title() %>% paste("Index", sep = " ")

  plot_palette <- c("#1F77B4FF", "#AEC7E8FF", "#FF7F0EFF",
                    "#FFBB78FF", "#2CA02CFF", "#98DF8AFF", "#D62728FF", "#FF9896FF",
                    "#9467BDFF", "#C5B0D5FF", "#8C564BFF", "#C49C94FF", "#E377C2FF",
                    "#F7B6D2FF", "#7F7F7FFF", "#C7C7C7FF", "#BCBD22FF", "#DBDB8DFF",
                    "#17BECFFF", "#9EDAE5FF")

  if (length(set_levels) <= 10) {
    plot_colors <- plot_palette[seq(1, 19, 2)][1:length(set_levels)]
    color_scale <- ggplot2::scale_color_manual(values = plot_colors)
  } else if (length(set_levels) <= 20) {
    plot_colors <- plot_palette[sort(c(seq(2, 20, 2)[1:(length(set_levels) - 5)],
                                       seq(1, 19, 2)))]
    color_scale <- ggplot2::scale_color_manual(values = plot_colors)
  } else {
    color_scale <- ggraph::scale_color_viridis(discrete = TRUE)
  }

  AD_plot_final <- ggplot2::ggplot(
    SummarizedExperiment::colData(AD_index),
    ggplot2::aes(SummarizedExperiment::colData(AD_index)[, var],
                 SummarizedExperiment::colData(AD_index)[, index])
  ) +
    ggplot2::geom_boxplot(lwd = 0.65, show.legend = FALSE, outlier.shape = NA) +
    ggplot2::geom_jitter(
      ggplot2::aes(color = SummarizedExperiment::colData(AD_index)[, var]),
      position = ggplot2::position_jitterdodge(0.2),
      alpha = 0.75, show.legend = FALSE
    ) +
    color_scale +
    ggplot2::theme_bw(base_size = 16, base_line_size = 1, base_rect_size = 1) +
    ggplot2::theme(
      axis.title.x = ggplot2::element_blank(),
      plot.title   = ggplot2::element_text(size = 17),
      panel.grid   = ggplot2::element_blank(),
      axis.text.x  = ggplot2::element_text(color = "Grey1", size = 12, angle = 45, hjust = 1, face = "plain"),
      axis.text.y  = ggplot2::element_text(color = "Grey1", size = 12, face = "plain"),
      axis.title.y = ggplot2::element_text(color = "Grey1", size = 14, face = "plain")
    ) +
    ggpubr::geom_pwc(
      label = "p.adj.signif",
      p.adjust.method = "BH",
      hide.ns = hide.ns,
      method.args = list(paired = paired)  # NEW: pass paired to tests
    ) +
    ggplot2::ggtitle(plot_title) +
    ggplot2::ylab(y_label)

  print(AD_plot_final)
  return(invisible(AD_plot_final))
}
```

#### Feces at T0 and T12

```{r AD of Feces at T0 and T12, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
AD_plot_fc_ch_csv <- AD_plot_paired_csv(tse = M_TSE_cf_UN,
                                var = 'Time',
                                var_levels = c('T0','T12'),
                                level = 'Species',
                                index = 'chao1',  
                                plot_title = 'Alpha diversity of Feces',
                                hide.ns = FALSE,
                                paired = TRUE,
                                export_csv = here::here("results", "alpha_feces_species_chao1.csv"))

AD_plot_fc_sh_csv <- AD_plot_paired_csv(tse = M_TSE_cf_UN,
                                var = 'Time',
                                var_levels = c('T0','T12'),
                                level = 'Species',
                                index = 'shannon',  
                                plot_title = 'Alpha diversity of Feces',
                                hide.ns = FALSE,
                                paired = TRUE,
                                export_csv = here::here("results", "alpha_feces_species_shannon.csv"))

AD_plot_fc_is_csv <- AD_plot_paired_csv(tse = M_TSE_cf_UN,
                                var = 'Time',
                                var_levels = c('T0','T12'),
                                level = 'Species',
                                index = 'inverse_simpson',
                                plot_title = 'Alpha diversity of Feces',
                                hide.ns = FALSE,
                                paired = TRUE,
                                export_csv = here::here("results", "alpha_feces_species_inverse_simpson.csv"))
```

##### Plots

###### Chao1

```{r Paper plots alpha diversity FECES Chao1, echo=TRUE, fig.show='show', message=FALSE, warning=FALSE}
# Load data
alpha_df_fc <- read_csv(here::here("results", "alpha_feces_species_chao1.csv")) %>%
  mutate(Time = factor(Time, levels = c("T0","T12"))) %>%
  group_by(PatientID) %>% filter(n_distinct(Time) == 2) %>% ungroup()

# Compute paired Wilcoxon test
wilc_fc <- alpha_df_fc %>%
  wilcox_test(AlphaValue ~ Time, paired = TRUE) %>%
  adjust_pvalue(method = "BH")

# Format p-value for label
p_value_label_fc <- ifelse(wilc_fc$p.adj < 0.001, "p.adj < 0.001",
                           paste0("p.adj = ", signif(wilc_fc$p.adj, 3)))

# Define colors for boxplots and points
col_box   <- c("T0" = "#F4E1A1", "T12" = "#A1E1E1")
col_point <- c("T0" = "#FAF3DC", "T12" = "#D6F5F5")

# Add color columns and numeric x-axis values for plotting
alpha_df_fc <- alpha_df_fc %>%
  mutate(
    fill_box   = col_box[Time],
    fill_point = col_point[Time],
    x          = as.numeric(Time)  
  )

# Data for p-value line and label
p_values_fc <- data.frame(
  x = 1, xend = 2,
  y = max(alpha_df_fc$AlphaValue) * 1.05,
  label = p_value_label_fc
)

# Raincloud-style plot with connecting lines
p_chao1_fc <- ggplot() +
  # Raincloud T0
  stat_slab(
    data = alpha_df_fc %>% filter(Time == "T0"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Raincloud T12
  stat_slab(
    data = alpha_df_fc %>% filter(Time == "T12"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Boxplots
  geom_boxplot(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, fill = fill_box),
    width = 0.5, outlier.shape = NA,
    color = "black", alpha = 0.9, show.legend = TRUE
  ) +
  # Points
  geom_point(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, fill = fill_point),
    shape = 21, size = 3, stroke = 0.3, color = "black",
    show.legend = FALSE
  ) +
  # Connecting lines
  geom_line(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, group = PatientID),
    color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
  ) +
  # p-value line and text
  geom_segment(
    data = p_values_fc,
    aes(x = x, xend = xend, y = y, yend = y),
    color = "black", linewidth = 0.5
  ) +
  geom_text(
    data = p_values_fc,
    aes(x = (x + xend) / 2, y = y + 0.02 * y, label = label),
    size = 4
  ) +
  scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
  scale_fill_identity(guide = "legend", name = "Time",
                      breaks = col_box, labels = names(col_box)) +
  labs(x = NULL, y = "Chao1 index - Feces") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 14)
  )

p_chao1_fc

# Save plot
ggsave(filename = here("results", "alpha_diversity_chao1_feces.svg"), plot = p_chao1_fc, width = 8, height = 6, dpi = 300, device = "svg")
```

###### Shannon

```{r Paper plots alpha diversity FECES Shannon, echo=TRUE, fig.show='show', message=FALSE, warning=FALSE}
# Load data
alpha_df_fc <- read_csv(here::here("results", "alpha_feces_species_shannon.csv")) %>%
  mutate(Time = factor(Time, levels = c("T0","T12"))) %>%
  group_by(PatientID) %>% filter(n_distinct(Time) == 2) %>% ungroup()

# Compute paired Wilcoxon test
wilc_fc <- alpha_df_fc %>%
  wilcox_test(AlphaValue ~ Time, paired = TRUE) %>%
  adjust_pvalue(method = "BH")

# Format p-value for label
p_value_label_fc <- ifelse(wilc_fc$p.adj < 0.001, "p.adj < 0.001",
                           paste0("p.adj = ", signif(wilc_fc$p.adj, 3)))

# Define colors for boxplots and points
col_box   <- c("T0" = "#F4E1A1", "T12" = "#A1E1E1")
col_point <- c("T0" = "#FAF3DC", "T12" = "#D6F5F5")

# Add color columns and numeric x-axis values for plotting
alpha_df_fc <- alpha_df_fc %>%
  mutate(
    fill_box   = col_box[Time],
    fill_point = col_point[Time],
    x          = as.numeric(Time)  
  )

# Data for p-value line and label
p_values_fc <- data.frame(
  x = 1, xend = 2,
  y = max(alpha_df_fc$AlphaValue) * 1.05,
  label = p_value_label_fc
)

# Raincloud-style plot with connecting lines
p_shannon_fc <- ggplot() +
  # Raincloud T0
  stat_slab(
    data = alpha_df_fc %>% filter(Time == "T0"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Raincloud T12
  stat_slab(
    data = alpha_df_fc %>% filter(Time == "T12"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Boxplots
  geom_boxplot(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, fill = fill_box),
    width = 0.5, outlier.shape = NA,
    color = "black", alpha = 0.9, show.legend = TRUE
  ) +
  # Points
  geom_point(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, fill = fill_point),
    shape = 21, size = 3, stroke = 0.3, color = "black",
    show.legend = FALSE
  ) +
  # Connecting lines
  geom_line(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, group = PatientID),
    color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
  ) +
  # p-value line and text
  geom_segment(
    data = p_values_fc,
    aes(x = x, xend = xend, y = y, yend = y),
    color = "black", linewidth = 0.5
  ) +
  geom_text(
    data = p_values_fc,
    aes(x = (x + xend) / 2, y = y + 0.02 * y, label = label),
    size = 4
  ) +
  scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
  scale_fill_identity(guide = "legend", name = "Time",
                      breaks = col_box, labels = names(col_box)) +
  labs(x = NULL, y = "Shannon index - Feces") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 14)
  )

p_shannon_fc

# Save plot
ggsave(filename = here("results", "alpha_diversity_shannon_feces.svg"), plot = p_shannon_fc, width = 8, height = 6, dpi = 300, device = "svg")
```

###### Inverse Simpson

```{r Paper plots alpha diversity FECES Inverse Simpson, echo=TRUE, fig.show='show', message=FALSE, warning=FALSE}
# Load data
alpha_df_fc <- read_csv(here::here("results", "alpha_feces_species_inverse_simpson.csv")) %>%
  mutate(Time = factor(Time, levels = c("T0","T12"))) %>%
  group_by(PatientID) %>% filter(n_distinct(Time) == 2) %>% ungroup()

# Compute paired Wilcoxon test
wilc_fc <- alpha_df_fc %>%
  wilcox_test(AlphaValue ~ Time, paired = TRUE) %>%
  adjust_pvalue(method = "BH")

# Format p-value for label
p_value_label_fc <- ifelse(wilc_fc$p.adj < 0.001, "p.adj < 0.001",
                           paste0("p.adj = ", signif(wilc_fc$p.adj, 3)))

# Define colors for boxplots and points
col_box   <- c("T0" = "#F4E1A1", "T12" = "#A1E1E1")
col_point <- c("T0" = "#FAF3DC", "T12" = "#D6F5F5")

# Add color columns and numeric x-axis values for plotting
alpha_df_fc <- alpha_df_fc %>%
  mutate(
    fill_box   = col_box[Time],
    fill_point = col_point[Time],
    x          = as.numeric(Time)  
  )

# Raincloud-style plot with connecting lines
p_values_fc <- data.frame(
  x = 1, xend = 2,
  y = max(alpha_df_fc$AlphaValue) * 1.05,
  label = p_value_label_fc
)

# Raincloud-style plot with connecting lines
p_invsimp_fc <- ggplot() +
  # Raincloud T0
  stat_slab(
    data = alpha_df_fc %>% filter(Time == "T0"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Raincloud T12
  stat_slab(
    data = alpha_df_fc %>% filter(Time == "T12"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Boxplots
  geom_boxplot(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, fill = fill_box),
    width = 0.5, outlier.shape = NA,
    color = "black", alpha = 0.9, show.legend = TRUE
  ) +
  # Points
  geom_point(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, fill = fill_point),
    shape = 21, size = 3, stroke = 0.3, color = "black",
    show.legend = FALSE
  ) +
  # Connecting lines
  geom_line(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, group = PatientID),
    color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
  ) +
  # p-value line and text
  geom_segment(
    data = p_values_fc,
    aes(x = x, xend = xend, y = y, yend = y),
    color = "black", linewidth = 0.5
  ) +
  geom_text(
    data = p_values_fc,
    aes(x = (x + xend) / 2, y = y + 0.02 * y, label = label),
    size = 4
  ) +
  scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
  scale_fill_identity(guide = "legend", name = "Time",
                      breaks = col_box, labels = names(col_box)) +
  labs(x = NULL, y = "Inverse Simpson index - Feces") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 14)
  )

p_invsimp_fc

# Save plot
ggsave(filename = here("results", "alpha_diversity_invsimp_feces.svg"), plot = p_invsimp_fc, width = 8, height = 6, dpi = 300, device = "svg")
```

##### PANEL PLOTS

```{r Panel AD of Feces at T0 and T12, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
p_AD <- (p_chao1_fc | p_shannon_fc | p_invsimp_fc) +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

p_AD <- p_AD + plot_annotation(tag_levels = "A")

ggsave(here("results","AD_feces_ped.svg"), p_AD, width = 10, height = 4, bg = "white")
```



# ------------------------------------------------------------------------------
# BETA DIVERSITY (BD)
# ------------------------------------------------------------------------------

A modification for the BD_plot() function from MADAV package was used in order to enable the paired analysis and export the results.

Dissimilarity Metrics used:

-   Bray-Curtis: species abundance dissimilarity

-   Weighted UniFrac: phylogenetic structure and abundance

### Modified function of BD_plot {MADAV}: BDplot_paired_csv

```{r BD_plot MODIFIED for paired results and CSV, message=FALSE, warning=FALSE}
# Paired version of BD_plot with CSV exports and optional dissimilarity plot
# Drop-in replacement of the original BD_plot:
# - Uses adonis2(..., strata = PatientID) for paired PERMANOVA (within-patient design)
# - Adds `export_prefix` argument to optionally save:
#     * PERMANOVA results       -> <export_prefix>_permanova.csv
#     * PERMUTEST (betadisper)  -> <export_prefix>_permutest.csv
#     * Ordination coordinates  -> <export_prefix>_coords.csv
#     * % variance of PCoA axes -> <export_prefix>_axes.csv
# - Preserves the `dissimilarity_plot` option from the original BD_plot:
#     when TRUE, also computes between-group distance boxplots and returns
#     the dissimilarity dataframe and the combined (ordination + boxplot) panel
# - Returns a list with the ordination plot, PERMANOVA and PERMUTEST objects
#   (and, if dissimilarity_plot = TRUE, also the dissimilarity plot, data and wrap)

BD_plot_paired_csv <- function(tse,
                               assay_type = "counts",
                               var,
                               var_levels = NULL,
                               level = NULL,
                               method = "bray",
                               weighted_UniFrac = TRUE,
                               ordination = "PCoA",
                               constraining = var,
                               plot_title = "Beta Diversity",
                               dissimilarity_plot = FALSE,
                               hide.ns = FALSE,
                               export_prefix = NULL) {

  set.seed(1234)
  `%>%` <- magrittr::`%>%`
  BD <- tse

  if (is.null(level)) {
    level <- mia::taxonomyRanks(tse)[length(mia::taxonomyRanks(tse))]
    warning(paste("No level was specified, beta-diversity analysis was performed on", level, "level"))
  }

  if (level %in% colnames(SummarizedExperiment::rowData(BD)) &
      !(level %in% mia::taxonomyRanks(BD))) {
    if ("domain" %in% colnames(SummarizedExperiment::rowData(BD))) {
      colnames(SummarizedExperiment::rowData(BD))[colnames(SummarizedExperiment::rowData(BD)) == "domain"] <- "true_domain_column"
    }
    colnames(SummarizedExperiment::rowData(BD))[colnames(SummarizedExperiment::rowData(BD)) == level] <- "domain"
    rank <- "domain"
  } else if (level %in% mia::taxonomyRanks(BD)) {
    rank <- level
  }

  BD_agg <- mia::agglomerateByRank(BD, rank, onRankOnly = TRUE,
                                   agglomerateTree = TRUE, na.rm = TRUE)

  if (length(var) == 1) var_levels <- list(var_levels)
  for (cov in 1:length(var)) {
    if (!is.null(var_levels[[cov]])) {
      SummarizedExperiment::colData(BD_agg)[, var[cov]] <- factor(
        SummarizedExperiment::colData(BD_agg)[, var[cov]],
        levels = var_levels[[cov]]
      )
    } else {
      SummarizedExperiment::colData(BD_agg)[, var[cov]] <- factor(
        SummarizedExperiment::colData(BD_agg)[, var[cov]]
      )
    }
  }

  set_levels <- lapply(var, function(x) levels(SummarizedExperiment::colData(BD_agg)[, x]))

  if (method %in% c("manhattan","euclidean","canberra","clark","bray","kulczynski","gower",
                    "altGower","morisita","horn","mountford","raup","binomial","chao","cao",
                    "mahalanobis","chisq","chord","hellinger","aitchison","robust.aitchison")) {
    BD_dist <- vegan::vegdist(as.matrix(t(SummarizedExperiment::assays(BD_agg)[[assay_type]])),
                              method = method, binary = FALSE)
  } else if (method == "jaccard") {
    BD_dist <- vegan::vegdist(as.matrix(t(SummarizedExperiment::assays(BD_agg)[[assay_type]])),
                              method = method, binary = TRUE)
  } else if (method == "UniFrac") {
    BD_dist <- calculateUnifrac(BD_agg, assay.type = assay_type, weighted = weighted_UniFrac)
  } else if (method == "JSD") {
    BD_dist <- calculateJSD(BD_agg, assay.type = assay_type)
  }

  # PERMANOVA (paired by PatientID)
  permanova <- vegan::adonis2(
    BD_dist ~ SummarizedExperiment::colData(BD_agg)[, var[1]],
    strata = SummarizedExperiment::colData(BD_agg)$PatientID
  ) %>% data.frame()
  rownames(permanova)[1] <- var[1]

  cat("Permutation test for adonis (paired by PatientID)\n\n")
  print(permanova)

  permanova_R2 <- signif(permanova[1, 3], 2)
  permanova_P  <- signif(permanova[1, 5], 2)
  permanova_plot_text <- bquote(atop(R^2:~.(permanova_R2),
                                     paste("     Pr(>F): ", .(permanova_P))))

  # Dispersion + permutest
  BD_dispr  <- vegan::betadisper(BD_dist, SummarizedExperiment::colData(BD_agg)[, var[1]])
  permutest <- vegan::permutest(BD_dispr, pairwise = TRUE)
  print(permutest)

  # Ordenación
  pct1 <- pct2 <- NA_real_

  if (ordination == "PCoA") {
    BD_ord <- stats::cmdscale(BD_dist, eig = TRUE)
    BD_ord_df <- data.frame(
      BD_ord$points[, 1],
      BD_ord$points[, 2],
      SummarizedExperiment::colData(BD_agg)[, var]
    )
    pct1 <- round(100 * BD_ord$eig[1] / sum(BD_ord$eig), 2)
    pct2 <- round(100 * BD_ord$eig[2] / sum(BD_ord$eig), 2)
    axis_title_x <- paste0("PCoA1 (", pct1, "%)")
    axis_title_y <- paste0("PCoA2 (", pct2, "%)")
  } else if (ordination == "NMDS") {
    BD_ord <- vegan::metaMDS(BD_dist, k = 2)
    BD_ord_df <- data.frame(BD_ord$points,
                            SummarizedExperiment::colData(BD_agg)[, var])
    axis_title_x <- "NMDS1"
    axis_title_y <- "NMDS2"
  } else if (ordination == "RDA") {
    BD_ord <- mia::calculateRDA(
      BD_agg,
      stats::formula(paste("data ~", paste(constraining, collapse = " + "))),
      method = method, assay.type = assay_type, homogeneity.test = "permanova"
    )
    BD_ord_df <- data.frame(BD_ord[, 1:2],
                            SummarizedExperiment::colData(BD_agg)[, var])
    axis_title_x <- paste0(
      "dbRDA1 (",
      round(100 * attr(BD_ord, "rda")$CCA$eig[1] /
              attr(BD_ord, "rda")$CCA$tot.chi, 2), "%)"
    )
    axis_title_y <- paste0(
      "dbRDA2 (",
      round(100 * attr(BD_ord, "rda")$CCA$eig[2] /
              attr(BD_ord, "rda")$CCA$tot.chi, 2), "%)"
    )
  } else if (ordination == "CCA") {
    BD_ord <- mia::calculateCCA(
      BD_agg,
      stats::formula(paste("data ~", paste(constraining, collapse = " + "))),
      method = method, assay.type = assay_type, homogeneity.test = "permanova"
    )
    BD_ord_df <- data.frame(BD_ord[, 1:2],
                            SummarizedExperiment::colData(BD_agg)[, var])
    axis_title_x <- paste0(
      "CCA1 (",
      round(100 * attr(BD_ord, "cca")$CCA$eig[1] /
              attr(BD_ord, "cca")$CCA$tot.chi, 2), "%)"
    )
    axis_title_y <- paste0(
      "CCA2 (",
      round(100 * attr(BD_ord, "cca")$CCA$eig[2] /
              attr(BD_ord, "cca")$CCA$tot.chi, 2), "%)"
    )
  } else if (ordination == "DPCoA") {
    BD_ord <- mia::calculateDPCoA(BD_agg)
    BD_ord_df <- data.frame(BD_ord[, 1:2],
                            SummarizedExperiment::colData(BD_agg)[, var])
    axis_title_x <- paste0(
      "DPCoA1 (",
      round(100 * attr(BD_ord, "eig")[1] /
              sum(attr(BD_ord, "eig")), 2), "%)"
    )
    axis_title_y <- paste0(
      "DPCoA2 (",
      round(100 * attr(BD_ord, "eig")[2] /
              sum(attr(BD_ord, "eig")), 2), "%)"
    )
  }

  if (length(var) > 1) {
    shape     <- BD_ord_df[, 4]
    shape_lab <- var[2]
  } else {
    shape     <- NULL
    shape_lab <- NULL
  }

  plot_palette <- c("#1F77B4FF","#AEC7E8FF","#FF7F0EFF","#FFBB78FF",
                    "#2CA02CFF","#98DF8AFF","#D62728FF","#FF9896FF",
                    "#9467BDFF","#C5B0D5FF","#8C564BFF","#C49C94FF",
                    "#E377C2FF","#F7B6D2FF","#7F7F7FFF","#C7C7C7FF",
                    "#BCBD22FF","#DBDB8DFF","#17BECFFF","#9EDAE5FF")

  if (length(set_levels[[1]]) <= 5) {
    plot_colors <- plot_palette[seq(1, 9, 2)][1:length(set_levels[[1]])]
    color_scale <- ggplot2::scale_color_manual(values = plot_colors)
  } else if (length(set_levels[[1]]) <= 10) {
    plot_colors <- plot_palette[
      sort(c(seq(2, 10, 2)[1:(length(set_levels[[1]]) - 5)],
             seq(1, 9, 2)))
    ]
    color_scale <- ggplot2::scale_color_manual(values = plot_colors)
  } else if (length(set_levels[[1]]) <= 20) {
    plot_colors <- plot_palette[
      sort(c(seq(2, 20, 2)[1:(length(set_levels[[1]]) - 10)],
             seq(1, 19, 2)))
    ]
    color_scale <- ggplot2::scale_color_manual(values = plot_colors)
  } else {
    color_scale <- ggraph::scale_color_viridis(discrete = TRUE)
  }

  BD_ord_plot <- ggplot2::ggplot(
    BD_ord_df,
    ggplot2::aes(BD_ord_df[, 1], BD_ord_df[, 2], color = BD_ord_df[, 3])
  ) +
    ggplot2::geom_point(ggplot2::aes(shape = shape), size = 2) +
    ggplot2::ggtitle(plot_title) +
    color_scale +
    ggplot2::stat_ellipse(
      ggplot2::aes(group = BD_ord_df[, 3]),
      linetype = 1, linewidth = 1
    ) +
    ggplot2::labs(
      caption = permanova_plot_text,
      x = axis_title_x, y = axis_title_y, shape = shape_lab
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      plot.title = ggplot2::element_text(size = 17),
      legend.text = ggplot2::element_text(size = 12),
      legend.title = ggplot2::element_text(size = 13),
      panel.grid = ggplot2::element_blank(),
      axis.text.x = ggplot2::element_text(color = "Black", size = 12),
      axis.text.y = ggplot2::element_text(color = "Black", size = 12),
      axis.title.x = ggplot2::element_text(color = "Black", size = 14),
      axis.title.y = ggplot2::element_text(color = "Black", size = 14),
      plot.caption.position = "plot",
      plot.caption = ggplot2::element_text(hjust = 0.08, size = 11)
    ) +
    ggplot2::guides(
      color = ggplot2::guide_legend(order = 1, title = var[1]),
      shape = ggplot2::guide_legend(order = 2, title = var[2])
    )

  # ---------- EXPORTS ----------
  if (!is.null(export_prefix)) {
    # 1) PERMANOVA
    readr::write_csv(
      permanova,
      file = here::here("results", paste0(export_prefix, "_permanova.csv"))
    )
    # 2) PERMUTEST
    readr::write_csv(
      as.data.frame(permutest$tab),
      file = here::here("results", paste0(export_prefix, "_permutest.csv"))
    )
    # 3) COORDINATES
    sample_ids <- rownames(
      if (ordination %in% c("PCoA", "NMDS"))
        BD_ord$points
      else
        as.matrix(BD_ord_df[, 1:2])
    )
    if (is.null(sample_ids)) {
      sample_ids <- colnames(SummarizedExperiment::assays(BD_agg)[[assay_type]])
    }
    coords_df <- data.frame(
      SampleID = sample_ids,
      Axis1    = BD_ord_df[[1]],
      Axis2    = BD_ord_df[[2]],
      Group    = BD_ord_df[[3]]
    )
    readr::write_csv(
      coords_df,
      file = here::here("results", paste0(export_prefix, "_coords.csv"))
    )
    # 4) % of axes (only PCoA)
    if (ordination == "PCoA" && is.finite(pct1) && is.finite(pct2)) {
      readr::write_csv(
        tibble::tibble(Axis = c("PCoA1","PCoA2"),
                       Percent = c(pct1, pct2)),
        here::here("results", paste0(export_prefix, "_axes.csv"))
      )
    }
  }
  # -----------------------------

  # Dissimilarity boxplots
  if (dissimilarity_plot == TRUE) {

    colData_frame <- data.frame(
      rownames(SummarizedExperiment::colData(BD_agg)),
      SummarizedExperiment::colData(BD_agg)[, var[1]]
    )
    colnames(colData_frame) <- c("Sample", var[1])

    BD_dissim <- as.data.frame(as.table(as.matrix(BD_dist))) %>%
      stats::setNames(c("Sample1", "Sample2", "distance")) %>%
      dplyr::right_join(colData_frame,
                        by = dplyr::join_by("Sample1" == "Sample")) %>%
      dplyr::right_join(colData_frame,
                        by = dplyr::join_by("Sample2" == "Sample")) %>%
      stats::setNames(c("Sample1", "Sample2", "distance",
                        "Group1", "Group2")) %>%
      dplyr::mutate(comparison = paste(Group1, Group2, sep = " - "))

    comp_levels <- NULL
    for (level_1 in 1:(length(set_levels[[1]]) - 1)) {
      lev_left <- set_levels[[1]][level_1:length(set_levels[[1]])]
      for (level_2 in 2:length(lev_left)) {
        comp <- paste(set_levels[[1]][level_1], lev_left[level_2], sep = " - ")
        comp_levels <- append(comp_levels, comp)
      }
    }

    BD_dissim <- BD_dissim[BD_dissim$comparison %in% comp_levels, ]
    BD_dissim$comparison <- factor(BD_dissim$comparison, levels = comp_levels)

    if (length(unique(BD_dissim$comparison)) <= 5) {
      plot_colors <- rev(plot_palette)[seq(2, 10, 2)][1:length(unique(BD_dissim$comparison))]
      color_scale_dis <- ggplot2::scale_color_manual(values = plot_colors)
    } else if (length(unique(BD_dissim$comparison)) <= 10) {
      plot_colors <- rev(plot_palette)[sort(
        c(seq(1, 9, 2)[1:(length(unique(BD_dissim$comparison)) - 5)],
          seq(2, 10, 2))
      )]
      color_scale_dis <- ggplot2::scale_color_manual(values = plot_colors)
    } else if (length(unique(BD_dissim$comparison)) <= 20) {
      plot_colors <- rev(plot_palette)[sort(
        c(seq(1, 19, 2)[1:(length(unique(BD_dissim$comparison)) - 10)],
          seq(2, 20, 2))
      )]
      color_scale_dis <- ggplot2::scale_color_manual(values = plot_colors)
    } else {
      color_scale_dis <- ggraph::scale_color_viridis(discrete = TRUE)
    }

    BD_dis_box <- ggplot2::ggplot(
      BD_dissim,
      ggplot2::aes(x = comparison, y = distance)
    ) +
      ggplot2::geom_boxplot(lwd = 0.65, show.legend = FALSE, outlier.shape = NA) +
      ggplot2::geom_jitter(
        ggplot2::aes(color = comparison),
        position = ggplot2::position_jitterdodge(0.5),
        cex = 0.1,
        show.legend = FALSE
      ) +
      color_scale_dis +
      ggplot2::theme_bw() +
      ggplot2::labs(color = var[1]) +
      ggplot2::theme(
        plot.title = ggplot2::element_text(size = 17),
        axis.title.y = ggplot2::element_text(color = "Grey1", size = 14),
        axis.title.x = ggplot2::element_blank(),
        axis.text.x  = ggplot2::element_text(color = "Grey1", size = 9,
                                             angle = 45, hjust = 1),
        axis.text.y  = ggplot2::element_text(color = "Grey1", size = 12),
        panel.grid   = ggplot2::element_blank()
      ) +
      ggpubr::geom_pwc(
        label = "p.adj.signif",
        p.adjust.method = "BH",
        hide.ns = hide.ns
      ) +
      ggplot2::ylab(paste("Dissimilarity", stringr::str_to_title(method)))

    wrap_BD_plot <- patchwork::wrap_plots(BD_ord_plot, BD_dis_box,
                                          ncol = 2, widths = c(5, 4))
    print(wrap_BD_plot)

    BD_output <- list(
      BD_ord_plot,
      BD_dis_box,
      permanova,
      permutest,
      BD_dissim,
      wrap_BD_plot
    )
    names(BD_output) <- c(
      "BD_plot",
      "BD_dissimilarity_box_plot",
      "PERMANOVA_results",
      "Permutest_results",
      "Dissimilarity_dataframe",
      "BD_wrap_plots"
    )

  } else {

    print(BD_ord_plot)
    BD_output <- list(BD_ord_plot, permanova, permutest)
    names(BD_output) <- c(
      "BD_plot",
      "PERMANOVA_results",
      "Permutest_results"
    )
  }

  invisible(BD_output)
}
```

#### Feces at T0 and T12

```{r Beta diversity paired CSV Feces at T0 and T12, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
BD_plot_cf_fc_csv <- BD_plot_paired_csv(
  tse = M_TSE_cf_UN,
  var = 'Time',
  var_levels = c('T0','T12'),
  level = 'Species',
  method = 'bray',
  weighted_UniFrac = TRUE,
  ordination = 'PCoA',
  constraining = var,
  plot_title = 'Beta diversity of Feces',
  dissimilarity_plot = FALSE,
  hide.ns = FALSE,
  export_prefix = "beta_feces_species_bray"
)

ggsave(filename = here("results", "BD_plot_cf_fc_csv.svg"), plot = BD_plot_cf_fc_csv$BD_plot, width = 10, height = 7, dpi = 300, device = "svg")
```

##### Plots

```{r Paper plot BD feces, echo=TRUE, fig.show='show', message=FALSE, warning=FALSE}
# ==== files prefixes ====
prefix <- "beta_feces_species_bray"
# =============================

# 1) Load data
coords <- read_csv(here("results", paste0(prefix, "_coords.csv"))) %>%
  mutate(Group = factor(Group, levels = c("T0","T12")))

perm <- read_csv(here("results", paste0(prefix, "_permanova.csv")))

# Helpers to extract numbers or text
get_num1 <- function(x) {
  if (is.numeric(x)) x[1] else readr::parse_number(as.character(x[1]))
}

# R2 and p
R2_val <- signif(get_num1(perm$R2), 3)

p_col  <- names(perm)[grepl("Pr\\(>F\\)|Pr|p", names(perm), ignore.case = TRUE)][1]
p_num  <- get_num1(perm[[p_col]])
p_txt  <- if (is.na(p_num)) "NA" else if (p_num < 1e-3) "< 0.001" else signif(p_num, 3)

cap_text <- paste0("R\u00B2 = ", R2_val, "   •   Pr(>F) = ", p_txt)

# % of axis
axes_file <- here("results", paste0(prefix, "_axes.csv"))
if (file.exists(axes_file)) {
  ax <- read_csv(axes_file)
  ax1 <- ax$Percent[ax$Axis %in% c("PCoA1","PC1")][1]
  ax2 <- ax$Percent[ax$Axis %in% c("PCoA2","PC2")][1]
  x_lab <- paste0("PCoA1 (", ax1, "%)")
  y_lab <- paste0("PCoA2 (", ax2, "%)")
} else {
  x_lab <- "PCoA1"; y_lab <- "PCoA2"
}

# Colours
col_ellipse <- c("T0" = "#F4E1A1", "T12" = "#A1E1E1")
col_point   <- c("T0" = "#FAF3DC", "T12" = "#D6F5F5")

# 3) Plot
p_beta_fc <- ggplot(coords, aes(Axis1, Axis2)) +
  stat_ellipse(aes(fill = Group, color = Group),
               geom = "polygon", level = 0.95,
               alpha = 0.35, linewidth = 0.6, show.legend = TRUE) +
  geom_point(
    data = coords %>% filter(Group == "T0"),
    shape = 21, size = 3, stroke = 0.3, color = "black",
    fill  = col_point["T0"]
  ) +
  geom_point(
    data = coords %>% filter(Group == "T12"),
    shape = 21, size = 3, stroke = 0.3, color = "black",
    fill  = col_point["T12"]
  ) +
  scale_fill_manual(values = col_ellipse, name = "Time") +
  scale_color_manual(values = col_ellipse, guide = "none") +
  guides(fill = guide_legend(override.aes = list(alpha = 0.35, linewidth = 0.6, color = col_ellipse))) +
  labs(
    x = x_lab, y = y_lab,
    title = "Beta diversity (PCoA) - Feces",
    caption = cap_text
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor      = element_blank(),
    legend.position       = "right",
    plot.title            = element_text(size = 14),
    plot.caption.position = "plot",
    plot.caption          = element_text(hjust = 0, size = 11)
  )

p_beta_fc

# 4) Save SVG
ggsave(here::here("results", "beta_diversity_pcoa_feces.svg"),
       p_beta_fc, width = 8, height = 6, units = "in")
```


# ------------------------------------------------------------------------------
# DIFFERENTIAL ABUNDANCE ANALYSIS (DDA)
# ------------------------------------------------------------------------------

A modification for the BD_plot() function from MADAV package was used in order to enable the paired analysis and export the results.

Differential abundance analysis used:

-   ANCOM-BC

#### Feces at T0 and T12

##### Run ANCOM-BC

```{r DAA ANCOMBC2 of Feces at T0 and T12 min_prevalence 0.10, results='hide', echo=TRUE, warning=FALSE, message=FALSE}
DAA_Metaphlan_cf_fc_abc <- DA_analysis(tse = M_TSE_cf_UN,
                               assay_type = 'counts',
                               var = 'Time',
                               main_var = NULL,
                               var_levels = c('T0','T12'),
                               pairwise = T,
                               level = 'Species',
                               min_prevalence = 0.10,
                               min_abundance = 0,
                               padj_method = 'BH',
                               padj_m = 'default',
                               rand_formula = NULL,
                               tools = 'ANCOM_BC2',
                               sign_cut = 0.05,
                               effect_cut = 0,
                               n_cores = 1,
                               verbose = T,
                               struc_zero = F,
                               neg_lb = F,
                               pseudo_sens = T,
                               s0_perc = 0.05)
```

##### Print the significant features

```{r Print significant features of DAA_Metaphlan_cf_fc, results='show', echo=TRUE, warning=FALSE, message=FALSE}
print(DAA_Metaphlan_cf_fc_abc$sign_features)
```

##### Save the results

```{r Save results feces T0 and T12, warning=FALSE, message=FALSE}
write.csv(DAA_Metaphlan_cf_fc_abc$sign_features$ANCOM_BC2,
          file = here("results", "DAA_ANCOMBC2_sign_species_feces.csv"),
          row.names = FALSE)
```

##### Bar plot

```{r Bar plot paper significat results feces, echo=TRUE, fig.show='show', warning=FALSE, message=FALSE}
sign_df_fc <- DAA_Metaphlan_cf_fc_abc$sign_features$ANCOM_BC2

DAA_Metaphlan_cf_fc_abc_bar <- ggplot(sign_df_fc, 
    aes(x = reorder(featureID, effect), y = effect, fill = effect)) +
  geom_bar(stat = "identity", color = "black", size = 0.2) +  
  coord_flip() +
  scale_fill_gradient2(low = "#F4E1A1", mid = "white", high = "#A1E1E1", midpoint = 0) +
  labs(title = "Differentially abundant species at T12 vs T0",
       x = "",
       y = "Effect size (log fold change)") +
  theme_minimal()

DAA_Metaphlan_cf_fc_abc_bar

ggsave(filename = here("results", "DAA_Metaphlan_cf_fc_abc_bar.svg"), plot = DAA_Metaphlan_cf_fc_abc_bar, width = 10, height = 7, dpi = 300, device = "svg")
```


#-------------------------------------
#### Relative abundance plot

A modification for the RA_plot() function from MADAV package was used in order to export the results.

##### Modified function of RA_plot {MADAV}: RA_plot_csv

```{r RA_plot MODIFIED for CSV, message=FALSE, warning=FALSE}
RA_plot_csv <- function (features, tse, var, var_levels = NULL, level = NULL, 
                         total_tools = 0, plot_title = "Relative Abundance", 
                         hide.ns = FALSE, panel_dim = c(4, 4),
                         export_csv = NULL) {  # NEW: optional CSV export path
  
  `%>%` <- magrittr::`%>%`
  RAP <- tse
  
  if (is.null(level)) {
    level <- mia::taxonomyRanks(tse)[length(mia::taxonomyRanks(tse))]
    warning(paste("No level was specified, DA analysis was performed on", level, "level"))
  }
  
  if (level %in% colnames(SummarizedExperiment::rowData(RAP)) & !(level %in% mia::taxonomyRanks(RAP))) {
    if ("domain" %in% colnames(SummarizedExperiment::rowData(RAP))) {
      colnames(SummarizedExperiment::rowData(RAP))[colnames(SummarizedExperiment::rowData(RAP)) == "domain"] <- "true_domain_column"
    }
    colnames(SummarizedExperiment::rowData(RAP))[colnames(SummarizedExperiment::rowData(RAP)) == level] <- "domain"
    rank <- "domain"
  } else if (level %in% mia::taxonomyRanks(RAP)) {
    rank <- level
  }
  
  RAP_agg <- mia::transformAssay(RAP, method = "relabundance") %>% 
    mia::agglomerateByRank(rank, onRankOnly = TRUE, na.rm = TRUE)
  
  if (!is.null(attr(features, "var"))) {
    feature_frame <- features$comparison[features$comparison$total_tools >= total_tools, c("featureID", "var")]
    if (nrow(feature_frame) != 0) {
      feature_frame <- feature_frame[!(duplicated(feature_frame)), ] %>% 
        dplyr::mutate(var = paste("sign", var, sep = "|"))
      feature_frame <- tidyr::pivot_wider(feature_frame, names_from = "var", values_from = "var")
      feature_frame[, 2:length(feature_frame)] <- apply(feature_frame[, 2:length(feature_frame)], 2, function(x) ifelse(is.na(x), FALSE, TRUE))
    } else {
      stop("No features were detected by the selected number of total_tools")
    }
  } else {
    feature_frame <- data.frame(matrix(ncol = length(var) + 1, nrow = length(features))) %>% 
      stats::setNames(c("featureID", paste("sign", var, sep = "|")))
    feature_frame$featureID <- features
    feature_frame[, paste("sign", var, sep = "|")] <- TRUE
  }
  
  if (all(feature_frame$featureID %in% rownames(RAP_agg))) {
    RAP_sel <- RAP_agg[rownames(RAP_agg) %in% feature_frame$featureID, ]
  } else {
    stop("Not all features are present in the tse object at the selected level")
  }
  
  RAP_meta <- tibble::rownames_to_column(data.frame(SummarizedExperiment::colData(RAP_sel)), var = "sampleID") %>% 
    dplyr::select(dplyr::all_of(c("sampleID", var)))
  
  RAP_melt <- as.data.frame(SummarizedExperiment::assays(RAP_sel)[["relabundance"]]) %>% 
    tibble::rownames_to_column(var = "featureID") %>% 
    reshape2::melt(id = "featureID") %>% 
    stats::setNames(c("featureID", "sampleID", "abundance")) %>% 
    dplyr::right_join(RAP_meta, by = "sampleID") %>% 
    dplyr::right_join(feature_frame, by = "featureID")
  
  # NEW: export relative abundance table to CSV (long format)
  if (!is.null(export_csv)) {
    dir.create(dirname(export_csv), recursive = TRUE, showWarnings = FALSE)
    
    rap_export <- as.data.frame(RAP_melt)
    first_cols <- c("featureID", "sampleID", "abundance", var)
    first_cols <- first_cols[first_cols %in% colnames(rap_export)]
    rap_export <- rap_export[, c(first_cols, setdiff(colnames(rap_export), first_cols)), drop = FALSE]
    
    names(rap_export)[names(rap_export) == "featureID"] <- "FeatureID"
    names(rap_export)[names(rap_export) == "sampleID"]  <- "SampleID"
    names(rap_export)[names(rap_export) == "abundance"] <- "RelAbundance"
    
    readr::write_csv(rap_export, export_csv)
    message("Relative abundance exported to: ", export_csv)
  }
  
  feature_order <- data.frame(SummarizedExperiment::assays(RAP_sel)[["relabundance"]]) %>% 
    rowSums() %>% sort(decreasing = TRUE) %>% names()
  RAP_melt$featureID <- factor(RAP_melt$featureID, levels = feature_order)
  
  RAP_out_list <- list()
  RAP_mean <- data.frame(feature_order) %>% stats::setNames("featureID")
  
  if (length(var) == 1) {
    var_levels <- list(var_levels)
  }
  
  for (cov in rev(var)) {
    if (paste("sign", cov, sep = "|") %in% colnames(RAP_melt)) {
      
      RAP_melt_sel <- RAP_melt[RAP_melt[, paste("sign", cov, sep = "|")] == TRUE, ] %>% 
        dplyr::select(all_of(c("featureID", "sampleID", "abundance", cov))) %>% 
        stats::setNames(c("featureID", "sampleID", "abundance", "variable"))
      
      RAP_cov_list <- list()
      
      if (is.character(RAP_melt_sel$variable) | is.factor(RAP_melt_sel$variable)) {
        RAP_melt_cov <- RAP_melt_sel
        
        if (!is.null(var_levels[[which(var == cov)]])) {
          RAP_melt_cov$variable <- factor(RAP_melt_cov$variable, levels = var_levels[[which(var == cov)]])
        } else {
          RAP_melt_cov$variable <- factor(RAP_melt_cov$variable)
        }
        
        plot_palette <- c("#1F77B4FF", "#AEC7E8FF", "#FF7F0EFF", 
                          "#FFBB78FF", "#2CA02CFF", "#98DF8AFF", "#D62728FF", 
                          "#FF9896FF", "#9467BDFF", "#C5B0D5FF", "#8C564BFF", 
                          "#C49C94FF", "#E377C2FF", "#F7B6D2FF", "#7F7F7FFF", 
                          "#C7C7C7FF", "#BCBD22FF", "#DBDB8DFF", "#17BECFFF", 
                          "#9EDAE5FF")
        
        if (length(levels(RAP_melt_cov$variable)) <= 10) {
          plot_colors <- plot_palette[seq(1, 19, 2)][1:length(levels(RAP_melt_cov$variable))]
          color_scale <- ggplot2::scale_color_manual(values = plot_colors)
        } else if (length(levels(RAP_melt_cov$variable)) <= 20) {
          # NOTE: kept logic; uses levels(variable) instead of undefined set_levels
          plot_colors <- plot_palette[sort(c(seq(2, 20, 2)[1:(length(levels(RAP_melt_cov$variable)) - 10)], seq(1, 19, 2)))]
          color_scale <- ggplot2::scale_color_manual(values = plot_colors)
        } else {
          color_scale <- ggraph::scale_color_viridis(discrete = TRUE)
        }
        
      } else {
        RAP_melt_cov <- suppressMessages(
          RAP_melt_sel %>% dplyr::group_by(featureID, variable) %>% 
            dplyr::summarise(mean = mean(abundance), std = plotrix::std.error(abundance))
        )
        
        lm_list <- lapply(unique(RAP_melt_cov$featureID), function(x) {
          summary(stats::lm(as.data.frame(RAP_melt_cov[RAP_melt_cov$featureID == x, ])$mean ~ 
                              as.data.frame(RAP_melt_cov[RAP_melt_cov$featureID == x, ])$variable))
        })
        
        RAP_lm_stats <- data.frame(
          as.character(unique(RAP_melt_cov$featureID)),
          unlist(lapply(lm_list, function(x) paste("R2:", signif(x$adj.r.squared, digits = 2)))),
          unlist(lapply(lm_list, function(x) paste("Pr(>|t|):", signif(x$coefficients[2, 4], digits = 2))))
        ) %>% stats::setNames(c("featureID", "R2", "Pr(>|t|)"))
      }
      
      for (part in (ceiling(length(unique(RAP_melt_sel$featureID)) / prod(panel_dim))):1) {
        feature_range <- (prod(panel_dim) * part - prod(panel_dim) + 1):(prod(panel_dim) * part)
        feature_sel <- unique(RAP_melt_sel$featureID)[feature_range]
        RAP_melt_panel <- RAP_melt_cov[RAP_melt_cov$featureID %in% feature_sel, ]
        
        if (is.numeric(RAP_melt_sel$variable) | is.double(RAP_melt_sel$variable)) {
          RAP_lm_stats_sel <- RAP_lm_stats[RAP_lm_stats$featureID %in% feature_sel, ]
        }
        
        if (length(unique(RAP_melt_sel$featureID)) > prod(panel_dim)) {
          plot_title_panel <- paste(plot_title, ": ", cov, " (", part, ")", sep = "")
        } else {
          plot_title_panel <- paste(plot_title, ": ", cov, sep = "")
        }
        
        if (is.factor(RAP_melt_cov$variable)) {
          RAP_out_panel <- ggplot2::ggplot(RAP_melt_panel, ggplot2::aes(x = variable, y = abundance)) +
            ggplot2::geom_boxplot(lwd = 0.5, show.legend = FALSE, outlier.shape = NA) +
            ggplot2::geom_jitter(ggplot2::aes(color = variable), size = 0.65,
                                 position = ggplot2::position_jitterdodge(0.2),
                                 alpha = 0.75, show.legend = FALSE) +
            ggpubr::geom_pwc(label = "p.adj.signif", p.adjust.method = "BH", hide.ns = hide.ns) +
            color_scale
        } else {
          RAP_out_panel <- ggplot2::ggplot(RAP_melt_panel, ggplot2::aes(x = variable, y = mean)) +
            ggplot2::geom_point(size = 2, color = "#26456E") +
            ggplot2::geom_line(linetype = 2, color = "#26456E") +
            ggplot2::geom_errorbar(ggplot2::aes(ymin = mean - std, ymax = mean + std), width = 0.2, color = "grey") +
            ggplot2::geom_text(data = RAP_lm_stats_sel, ggplot2::aes(x = -Inf, y = Inf, label = R2),
                               size = 2.5, vjust = 1.2, hjust = -0.1) +
            ggplot2::geom_text(data = RAP_lm_stats_sel, ggplot2::aes(x = -Inf, y = Inf, label = `Pr(>|t|)`),
                               size = 2.5, vjust = 2.4, hjust = -0.07)
        }
        
        RAP_out_panel <- RAP_out_panel +
          ggplot2::labs(x = "", y = "Relative abundance", title = plot_title_panel, fill = cov) +
          ggplot2::theme_bw() +
          ggplot2::theme(plot.title = ggplot2::element_text(size = 20, face = "plain"),
                         axis.title.y = ggplot2::element_text(size = 15, face = "plain"),
                         strip.text.x = ggplot2::element_text(size = 9, face = "italic"),
                         panel.grid = ggplot2::element_blank()) +
          ggplot2::facet_wrap(~featureID, scales = "free", ncol = panel_dim[2])
        
        RAP_cov_list <- append(RAP_cov_list, list(RAP_out_panel))
        names(RAP_cov_list)[length(RAP_cov_list)] <- paste(cov, part, sep = "_")
      }
      
      print(RAP_cov_list)
      
      for (feat in unique(RAP_melt_sel$featureID)) {
        RAP_melt_feat <- RAP_melt_cov[RAP_melt_cov$featureID == feat, ]
        
        if (is.numeric(RAP_melt_sel$variable) | is.double(RAP_melt_sel$variable)) {
          RAP_lm_stats_feat <- RAP_lm_stats[RAP_lm_stats$featureID == feat, ]
        }
        
        if (is.factor(RAP_melt_cov$variable)) {
          RAP_out_feat <- ggplot2::ggplot(RAP_melt_feat, ggplot2::aes(x = variable, y = abundance)) +
            ggplot2::geom_boxplot(lwd = 0.65, show.legend = FALSE, outlier.shape = NA) +
            ggplot2::geom_jitter(ggplot2::aes(color = variable),
                                 position = ggplot2::position_jitterdodge(0.2),
                                 alpha = 0.75, show.legend = FALSE) +
            ggpubr::geom_pwc(label = "p.adj.signif", p.adjust.method = "BH",
                             hide.ns = hide.ns, label.size = 2.8) +
            color_scale
        } else {
          RAP_out_feat <- ggplot2::ggplot(RAP_melt_feat, ggplot2::aes(x = variable, y = mean)) +
            ggplot2::geom_point(size = 2, color = "#26456E") +
            ggplot2::geom_line(linetype = 2, color = "#26456E") +
            ggplot2::geom_errorbar(ggplot2::aes(ymin = mean - std, ymax = mean + std), width = 0.2, color = "grey") +
            ggplot2::geom_text(data = RAP_lm_stats_feat, ggplot2::aes(x = -Inf, y = Inf, label = R2),
                               size = 4, vjust = 1.2, hjust = -0.1) +
            ggplot2::geom_text(data = RAP_lm_stats_feat, ggplot2::aes(x = -Inf, y = Inf, label = `Pr(>|t|)`),
                               size = 4, vjust = 2.4, hjust = -0.07)
        }
        
        RAP_out_feat <- RAP_out_feat +
          ggplot2::labs(x = "", y = "", fill = cov, title = feat) +
          ggplot2::theme_bw() +
          ggplot2::theme(plot.title = ggplot2::element_text(size = 13, face = "italic"),
                         axis.title.y = ggplot2::element_text(size = 12, face = "plain"),
                         strip.text.x = ggplot2::element_text(size = 9, face = "italic"),
                         panel.grid = ggplot2::element_blank())
        
        RAP_cov_list$features <- append(RAP_cov_list$features, list(RAP_out_feat))
        names(RAP_cov_list$features)[length(RAP_cov_list$features)] <- feat
      }
      
      RAP_out_list <- append(RAP_out_list, list(RAP_cov_list))
      names(RAP_out_list)[length(RAP_out_list)] <- cov
      
      RAP_mean_add <- stats::aggregate(stats::as.formula(paste("abundance ~ featureID +", cov)), RAP_melt, mean) %>% 
        tidyr::pivot_wider(names_from = dplyr::all_of(cov), values_from = abundance)
      colnames(RAP_mean_add) <- c("featureID", paste(cov, colnames(RAP_mean_add)[-1], sep = ""))
      RAP_mean <- dplyr::right_join(RAP_mean, RAP_mean_add, by = "featureID")
      
    } else {
      warning(paste("No significant comparisons were identified in the DA_analysis object for selected conditions for variable:", cov))
    }
  }
  
  RAP_out_list <- append(RAP_out_list, list(RAP_mean))
  names(RAP_out_list)[length(RAP_out_list)] <- "mean_relabundance"
  
  return(invisible(RAP_out_list))
}
```

###### Export CSV

```{r RA CSV of DAA of Feces at T0 and T12, echo=TRUE, fig.show='hide', warning=FALSE, message=FALSE}
RAP_out <- RA_plot_csv(
  features = DAA_Metaphlan_cf_fc_abc,
  tse = M_TSE_cf_UN,
  var = c("Time"),
  var_levels = c("T0","T12"),
  level = "Species",
  plot_title = "Relative Abundance",
  export_csv = here::here("results", "RA_plot_relabundance_long.csv")
)
```

#### Function to export p- values and p-adjusted values from RA
```{r Function to export p values and p adj values from RA, echo=TRUE, warning=FALSE, message=FALSE}
RA_padj_from_RA <- function(ra_long,
                            feature_col  = "FeatureID",
                            value_col    = "RelAbundance",
                            group_col    = "Time",
                            group_levels = c("T0","T12"),
                            paired       = FALSE,
                            id_col       = "PatientID",
                            padj_method  = "BH") {

  df <- ra_long %>%
    dplyr::mutate(
      .feature = .data[[feature_col]],
      .value   = .data[[value_col]],
      .group   = factor(.data[[group_col]], levels = group_levels)
    ) %>%
    dplyr::filter(!is.na(.group), !is.na(.value))

  if (paired) {
    if (!(id_col %in% colnames(df))) {
      stop(sprintf("paired=TRUE: la columna '%s' no existe en ra_long", id_col))
    }

    df <- df %>%
      dplyr::group_by(.feature, .data[[id_col]], .group) %>%
      dplyr::summarise(.value = stats::median(.value, na.rm = TRUE), .groups = "drop") %>%
      dplyr::group_by(.feature, .data[[id_col]]) %>%
      dplyr::filter(all(group_levels %in% .group)) %>%
      dplyr::ungroup()

    out <- df %>%
      tidyr::pivot_wider(names_from = .group, values_from = .value) %>%
      dplyr::group_by(.feature) %>%
      dplyr::summarise(
        n_pairs = sum(!is.na(.data[[group_levels[1]]]) & !is.na(.data[[group_levels[2]]])),
        n_T0  = n_pairs,
        n_T12 = n_pairs,
        p = tryCatch(
          stats::wilcox.test(
            .data[[group_levels[1]]],
            .data[[group_levels[2]]],
            paired = TRUE,
            exact = FALSE
          )$p.value,
          error = function(e) NA_real_
        ),
        .groups = "drop"
      )

  } else {

    out <- df %>%
      dplyr::group_by(.feature) %>%
      dplyr::summarise(
        n_T0  = sum(.group == group_levels[1]),
        n_T12 = sum(.group == group_levels[2]),
        p = tryCatch(
          stats::wilcox.test(.value ~ .group, paired = FALSE, exact = FALSE)$p.value,
          error = function(e) NA_real_
        ),
        .groups = "drop"
      )
  }

  out %>%
    dplyr::mutate(
      padj = stats::p.adjust(p, method = padj_method),
      padj_signif = dplyr::case_when(
        is.na(padj)      ~ NA_character_,
        padj <= 0.0001   ~ "****",
        padj <= 0.001    ~ "***",
        padj <= 0.01     ~ "**",
        padj <= 0.05     ~ "*",
        TRUE             ~ "ns"
      )
    ) %>%
    dplyr::rename(FeatureID = .feature)
}
```

##### Run function to export p-values and p-adjusted values from RA PAIRED

```{r Run function to export pval and p adj values from RA PAIRED, echo=TRUE, warning=FALSE, message=FALSE}
ra_df <- read_csv(here::here("results","RA_plot_relabundance_long.csv"), show_col_types = FALSE) %>%
  mutate(
    PatientID = str_extract(SampleID, "^[A-Za-z]+\\d+")
  )

ra_padj <- RA_padj_from_RA(
  ra_long = ra_df,
  paired = TRUE
)

write_csv(ra_padj, here::here("results","RA_padj_from_RA_Time.csv"))
```

### Plots

```{r Plot of RA of DAA of Feces at T0 and T12, echo=TRUE, fig.show='hide', warning=FALSE, message=FALSE}
# Load ANCOM-BC2 significant features
ancom_tab <- read_csv(here::here("results", "DAA_ANCOMBC2_sign_species_feces.csv"),
                      show_col_types = FALSE) %>%
  rename(FeatureID = featureID)

sig_feats <- ancom_tab %>%
  filter(padj < 0.05) %>%
  pull(FeatureID) %>%
  unique()

# Load RA long CSV
ra_df <- read_csv(here::here("results", "RA_plot_relabundance_long.csv"),
                  show_col_types = FALSE) %>%
  mutate(
    PatientID = str_extract(SampleID, "^[A-Za-z]+\\d+"),
    Time = factor(Time, levels = c("T0", "T12"))
  )

# Keep only sig features + paired patients (T0 and T12)
ra_sig <- ra_df %>%
  filter(FeatureID %in% sig_feats) %>%
  group_by(FeatureID, PatientID) %>%
  filter(n_distinct(Time) == 2) %>%
  ungroup()

# ---- Compute Wilcoxon paired p-values on Relative Abundance (for labels) ----
ra_pvals <- ra_sig %>%
  group_by(FeatureID, PatientID, Time) %>%
  summarise(RelAbundance = median(RelAbundance, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Time, values_from = RelAbundance) %>%
  group_by(FeatureID) %>%
  summarise(
    p_RA = tryCatch(
      wilcox.test(T0, T12, paired = TRUE, exact = FALSE)$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  )

# Colors
col_box   <- c("T0" = "#F4E1A1", "T12" = "#A1E1E1")
col_point <- c("T0" = "#FAF3DC", "T12" = "#D6F5F5")

# Paired plot for ONE feature
plot_ra_feature <- function(feat) {

  df <- ra_sig %>%
    filter(FeatureID == feat) %>%
    mutate(
      fill_box   = col_box[Time],
      fill_point = col_point[Time],
      x          = as.numeric(Time)
    )

  # ---- Label using RA Wilcoxon paired p-value ----
  p_val <- ra_pvals %>%
    filter(FeatureID == feat) %>%
    slice(1) %>%
    pull(p_RA)

  lab <- ifelse(is.na(p_val), "Wilcoxon p = NA",
                ifelse(p_val < 0.001, "Wilcoxon p < 0.001",
                       paste0("p = ", signif(p_val, 3))))

  y_max <- max(df$RelAbundance, na.rm = TRUE)
  if (!is.finite(y_max) || y_max == 0) y_max <- 1e-6

  y_bracket <- y_max * 1.15
  y_text    <- y_max * 1.22
  y_lim     <- y_max * 1.30

  p_values <- data.frame(x = 1, xend = 2, y = y_bracket, label = lab)

  ggplot() +
    stat_slab(
      data = df %>% filter(Time == "T0"),
      aes(x = x, y = RelAbundance, fill = fill_box),
      side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    stat_slab(
      data = df %>% filter(Time == "T12"),
      aes(x = x, y = RelAbundance, fill = fill_box),
      side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    geom_boxplot(
      data = df,
      aes(x = x, y = RelAbundance, fill = fill_box),
      width = 0.5, outlier.shape = NA,
      color = "black", alpha = 0.9, show.legend = TRUE
    ) +
    geom_point(
      data = df,
      aes(x = x, y = RelAbundance, fill = fill_point),
      shape = 21, size = 3, stroke = 0.3, color = "black",
      show.legend = FALSE
    ) +
    geom_line(
      data = df,
      aes(x = x, y = RelAbundance, group = PatientID),
      color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
    ) +
    geom_segment(
      data = p_values,
      aes(x = x, xend = xend, y = y, yend = y),
      color = "black", linewidth = 0.5
    ) +
    geom_text(
      data = p_values,
      aes(x = 1.5, y = y_text, label = label),
      size = 4
    ) +
    scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
    scale_fill_identity(
      guide = "legend", name = "Time",
      breaks = col_box, labels = names(col_box)
    ) +
    coord_cartesian(ylim = c(0, y_lim)) +
    labs(x = NULL, y = "Relative abundance", title = feat) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "right",
      axis.text.x = element_text(size = 14)
    )
}

# Save ONLY the ANCOM-BC2 significant features into results/
for (feat in sig_feats) {
  p <- plot_ra_feature(feat)

  out_file <- here::here(
    "results",
    paste0("RA_", gsub("[^A-Za-z0-9]+", "_", feat), ".svg")
  )

  ggsave(out_file, plot = p, width = 8, height = 6, dpi = 300, device = "svg")
  message("Saved: ", out_file)
}
```

##### Only one microorganism

```{r RA function of DAA at T0 and T12, echo=TRUE, fig.show='hide', warning=FALSE, message=FALSE}
RA_one_microbe <- function(tse,
                           microbe,
                           var = "Time",
                           var_levels = c("T0","T12"),
                           level = "Species",
                           subset_var = NULL,
                           subset_levels = NULL,
                           export_csv = NULL,
                           save_svg = NULL,
                           w = 8, h = 6) {

  `%>%` <- magrittr::`%>%`

  # Rel. abundance + agglomerate
  rank <- level
  RA_agg <- mia::transformAssay(tse, method = "relabundance") %>%
    mia::agglomerateByRank(rank, onRankOnly = TRUE, na.rm = TRUE)

  # Metadata + SampleID
  cd <- as.data.frame(SummarizedExperiment::colData(RA_agg))
  cd$SampleID <- rownames(cd)

  # Optional subset
  if (!is.null(subset_var) && !is.null(subset_levels)) {
    keep <- cd[[subset_var]] %in% subset_levels
    RA_agg <- RA_agg[, keep]
    cd <- cd[keep, , drop = FALSE]
  }

  # Robust feature matching (spaces/underscores/case)
  feat_names <- rownames(RA_agg)
  norm <- function(x) tolower(gsub("[ _]", "", x))
  idx <- which(norm(feat_names) == norm(microbe))

  if (length(idx) == 0) {
    stop("Microbe not found at level = ", level, ": ", microbe)
  }

  # Build tidy table for the microbe
  ra_vec <- as.numeric(SummarizedExperiment::assay(RA_agg, "relabundance")[idx[1], ])

  df <- dplyr::tibble(
    FeatureID = feat_names[idx[1]],
    SampleID = cd$SampleID,
    Time = cd[[var]],
    RelAbundance = ra_vec
  ) %>%
    dplyr::mutate(
      # PatientID from SampleID (e.g., P2H0 -> P2; A10E12 -> A10)
      PatientID = stringr::str_extract(SampleID, "^[A-Za-z]+\\d+"),
      Time = factor(Time, levels = var_levels)
    ) %>%
    dplyr::group_by(PatientID) %>%
    dplyr::filter(dplyr::n_distinct(Time) == 2) %>%
    dplyr::ungroup()

  if (nrow(df) == 0) {
    stop("No paired T0/T12 data found for: ", microbe, " (after filtering/subsetting).")
  }

  # Optional export values used in the plot
  if (!is.null(export_csv)) {
    readr::write_csv(df, export_csv)
    message("CSV exported to: ", export_csv)
  }

  # Colors
  col_box   <- c("T0" = "#F4E1A1", "T12" = "#A1E1E1")
  col_point <- c("T0" = "#FAF3DC", "T12" = "#D6F5F5")

  df <- df %>%
    dplyr::mutate(
      fill_box   = col_box[Time],
      fill_point = col_point[Time],
      x          = as.numeric(Time)
    )

  # --- per-microbe y scaling ---
  y_max <- max(df$RelAbundance, na.rm = TRUE)
  if (!is.finite(y_max) || y_max == 0) y_max <- 1e-6
  y_lim <- y_max * 1.30

  # --- fancy plot ---
  p <- ggplot2::ggplot() +
    ggdist::stat_slab(
      data = df %>% dplyr::filter(Time == "T0"),
      ggplot2::aes(x = x, y = RelAbundance, fill = fill_box),
      side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    ggdist::stat_slab(
      data = df %>% dplyr::filter(Time == "T12"),
      ggplot2::aes(x = x, y = RelAbundance, fill = fill_box),
      side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    ggplot2::geom_boxplot(
      data = df,
      ggplot2::aes(x = x, y = RelAbundance, fill = fill_box),
      width = 0.5, outlier.shape = NA,
      color = "black", alpha = 0.9, show.legend = TRUE
    ) +
    ggplot2::geom_point(
      data = df,
      ggplot2::aes(x = x, y = RelAbundance, fill = fill_point),
      shape = 21, size = 3, stroke = 0.3, color = "black",
      show.legend = FALSE
    ) +
    ggplot2::geom_line(
      data = df,
      ggplot2::aes(x = x, y = RelAbundance, group = PatientID),
      color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
    ) +
    ggplot2::scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
    ggplot2::scale_fill_identity(
      guide = "legend", name = "Time",
      breaks = col_box, labels = names(col_box)
    ) +
    ggplot2::coord_cartesian(ylim = c(0, y_lim)) +
    ggplot2::labs(x = NULL, y = "Relative abundance", title = microbe) +
    ggplot2::theme_minimal(base_size = 14) +
    ggplot2::theme(
      legend.position = "right",
      axis.text.x = ggplot2::element_text(size = 14)
    )

  if (!is.null(save_svg)) {
    ggplot2::ggsave(save_svg, plot = p, width = w, height = h, dpi = 300, device = "svg")
    message("SVG saved to: ", save_svg)
  }

  return(p)
}
```

###### Microorganism 1

```{r RA plot only one microorganism at T0 and T12, echo=TRUE, fig.show='show', warning=FALSE, message=FALSE}
p_Aliq <- RA_one_microbe(
  tse = M_TSE_cf_UN,
  microbe = "Adlercreutzia_equolifaciens", # Change here the microorganism
  var = "Time",
  var_levels = c("T0","T12"),
  level = "Species",
  subset_var = NULL,
  subset_levels = NULL,
  save_svg = here::here("results", "RA_Adlercreutzia_equolifaciens.svg")
)
p_Aliq
```

###### Microorganism 2

```{r RA plot only one microorganism at T0 and T12, echo=TRUE, fig.show='show', warning=FALSE, message=FALSE}
p_Clos <- RA_one_microbe(
  tse = M_TSE_cf_UN,
  microbe = "Clostridium_SGB6179", # Change here the microorganism
  var = "Time",
  var_levels = c("T0","T12"),
  level = "Species",
  subset_var = NULL,
  subset_levels = NULL,
  save_svg = here::here("results", "RA_Clostridium_SGB6179.svg")
)
p_Clos
```

###### Microorganism 3

```{r RA plot only one microorganism at T0 and T12, echo=TRUE, fig.show='show', warning=FALSE, message=FALSE}
p_Fpla <- RA_one_microbe(
  tse = M_TSE_cf_UN,
  microbe = "Flavonifractor_plautii", # Change here the microorganism
  var = "Time",
  var_levels = c("T0","T12"),
  level = "Species",
  subset_var = NULL,
  subset_levels = NULL,
  save_svg = here::here("results", "RA_Flavonifractor_plautii.svg")
)
p_Fpla
```

#### Prevalence of zeros (%) by taxon and by time

```{r Prevalence of zeros, echo=TRUE, fig.show='hide', warning=FALSE, message=FALSE}
# ---- function: percent zeros (prevalence of zeros) by Feature and Time ----
zeros_by_feature_time <- function(ra_long,
                                  feature_col  = "FeatureID",
                                  value_col    = "RelAbundance",
                                  group_col    = "Time",
                                  group_levels = c("T0","T12"),
                                  zero_eps     = 0) {

  ra_long %>%
    mutate(
      .feature = .data[[feature_col]],
      .value   = .data[[value_col]],
      .group   = factor(.data[[group_col]], levels = group_levels)
    ) %>%
    filter(!is.na(.feature), !is.na(.group), !is.na(.value)) %>%
    group_by(.feature, .group) %>%
    summarise(
      n = n(),
      n_zero = sum(.value <= zero_eps),
      pct_zero = 100 * n_zero / n,
      .groups = "drop"
    ) %>%
    pivot_wider(
      names_from  = .group,
      values_from = c(n, n_zero, pct_zero),
      names_sep   = "_"
    ) %>%
    rename(FeatureID = .feature)
}
```

##### Load data and export results

```{r Load in data and export results, echo=TRUE, fig.show='hide', warning=FALSE, message=FALSE}
# ---- load your long RA table ----
ra_df <- read_csv(
  here::here("results", "RA_plot_relabundance_long.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    PatientID = str_extract(SampleID, "^[A-Za-z]+\\d+"),
    Time = factor(Time, levels = c("T0","T12"))
  )

# ---- parameters ----
group_levels <- c("T0","T12")
zero_eps <- 0   # your summary contains exact zeros

# ---- 1) Prevalence of zeros by feature and time ----
zero_df <- ra_df %>%
  filter(!is.na(FeatureID), !is.na(Time), !is.na(RelAbundance)) %>%
  group_by(FeatureID, Time) %>%
  summarise(
    n = n(),
    n_zero = sum(RelAbundance <= zero_eps),
    pct_zero = 100 * n_zero / n,
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from  = Time,
    values_from = c(n, n_zero, pct_zero),
    names_sep   = "_"
  )

# ---- 2) Paired Wilcoxon test comparing T0 vs T12 by feature (p-value) ----
p_df <- ra_df %>%
  filter(!is.na(FeatureID), !is.na(Time), !is.na(RelAbundance), !is.na(PatientID)) %>%
  # in case there are duplicates (there should not be any), collapse to 1 value per patient and time point
  group_by(FeatureID, PatientID, Time) %>%
  summarise(RelAbundance = median(RelAbundance, na.rm = TRUE), .groups = "drop") %>%
# requires complete T0 and T12 pairs
  group_by(FeatureID, PatientID) %>%
  filter(all(group_levels %in% Time)) %>%
  ungroup() %>%
  pivot_wider(names_from = Time, values_from = RelAbundance) %>%
  group_by(FeatureID) %>%
  summarise(
    n_pairs = sum(!is.na(.data[[group_levels[1]]]) & !is.na(.data[[group_levels[2]]])),
    p = tryCatch(
      wilcox.test(
        .data[[group_levels[1]]],
        .data[[group_levels[2]]],
        paired = TRUE,
        exact  = FALSE
      )$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) %>%
  mutate(
    padj = p.adjust(p, method = "BH"),
    p_signif = case_when(
      is.na(p) ~ NA_character_,
      p <= 0.0001 ~ "****",
      p <= 0.001  ~ "***",
      p <= 0.01   ~ "**",
      p <= 0.05   ~ "*",
      TRUE ~ "ns"
    ),
    padj_signif = case_when(
      is.na(padj) ~ NA_character_,
      padj <= 0.0001 ~ "****",
      padj <= 0.001  ~ "***",
      padj <= 0.01   ~ "**",
      padj <= 0.05   ~ "*",
      TRUE ~ "ns"
    )
  )

# ---- 3) Join everything ----
ra_zero_p <- p_df %>%
  left_join(zero_df, by = "FeatureID") %>%
  arrange(p)

# ---- export ----
write_csv(ra_zero_p, here::here("results", "RA_pvalues_and_percent_zeros_by_Time.csv"))
```

#-------------------------------------------------------------------------------
##### PANEL PLOTS
#-------------------------------------------------------------------------------

```{r Panel RA of DAA of Feces at T0 and T12, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
p_RA_DAA <- (p_Aliq | p_Clos | p_Fpla) +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

p_RA_DAA <- p_RA_DAA + plot_annotation(tag_levels = "A")

ggsave(here("results","RA_DAA_feces_ped.svg"), p_RA_DAA, width = 10, height = 4, bg = "white")
```

#-------------------------------------------------------------------------------
##### PANEL PLOTS AD + BD + DAA + RA_DAA
#-------------------------------------------------------------------------------

```{r Panel AD and BD of Feces at T0 and T12, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
p_microbiota <- (p_chao1_fc | p_shannon_fc | p_invsimp_fc) /
  (p_beta_fc | DAA_Metaphlan_cf_fc_abc_bar) /
  (p_Aliq | p_Clos | p_Fpla) +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

#p_microbiota <- p_microbiota + plot_annotation(tag_levels = "A")

ggsave(here("results","Microbiota_feces_ped.svg"), p_microbiota, width = 10, height = 11, bg = "white")
```
