---
title: "cf_metagenomics_ARGs_thesis_chapter_II"
author: "Natalia Bastón-Paz"
date: "2026"
output: html_document
---

# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require("knitr")
opts_knit$set(root.dir = "C:/RProjects/cf_metagenomics_ARGs_thesis_chapter_II")
```

# Clean environments

```{#r Clean environments, message=FALSE, warning=FALSE}
rm(list = ls())
```

# ------------------------------------------------------------------------------
# LIBRARIES
# ------------------------------------------------------------------------------

```{r Libraries, message=FALSE, warning=FALSE}
library(mia)     
library(miaViz)
library(tidyverse)         
library(castor)
library(readr)
library(ggpubr)
library(utils)
library(ggplot2)          
library(patchwork)
library(plotrix)
library(ComplexUpset)
library(vegan)
library(stats)
library(rstatix)
library(data.table)
library(Maaslin2)
library(MicrobiomeStat)
library(ANCOMBC)
library(ALDEx2)
library(MADAV)
library(SummarizedExperiment)
library(dplyr)
library(tibble)
library(TreeSummarizedExperiment)
library(stringr)
library(here)
library(pheatmap)
library(ggrepel)
library(RColorBrewer)
library(ggtree)
library(ape)
library(ggnewscale)
# library(Polychrome)
library(ggsci)
library(ComplexHeatmap)
library(circlize)
library(patchwork)
library(tidyr)
library(exact2x2)
library(PropCIs)
library(binom)
library(scales)
library(patchwork)

# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("Polychrome")
```

# ------------------------------------------------------------------------------

# METADATA FILE

```{r Load in metadata file, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
arg_metadata_cf_fc <- read_delim(here("data", "cf_metadata_profiles_pediatrics.txt"), 
                                "\t", 
                                escape_double = FALSE, 
                                trim_ws = TRUE)
```

# RESISTANCE FILE

```{r Load in resistance file, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
arg_data_cf_fc <- read_delim(here("data", "ARGs_feces_pediatrics.txt"), 
                                "\t", 
                                escape_double = FALSE, 
                                trim_ws = TRUE)
```

##### Remove samples

```{r Exclude samples, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
samples_to_remove <- c("A10E0","A10E12","A10H0","A10H12","A12E0","A12E12","A12H0","A12H12","A13E0","A13E12","A13H0","A13H12","A16E0","A16E12","A16H0","A16H12","A17E0","A17E12","A17H0","A17H12","A19E0","A19E12","A19H0","A19H12","A1E0","A1E12","A1H0","A1H12","A20E0","A20E12","A20H0","A20H12","A22E0","A22E12","A22H0","A22H12","A24E0","A24E12","A24H0","A24H12","A25E0","A25E12","A25H0","A25H12","A26E0","A26E12","A26H0","A26H12","A27E0","A27E12","A27H0","A27H12","A28E0","A28E12","A28H0","A28H12","A30E0","A30E12","A30H0","A30H12","A3E0","A3E12","A3H0","A3H12","A4E0","A4E12","A4H0","A4H12","A6E0","A6E12","A6H0","A6H12","A7E0","A7E12","A7H0","A7H12","A8E0","A8E12","A8H0","A8H12","A9E0","A9E12","A9H0","A9H12","P11E0","P11E12","P14E0","P14E12","P16E0","P16E12","P17E0","P17E12","P20E0","P20E12","P20H0","P20H12","P21E0","P21E12","P25E0","P25E12","P25H0","P25H12","P26E0","P26E12","P27E0","P27E12","P2E0","P2E12","P3E0","P3E12","P3H0","P3H12","P42E0","P42E12","P42H0","P42H12","P4E0","P4E12","P5E0","P5E12","P5H0","P5H12")
arg_data_cf_fc <- arg_data_cf_fc %>% filter(!SampleID %in% samples_to_remove)
arg_data_cf_fc <- arg_data_cf_fc %>% filter(!SampleID %in% samples_to_remove)

colnames(arg_data_cf_fc)
```

##### Filter resistance genes

```{r Filter resistance genes, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
resistance_genes <- arg_data_cf_fc %>% 
  filter(Data_Type == 'Resistance') %>%
  dplyr::select(SampleID, Gene)
```

##### Convert to binary

```{r Convert to binary, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
binary_data <- resistance_genes %>%
  mutate(Presence = 1) %>%
  distinct(SampleID, Gene, .keep_all = TRUE) %>%  
  pivot_wider(names_from = Gene, values_from = Presence, values_fill = 0)
```

##### Join with metadata file

```{r Join with metadata, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
binary_data <- binary_data %>%
  left_join(arg_metadata_cf_fc %>% dplyr::select(SampleID, PatientID, Time), by = "SampleID")
```

##### Separate data by Time points

```{r Separate data by time points, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
binary_T0 <- binary_data %>% filter(Time == "T0")
binary_T12 <- binary_data %>% filter(Time == "T12")
```

# ------------------------------------------------------------------------------
# PLOTTING
# ------------------------------------------------------------------------------
##### Prepare data for plotting

```{r Prepare data for plotting, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
binary_long_T0 <- binary_T0 %>%
  pivot_longer(cols = -c(SampleID, PatientID, Time), names_to = "Gene", values_to = "Presence") %>%
  mutate(Time = "T0")

binary_long_T12 <- binary_T12 %>%
  pivot_longer(cols = -c(SampleID, PatientID, Time), names_to = "Gene", values_to = "Presence") %>%
  mutate(Time = "T12")

binary_long <- bind_rows(binary_long_T0, binary_long_T12)
```


##### Sort by PatientID

```{r Sort by PatientID, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Create Sample_order right after binding T0 and T12
binary_long <- binary_long %>%
  mutate(Sample_order = paste0(PatientID, "_", Time))

# Sort PatientID alphabetically and numerically
patient_levels <- binary_long %>%
  distinct(PatientID) %>%
  mutate(letter = str_extract(PatientID, "^[A-Za-z]+"),
         num = as.numeric(str_extract(PatientID, "\\d+"))) %>%
  arrange(letter, num) %>%
  pull(PatientID)

# Build ordered Sample_order: A1_T0, A1_T12, A2_T0, A2_T12, etc.
sample_order <- expand.grid(PatientID = patient_levels,
                          Time = c("T0", "T12")) %>%
  arrange(PatientID, Time) %>%
  mutate(Sample_order = paste0(PatientID, "_", Time)) %>%
  pull(Sample_order)

# Apply factor levels with the correct order
binary_long$Sample_order <- factor(binary_long$Sample_order, levels = sample_order)
```

#### Vertical plot

```{r Vertical plot, echo=TRUE, results='show', message=FALSE, warning=FALSE}
palette20 <- pal_d3("category20")(20)

# === Filter only variable genes ===
n_samples <- length(unique(binary_long$SampleID))

variable_genes <- binary_long %>%
  group_by(Gene) %>%
  summarise(total = sum(Presence), .groups = "drop") %>%
  filter(total > 0 & total < n_samples) %>%
  pull(Gene)


filtered_binary <- binary_long %>%
  filter(Gene %in% variable_genes) %>%
  mutate(
    Pres_Time = case_when(
      Presence == 0 ~ "Absent",
      Presence == 1 & Time == "T0"  ~ "Present_T0",
      Presence == 1 & Time == "T12" ~ "Present_T12"
    )
  )


pal_heat <- c("Absent" = "white", 
              "Present_T0" = "#F4E1A1",
              "Present_T12" = "#A1E1E1")

# === Heatmap (genes in rows, pacients in columns) ===
heatmap_plot_fc <- ggplot(filtered_binary,
                       aes(x = Sample_order, y = Gene, fill = Pres_Time)) +
  geom_tile(color = "grey85", linewidth = 0.25) +
  scale_fill_manual(values = pal_heat,
                    name = "Status",
                    labels = c("Absent", "Present at T0", "Present at T12")) +
  labs(title = "ARGs that change between T0 and T12 in feces",
       x = "PatientID_Time", y = "Gene") +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
    axis.text.y = element_text(size = 7)
  )

# === Calculate proportions per gene ===
proportions <- filtered_binary %>%
  group_by(Gene, Time) %>%
  summarise(Proportion = mean(Presence), .groups = "drop")

# === Barplot lateral (genes in Y, proportion in X) ===
barplot_fc <- ggplot(proportions, aes(x = Proportion, y = Gene, fill = Time)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("T0" = "#F4E1A1", "T12" = "#A1E1E1")) +
  labs(title = "Prevalence", x = "Proportion of samples", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank())

# === Combine both ===
final_plot_fc <- heatmap_plot_fc + barplot_fc + plot_layout(widths = c(3, 1))

final_plot_fc

ggsave(here::here("results","ARG_heatmap_bar_vertical_feces_pediatrics.svg"),
       heatmap_plot_fc, width = 8, height = 10, dpi = 300)

ggsave(here::here("results","ARG_heatmap_bar_vertical_combined_feces_pediatrics.svg"),
       final_plot_fc, width = 10, height = 10, dpi = 300)
```

##### Final plot

```{r Final plot, echo=TRUE, results='show', message=FALSE, warning=FALSE}
# -------------------------------
# Objects required:
# - filtered_binary (with Gene, Sample_order, Pres_Time, Presence, Time)
# - heatmap_plot_fc
# - barplot_fc
# - palette20 (named vector of colors for ATB classes)
# - arg_data_cf_fc (raw ARG table with ATB_class)
# -------------------------------

# Ensure Gene order is identical across plots (use the order you already plot)
gene_levels <- ggplot_build(heatmap_plot_fc)$layout$panel_params[[1]]$y$get_limits()
filtered_binary <- filtered_binary %>%
  mutate(Gene = factor(Gene, levels = gene_levels))

# Build Gene -> ATB_class mapping (1 class per gene; otherwise "Multi-class")
gene_col <- intersect(c("Gene","gene","ARG","arg","feature","featureID","FeatureID","ID","id"), names(arg_data_cf_fc))[1]
if (is.na(gene_col)) stop("I couldn't find the gene column in arg_data_cf_fc (expected something like Gene/ARG/FeatureID).")

gene_class <- arg_data_cf_fc %>%
  filter(Data_Type == "Resistance") %>%
  filter(!is.na(ATB_class) & ATB_class != "") %>%
  transmute(
    Gene = .data[[gene_col]],
    ATB_class
  ) %>%
  distinct() %>%
  group_by(Gene) %>%
  summarise(
    ATB_class = if (n_distinct(ATB_class) == 1) first(ATB_class) else "Multi-class",
    .groups = "drop"
  ) %>%
  filter(Gene %in% levels(filtered_binary$Gene)) %>%
  mutate(
    Gene = factor(Gene, levels = levels(filtered_binary$Gene)),
    ATB_class = as.factor(ATB_class)
  )

# Optional: legend order to follow palette20
if (!is.null(names(palette20)) && any(nzchar(names(palette20)))) {
  lvl <- unique(c(intersect(names(palette20), levels(gene_class$ATB_class)),
                  setdiff(levels(gene_class$ATB_class), names(palette20))))
  gene_class$ATB_class <- factor(gene_class$ATB_class, levels = lvl)
}

# Sidebar plot: one tile per gene indicating ATB class
p_class <- ggplot(gene_class, aes(x = 1, y = Gene, fill = ATB_class)) +
  geom_tile() +
  scale_fill_manual(values = palette20, name = "ATB Class") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_discrete(drop = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = margin(t = 5, r = 4, b = 5, l = 5)
  )

# Combine: ATB class sidebar + heatmap + prevalence barplot
final_plot_fc2 <- p_class + heatmap_plot_fc + barplot_fc +
  plot_layout(widths = c(0.25, 3, 1), guides = "collect") &
  theme(legend.position = "right")

final_plot_fc2

ggsave(here::here("results","ARG_heatmap_ATBclass_sidebar_bar_vertical_combined_feces_pediatrics.svg"),
       final_plot_fc2, width = 8, height = 10, dpi = 300)
```

##### Final vertical plot separating by pair
```{r Final plot separating by pair, echo=TRUE, results='show', message=FALSE, warning=FALSE}
# Filter only variable genes
n_samples <- length(unique(binary_long$SampleID))

variable_genes <- binary_long %>%
  group_by(Gene) %>%
  summarise(total = sum(Presence), .groups = "drop") %>%
  filter(total > 0 & total < n_samples) %>%
  pull(Gene)


filtered_binary <- binary_long %>%
  filter(Gene %in% variable_genes) %>%
  mutate(
    Pres_Time = case_when(
      Presence == 0 ~ "Absent",
      Presence == 1 & Time == "T0"  ~ "Present_T0",
      Presence == 1 & Time == "T12" ~ "Present_T12"
    )
  )

# Add spacing between patient pairs (T0/T12)
gap <- 0.6

# Build patient order
patient_levels2 <- filtered_binary %>%
  distinct(PatientID) %>%
  mutate(letter = stringr::str_extract(PatientID, "^[A-Za-z]+"),
         num    = as.numeric(stringr::str_extract(PatientID, "\\d+"))) %>%
  arrange(letter, num) %>%
  pull(PatientID)

# Map each Sample_order to a numeric x position with a gap after each patient pair
x_map <- tidyr::expand_grid(
  PatientID = patient_levels2,
  Time      = c("T0","T12")
) %>%
  mutate(
    pid_i = match(PatientID, patient_levels2),
    Sample_order = paste0(PatientID, "_", Time),
    x_pos = (pid_i - 1) * (2 + gap) + ifelse(Time == "T0", 1, 2)
  ) %>%
  arrange(pid_i, Time)

filtered_binary <- filtered_binary %>%
  dplyr::left_join(x_map %>% dplyr::select(Sample_order, x_pos), by = "Sample_order")


pal_heat <- c("Absent" = "white", 
              "Present_T0" = "#F4E1A1",
              "Present_T12" = "#A1E1E1")

# Heatmap (genes in rows, patients in columns)
heatmap_plot_fc <- ggplot(filtered_binary,
                          aes(x = x_pos, y = Gene, fill = Pres_Time)) +
  geom_tile(color = "grey85", linewidth = 0.25, width = 0.95, height = 0.95) +
  scale_fill_manual(values = pal_heat,
                    name = "Status",
                    labels = c("Absent", "Present at T0", "Present at T12")) +
  scale_x_continuous(
    breaks = x_map$x_pos,
    labels = gsub("_", "\n", x_map$Sample_order),
    expand = c(0, 0)
  ) +
  labs(title = "ARGs that change between T0 and T12 in feces",
       x = "PatientID_Time", y = "Gene") +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
    axis.text.y = element_text(size = 7)
  )

# Calculate proportions per gene
proportions <- filtered_binary %>%
  dplyr::group_by(Gene, Time) %>%
  dplyr::summarise(Proportion = mean(Presence), .groups = "drop") %>%
  dplyr::mutate(Time = factor(Time, levels = c("T0", "T12")))

# Barplot lateral (genes in Y, proportion in X)
barplot_fc <- ggplot2::ggplot(proportions, ggplot2::aes(x = Proportion, y = Gene, fill = Time)) +
  ggplot2::geom_bar(
    stat = "identity",
    position = ggplot2::position_dodge2(width = 0.9, preserve = "single", reverse = TRUE),
    color = "black", linewidth = 0.1
  ) +
  ggplot2::scale_fill_manual(
    values = c("T0" = "#F4E1A1", "T12" = "#A1E1E1"),
    breaks = c("T0", "T12")
  ) +
  ggplot2::labs(title = "Prevalence", x = "Proportion of samples", y = NULL) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    axis.text.y = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    panel.grid.major.y = ggplot2::element_blank()
  )

# Ensure Gene order is identical across plots (use the order you already plot)
gene_levels <- ggplot_build(heatmap_plot_fc)$layout$panel_params[[1]]$y$get_limits()
filtered_binary <- filtered_binary %>%
  mutate(Gene = factor(Gene, levels = gene_levels))

# Build Gene -> ATB_class mapping (1 class per gene; otherwise "Multi-class")
gene_col <- intersect(c("Gene","gene","ARG","arg","feature","featureID","FeatureID","ID","id"), names(arg_data_cf_fc))[1]
if (is.na(gene_col)) stop("I couldn't find the gene column in arg_data_cf_fc (expected something like Gene/ARG/FeatureID).")

gene_class <- arg_data_cf_fc %>%
  filter(Data_Type == "Resistance") %>%
  filter(!is.na(ATB_class) & ATB_class != "") %>%
  transmute(
    Gene = .data[[gene_col]],
    ATB_class
  ) %>%
  distinct() %>%
  group_by(Gene) %>%
  summarise(
    ATB_class = if (n_distinct(ATB_class) == 1) first(ATB_class) else "Multi-class",
    .groups = "drop"
  ) %>%
  filter(Gene %in% levels(filtered_binary$Gene)) %>%
  mutate(
    Gene = factor(Gene, levels = levels(filtered_binary$Gene)),
    ATB_class = as.factor(ATB_class)
  )

# Optional: legend order to follow palette20
if (!is.null(names(palette20)) && any(nzchar(names(palette20)))) {
  lvl <- unique(c(intersect(names(palette20), levels(gene_class$ATB_class)),
                  setdiff(levels(gene_class$ATB_class), names(palette20))))
  gene_class$ATB_class <- factor(gene_class$ATB_class, levels = lvl)
}

# Sidebar plot: one tile per gene indicating ATB class
p_class <- ggplot(gene_class, aes(x = 1, y = Gene, fill = ATB_class)) +
  geom_tile() +
  scale_fill_manual(values = palette20, name = "ATB Class") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_discrete(drop = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = margin(t = 5, r = 4, b = 5, l = 5)
  )

# Combine: ATB class sidebar + heatmap + prevalence barplot
final_plot_fc2 <- p_class + heatmap_plot_fc + barplot_fc +
  plot_layout(widths = c(0.25, 3, 1), guides = "collect") &
  theme(legend.position = "right")

final_plot_fc2

ggsave(here::here("results","ARG_heatmap_ATBclass_sidebar_bar_vertical_combined_feces_pediatrics_wider.svg"),
       final_plot_fc2, width = 10, height = 10, dpi = 300)
```


# ------------------------------------------------------------------------------
# STATISTICS
# ------------------------------------------------------------------------------

## McNemar exact per gene

```{r McNemar exact test per gene, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Pair T0/T12 per patient for each gene (fill missing as 0, and if duplicates, take max)
paired_data <- filtered_binary %>%
  dplyr::select(PatientID, Gene, Time, Presence) %>%
  dplyr::group_by(PatientID, Gene, Time) %>%
  dplyr::summarise(Presence = max(Presence, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(
    names_from  = Time,
    values_from = Presence,
    values_fill = list(Presence = 0)
  )

# Quick sanity check: should be 9 rows per gene if all patients are paired
check_n <- paired_data %>%
  dplyr::count(Gene, name = "n_patients") %>%
  dplyr::summarise(min_n = min(n_patients), max_n = max(n_patients))
print(check_n)

# Exact McNemar via binomial test on discordants:
# b = (T0=1,T12=0), c = (T0=0,T12=1), under H0: b ~ Binom(b+c, 0.5)
mcnemar_results <- paired_data %>%
  dplyr::mutate(
    Lost_1to0   = (T0 == 1 & T12 == 0),
    Gained_0to1 = (T0 == 0 & T12 == 1)
  ) %>%
  dplyr::group_by(Gene) %>%
  dplyr::summarise(
    n_paired = dplyr::n(),
    Prev_T0  = mean(T0),
    Prev_T12 = mean(T12),
    DeltaPrev = Prev_T12 - Prev_T0,
    Lost_1to0   = sum(Lost_1to0),
    Gained_0to1 = sum(Gained_0to1),
    Total_Discordant = Lost_1to0 + Gained_0to1,
    P_value = {
      b <- sum(T0 == 1 & T12 == 0)
      c <- sum(T0 == 0 & T12 == 1)
      if ((b + c) == 0) 1 else stats::binom.test(b, b + c, p = 0.5, alternative = "two.sided")$p.value
    },
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    P_adj = p.adjust(P_value, method = "BH")
  ) %>%
  dplyr::arrange(P_value)

print(mcnemar_results)

readr::write_excel_csv(mcnemar_results, here("results", "ARG_mcnemar_results_ped.csv"))
```

## Normality and statistical test function

```{r Shapiro on paired differences for t-test or Wilcoxon, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
paired_test_auto <- function(df, col_T0 = "T0", col_T12 = "T12",
                             id_col = NULL, alpha = 0.05,
                             shapiro_min_unique = 3, verbose = TRUE) {

  stopifnot(all(c(col_T0, col_T12) %in% names(df)))

  dat <- df %>%
    dplyr::select(dplyr::all_of(c(id_col, col_T0, col_T12))) %>%
    dplyr::filter(!is.na(.data[[col_T0]]) & !is.na(.data[[col_T12]])) %>%
    dplyr::mutate(delta = .data[[col_T12]] - .data[[col_T0]])

  n <- nrow(dat)
  if (n < 3) stop("Need at least 3 paired observations.")

  can_shapiro <- length(unique(dat$delta)) >= shapiro_min_unique && stats::sd(dat$delta) > 0

  shapiro_p <- NA_real_
  shapiro_W <- NA_real_
  chosen <- NA_character_

  if (can_shapiro) {
    sh <- stats::shapiro.test(dat$delta)
    shapiro_p <- unname(sh$p.value)
    shapiro_W <- unname(sh$statistic)
    use_t <- shapiro_p > alpha
  } else {
    # If delta has too little variation, default to Wilcoxon (robust)
    use_t <- FALSE
  }

  if (use_t) {
    test <- stats::t.test(dat[[col_T0]], dat[[col_T12]], paired = TRUE)
    chosen <- "paired_t_test"
    out <- tibble::tibble(
      test_used = chosen,
      n_pairs = n,
      shapiro_W = shapiro_W,
      shapiro_p = shapiro_p,
      statistic = unname(test$statistic),
      df = unname(test$parameter),
      p_value = unname(test$p.value),
      estimate_mean_delta = unname(test$estimate),
      conf_low = unname(test$conf.int[1]),
      conf_high = unname(test$conf.int[2]),
      mean_delta_T12_minus_T0 = mean(dat$delta)
    )
  } else {
    test <- stats::wilcox.test(dat[[col_T0]], dat[[col_T12]], paired = TRUE, exact = FALSE)
    chosen <- "paired_wilcoxon"
    out <- tibble::tibble(
      test_used = chosen,
      n_pairs = n,
      shapiro_W = shapiro_W,
      shapiro_p = shapiro_p,
      statistic = unname(test$statistic),
      p_value = unname(test$p.value),
      mean_delta_T12_minus_T0 = mean(dat$delta),
      median_delta_T12_minus_T0 = stats::median(dat$delta)
    )
  }

  if (verbose) {
    message("Auto test: ", chosen,
            if (!is.na(shapiro_p)) paste0(" (Shapiro p=", signif(shapiro_p, 3), ")") else " (Shapiro n/a)")
  }

  list(
    summary = out,
    data_used = dat
  )
}
```


### ARG counts

```{r ARG counts, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
arg_counts <- binary_long %>%
  dplyr::group_by(PatientID, Time) %>%
  dplyr::summarise(n_ARGs = sum(Presence, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = Time, values_from = n_ARGs)

print(arg_counts)
```

### Run function per gene

```{r Run function per gene, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
res_auto_ARG <- paired_test_auto(arg_counts, col_T0 = "T0", col_T12 = "T12", alpha = 0.05)

res_auto_ARG$summary

readr::write_excel_csv(res_auto_ARG$summary,
  here::here("results", "ARG_counts_paired_test_auto.csv"))
```
