---
title: "cf_SCFA_thesis_chapter_II"
author: "Natalia Bastón-Paz"
date: '2026'
output: html_document
---

# Set up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require("knitr")
opts_knit$set(root.dir = "C:/RProjects/cf_SCFA_thesis/cf_SCFA_thesis_chapter_II")
```

# Clean environments
```{# r Clean environments, message=FALSE, warning=FALSE}
rm(list = ls())
```

## General introduction

Cystic fibrosis (CF) is a genetic disorder caused by pathogenic variants in the CFTR (Cystic Fibrosis Transmembrane Conductance Regulator) gene. CFTR dysfunction leads to impaired ion transport and the production of thick mucus, primarily affecting the lungs and gastrointestinal tract. People with CF (pwCF) commonly develop chronic airway infections and gastrointestinal complications.

In recent years, CFTR modulator therapies, particularly the triple combination elexacaftor/tezacaftor/ivacaftor (ETI), have markedly improved clinical outcomes, especially respiratory function and nutritional status. Short-chain fatty acids (SCFAs) are produced through the fermentation of dietary fiber by colonic bacteria. Among them, butyrate has been associated with improved gut function. SCFAs can be quantified using gas chromatography–mass spectrometry (GC–MS).

In this script, we analyze SCFA concentrations in 35 people with CF aged 6–66 years, assessed before ETI treatment (T0) and after 12 months of treatment (T12).

*Main question:*

-   *Do fecal and sputum SCFA concentrations change after one year of ETI treatment?*

## Libraries
```{r Libraries, message=FALSE, warning=FALSE}
library(here)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggdist)
library(purrr)
library(stringr)
library(patchwork)
library(ggpubr)
library(ggridges)
library(ppcor)
library(lme4)
library(lmerTest)
library(grid)
library(stringr)
library(forcats)
library(scales)
library(gtools)
library(emmeans)
library(tibble)
library(broom.mixed)
library(ggbreak)
library(cowplot)
library(lme4)
library(lmerTest)
```

## Directories
```{r Create directory, message=FALSE, warning=FALSE}
dir.create(here("results"), showWarnings = FALSE)
```


# ------------------------------------------------------------------------------
# SCFA FECES
# ------------------------------------------------------------------------------

### Load in the data
```{r Load in data SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_scfa <- read_delim(here("data", "SCFA_nutri_ped.csv"))
```

#### Descriptive stats
```{r Descriptive stats SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
range(df_scfa$acetate, na.rm = TRUE)
min_val <- min(df_scfa$acetate, na.rm = TRUE)
max_val <- max(df_scfa$acetate, na.rm = TRUE)

range(df_scfa$propionate, na.rm = TRUE)
min_val <- min(df_scfa$propionate, na.rm = TRUE)
max_val <- max(df_scfa$propionate, na.rm = TRUE)

range(df_scfa$butyrate, na.rm = TRUE)
min_val <- min(df_scfa$butyrate, na.rm = TRUE)
max_val <- max(df_scfa$butyrate, na.rm = TRUE)
```

### Filter data by time
```{r Filter data by time SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
data_scfa_T0 <- subset(df_scfa, Time == "T0")
data_scfa_T12 <- subset(df_scfa, Time == "T12")
```

### Function for statistic test
```{r Function for statistic test based on shapiro results SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
auto_paired_test_scfa <- function(df,
                                   id_col   = "PatientID",
                                   time_col = "Time",
                                   value_col = "acetate",
                                   levels   = c("T0","T12")) {
  stopifnot(all(c(id_col, time_col, value_col) %in% names(df)))
  
  dfw <- df[, c(id_col, time_col, value_col)]
  dfw <- tidyr::pivot_wider(
    dfw,
    names_from  = !!rlang::sym(time_col),
    values_from = !!rlang::sym(value_col)
  )
  if (!all(levels %in% names(dfw))) {
    stop("No columns found ", paste(levels, collapse = " and "), " after pivot_wider().")
  }
  dfw <- dfw[stats::complete.cases(dfw[, levels]), ]
  
  x0   <- dfw[[levels[1]]]
  x12  <- dfw[[levels[2]]]
  diff <- x12 - x0
  n    <- length(diff)
  if (n < 3) stop("Very few valid pairs (n < 3).")
  
  sh     <- shapiro.test(diff)
  normal <- sh$p.value >= 0.05
  
  if (normal) {
    tt        <- t.test(x0, x12, paired = TRUE)
    mean_diff <- mean(diff, na.rm = TRUE)
    sd_diff   <- stats::sd(diff, na.rm = TRUE)
    d_paired  <- mean_diff / sd_diff
    
    mean_T0  <- mean(x0,  na.rm = TRUE); sd_T0  <- stats::sd(x0,  na.rm = TRUE)
    mean_T12 <- mean(x12, na.rm = TRUE); sd_T12 <- stats::sd(x12, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Paired t-test\n")
    cat("n =", n, "| t =", round(tt$statistic, 4), "| p =", signif(tt$p.value, 4), "\n")
    cat("T0:   mean ± SD = ", round(mean_T0, 3), " ± ", round(sd_T0, 3), "\n", sep = "")
    cat("T12:  mean ± SD = ", round(mean_T12, 3), " ± ", round(sd_T12, 3), "\n", sep = "")
    cat("Mean difference (T12−T0) =", round(mean_diff, 3),
        " | 95% CI [", round(tt$conf.int[1], 3), ", ", round(tt$conf.int[2], 3), "]\n", sep = "")
    cat("Effect size (Cohen's d paired) =", round(d_paired, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "paired_t",
      n = n,
      shapiro = sh,
      test = tt,
      group_summary = list(
        T0  = c(mean = mean_T0, sd = sd_T0),
        T12 = c(mean = mean_T12, sd = sd_T12)
      ),
      mean_diff = mean_diff,
      ci = tt$conf.int,
      effect_size = list(type = "cohen_d_paired", value = d_paired)
    ))
    
  } else {
    wt   <- wilcox.test(x0, x12, paired = TRUE, exact = FALSE, correct = FALSE)
    V    <- as.numeric(wt$statistic)
    r_rb <- 1 - (2 * V) / (n * (n + 1))
    
    med_T0   <- stats::median(x0,  na.rm = TRUE)
    q1_T0    <- stats::quantile(x0,  0.25, na.rm = TRUE)
    q3_T0    <- stats::quantile(x0,  0.75, na.rm = TRUE)
    med_T12  <- stats::median(x12, na.rm = TRUE)
    q1_T12   <- stats::quantile(x12, 0.25, na.rm = TRUE)
    q3_T12   <- stats::quantile(x12, 0.75, na.rm = TRUE)
    med_diff <- stats::median(diff, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Not normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Wilcoxon signed-rank (paired)\n")
    cat("n =", n, "| V =", V, "| p =", signif(wt$p.value, 4), "\n")
    cat("T0:   median [IQR] = ", round(med_T0, 3), " [",
        round(q1_T0, 3), "–", round(q3_T0, 3), "]\n", sep = "")
    cat("T12:  median [IQR] = ", round(med_T12, 3), " [",
        round(q1_T12, 3), "–", round(q3_T12, 3), "]\n", sep = "")
    cat("Median difference (T12−T0) =", round(med_diff, 3), "\n")
    cat("Effect size (rank-biserial r) =", round(r_rb, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "wilcoxon_paired",
      n = n,
      shapiro = sh,
      test = wt,
      group_summary = list(
        T0  = c(median = med_T0, Q1 = q1_T0, Q3 = q3_T0),
        T12 = c(median = med_T12, Q1 = q1_T12, Q3 = q3_T12)
      ),
      median_diff = med_diff,
      effect_size = list(type = "rank_biserial_r", value = r_rb, V = V)
    ))
  }
}
# Wrapper "safe": if n < 3, returns a valid res_obj object with p = NA instead of stopping
auto_paired_test_scfa_safe <- function(df,
                                       id_col   = "PatientID",
                                       time_col = "Time",
                                       value_col = "acetate",
                                       levels   = c("T0","T12")) {
  tryCatch(
    auto_paired_test_scfa(df,
                          id_col = id_col,
                          time_col = time_col,
                          value_col = value_col,
                          levels = levels),
    error = function(e) {
      if (grepl("Very few valid pairs", e$message) || grepl("n < 3", e$message)) {
        list(
          variable = value_col,
          method   = "too_few_pairs",
          n        = NA_integer_,
          shapiro  = list(p.value = NA_real_),
          test     = list(statistic = NA_real_, p.value = NA_real_),
          effect_size = list(type = NA_character_, value = NA_real_)
        )
      } else {
        stop(e)
      }
    }
  )
}
```

### Perform  the statistical test
#### Acetate
```{r Execute stats function SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_acetate <- auto_paired_test_scfa(df_scfa, value_col = "acetate")
```

#### Propionate
```{r Execute stats function SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_propionate <- auto_paired_test_scfa(df_scfa, value_col = "propionate")
```

#### Butyrate
```{r Execute stats function SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_butyrate <- auto_paired_test_scfa(df_scfa, value_col = "butyrate")
```


### Function for plots
```{r Function for plot SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
make_paired_plot_scfa <- function(df, value_var, res_obj, y_lab, outfile) {
  stopifnot(all(c("PatientID","Time", value_var) %in% names(df)))
  
  d <- df %>%
    mutate(Time = factor(Time, levels = c("T0","T12"))) %>%
    group_by(PatientID) %>% filter(n_distinct(Time) == 2) %>% ungroup() %>%
    mutate(x = as.numeric(Time))
  
  col_box   <- c("T0" = "#F4E1A1", "T12" = "#A1E1E1")
  col_point <- c("T0" = "#FAF3DC", "T12" = "#D6F5F5")
  
  d <- d %>%
    mutate(
      fill_box   = col_box[Time],
      fill_point = col_point[Time]
    )
  
  p_val   <- res_obj$test$p.value
  p_label <- ifelse(p_val < 0.001, "p < 0.001", paste0("p = ", signif(p_val, 3)))
  
  y_max    <- max(d[[value_var]], na.rm = TRUE) * 1.05
  p_values <- data.frame(x = 1, xend = 2, y = y_max, label = p_label)
  
  p <- ggplot() +
    stat_slab(
      data = d %>% filter(Time == "T0"),
      aes(x = x, y = .data[[value_var]], fill = fill_box),
      side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    stat_slab(
      data = d %>% filter(Time == "T12"),
      aes(x = x, y = .data[[value_var]], fill = fill_box),
      side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    geom_boxplot(
      data = d,
      aes(x = x, y = .data[[value_var]], fill = fill_box),
      width = 0.5, outlier.shape = NA,
      color = "black", alpha = 0.9, show.legend = TRUE
    ) +
    geom_point(
      data = d,
      aes(x = x, y = .data[[value_var]], fill = fill_point),
      shape = 21, size = 3, stroke = 0.3, color = "black",
      show.legend = FALSE
    ) +
    geom_line(
      data = d,
      aes(x = x, y = .data[[value_var]], group = PatientID),
      color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
    ) +
    geom_segment(
      data = p_values,
      aes(x = x, xend = xend, y = y, yend = y),
      color = "black", linewidth = 0.5
    ) +
    geom_text(
      data = p_values,
      aes(x = (x + xend) / 2, y = y * 1.02, label = label),
      size = 4
    ) +
    scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
    scale_fill_identity(guide = "legend", name = "Time",
                        breaks = col_box, labels = names(col_box)) +
    labs(x = NULL, y = y_lab) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "right",
          axis.text.x = element_text(size = 14))
  
  if (!dir.exists(here("results"))) dir.create(here("results"))
  ggsave(here("results", outfile), p, width = 8, height = 6, units = "in", dpi = 300)
  
  invisible(p)
}
```


#### Panel plots
```{r Panel plots lung function, echo=TRUE, results='show', message=FALSE, warning=FALSE}
polish_panel_plot <- function(p) {
  p +
    theme(
      legend.position = "bottom",
      plot.margin = margin(5.5, 5.5, 5.5, 5.5),
      axis.title.y = element_text(margin = margin(r = 6))
    )
}

p_ace <- polish_panel_plot(
  make_paired_plot_scfa(df_scfa, "acetate", res_acetate, "Acetate (mM)", "SCFA_acetate_feces.svg")
)

p_pro <- polish_panel_plot(
  make_paired_plot_scfa(df_scfa, "propionate", res_propionate, "Propionate (mM)", "SCFA_propionate_feces.svg")
)

p_but <- polish_panel_plot(
  make_paired_plot_scfa(df_scfa, "butyrate", res_butyrate, "Butyrate (mM)", "SCFA_butyrate_feces.svg")
)

panel_scfa <- (p_ace | p_pro | p_but) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

panel_scfa <- panel_scfa + plot_annotation(tag_levels = "A")

ggsave(here("results","SCFA_without_ratio_panel.svg"),
       panel_scfa, width = 10, height = 5.5, bg = "white")
```


### Save stats results
```{r Save stats results SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# === Export: acetate, propionate, butyrate ===

.build_row_scfa <- function(df, value_col, res_obj, p_fdr = NA_real_) {
  stopifnot(all(c("PatientID","Time", value_col) %in% names(df)))
  is_t <- identical(res_obj$method, "paired_t")

  paired <- df %>%
    dplyr::select(PatientID, Time, !!rlang::sym(value_col)) %>%
    tidyr::pivot_wider(names_from = Time, values_from = !!rlang::sym(value_col)) %>%
    dplyr::filter(stats::complete.cases(T0, T12))

  mean_T0  <- mean(paired$T0);  sd_T0  <- sd(paired$T0)
  mean_T12 <- mean(paired$T12); sd_T12 <- sd(paired$T12)

  med_T0 <- median(paired$T0)
  q1_T0  <- as.numeric(quantile(paired$T0,  0.25))
  q3_T0  <- as.numeric(quantile(paired$T0,  0.75))

  med_T12 <- median(paired$T12)
  q1_T12  <- as.numeric(quantile(paired$T12, 0.25))
  q3_T12  <- as.numeric(quantile(paired$T12, 0.75))

  fmt_group_T0  <- if (is_t) sprintf("%.3f ± %.3f", mean_T0, sd_T0)  else sprintf("%.3f [%0.3f-%0.3f]", med_T0,  q1_T0,  q3_T0)
  fmt_group_T12 <- if (is_t) sprintf("%.3f ± %.3f", mean_T12, sd_T12) else sprintf("%.3f [%0.3f-%0.3f]", med_T12, q1_T12, q3_T12)

  tibble::tibble(
    variable      = res_obj$variable,
    method        = res_obj$method,
    n_pairs       = res_obj$n,
    shapiro_p     = unname(res_obj$shapiro$p.value),
    stat          = unname(res_obj$test$statistic),
    p_value       = res_obj$test$p.value,
    mean_T0  = mean_T0,
    sd_T0    = sd_T0,
    median_T0= med_T0,
    q1_T0    = q1_T0,
    q3_T0    = q3_T0,
    min_T0   = min(paired$T0),
    max_T0   = max(paired$T0),
    range_T0 = max(paired$T0) - min(paired$T0),
    mean_T12   = mean_T12,
    sd_T12     = sd_T12,
    median_T12 = med_T12,
    q1_T12     = q1_T12,
    q3_T12     = q3_T12,
    min_T12    = min(paired$T12),
    max_T12    = max(paired$T12),
    range_T12  = max(paired$T12) - min(paired$T12),
    group_T0      = fmt_group_T0,
    group_T12     = fmt_group_T12,
    diff_estimate = if (is_t) res_obj$mean_diff else res_obj$median_diff,
    ci_low        = if (is_t) res_obj$ci[1] else NA_real_,
    ci_high       = if (is_t) res_obj$ci[2] else NA_real_,
    effect_type   = res_obj$effect_size$type,
    effect_value  = res_obj$effect_size$value
  )
}

row_acetate <- .build_row_scfa(df_scfa, "acetate", res_acetate)
row_propionate <- .build_row_scfa(df_scfa, "propionate", res_propionate)
row_butyrate <- .build_row_scfa(df_scfa, "butyrate", res_butyrate)

scfa_stats <- dplyr::bind_rows(row_acetate, row_propionate, row_butyrate)

scfa_stats$p_fdr <- p.adjust(scfa_stats$p_value, method = "BH")

if (!dir.exists(here("results"))) dir.create(here("results"))
readr::write_excel_csv(scfa_stats, here("results", "SCFA_without_ratios_feces_stats.csv"))
```


# -----------------------------------------------------------------------------
# GLOBAL FDR
# -----------------------------------------------------------------------------

### FDR SCFAs feces
```{r Stats results FDR, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# 1) Put all your test objects in a single nested list (system -> variables -> res_object)
clinical_tests <- list(
  `Fecal SCFAs` = list(
    acetate = res_acetate,
    propionate = res_propionate,
    butyrate = res_butyrate
  )
)

# 2) Helper to extract p-value (adapt if your object structure differs)
get_p <- function(x) x$test$p.value

# 3) Build one table with all clinical p-values
clinical_pvals <- purrr::imap_dfr(clinical_tests, \(tests, sys) {
  tibble::tibble(
    system   = sys,
    variable = names(tests),
    p_value  = purrr::map_dbl(tests, get_p)
  )
})

# 4) Global BH-FDR across ALL clinical variables together
clinical_pvals_globalFDR <- clinical_pvals %>%
  dplyr::mutate(
    p_fdr_global = p.adjust(p_value, method = "BH"),
    signif_fdr   = p_fdr_global < 0.05
  ) %>%
  dplyr::arrange(p_fdr_global)

print(clinical_pvals_globalFDR, n = Inf)

# 5) Save
if (!dir.exists(here::here("results"))) dir.create(here::here("results"))
readr::write_excel_csv(clinical_pvals_globalFDR, here::here("results", "clinical_globalFDR_pvalues_SCFAs_without_ratios.csv"))
```

