---
title: "cf_SCFA_thesis_chapter_II"
author: "Natalia Bastón-Paz"
date: '2026'
output: html_document
---

# Set up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require("knitr")
opts_knit$set(root.dir = "C:/RProjects/cf_SCFA_thesis/cf_SCFA_thesis_chapter_II")
```

# Clean environments
```{# r Clean environments, message=FALSE, warning=FALSE}
rm(list = ls())
```

## General introduction

Cystic fibrosis (CF) is a genetic disorder caused by pathogenic variants in the CFTR (Cystic Fibrosis Transmembrane Conductance Regulator) gene. CFTR dysfunction leads to impaired ion transport and the production of thick mucus, primarily affecting the lungs and gastrointestinal tract. People with CF (pwCF) commonly develop chronic airway infections and gastrointestinal complications.

In recent years, CFTR modulator therapies, particularly the triple combination elexacaftor/tezacaftor/ivacaftor (ETI), have markedly improved clinical outcomes, especially respiratory function and nutritional status. Short-chain fatty acids (SCFAs) are produced through the fermentation of dietary fiber by colonic bacteria. Among them, butyrate has been associated with improved gut function. SCFAs can be quantified using gas chromatography–mass spectrometry (GC–MS).

In this script, we analyze SCFA concentrations in 35 people with CF aged 6–66 years, assessed before ETI treatment (T0) and after 12 months of treatment (T12).

*Main question:*

-   *Do fecal SCFA concentrations change after one year of ETI treatment?*

## Libraries
```{r Libraries, message=FALSE, warning=FALSE}
library(here)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggdist)
library(purrr)
library(stringr)
library(patchwork)
library(ggpubr)
library(ggridges)
library(ppcor)
library(lme4)
library(lmerTest)
library(grid)
library(stringr)
library(forcats)
library(scales)
library(gtools)
library(emmeans)
library(tibble)
library(broom.mixed)
library(ggbreak)
library(cowplot)
library(lme4)
library(lmerTest)
```

## Directories
```{r Create directory, message=FALSE, warning=FALSE}
dir.create(here("results"), showWarnings = FALSE)
```


# ------------------------------------------------------------------------------
# SCFA FECES
# ------------------------------------------------------------------------------

### Load in the data
```{r Load in data SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_scfa <- read_delim(here("data", "SCFA_nutri_ped.csv"))
```

#### Descriptive stats
```{r Descriptive stats SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
range(df_scfa$acetate, na.rm = TRUE)
min_val <- min(df_scfa$acetate, na.rm = TRUE)
max_val <- max(df_scfa$acetate, na.rm = TRUE)

range(df_scfa$propionate, na.rm = TRUE)
min_val <- min(df_scfa$propionate, na.rm = TRUE)
max_val <- max(df_scfa$propionate, na.rm = TRUE)

range(df_scfa$butyrate, na.rm = TRUE)
min_val <- min(df_scfa$butyrate, na.rm = TRUE)
max_val <- max(df_scfa$butyrate, na.rm = TRUE)
```

### Filter data by time
```{r Filter data by time SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
data_scfa_T0 <- subset(df_scfa, Time == "T0")
data_scfa_T12 <- subset(df_scfa, Time == "T12")
```

### Function for statistic test
```{r Function for statistic test based on shapiro results SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
auto_paired_test_scfa <- function(df,
                                   id_col   = "PatientID",
                                   time_col = "Time",
                                   value_col = "acetate",
                                   levels   = c("T0","T12")) {
  stopifnot(all(c(id_col, time_col, value_col) %in% names(df)))
  
  dfw <- df[, c(id_col, time_col, value_col)]
  dfw <- tidyr::pivot_wider(
    dfw,
    names_from  = !!rlang::sym(time_col),
    values_from = !!rlang::sym(value_col)
  )
  if (!all(levels %in% names(dfw))) {
    stop("No columns found ", paste(levels, collapse = " and "), " after pivot_wider().")
  }
  dfw <- dfw[stats::complete.cases(dfw[, levels]), ]
  
  x0   <- dfw[[levels[1]]]
  x12  <- dfw[[levels[2]]]
  diff <- x12 - x0
  n    <- length(diff)
  if (n < 3) stop("Very few valid pairs (n < 3).")
  
  sh     <- shapiro.test(diff)
  normal <- sh$p.value >= 0.05
  
  if (normal) {
    tt        <- t.test(x0, x12, paired = TRUE)
    mean_diff <- mean(diff, na.rm = TRUE)
    sd_diff   <- stats::sd(diff, na.rm = TRUE)
    d_paired  <- mean_diff / sd_diff
    
    mean_T0  <- mean(x0,  na.rm = TRUE); sd_T0  <- stats::sd(x0,  na.rm = TRUE)
    mean_T12 <- mean(x12, na.rm = TRUE); sd_T12 <- stats::sd(x12, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Paired t-test\n")
    cat("n =", n, "| t =", round(tt$statistic, 4), "| p =", signif(tt$p.value, 4), "\n")
    cat("T0:   mean ± SD = ", round(mean_T0, 3), " ± ", round(sd_T0, 3), "\n", sep = "")
    cat("T12:  mean ± SD = ", round(mean_T12, 3), " ± ", round(sd_T12, 3), "\n", sep = "")
    cat("Mean difference (T12−T0) =", round(mean_diff, 3),
        " | 95% CI [", round(tt$conf.int[1], 3), ", ", round(tt$conf.int[2], 3), "]\n", sep = "")
    cat("Effect size (Cohen's d paired) =", round(d_paired, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "paired_t",
      n = n,
      shapiro = sh,
      test = tt,
      group_summary = list(
        T0  = c(mean = mean_T0, sd = sd_T0),
        T12 = c(mean = mean_T12, sd = sd_T12)
      ),
      mean_diff = mean_diff,
      ci = tt$conf.int,
      effect_size = list(type = "cohen_d_paired", value = d_paired)
    ))
    
  } else {
    wt   <- wilcox.test(x0, x12, paired = TRUE, exact = FALSE, correct = FALSE)
    V    <- as.numeric(wt$statistic)
    r_rb <- 1 - (2 * V) / (n * (n + 1))
    
    med_T0   <- stats::median(x0,  na.rm = TRUE)
    q1_T0    <- stats::quantile(x0,  0.25, na.rm = TRUE)
    q3_T0    <- stats::quantile(x0,  0.75, na.rm = TRUE)
    med_T12  <- stats::median(x12, na.rm = TRUE)
    q1_T12   <- stats::quantile(x12, 0.25, na.rm = TRUE)
    q3_T12   <- stats::quantile(x12, 0.75, na.rm = TRUE)
    med_diff <- stats::median(diff, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Not normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Wilcoxon signed-rank (paired)\n")
    cat("n =", n, "| V =", V, "| p =", signif(wt$p.value, 4), "\n")
    cat("T0:   median [IQR] = ", round(med_T0, 3), " [",
        round(q1_T0, 3), "–", round(q3_T0, 3), "]\n", sep = "")
    cat("T12:  median [IQR] = ", round(med_T12, 3), " [",
        round(q1_T12, 3), "–", round(q3_T12, 3), "]\n", sep = "")
    cat("Median difference (T12−T0) =", round(med_diff, 3), "\n")
    cat("Effect size (rank-biserial r) =", round(r_rb, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "wilcoxon_paired",
      n = n,
      shapiro = sh,
      test = wt,
      group_summary = list(
        T0  = c(median = med_T0, Q1 = q1_T0, Q3 = q3_T0),
        T12 = c(median = med_T12, Q1 = q1_T12, Q3 = q3_T12)
      ),
      median_diff = med_diff,
      effect_size = list(type = "rank_biserial_r", value = r_rb, V = V)
    ))
  }
}
# Wrapper "safe": if n < 3, returns a valid res_obj object with p = NA instead of stopping
auto_paired_test_scfa_safe <- function(df,
                                       id_col   = "PatientID",
                                       time_col = "Time",
                                       value_col = "acetate",
                                       levels   = c("T0","T12")) {
  tryCatch(
    auto_paired_test_scfa(df,
                          id_col = id_col,
                          time_col = time_col,
                          value_col = value_col,
                          levels = levels),
    error = function(e) {
      if (grepl("Very few valid pairs", e$message) || grepl("n < 3", e$message)) {
        list(
          variable = value_col,
          method   = "too_few_pairs",
          n        = NA_integer_,
          shapiro  = list(p.value = NA_real_),
          test     = list(statistic = NA_real_, p.value = NA_real_),
          effect_size = list(type = NA_character_, value = NA_real_)
        )
      } else {
        stop(e)
      }
    }
  )
}
```

### Perform  the statistical test
#### Acetate
```{r Execute stats function SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_acetate <- auto_paired_test_scfa(df_scfa, value_col = "acetate")
```

#### Propionate
```{r Execute stats function SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_propionate <- auto_paired_test_scfa(df_scfa, value_col = "propionate")
```

#### Butyrate
```{r Execute stats function SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_butyrate <- auto_paired_test_scfa(df_scfa, value_col = "butyrate")
```


### Function for plots
```{r Function for plot SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
make_paired_plot_scfa <- function(df, value_var, res_obj, y_lab, outfile) {
  stopifnot(all(c("PatientID","Time", value_var) %in% names(df)))
  
  d <- df %>%
    mutate(Time = factor(Time, levels = c("T0","T12"))) %>%
    group_by(PatientID) %>% filter(n_distinct(Time) == 2) %>% ungroup() %>%
    mutate(x = as.numeric(Time))
  
  col_box   <- c("T0" = "#F4E1A1", "T12" = "#A1E1E1")
  col_point <- c("T0" = "#FAF3DC", "T12" = "#D6F5F5")
  
  d <- d %>%
    mutate(
      fill_box   = col_box[Time],
      fill_point = col_point[Time]
    )
  
  p_val   <- res_obj$test$p.value
  p_label <- ifelse(p_val < 0.001, "p < 0.001", paste0("p = ", signif(p_val, 3)))
  
  y_max    <- max(d[[value_var]], na.rm = TRUE) * 1.05
  p_values <- data.frame(x = 1, xend = 2, y = y_max, label = p_label)
  
  p <- ggplot() +
    stat_slab(
      data = d %>% filter(Time == "T0"),
      aes(x = x, y = .data[[value_var]], fill = fill_box),
      side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    stat_slab(
      data = d %>% filter(Time == "T12"),
      aes(x = x, y = .data[[value_var]], fill = fill_box),
      side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    geom_boxplot(
      data = d,
      aes(x = x, y = .data[[value_var]], fill = fill_box),
      width = 0.5, outlier.shape = NA,
      color = "black", alpha = 0.9, show.legend = TRUE
    ) +
    geom_point(
      data = d,
      aes(x = x, y = .data[[value_var]], fill = fill_point),
      shape = 21, size = 3, stroke = 0.3, color = "black",
      show.legend = FALSE
    ) +
    geom_line(
      data = d,
      aes(x = x, y = .data[[value_var]], group = PatientID),
      color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
    ) +
    geom_segment(
      data = p_values,
      aes(x = x, xend = xend, y = y, yend = y),
      color = "black", linewidth = 0.5
    ) +
    geom_text(
      data = p_values,
      aes(x = (x + xend) / 2, y = y * 1.02, label = label),
      size = 4
    ) +
    scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
    scale_fill_identity(guide = "legend", name = "Time",
                        breaks = col_box, labels = names(col_box)) +
    labs(x = NULL, y = y_lab) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "right",
          axis.text.x = element_text(size = 14))
  
  if (!dir.exists(here("results"))) dir.create(here("results"))
  ggsave(here("results", outfile), p, width = 8, height = 6, units = "in", dpi = 300)
  
  invisible(p)
}
```


#### Panel plots
```{r Panel plots lung function, echo=TRUE, results='show', message=FALSE, warning=FALSE}
polish_panel_plot <- function(p) {
  p +
    theme(
      legend.position = "bottom",
      plot.margin = margin(5.5, 5.5, 5.5, 5.5),
      axis.title.y = element_text(margin = margin(r = 6))
    )
}

p_ace <- polish_panel_plot(
  make_paired_plot_scfa(df_scfa, "acetate", res_acetate, "Acetate (mM)", "SCFA_acetate_feces.svg")
)

p_pro <- polish_panel_plot(
  make_paired_plot_scfa(df_scfa, "propionate", res_propionate, "Propionate (mM)", "SCFA_propionate_feces.svg")
)

p_but <- polish_panel_plot(
  make_paired_plot_scfa(df_scfa, "butyrate", res_butyrate, "Butyrate (mM)", "SCFA_butyrate_feces.svg")
)

panel_scfa <- (p_ace | p_pro | p_but) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

panel_scfa <- panel_scfa + plot_annotation(tag_levels = "A")

ggsave(here("results","SCFA_without_ratio_panel.svg"),
       panel_scfa, width = 10, height = 5.5, bg = "white")
```


### Save stats results
```{r Save stats results SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# === Export: acetate, propionate, butyrate ===

.build_row_scfa <- function(df, value_col, res_obj, p_fdr = NA_real_) {
  stopifnot(all(c("PatientID","Time", value_col) %in% names(df)))
  is_t <- identical(res_obj$method, "paired_t")

  paired <- df %>%
    dplyr::select(PatientID, Time, !!rlang::sym(value_col)) %>%
    tidyr::pivot_wider(names_from = Time, values_from = !!rlang::sym(value_col)) %>%
    dplyr::filter(stats::complete.cases(T0, T12))

  mean_T0  <- mean(paired$T0);  sd_T0  <- sd(paired$T0)
  mean_T12 <- mean(paired$T12); sd_T12 <- sd(paired$T12)

  med_T0 <- median(paired$T0)
  q1_T0  <- as.numeric(quantile(paired$T0,  0.25))
  q3_T0  <- as.numeric(quantile(paired$T0,  0.75))

  med_T12 <- median(paired$T12)
  q1_T12  <- as.numeric(quantile(paired$T12, 0.25))
  q3_T12  <- as.numeric(quantile(paired$T12, 0.75))

  fmt_group_T0  <- if (is_t) sprintf("%.3f ± %.3f", mean_T0, sd_T0)  else sprintf("%.3f [%0.3f-%0.3f]", med_T0,  q1_T0,  q3_T0)
  fmt_group_T12 <- if (is_t) sprintf("%.3f ± %.3f", mean_T12, sd_T12) else sprintf("%.3f [%0.3f-%0.3f]", med_T12, q1_T12, q3_T12)

  tibble::tibble(
    variable      = res_obj$variable,
    method        = res_obj$method,
    n_pairs       = res_obj$n,
    shapiro_p     = unname(res_obj$shapiro$p.value),
    stat          = unname(res_obj$test$statistic),
    p_value       = res_obj$test$p.value,
    mean_T0  = mean_T0,
    sd_T0    = sd_T0,
    median_T0= med_T0,
    q1_T0    = q1_T0,
    q3_T0    = q3_T0,
    min_T0   = min(paired$T0),
    max_T0   = max(paired$T0),
    range_T0 = max(paired$T0) - min(paired$T0),
    mean_T12   = mean_T12,
    sd_T12     = sd_T12,
    median_T12 = med_T12,
    q1_T12     = q1_T12,
    q3_T12     = q3_T12,
    min_T12    = min(paired$T12),
    max_T12    = max(paired$T12),
    range_T12  = max(paired$T12) - min(paired$T12),
    group_T0      = fmt_group_T0,
    group_T12     = fmt_group_T12,
    diff_estimate = if (is_t) res_obj$mean_diff else res_obj$median_diff,
    ci_low        = if (is_t) res_obj$ci[1] else NA_real_,
    ci_high       = if (is_t) res_obj$ci[2] else NA_real_,
    effect_type   = res_obj$effect_size$type,
    effect_value  = res_obj$effect_size$value
  )
}

row_acetate <- .build_row_scfa(df_scfa, "acetate", res_acetate)
row_propionate <- .build_row_scfa(df_scfa, "propionate", res_propionate)
row_butyrate <- .build_row_scfa(df_scfa, "butyrate", res_butyrate)

scfa_stats <- dplyr::bind_rows(row_acetate, row_propionate, row_butyrate)

scfa_stats$p_fdr <- p.adjust(scfa_stats$p_value, method = "BH")

if (!dir.exists(here("results"))) dir.create(here("results"))
readr::write_excel_csv(scfa_stats, here("results", "SCFA_without_ratios_feces_stats.csv"))
```


# -----------------------------------------------------------------------------
# GLOBAL FDR
# -----------------------------------------------------------------------------

### FDR SCFAs feces
```{r Stats results FDR, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# 1) Put all your test objects in a single nested list (system -> variables -> res_object)
clinical_tests <- list(
  `Fecal SCFAs` = list(
    acetate = res_acetate,
    propionate = res_propionate,
    butyrate = res_butyrate
  )
)

# 2) Helper to extract p-value (adapt if your object structure differs)
get_p <- function(x) x$test$p.value

# 3) Build one table with all clinical p-values
clinical_pvals <- purrr::imap_dfr(clinical_tests, \(tests, sys) {
  tibble::tibble(
    system   = sys,
    variable = names(tests),
    p_value  = purrr::map_dbl(tests, get_p)
  )
})

# 4) Global BH-FDR across ALL clinical variables together
clinical_pvals_globalFDR <- clinical_pvals %>%
  dplyr::mutate(
    p_fdr_global = p.adjust(p_value, method = "BH"),
    signif_fdr   = p_fdr_global < 0.05
  ) %>%
  dplyr::arrange(p_fdr_global)

print(clinical_pvals_globalFDR, n = Inf)

# 5) Save
if (!dir.exists(here::here("results"))) dir.create(here::here("results"))
readr::write_excel_csv(clinical_pvals_globalFDR, here::here("results", "clinical_globalFDR_pvalues_SCFAs_without_ratios.csv"))
```

#-------------------------------------------------------------------------------
#-------------------------------------------------------------------------------
# MICROBIOME CORRELATIONS
#-------------------------------------------------------------------------------
#-------------------------------------------------------------------------------

## 0.1. Function: run LMM z(outcome) ~ z(taxon) + Time + (1|PatientID) or z(outcome) ~ z(taxon) + Time + PatientID if singular fit
```{r Function LMM, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
run_lmm_taxa_vs_outcomes <- function(dat_long,
                                    taxa_keep,
                                    outcomes,
                                    taxon_col = "taxon_val",
                                    id_col = "PatientID",
                                    time_col = "Time",
                                    tol_singular = 1e-4) {

  # safety check
  missing_out <- setdiff(outcomes, colnames(dat_long))
  if (length(missing_out) > 0) {
    stop("These outcomes are missing in dat_long: ", paste(missing_out, collapse = ", "))
  }

  purrr::map_dfr(outcomes, function(out) {

    out <- as.character(out)[1]

    purrr::map_dfr(taxa_keep, function(tx) {

      tx <- as.character(tx)[1]

      dat_tx <- dat_long %>%
        dplyr::filter(FeatureID == tx) %>%
        dplyr::select(FeatureID, dplyr::all_of(c(taxon_col, out, time_col, id_col))) %>%
        dplyr::filter(stats::complete.cases(.)) %>%
        dplyr::mutate(
          !!time_col := factor(.data[[time_col]], levels = c("T0","T12")),
          !!id_col   := factor(.data[[id_col]])
        )

      # skip if too few points or no variance
      if (nrow(dat_tx) < 6 ||
          sd(dat_tx[[taxon_col]], na.rm = TRUE) == 0 ||
          sd(dat_tx[[out]],       na.rm = TRUE) == 0) {
        return(tibble::tibble(
          Outcome = out, FeatureID = tx, n = nrow(dat_tx),
          beta_SD = NA_real_, conf.low = NA_real_, conf.high = NA_real_, p = NA_real_,
          singular_fit = NA, model_used = NA_character_
        ))
      }

      # standardize within this model dataset
      dat_tx <- dat_tx %>%
        dplyr::mutate(
          taxon_z = as.numeric(scale(.data[[taxon_col]])),
          y_z     = as.numeric(scale(.data[[out]]))
        )

      # 1) Fit LMM first
      fml_lmm <- stats::as.formula(paste0("y_z ~ taxon_z + ", time_col, " + (1|", id_col, ")"))
      fit_lmm <- lmerTest::lmer(fml_lmm, data = dat_tx)

      sing <- lme4::isSingular(fit_lmm, tol = tol_singular)

      # 2) If singular, refit with PatientID as fixed effect
      if (isTRUE(sing)) {

        fml_fe <- stats::as.formula(paste0("y_z ~ taxon_z + ", time_col, " + ", id_col))
        fit_fe <- stats::lm(fml_fe, data = dat_tx)

        co <- summary(fit_fe)$coefficients
        ci <- suppressWarnings(stats::confint(fit_fe, parm = "taxon_z"))

        return(tibble::tibble(
          Outcome   = out,
          FeatureID = tx,
          n         = nrow(dat_tx),
          beta_SD   = unname(co["taxon_z","Estimate"]),
          conf.low  = ci[1],
          conf.high = ci[2],
          p         = unname(co["taxon_z","Pr(>|t|)"]),
          singular_fit = TRUE,
          model_used   = "FE_PatientID"
        ))
      }

      # 3) Otherwise keep LMM results
      co <- summary(fit_lmm)$coefficients
      ci <- suppressWarnings(confint(fit_lmm, parm = "taxon_z", method = "Wald"))

      tibble::tibble(
        Outcome   = out,
        FeatureID = tx,
        n         = nrow(dat_tx),
        beta_SD   = unname(co["taxon_z","Estimate"]),
        conf.low  = ci[1],
        conf.high = ci[2],
        p         = unname(co["taxon_z","Pr(>|t|)"]),
        singular_fit = FALSE,
        model_used   = "LMM"
      )
    })

  }) %>%
    dplyr::group_by(Outcome) %>%
    dplyr::mutate(p_fdr_outcome = p.adjust(p, method = "BH")) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(p_fdr_global = p.adjust(p, method = "BH"))
}
```

## 0.2. Helper: build forest dataframe for a given group
```{r Helper forest, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
prep_forest_df <- function(res_table, outcomes_vec, labels_vec, taxa_order) {

  res_table %>%
    dplyr::filter(Outcome %in% outcomes_vec, FeatureID %in% taxa_order) %>%
    dplyr::mutate(
      Outcome_label = dplyr::recode(Outcome, !!!as.list(labels_vec)),
      Outcome_label = factor(Outcome_label,
                             levels = rev(unname(labels_vec[outcomes_vec]))),
      FeatureID = factor(FeatureID, levels = taxa_order),
      sig = !is.na(.data[[FDR_COL]]) & .data[[FDR_COL]] < 0.05
    ) %>%
    dplyr::filter(!is.na(beta_SD), !is.na(conf.low), !is.na(conf.high))
}
```


## 0.3. Taxa deltas for microbiota
```{r Taxa deltas, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Load the long RA table
ra <- read_delim(here("data", "RA_plot_relabundance_long.csv"))
colnames(ra)

# Standardize + keep only the 3 taxa
taxa_keep <- c("Adlercreutzia_equolifaciens", "Clostridium_SGB6179", "Flavonifractor_plautii")

ra2 <- ra %>%
  mutate(
    PatientID = sub("H\\d+$", "", SampleID),
    Time = case_when(
      Time %in% c("T0","time0","Time0") ~ "T0",
      Time %in% c("T12","time12","Time12") ~ "T12",
      TRUE ~ as.character(Time)
    ),
    RelAbundance = as.numeric(RelAbundance)
  ) %>%
  filter(FeatureID %in% taxa_keep)

# Build Δ per taxon (log1p handles zeros nicely)
taxa_delta <- ra2 %>%
  dplyr::mutate(val = log1p(RelAbundance)) %>%
  dplyr::select(PatientID, FeatureID, Time, val) %>%
  tidyr::pivot_wider(names_from = Time, values_from = val) %>%
  dplyr::filter(complete.cases(T0, T12)) %>%
  dplyr::mutate(delta = T12 - T0) %>%
  dplyr::select(PatientID, FeatureID, delta) %>%
  tidyr::pivot_wider(
    names_from = FeatureID,
    values_from = delta,
    names_glue = "{FeatureID}_delta"
  )

# Start with taxa deltas; SCFA deltas will be added later
delta_all_taxa <- taxa_delta

colnames(taxa_delta)
summary(delta_all_taxa[, grep("_delta$", colnames(delta_all_taxa))])
```

#----------------


# 1. Microbiome composition and SCFAs
## Deltas
```{r Microbiota composition and SCFAs, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# SCFA deltas (Δ = T12 - T0)
scfa_vars <- c("acetate", "propionate", "butyrate")

scfa_delta <- purrr::map(scfa_vars, \(v) {
  df_scfa %>%
    dplyr::select(PatientID, Time, dplyr::all_of(v)) %>%
    tidyr::pivot_wider(names_from = Time, values_from = dplyr::all_of(v)) %>%
    dplyr::filter(stats::complete.cases(T0, T12)) %>%
    dplyr::mutate(delta = T12 - T0) %>%
    dplyr::select(PatientID, delta) %>%
    dplyr::rename(!!paste0(v, "_delta") := delta)
}) %>%
  purrr::reduce(dplyr::left_join, by = "PatientID")

# ---- join to your existing delta_all_taxa ----
delta_all_taxa <- delta_all_taxa %>%
  dplyr::left_join(scfa_delta, by = "PatientID")

# check
colnames(delta_all_taxa)
```

## Spearman
```{r Spearman, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
taxa_y <- c(
  "Adlercreutzia_equolifaciens_delta",
  "Clostridium_SGB6179_delta",
  "Flavonifractor_plautii_delta"
)

vars_x <- c(
  "acetate_delta",
  "propionate_delta",
  "butyrate_delta"
)

min_n <- 6

res_spear_taxa_scfa <- purrr::map_dfr(taxa_y, \(ty) {
  purrr::map_dfr(vars_x, \(vx) {
    tmp <- delta_all_taxa %>%
      dplyr::select(dplyr::all_of(c(ty, vx))) %>%
      dplyr::filter(stats::complete.cases(.))

    if (nrow(tmp) < min_n) {
      return(tibble::tibble(taxon = ty, var = vx, n = nrow(tmp), rho = NA_real_, p = NA_real_))
    }

    ct <- suppressWarnings(cor.test(tmp[[1]], tmp[[2]], method = "spearman", exact = FALSE))
    tibble::tibble(taxon = ty, var = vx, n = nrow(tmp), rho = unname(ct$estimate), p = ct$p.value)
  })
}) %>%
  dplyr::mutate(p_fdr = p.adjust(p, method = "fdr")) %>%
  dplyr::arrange(p_fdr)

res_spear_taxa_scfa

write.csv(res_spear_taxa_scfa, "res_spearman_taxa_scfa_without_ratios.csv", row.names = FALSE)
```

### Heatmap
```{r Heatmap, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
vars_heat <- c(
  "Adlercreutzia_equolifaciens_delta",
  "Clostridium_SGB6179_delta",
  "Flavonifractor_plautii_delta",
  "acetate_delta",
  "propionate_delta", 
  "butyrate_delta"
)

dfh <- delta_all_taxa %>% dplyr::select(dplyr::all_of(vars_heat))

# Spearman rho matrix
rho_mat <- cor(dfh, method = "spearman", use = "pairwise.complete.obs")

# P-value matrix (pairwise cor.test), only if >= min_n pairs
min_n <- 6
p_mat <- matrix(NA_real_, nrow = ncol(dfh), ncol = ncol(dfh),
                dimnames = list(colnames(dfh), colnames(dfh)))

for (r in seq_len(ncol(dfh))) {
  for (c in seq_len(ncol(dfh))) {
    x <- dfh[[r]]
    y <- dfh[[c]]
    ok <- complete.cases(x, y)
    if (sum(ok) >= min_n) {
      p_mat[r, c] <- suppressWarnings(
        cor.test(x[ok], y[ok], method = "spearman", exact = FALSE)$p.value
      )
    }
  }
}

# Convert to long format
idx <- setNames(seq_along(vars_heat), vars_heat)

rho_long <- as.data.frame(as.table(rho_mat)) %>%
  dplyr::rename(var_row = Var1, var_col = Var2, rho = Freq)

p_long <- as.data.frame(as.table(p_mat)) %>%
  dplyr::rename(var_row = Var1, var_col = Var2, p = Freq)

plot_df <- rho_long %>%
  dplyr::left_join(p_long, by = c("var_row", "var_col")) %>%
  dplyr::mutate(
    i = idx[var_row],
    j = idx[var_col]
  ) %>%
  dplyr::filter(i > j) %>%
  dplyr::mutate(pair_id = paste0(pmin(i, j), "_", pmax(i, j)))

# GLOBAL FDR across all unique pairs shown in the heatmap
p_unique <- plot_df %>%
  dplyr::select(pair_id, p) %>%
  dplyr::distinct() %>%
  dplyr::mutate(p_fdr = p.adjust(p, method = "fdr"))

plot_df <- plot_df %>%
  dplyr::left_join(p_unique, by = "pair_id") %>%
  dplyr::mutate(
    sig = dplyr::case_when(
      is.na(p_fdr) ~ "",
      p_fdr < 0.001 ~ "***",
      p_fdr < 0.01  ~ "**",
      p_fdr < 0.05  ~ "*",
      TRUE ~ ""
    ),
    label = paste0(sprintf("%.2f", rho), sig),
    var_col = factor(var_col, levels = vars_heat),
    var_row = factor(var_row, levels = rev(vars_heat))
  )

# Plot heatmap
p_heat_taxa_scfa <- ggplot(plot_df, aes(x = var_col, y = var_row, fill = rho)) +
  geom_tile() +
  geom_text(aes(label = label), size = 3) +
  scale_fill_gradient2(
    low = "#2C7BB6",
    mid = "grey95",
    high = "#D7191C",
    midpoint = 0,
    limits = c(-1, 1),
    name = "Spearman ρ"
  ) +
  coord_fixed() +
  labs(title = "Spearman correlations of changes (ΔT12–T0): taxa vs SCFAs") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

p_heat_taxa_scfa

# --- Save as SVG ---
ggsave("heatmap_spearman_deltas_FDR_taxa_SCFAs_without_ratios.svg",
       plot = p_heat_taxa_scfa, device = "svg", width = 11, height = 9, units = "in")
```

## LMER
```{r LMER, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
taxa_keep <- c("Adlercreutzia_equolifaciens", "Clostridium_SGB6179", "Flavonifractor_plautii")
scfa_outcomes <- c("acetate", "propionate", "butyrate")

# Harmonize Time to "T0"/"T12"
fix_time <- function(x) {
  x <- as.character(x)
  dplyr::case_when(
    x %in% c("T0","0","Time0","time0","baseline") ~ "T0",
    x %in% c("T12","12","Time12","time12","followup") ~ "T12",
    TRUE ~ x
  )
}

# ONE row per PatientID-Time for SCFAs
meta_scfa <- df_scfa %>%
  dplyr::mutate(Time = fix_time(Time)) %>%
  dplyr::select(PatientID, Time, dplyr::all_of(scfa_outcomes)) %>%
  dplyr::group_by(PatientID, Time) %>%
  dplyr::summarise(
    dplyr::across(dplyr::all_of(scfa_outcomes), ~ if (all(is.na(.x))) NA_real_ else mean(.x, na.rm = TRUE)),
    .groups = "drop"
    )

# Microbiome long table
dat_long <- ra2 %>%
  dplyr::select(PatientID, Time, FeatureID, RelAbundance) %>%
  dplyr::mutate(
    taxon_val = log1p(RelAbundance)
  )

# Join SCFAs into microbiome long table (dat_long)
dat_long_scfa_full <- dat_long %>%
  dplyr::mutate(
    Time = fix_time(Time),
    PatientID = as.character(PatientID)
  ) %>%
  dplyr::left_join(meta_scfa, by = c("PatientID","Time")) %>%
  dplyr::mutate(
    Time = factor(Time, levels = c("T0","T12")),
    PatientID = factor(PatientID)
  )

# Run your SAME function (LMM with FE fallback if singular)
res_lmm_scfa <- run_lmm_taxa_vs_outcomes(
  dat_long = dat_long_scfa_full,
  taxa_keep = taxa_keep,
  outcomes = scfa_outcomes
)

res_lmm_scfa
write.csv(res_lmm_scfa, "res_lmm_taxa_vs_SCFAs_LMM_SD_without_ratios.csv", row.names = FALSE)
```

### Forest plot
```{r Forest plot, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
FDR_COL <- "p_fdr_global"   # or "p_fdr_outcome"

scfa_labels <- c(
  acetate            = "Acetate",
  propionate         = "Propionate",
  butyrate           = "Butyrate"
)

scfa_outcomes <- names(scfa_labels)

prep_forest_df <- function(res_table, outcomes_vec, labels_vec, taxa_order) {

  res_table %>%
    dplyr::filter(Outcome %in% outcomes_vec, FeatureID %in% taxa_order) %>%
    dplyr::mutate(
      Outcome_label = dplyr::recode(Outcome, !!!as.list(labels_vec)),
      Outcome_label = factor(Outcome_label,
                             levels = rev(unname(labels_vec[outcomes_vec]))),
      FeatureID = factor(FeatureID, levels = taxa_order),
      sig = !is.na(.data[[FDR_COL]]) & .data[[FDR_COL]] < 0.05
    ) %>%
    dplyr::filter(!is.na(beta_SD), !is.na(conf.low), !is.na(conf.high))
}

df_scfa_forest <- prep_forest_df(
  res_table    = res_lmm_scfa,
  outcomes_vec = scfa_outcomes,
  labels_vec   = scfa_labels,
  taxa_order   = taxa_keep
) %>%
  dplyr::mutate(FeatureID = factor(FeatureID, levels = taxa_keep))

p_taxa_scfa <- ggplot(df_scfa_forest, aes(x = beta_SD, y = Outcome_label)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high),
                width = 0.2, linewidth = 0.9, orientation = "y") +
  geom_point(aes(fill = sig), shape = 21, size = 3.2, color = "black") +
  facet_grid(. ~ FeatureID) +
  scale_fill_manual(
    name = "BH-FDR (q < 0.05)",
    values = c(`TRUE` = "black", `FALSE` = "white"),
    breaks = c(TRUE, FALSE),
    labels = c("Significant", "Not significant")
  ) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE)) +
  labs(
    title = "Association between fecal taxa and SCFAs (standardized)",
    subtitle = "Models adjusted for Time; effect sizes are standardized (SD outcome per 1 SD taxon)",
    x = "Standardized effect (SD outcome per 1 SD taxon)",
    y = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "grey85", linewidth = 0.6),
    panel.grid.minor = element_line(color = "grey92", linewidth = 0.4),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    strip.text  = element_text(face = "plain"),
    legend.position = "bottom"
  )

p_taxa_scfa

ggsave("forest_SCFAs_taxa_SD_without_ratios.svg", plot = p_taxa_scfa, device = "svg",
       width = 9, height = 6, units = "in")
```
