---
title: "cf_clinical_outcomes_associations_thesis_chapter_I"
author: "Natalia Bastón-Paz"
date: "2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require("knitr")
opts_knit$set(root.dir = "C:/RProjects/cf_clinical_outcomes_associations_thesis_chapter_I")
```

```{#r Clean environments, message=FALSE, warning=FALSE}
rm(list = ls())
```

## General introduction

Cystic fibrosis (CF) is a genetic disorder caused by mutations in the CFTR (Cystic Fibrosis Transmembrane Conductance Regulator) gene, leading to impaired chloride and bicarbonate transport across epithelial surfaces. This defect results in thick, viscous secretions in the lungs, gut, pancreas and other organs, and predisposes people with CF (pwCF) to chronic respiratory infections, malabsorption, and systemic complications. The CF lung microbiome is typically characterized by chronic colonisation with pathogens such as *Pseudomonas aeruginosa* and *Staphylococcus aureus*, while the gut microbiota is often altered by pancreatic insufficiency, antibiotic exposure and nutritional factors.
The introduction of highly effective CFTR modulators, particularly the triple therapy elexacaftor–tezacaftor–ivacaftor (ETI), has markedly improved lung function, nutritional status and quality of life in pwCF. However, the relationships between gut and lung microbiota and clinical markers under ETI treatment remain incompletely understood. In particular, it is relevant to identify which bacterial taxa are associated with lung function and markers of liver injury and intestinal inflammation before and after one year of ETI.
In this script, we use ANCOM-BC2 on fecal and sputum microbiota data (TreeSummarizedExperiment subsets for feces and sputum at T0 and T12), adjusting for sex and age, to test associations between bacterial taxa and selected clinical variables (ppFEV1, ALT, AST, GGT, fecal elastase and fecal calprotectin).

*Questions addressed:*

- *Which bacterial taxa in fecal and sputum samples are associated with lung function (ppFEV$_1$) at T0 and T12 under ETI treatment?*  
- *Which bacterial taxa are associated with liver enzymes (ALT, AST, GGT) and markers of intestinal function/inflammation (fecal elastase and fecal calprotectin) at T0 and T12?*

## Libraries

```{r Libraries, message=FALSE, warning=FALSE}
library(mia)     
library(miaViz)
library(tidyverse)         
library(castor)
library(readr)
library(ggpubr)
library(utils)
library(ggplot2)          
library(patchwork)
library(plotrix)
library(ComplexUpset)
library(vegan)
library(stats)
library(rstatix)
library(data.table)
library(Maaslin2)
library(MicrobiomeStat)
library(ANCOMBC)
library(ALDEx2)
library(MADAV)
library(SummarizedExperiment)
library(tibble)
library(TreeSummarizedExperiment)
library(stringr)
library(here)
library(svglite)
library(aplot)
library(patchwork)
library(tidyr)
library(ggsci)
library(purrr)
library(openxlsx)
library(broom)
library(ggthemes)
library(RColorBrewer)
library(gghalves)
library(scales)
library(ggdist)
library(forcats)
library(eulerr)
library(dplyr)
```

## Load the data

Optional: If we want to know which samples have 0 counts:

```{r Samples with 0 counts, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df <- read.table(here("data", "cf_metaphlan_counts.txt"),
                 header = TRUE, sep = "\t", 
                 row.names = 1, stringsAsFactors = FALSE)

print(names(which(colSums(df) == 100000000)))
```

#### Metadata file

Load in sample metadata file --\> colData_Metaphlan object:

```{r Load in Sample Metadata file, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
Metaphlan_data_cf <- read_delim(here("data", "cf_metadata_models.txt"), 
                                "\t", 
                                escape_double = FALSE, 
                                trim_ws = TRUE)
# Fix column names
colnames(Metaphlan_data_cf)[1] <- 'SampleID'
```

Remove the samples that have 0 counts and others of interest:

```{r Remove samples of choice, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
samples_to_remove <- c("A12E0", "A12E12", "A20E0", "A20E12", "A26E0", "A26E12", "P16E0", "P16E12", "P21E0", "P21E12", "P3H0", "P3H12")  # Replace with actual SampleIDs
Metaphlan_data_cf <- Metaphlan_data_cf[!Metaphlan_data_cf$SampleID %in% samples_to_remove, ]
```

Convert the column SampleID into row names:

```{r Convert col SampleID into row names, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
Metaphlan_data_cf <- Metaphlan_data_cf %>%
  column_to_rownames(var = 'SampleID') %>%                                          
  data.frame()
```

#### Abundance file

Load in abundance file (change file name and FeatureID_column) --> assay_Metaphlan object:

```{r Load in abundance file, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
M_assay_cf <- read_delim(here("data", "cf_metaphlan_counts.txt"), "\t", escape_double = FALSE, trim_ws = TRUE)
colnames(M_assay_cf)[1] <- 'FeatureID'
```

Get the taxonomy data from the FeatureID column of the assay_Metaphlan object --> rowData_Metaphlan object:

```{r Get the taxonomy data from FeatureID, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
M_rowData_cf <- M_assay_cf[,'FeatureID'] %>% 
  mutate(split = FeatureID) %>%  
  separate(col = split, into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species",'Type'), sep = "\\|.__") %>%
  mutate(Kingdom = str_remove_all(Kingdom,'^k__')) %>%  
  column_to_rownames(var = "FeatureID") %>%        
  data.frame()
```

Filter the assay_Metaphlan object (M_assay_cf) using the rowData_Metaphlan (M_rowData_cf) & colData_Metaphlan (Metaphlan_data_cf) objects and create the TSE:

```{r Filter and create the TSE, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
M_assay_cf <- M_assay_cf[,c('FeatureID',rownames(Metaphlan_data_cf))] %>% # The assay data columns are reorder to match the colData
  column_to_rownames(var = 'FeatureID') %>%        
  as.matrix()
M_TSE_cf_unfiltered <- TreeSummarizedExperiment(assays = M_assay_cf, # To create TreeSummarizedExperiment object                                   
                                                    rowData = M_rowData_cf,
                                                     colData = Metaphlan_data_cf)
assayNames(M_TSE_cf_unfiltered) <-'counts'
```

Rename NA's in 'UNCLASSIFIED' row of rowData(TSE_Metaphlan) --> Replaces NA values (and any other values) with "UNCLASSIFIED" in the UNCLASSIFIED row only:

```{r Rename UNCLASSIFIED, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
rowData(M_TSE_cf_unfiltered)[rowData(M_TSE_cf_unfiltered)$Kingdom == 'UNCLASSIFIED',1:ncol(rowData(M_TSE_cf_unfiltered))] <- 'UNCLASSIFIED' 
```

Filter to 'Species' level and retain 'UNCLASSIFIED':

```{r Filter to Species and retain UNCLASSIFIED, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
M_TSE_cf_unfiltered <- M_TSE_cf_unfiltered[
  (!is.na(rowData(M_TSE_cf_unfiltered)$Species) & is.na(rowData(M_TSE_cf_unfiltered)$Type)) | 
    (rowData(M_TSE_cf_unfiltered)$Kingdom == 'UNCLASSIFIED'),
]
```

Optional filtering steps: Remove all non-bacteria taxa (check first whether Bacteria are in the 'Kingdom' or 'Domain' class). Here we also keep unclassified Bacteria and UNCLASSIFIED

```{r Remove non-bacteria taxa, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
M_TSE_cf_bacteria <- M_TSE_cf_unfiltered[rowData(M_TSE_cf_unfiltered)$Kingdom %in% c('UNCLASSIFIED','Bacteria'), ]
```

Remove all singleton features:

```{r Remove singletons, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
M_TSE_cf_singletons <- M_TSE_cf_bacteria[apply(assay(M_TSE_cf_bacteria), 1, function(x) { sum(x > 0) > 1 }), ]
```

Remove 'UNCLASSIFIED' feature:

```{r Remove UNCLASSIFIED, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
M_TSE_cf_UN <- M_TSE_cf_singletons
M_TSE_cf_no_UN <- M_TSE_cf_singletons[rownames(M_TSE_cf_singletons) != 'UNCLASSIFIED',]
```

## Create subsets

Subsets by Time:

```{r Subset by Time, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
subset_T0 <- M_TSE_cf_UN[, M_TSE_cf_UN$Time %in% 'T0']
subset_T12 <- M_TSE_cf_UN[, M_TSE_cf_UN$Time %in% 'T12']
subset_T0_no_UN <- M_TSE_cf_no_UN[, M_TSE_cf_no_UN$Time %in% 'T0']
subset_T12_no_UN <- M_TSE_cf_no_UN[, M_TSE_cf_no_UN$Time %in% 'T12']
```

Subsets by SampleType:

```{r Subset by SampleType, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
subset_feces <- M_TSE_cf_UN[, M_TSE_cf_UN$SampleType %in% 'feces']
subset_sputum <- M_TSE_cf_UN[, M_TSE_cf_UN$SampleType %in% 'sputum']
subset_feces_no_UN <- M_TSE_cf_no_UN[, M_TSE_cf_no_UN$SampleType %in% 'feces']
subset_sputum_no_UN <- M_TSE_cf_no_UN[, M_TSE_cf_no_UN$SampleType %in% 'sputum']
```

Subsets by Time AND SampleType

```{r Subset by SampleType and Time, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
subset_feces_T0 <- subset_feces[, subset_feces$Time %in% 'T0']
subset_feces_T12 <- subset_feces[, subset_feces$Time %in% 'T12']
subset_sputum_T0 <- subset_sputum[, subset_sputum$Time %in% 'T0']
subset_sputum_T12 <- subset_sputum[, subset_sputum$Time %in% 'T12']
subset_feces_T0_no_UN <- subset_feces_no_UN[, subset_feces_no_UN$Time %in% 'T0']
subset_feces_T12_no_UN <- subset_feces_no_UN[, subset_feces_no_UN$Time %in% 'T12']
subset_sputum_T0_no_UN <- subset_sputum_no_UN[, subset_sputum_no_UN$Time %in% 'T0']
subset_sputum_T12_no_UN <- subset_sputum_no_UN[, subset_sputum_no_UN$Time %in% 'T12']
```

## Directories

```{r Directories, message=FALSE, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
dir.create(here("results_ancombc_adjusted_feces", "tables"), showWarnings = FALSE, recursive = TRUE)
dir.create(here("results_ancombc_adjusted_feces", "figures"), showWarnings = FALSE, recursive = TRUE)
dir.create(here("results_ancombc_adjusted_sputum", "tables"), showWarnings = FALSE, recursive = TRUE)
dir.create(here("results_ancombc_adjusted_sputum", "figures"), showWarnings = FALSE, recursive = TRUE)
```

# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------
# FECAL SAMPLES ANALYSIS
# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------

# ------------------------------------------------------------------------------
# LUNG FUNCTION
# ------------------------------------------------------------------------------

### Run ANCOM-BC2 at T0 and T12
```{r ppFEV1 feces, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
out_ppFEV1_fc_T0 <- ancombc2(
  data         = subset_feces_T0_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "ppFEV1 + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)

# T12
out_ppFEV1_fc_T12 <- ancombc2(
  data         = subset_feces_T12_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "ppFEV1 + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)
```

### Export the results
```{r Results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
## All
res_all_ppFEV1_fc_T0 <- out_ppFEV1_fc_T0$res
write_csv(res_all_ppFEV1_fc_T0, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_all_ppFEV1_T0_feces.csv"))

## Significant
res_sig_ppFEV1_fc_T0 <- res_all_ppFEV1_fc_T0 %>%
  filter(diff_ppFEV1 == TRUE, q_ppFEV1 < 0.05) %>%
  arrange(lfc_ppFEV1)
write_csv(res_sig_ppFEV1_fc_T0, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_sig_ppFEV1_T0_feces.csv"))

# T12
## All
res_all_ppFEV1_fc_T12 <- out_ppFEV1_fc_T12$res
write_csv(res_all_ppFEV1_fc_T12, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_all_ppFEV1_T12_feces.csv"))

## Significant
res_sig_ppFEV1_fc_T12 <- res_all_ppFEV1_fc_T12 %>%
  filter(diff_ppFEV1 == TRUE, q_ppFEV1 < 0.05) %>%
  arrange(lfc_ppFEV1)
write_csv(res_sig_ppFEV1_fc_T12, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_sig_ppFEV1_T12_feces.csv"))
```


# ------------------------------------------------------------------------------
# TRANSAMINASES
# ------------------------------------------------------------------------------

## ALT

### Run ANCOM-BC2 at T0 and T12
```{r ALT feces, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
out_ALT_fc_T0 <- ancombc2(
  data         = subset_feces_T0_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "ALT + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)

# T12
out_ALT_fc_T12 <- ancombc2(
  data         = subset_feces_T12_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "ALT + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)
```

### Export the results
```{r Results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
## All
res_all_ALT_fc_T0 <- out_ALT_fc_T0$res
write_csv(res_all_ALT_fc_T0, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_all_ALT_T0_feces.csv"))

## Significant
res_sig_ALT_fc_T0 <- res_all_ALT_fc_T0 %>%
  filter(diff_ALT == TRUE, q_ALT < 0.05) %>%
  arrange(lfc_ALT)
write_csv(res_sig_ALT_fc_T0, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_sig_ALT_T0_feces.csv"))

# T12
## All
res_all_ALT_fc_T12 <- out_ALT_fc_T12$res
write_csv(res_all_ALT_fc_T12, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_all_ALT_T12_feces.csv"))

## Significant
res_sig_ALT_fc_T12 <- res_all_ALT_fc_T12 %>%
  filter(diff_ALT == TRUE, q_ALT < 0.05) %>%
  arrange(lfc_ALT)
write_csv(res_sig_ALT_fc_T12, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_sig_ALT_T12_feces.csv"))
```

#------
## AST

### Run ANCOM-BC2 at T0 and T12
```{r AST feces, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
out_AST_fc_T0 <- ancombc2(
  data         = subset_feces_T0_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "AST + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10,
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)

# T12
out_AST_fc_T12 <- ancombc2(
  data         = subset_feces_T12_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "AST + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10,
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)
```

### Export the results
```{r Results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
## All
res_all_AST_fc_T0 <- out_AST_fc_T0$res
write_csv(res_all_AST_fc_T0, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_all_AST_T0_feces.csv"))

## Significant
res_sig_AST_fc_T0 <- res_all_AST_fc_T0 %>%
  filter(diff_AST == TRUE, q_AST < 0.05) %>%
  arrange(lfc_AST)
write_csv(res_sig_AST_fc_T0, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_sig_AST_T0_feces.csv"))

# T12
## All
res_all_AST_fc_T12 <- out_AST_fc_T12$res
write_csv(res_all_AST_fc_T12, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_all_AST_T12_feces.csv"))

## Significant
res_sig_AST_fc_T12 <- res_all_AST_fc_T12 %>%
  filter(diff_AST == TRUE, q_AST < 0.05) %>%
  arrange(lfc_AST)
write_csv(res_sig_AST_fc_T12, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_sig_AST_T12_feces.csv"))
```

#------
## GGT

### Run ANCOM-BC2 at T0 and T12
```{r GGT feces, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
out_GGT_fc_T0 <- ancombc2(
  data         = subset_feces_T0_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "GGT + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)

# T12
out_GGT_fc_T12 <- ancombc2(
  data         = subset_feces_T12_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "GGT + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)
```

### Export the results
```{r Results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
#T0
## All
res_all_GGT_fc_T0 <- out_GGT_fc_T0$res
write_csv(res_all_GGT_fc_T0, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_all_GGT_T0_feces.csv"))

## Significant
res_sig_GGT_fc_T0 <- res_all_GGT_fc_T0 %>%
  filter(diff_GGT == TRUE, q_GGT < 0.05) %>%
  arrange(lfc_GGT)
write_csv(res_sig_GGT_fc_T0, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_sig_GGT_T0_feces.csv"))

# T12
## All
res_all_GGT_fc_T12 <- out_GGT_fc_T12$res
write_csv(res_all_GGT_fc_T12, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_all_GGT_T12_feces.csv"))

## Significant
res_sig_GGT_fc_T12 <- res_all_GGT_fc_T12 %>%
  filter(diff_GGT == TRUE, q_GGT < 0.05) %>%
  arrange(lfc_GGT)
write_csv(res_sig_GGT_fc_T12, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_sig_GGT_T12_feces.csv"))
```


# ------------------------------------------------------------------------------
# ELASTASE
# ------------------------------------------------------------------------------

### Run ANCOM-BC2 at T0 and T12
```{r Elastase feces, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
out_elas_fc_T0 <- ancombc2(
  data         = subset_feces_T0_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "Elastase + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)

# T12
out_elas_fc_T12 <- ancombc2(
  data         = subset_feces_T12_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "Elastase + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)
```

### Export the results
```{r Results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
## All
res_all_elas_fc_T0 <- out_elas_fc_T0$res
write_csv(res_all_elas_fc_T0, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_all_elas_T0_feces.csv"))

## Significant
res_sig_elas_fc_T0 <- res_all_elas_fc_T0 %>%
  filter(diff_Elastase == TRUE, q_Elastase < 0.05) %>%
  arrange(lfc_Elastase)
write_csv(res_sig_elas_fc_T0, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_sig_elas_T0_feces.csv"))

# T12
## All
res_all_elas_fc_T12 <- out_elas_fc_T12$res
write_csv(res_all_elas_fc_T12, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_all_elas_T12_feces.csv"))

## Significant
res_sig_elas_fc_T12 <- res_all_elas_fc_T12 %>%
  filter(diff_Elastase == TRUE, q_Elastase < 0.05) %>%
  arrange(lfc_Elastase)
write_csv(res_sig_elas_fc_T12, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_sig_elas_T12_feces.csv"))
```


# ------------------------------------------------------------------------------
# CALPROTECTIN
# ------------------------------------------------------------------------------

### Run ANCOM-BC2 at T0 and T12
```{r Calprotectin feces, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
out_calpro_fc_T0 <- ancombc2(
  data         = subset_feces_T0_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "Calprotectin + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)

# T12
out_calpro_fc_T12 <- ancombc2(
  data         = subset_feces_T12_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "Calprotectin + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)
```

### Export the results
```{r Results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
## All
res_all_calpro_fc_T0 <- out_calpro_fc_T0$res
write_csv(res_all_calpro_fc_T0, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_all_calpro_T0_feces.csv"))

## Significant
res_sig_calpro_fc_T0 <- res_all_calpro_fc_T0 %>%
  filter(diff_Calprotectin == TRUE, q_Calprotectin < 0.05) %>%
  arrange(lfc_Calprotectin)
write_csv(res_sig_calpro_fc_T0, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_sig_calpro_T0_feces.csv"))

# T12
## All
res_all_calpro_fc_T12 <- out_calpro_fc_T12$res
write_csv(res_all_calpro_fc_T12, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_all_calpro_T12_feces.csv"))

## Significant
res_sig_calpro_fc_T12 <- res_all_calpro_fc_T12 %>%
  filter(diff_Calprotectin == TRUE, q_Calprotectin < 0.05) %>%
  arrange(lfc_Calprotectin)
write_csv(res_sig_calpro_fc_T12, here("results_ancombc_adjusted_feces", "tables", "ANCOMBC2_sig_calpro_T12_feces.csv"))
```


# ------------------------------------------------------------------------------
## Master Table Consolidation
# ------------------------------------------------------------------------------

### Create Master table
```{r Master table, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# List of result objects to consolidate
items_fc <- list(
  list(obj = "out_ppFEV1_fc_T0", term = "ppFEV1",      time = "T0",  clin = "ppFEV1"),
  list(obj = "out_ppFEV1_fc_T12",term = "ppFEV1",      time = "T12", clin = "ppFEV1"),
  list(obj = "out_ALT_fc_T0",    term = "ALT",         time = "T0",  clin = "ALT"),
  list(obj = "out_ALT_fc_T12",   term = "ALT",         time = "T12", clin = "ALT"),
  list(obj = "out_AST_fc_T0",    term = "AST",         time = "T0",  clin = "AST"),
  list(obj = "out_AST_fc_T12",   term = "AST",         time = "T12", clin = "AST"),
  list(obj = "out_GGT_fc_T0",    term = "GGT",         time = "T0",  clin = "GGT"),
  list(obj = "out_GGT_fc_T12",   term = "GGT",         time = "T12", clin = "GGT"),
  list(obj = "out_elas_fc_T0",   term = "Elastase",    time = "T0",  clin = "Elastase"),
  list(obj = "out_elas_fc_T12",  term = "Elastase",    time = "T12", clin = "Elastase"),
  list(obj = "out_calpro_fc_T0", term = "Calprotectin",time = "T0",  clin = "Calprotectin"),
  list(obj = "out_calpro_fc_T12",term = "Calprotectin",time = "T12", clin = "Calprotectin")
)

master_all_fc <- list()   
master_sig_fc <- list()  
alpha <- 0.05             

for (it in items_fc) {
  if (!exists(it$obj)) next
  out <- get(it$obj)
  res <- out$res

  lfc_col  <- paste0("lfc_",  it$term)
  q_col    <- paste0("q_",    it$term)
  diff_col <- paste0("diff_", it$term)

  if (!all(c(lfc_col, q_col, diff_col) %in% names(res))) next

  tab_fc <- data.frame(
    Taxon = res$taxon,
    Species = if ("taxon" %in% names(res)) sub(".*\\|s__", "", res$taxon) else NA_character_,
    Time = it$time,
    ClinVar = it$clin,
    lfc = res[[lfc_col]],
    q   = res[[q_col]],
    diff = as.logical(res[[diff_col]]),
    stringsAsFactors = FALSE
  )
  tab_fc$sign <- ifelse(is.na(tab_fc$lfc), NA, ifelse(tab_fc$lfc > 0, "+", ifelse(tab_fc$lfc < 0, "-", "0")))

  master_all_fc[[paste(it$clin, it$time, sep = "_")]] <- tab_fc

  sig_fc <- tab_fc[!is.na(tab_fc$q) & tab_fc$q < alpha & tab_fc$diff, , drop = FALSE]
  if (nrow(sig_fc) > 0) sig_fc <- sig_fc[order(abs(sig_fc$lfc), decreasing = TRUE), ]
  master_sig_fc[[paste(it$clin, it$time, sep = "_")]] <- sig_fc
}

master_all_df_fc <- if (length(master_all_fc)) do.call(rbind, master_all_fc) else data.frame()
master_sig_df_fc <- if (length(master_sig_fc)) do.call(rbind, master_sig_fc) else data.frame()

if (nrow(master_all_df_fc)) write.csv(master_all_df_fc, "ANCOMBC2_master_all_fc.csv", row.names = FALSE)
if (nrow(master_sig_df_fc)) write.csv(master_sig_df_fc, "ANCOMBC2_master_sig_fc.csv", row.names = FALSE)

cat("Rows in master_all_df_fc:", nrow(master_all_df_fc), "\n")
cat("Rows in master_sig_df_fc:", nrow(master_sig_df_fc), "\n")
```

### Summarize top taxa
```{r Top taxa by clinical variable, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# A) Top taxa by clinical variable (T0 and T12 combined)
top_by_variable_fc <- list()
if (nrow(master_sig_df_fc)) {
  vars <- unique(master_sig_df_fc$ClinVar)
  for (v in vars) {
    sub <- master_sig_df_fc[master_sig_df_fc$ClinVar == v, , drop = FALSE]
    if (nrow(sub) == 0) next
    sub_ord <- sub[order(abs(sub$lfc), decreasing = TRUE), ]
    top_by_variable_fc[[v]] <- head(sub_ord, 20)
  }
}
top_by_variable_df_fc <- if (length(top_by_variable_fc)) do.call(rbind, top_by_variable_fc) else data.frame()

# B) Top taxa by clinical variable AND timepoint (T0 and T12 separately)
top_by_variable_time_fc <- list()
if (nrow(master_sig_df_fc)) {
  combos <- unique(master_sig_df_fc[, c("ClinVar","Time")])
  for (i in seq_len(nrow(combos))) {
    v <- combos$ClinVar[i]; t <- combos$Time[i]
    sub <- master_sig_df_fc[master_sig_df_fc$ClinVar == v & master_sig_df_fc$Time == t, , drop = FALSE]
    if (nrow(sub) == 0) next
    sub_ord <- sub[order(abs(sub$lfc), decreasing = TRUE), ]
    top_by_variable_time_fc[[paste(v, t, sep = "_")]] <- head(sub_ord, 20)
  }
}
top_by_variable_time_df_fc <- if (length(top_by_variable_time_fc)) do.call(rbind, top_by_variable_time_fc) else data.frame()

# Save summaries to CSV
if (nrow(top_by_variable_df_fc))       write.csv(top_by_variable_df_fc,       "ANCOMBC2_top_by_variable_fc.csv", row.names = FALSE)
if (nrow(top_by_variable_time_df_fc)) write.csv(top_by_variable_time_df_fc, "ANCOMBC2_top_by_variable_time_fc.csv", row.names = FALSE)
```

### Detailed T0 vs. T12 comparison
```{r T0 vs T12 comparison, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
comparisons_list_fc <- list()
summary_counts_fc <- list()

# Define the variable to iterate over from the feces master table
vars_fc <- sort(unique(master_sig_df_fc$ClinVar))

for (v in vars_fc) {
  sig_T0  <- master_sig_df_fc %>% dplyr::filter(ClinVar==v, Time=="T0")
  sig_T12 <- master_sig_df_fc %>% dplyr::filter(ClinVar==v, Time=="T12")

  sp_T0  <- unique(sig_T0$Species)
  sp_T12 <- unique(sig_T12$Species)

  if (length(sp_T0)==0 && length(sp_T12)==0) {
    summary_counts_fc[[v]] <- data.frame(
      ClinVar=v, n_T0_sig=0, n_T12_sig=0,
      n_common_same=0, n_common_flip=0,
      n_only_T0=0, n_only_T12=0
    )
    next
  }

  common <- dplyr::inner_join(
    dplyr::select(sig_T0,  Species, lfc_T0 = lfc,  q_T0 = q,  sign_T0 = sign),
    dplyr::select(sig_T12, Species, lfc_T12 = lfc, q_T12 = q, sign_T12 = sign),
    by = "Species"
  )

  common_same <- common %>% dplyr::filter(sign_T0 %in% c("+","-") & sign_T0 == sign_T12)
  common_flip <- common %>% dplyr::filter(sign_T0 %in% c("+","-") & sign_T12 %in% c("+","-") & sign_T0 != sign_T12)

  only_T0  <- setdiff(sp_T0,  sp_T12)
  only_T12 <- setdiff(sp_T12, sp_T0)

  # --- Creating the detailed comparison tables ---
  only_T0_tab <- sig_T0  %>% dplyr::filter(Species %in% only_T0) %>%
    dplyr::transmute(ClinVar=v, Category="only_T0", Species,
                     lfc_T0=lfc, q_T0=q, sign_T0=sign,
                     lfc_T12=NA_real_, q_T12=NA_real_, sign_T12=NA_character_)

  only_T12_tab <- sig_T12 %>% dplyr::filter(Species %in% only_T12) %>%
    dplyr::transmute(ClinVar=v, Category="only_T12", Species,
                     lfc_T0=NA_real_, q_T0=NA_real_, sign_T0=NA_character_,
                     lfc_T12=lfc, q_T12=q, sign_T12=sign)

  common_same_tab <- common_same %>% dplyr::mutate(ClinVar=v, Category="common_same") %>%
    dplyr::select(ClinVar, Category, Species, lfc_T0, q_T0, sign_T0, lfc_T12, q_T12, sign_T12)

  common_flip_tab <- common_flip %>% dplyr::mutate(ClinVar=v, Category="common_flip") %>%
    dplyr::select(ClinVar, Category, Species, lfc_T0, q_T0, sign_T0, lfc_T12, q_T12, sign_T12)

  comparisons_list_fc[[v]] <- dplyr::bind_rows(common_same_tab, common_flip_tab, only_T0_tab, only_T12_tab)

  summary_counts_fc[[v]] <- data.frame(
    ClinVar = v,
    n_T0_sig  = length(sp_T0),
    n_T12_sig = length(sp_T12),
    n_common_same = nrow(common_same),
    n_common_flip = nrow(common_flip),
    n_only_T0  = length(only_T0),
    n_only_T12 = length(only_T12)
  )
}

# Combine the lists into final data frames
comparisons_df_fc   <- if (length(comparisons_list_fc)) dplyr::bind_rows(comparisons_list_fc) else data.frame()
summary_counts_df_fc <- if (length(summary_counts_fc))   dplyr::bind_rows(summary_counts_fc)   else data.frame()

# Save the results to CSV files
if (nrow(comparisons_df_fc)) write.csv(comparisons_df_fc, "ANCOMBC2_T0vsT12_comparisons_fc.csv", row.names=FALSE)
if (nrow(summary_counts_df_fc)) write.csv(summary_counts_df_fc, "ANCOMBC2_T0vsT12_summary_counts_fc.csv", row.names=FALSE)
```

### Visualization
#### Forest plots
##### Prepare data for forest plots
```{r Prepare data, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
forest_plot_data_list_fc <- list()
alpha <- 0.05

for (it in items_fc) {
  if (!exists(it$obj)) next
  out <- get(it$obj)
  res <- out$res

  lfc_col  <- paste0("lfc_",  it$term)
  q_col    <- paste0("q_",    it$term)
  se_col   <- paste0("se_",   it$term)
  diff_col <- paste0("diff_", it$term)

  if (!all(c(lfc_col, q_col, se_col, diff_col) %in% names(res))) next

  temp_df <- data.frame(
    Taxon = res$taxon,
    Time = it$time,
    ClinVar = it$clin,
    lfc = res[[lfc_col]],
    se  = res[[se_col]],
    q   = res[[q_col]],
    diff = as.logical(res[[diff_col]])
  )
  
  forest_plot_data_list_fc[[it$obj]] <- temp_df
}

forest_plot_data_full_fc <- bind_rows(forest_plot_data_list_fc) %>%
  mutate(Species = sub(".*\\|s__", "", Taxon))
```

##### Generate and save Forest plots
```{r Generate and save Forest plots, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
dir.create("figures_svg_fc", showWarnings = FALSE)

for (variable_to_plot in unique(forest_plot_data_full_fc$ClinVar)) {

  top_species_for_var_fc <- forest_plot_data_full_fc %>%
    filter(ClinVar == variable_to_plot) %>%
    filter(!is.na(q) & q < alpha & diff == TRUE) %>%
    arrange(q) %>%
    distinct(Species) %>%
    slice_head(n = 20) %>%
    pull(Species)

  if (length(top_species_for_var_fc) == 0) {
    print(paste("No significant species found for", variable_to_plot, "- skipping plot."))
    next
  }
  
  plot_data_fc <- forest_plot_data_full_fc %>%
    filter(ClinVar == variable_to_plot, Species %in% top_species_for_var_fc) %>%
    mutate(
      lfc_lower = lfc - 1.96 * se,
      lfc_upper = lfc + 1.96 * se,
      is_significant = !is.na(q) & q < alpha & diff == TRUE,
      Species = factor(Species, levels = rev(unique(Species)))
    )

  p_fc <- ggplot(plot_data_fc, aes(y = Species, x = lfc, xmin = lfc_lower, xmax = lfc_upper, color = is_significant)) +
    geom_vline(xintercept = 0, linetype="dashed", color="grey70") +
    geom_errorbar(width = 0.2, linewidth = 0.8) +
    geom_point(size = 2.5) +
    facet_grid(. ~ Time) +
    scale_color_manual(
      name = "Significance", 
      values = c(`TRUE` = "#d95f02", `FALSE` = "grey70"),
      labels = c(`TRUE` = "Significative", `FALSE` = "Not significative")
    ) +
    labs(
      title = paste("Forest Plot:", variable_to_plot, "in Feces"),
      x = "LFC (coefficient)", 
      y = NULL
    ) +
    theme_bw(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.y = element_text(face = "italic"),
      strip.background = element_rect(fill = "grey90"),
      legend.position = "right"
    )

  plot_height_fc <- max(4, 0.3 * length(top_species_for_var_fc))
  
  ggsave(
    filename = file.path("figures_svg_fc", paste0("Faceted_ForestPlot_", variable_to_plot, "_fc.svg")),
    plot = p_fc,
    device = svglite::svglite,
    width = 8,
    height = plot_height_fc
  )
}
```



# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------
# SPUTUM SAMPLES ANALYSIS
# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------

# ------------------------------------------------------------------------------
# LUNG FUNCTION
# ------------------------------------------------------------------------------

### Run ANCOM-BC2 at T0 and T12
```{r ppFEV1 sputum, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
out_ppFEV1_sp_T0 <- ancombc2(
  data         = subset_sputum_T0_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "ppFEV1 + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)

# T12
out_ppFEV1_sp_T12 <- ancombc2(
  data         = subset_sputum_T12_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "ppFEV1 + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)
```

### Export the results
```{r Results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
## All
res_all_ppFEV1_sp_T0 <- out_ppFEV1_sp_T0$res
write_csv(res_all_ppFEV1_sp_T0, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_all_ppFEV1_T0_sputum.csv"))

## Significant
res_sig_ppFEV1_sp_T0 <- res_all_ppFEV1_sp_T0 %>%
  filter(diff_ppFEV1 == TRUE, q_ppFEV1 < 0.05) %>%
  arrange(lfc_ppFEV1)
write_csv(res_sig_ppFEV1_sp_T0, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_sig_ppFEV1_T0_sputum.csv"))

# T12
## All
res_all_ppFEV1_sp_T12 <- out_ppFEV1_sp_T12$res
write_csv(res_all_ppFEV1_sp_T12, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_all_ppFEV1_T12_sputum.csv"))

## Significant
res_sig_ppFEV1_sp_T12 <- res_all_ppFEV1_sp_T12 %>%
  filter(diff_ppFEV1 == TRUE, q_ppFEV1 < 0.05) %>%
  arrange(lfc_ppFEV1)
write_csv(res_sig_ppFEV1_sp_T12, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_sig_ppFEV1_T12_sputum.csv"))
```


# ------------------------------------------------------------------------------
# TRANSAMINASES
# ------------------------------------------------------------------------------

## ALT

### Run ANCOM-BC2 at T0 and T12
```{r ALT sputum, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
out_ALT_sp_T0 <- ancombc2(
  data         = subset_sputum_T0_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "ALT + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)

# T12
out_ALT_sp_T12 <- ancombc2(
  data         = subset_sputum_T12_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "ALT + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)
```

### Export the results
```{r Results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
## All
res_all_ALT_sp_T0 <- out_ALT_sp_T0$res
write_csv(res_all_ALT_sp_T0, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_all_ALT_T0_sputum.csv"))

## Significant
res_sig_ALT_sp_T0 <- res_all_ALT_sp_T0 %>%
  filter(diff_ALT == TRUE, q_ALT < 0.05) %>%
  arrange(lfc_ALT)
write_csv(res_sig_ALT_sp_T0, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_sig_ALT_T0_sputum.csv"))

# T12
## All
res_all_ALT_sp_T12 <- out_ALT_sp_T12$res
write_csv(res_all_ALT_sp_T12, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_all_ALT_T12_sputum.csv"))

## Significant
res_sig_ALT_sp_T12 <- res_all_ALT_sp_T12 %>%
  filter(diff_ALT == TRUE, q_ALT < 0.05) %>%
  arrange(lfc_ALT)
write_csv(res_sig_ALT_sp_T12, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_sig_ALT_T12_sputum.csv"))
```

#------
## AST

### Run ANCOM-BC2 at T0 and T12
```{r AST sputum, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
out_AST_sp_T0 <- ancombc2(
  data         = subset_sputum_T0_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "AST + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10,
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)

# T12
out_AST_sp_T12 <- ancombc2(
  data         = subset_sputum_T12_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "AST + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10,
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)
```

### Export the results
```{r Results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
## All
res_all_AST_sp_T0 <- out_AST_sp_T0$res
write_csv(res_all_AST_sp_T0, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_all_AST_T0_sputum.csv"))

## Significant
res_sig_AST_sp_T0 <- res_all_AST_sp_T0 %>%
  filter(diff_AST == TRUE, q_AST < 0.05) %>%
  arrange(lfc_AST)
write_csv(res_sig_AST_sp_T0, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_sig_AST_T0_sputum.csv"))

# T12
## All
res_all_AST_sp_T12 <- out_AST_sp_T12$res
write_csv(res_all_AST_sp_T12, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_all_AST_T12_sputum.csv"))

## Significant
res_sig_AST_sp_T12 <- res_all_AST_sp_T12 %>%
  filter(diff_AST == TRUE, q_AST < 0.05) %>%
  arrange(lfc_AST)
write_csv(res_sig_AST_sp_T12, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_sig_AST_T12_sputum.csv"))
```

#------
## GGT

### Run ANCOM-BC2 at T0 and T12
```{r GGT sputum, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
out_GGT_sp_T0 <- ancombc2(
  data         = subset_sputum_T0_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "GGT + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)

# T12
out_GGT_sp_T12 <- ancombc2(
  data         = subset_sputum_T12_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "GGT + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)
```

### Export the results
```{r Results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
#T0
## All
res_all_GGT_sp_T0 <- out_GGT_sp_T0$res
write_csv(res_all_GGT_sp_T0, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_all_GGT_T0_sputum.csv"))

## Significant
res_sig_GGT_sp_T0 <- res_all_GGT_sp_T0 %>%
  filter(diff_GGT == TRUE, q_GGT < 0.05) %>%
  arrange(lfc_GGT)
write_csv(res_sig_GGT_sp_T0, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_sig_GGT_T0_sputum.csv"))

# T12
## All
res_all_GGT_sp_T12 <- out_GGT_sp_T12$res
write_csv(res_all_GGT_sp_T12, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_all_GGT_T12_sputum.csv"))

## Significant
res_sig_GGT_sp_T12 <- res_all_GGT_sp_T12 %>%
  filter(diff_GGT == TRUE, q_GGT < 0.05) %>%
  arrange(lfc_GGT)
write_csv(res_sig_GGT_sp_T12, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_sig_GGT_T12_sputum.csv"))
```


# ------------------------------------------------------------------------------
# ELASTASE
# ------------------------------------------------------------------------------

### Run ANCOM-BC2 at T0 and T12
```{r Elastase sputum, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
out_elas_sp_T0 <- ancombc2(
  data         = subset_sputum_T0_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "Elastase + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)

# T12
out_elas_sp_T12 <- ancombc2(
  data         = subset_sputum_T12_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "Elastase + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)
```

### Export the results
```{r Results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
## All
res_all_elas_sp_T0 <- out_elas_sp_T0$res
write_csv(res_all_elas_sp_T0, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_all_elas_T0_sputum.csv"))

## Significant
res_sig_elas_sp_T0 <- res_all_elas_sp_T0 %>%
  filter(diff_Elastase == TRUE, q_Elastase < 0.05) %>%
  arrange(lfc_Elastase)
write_csv(res_sig_elas_sp_T0, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_sig_elas_T0_sputum.csv"))

# T12
## All
res_all_elas_sp_T12 <- out_elas_sp_T12$res
write_csv(res_all_elas_sp_T12, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_all_elas_T12_sputum.csv"))

## Significant
res_sig_elas_sp_T12 <- res_all_elas_sp_T12 %>%
  filter(diff_Elastase == TRUE, q_Elastase < 0.05) %>%
  arrange(lfc_Elastase)
write_csv(res_sig_elas_sp_T12, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_sig_elas_T12_sputum.csv"))
```


# ------------------------------------------------------------------------------
# CALPROTECTIN
# ------------------------------------------------------------------------------

### Run ANCOM-BC2 at T0 and T12
```{r Calprotectin sputum, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
out_calpro_sp_T0 <- ancombc2(
  data         = subset_sputum_T0_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "Calprotectin + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)

# T12
out_calpro_sp_T12 <- ancombc2(
  data         = subset_sputum_T12_no_UN,
  assay_name   = "counts",
  tax_level    = "Species",
  fix_formula  = "Calprotectin + Age + Sex", 
  rand_formula = NULL, 
  prv_cut      = 0.10, 
  lib_cut      = 0,
  p_adj_method = "BH",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  pseudo_sens  = TRUE,
  s0_perc      = 0.05,
  verbose      = TRUE
)
```

### Export the results
```{r Results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# T0
## All
res_all_calpro_sp_T0 <- out_calpro_sp_T0$res
write_csv(res_all_calpro_sp_T0, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_all_calpro_T0_sputum.csv"))

## Significant
res_sig_calpro_sp_T0 <- res_all_calpro_sp_T0 %>%
  filter(diff_Calprotectin == TRUE, q_Calprotectin < 0.05) %>%
  arrange(lfc_Calprotectin)
write_csv(res_sig_calpro_sp_T0, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_sig_calpro_T0_sputum.csv"))

# T12
## All
res_all_calpro_sp_T12 <- out_calpro_sp_T12$res
write_csv(res_all_calpro_sp_T12, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_all_calpro_T12_sputum.csv"))

## Significant
res_sig_calpro_sp_T12 <- res_all_calpro_sp_T12 %>%
  filter(diff_Calprotectin == TRUE, q_Calprotectin < 0.05) %>%
  arrange(lfc_Calprotectin)
write_csv(res_sig_calpro_sp_T12, here("results_ancombc_adjusted_sputum", "tables", "ANCOMBC2_sig_calpro_T12_sputum.csv"))
```


# ------------------------------------------------------------------------------
## Master Table Consolidation
# ------------------------------------------------------------------------------

### Create Master table
```{r Master Table, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# List of result objects to consolidate
items_sp <- list(
  list(obj = "out_ppFEV1_sp_T0", term = "ppFEV1",      time = "T0",  clin = "ppFEV1"),
  list(obj = "out_ppFEV1_sp_T12",term = "ppFEV1",      time = "T12", clin = "ppFEV1"),
  list(obj = "out_ALT_sp_T0",    term = "ALT",         time = "T0",  clin = "ALT"),
  list(obj = "out_ALT_sp_T12",   term = "ALT",         time = "T12", clin = "ALT"),
  list(obj = "out_AST_sp_T0",    term = "AST",         time = "T0",  clin = "AST"),
  list(obj = "out_AST_sp_T12",   term = "AST",         time = "T12", clin = "AST"),
  list(obj = "out_GGT_sp_T0",    term = "GGT",         time = "T0",  clin = "GGT"),
  list(obj = "out_GGT_sp_T12",   term = "GGT",         time = "T12", clin = "GGT"),
  list(obj = "out_elas_sp_T0",   term = "Elastase",    time = "T0",  clin = "Elastase"),
  list(obj = "out_elas_sp_T12",  term = "Elastase",    time = "T12", clin = "Elastase"),
  list(obj = "out_calpro_sp_T0", term = "Calprotectin",time = "T0",  clin = "Calprotectin"),
  list(obj = "out_calpro_sp_T12",term = "Calprotectin",time = "T12", clin = "Calprotectin")
)

master_all_sp <- list()  
master_sig_sp <- list()   
alpha <- 0.05         

for (it in items_sp) {
  if (!exists(it$obj)) next
  out <- get(it$obj)
  res <- out$res

  lfc_col  <- paste0("lfc_",  it$term)
  q_col    <- paste0("q_",    it$term)
  diff_col <- paste0("diff_", it$term)

  # Skip if expected columns aren't present
  if (!all(c(lfc_col, q_col, diff_col) %in% names(res))) next

  # Create a standardized table
  tab_sp <- data.frame(
    Taxon = res$taxon,
    Species = if ("taxon" %in% names(res)) sub(".*\\|s__", "", res$taxon) else NA_character_,
    Time = it$time,
    ClinVar = it$clin,
    lfc = res[[lfc_col]],
    q   = res[[q_col]],
    diff = as.logical(res[[diff_col]]),
    stringsAsFactors = FALSE
  )
  tab_sp$sign <- ifelse(is.na(tab_sp$lfc), NA, ifelse(tab_sp$lfc > 0, "+", ifelse(tab_sp$lfc < 0, "-", "0")))

  master_all_sp[[paste(it$clin, it$time, sep = "_")]] <- tab_sp

  # Filter for significant results
  sig_sp <- tab_sp[!is.na(tab_sp$q) & tab_sp$q < alpha & tab_sp$diff, , drop = FALSE]
  if (nrow(sig_sp) > 0) sig_sp <- sig_sp[order(abs(sig_sp$lfc), decreasing = TRUE), ]
  master_sig_sp[[paste(it$clin, it$time, sep = "_")]] <- sig_sp
}

# Combine lists into single data frames
master_all_df_sp <- if (length(master_all_sp)) do.call(rbind, master_all_sp) else data.frame()
master_sig_df_sp <- if (length(master_sig_sp)) do.call(rbind, master_sig_sp) else data.frame()

# (Optional) Save to CSV files
if (nrow(master_all_df_sp)) write.csv(master_all_df_sp, "ANCOMBC2_master_all_sp.csv", row.names = FALSE)
if (nrow(master_sig_df_sp)) write.csv(master_sig_df_sp, "ANCOMBC2_master_sig_sp.csv", row.names = FALSE)

# Quick summary
cat("Rows in master_all_df_sp:", nrow(master_all_df_sp), "\n")
cat("Rows in master_sig_df_sp:", nrow(master_sig_df_sp), "\n")
```

### Summarize top taxa
```{r Top taxa by clinical variable, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# A) Top taxa by clinical variable (T0 and T12 combined)
top_by_variable_sp <- list()
if (nrow(master_sig_df_sp)) {
  vars <- unique(master_sig_df_sp$ClinVar)
  for (v in vars) {
    sub <- master_sig_df_sp[master_sig_df_sp$ClinVar == v, , drop = FALSE]
    if (nrow(sub) == 0) next
    sub_ord <- sub[order(abs(sub$lfc), decreasing = TRUE), ]
    top_by_variable_sp[[v]] <- head(sub_ord, 20)
  }
}
top_by_variable_df_sp <- if (length(top_by_variable_sp)) do.call(rbind, top_by_variable_sp) else data.frame()

# B) Top taxa by clinical variable AND timepoint (T0 and T12 separately)
top_by_variable_time_sp <- list()
if (nrow(master_sig_df_sp)) {
  combos <- unique(master_sig_df_sp[, c("ClinVar","Time")])
  for (i in seq_len(nrow(combos))) {
    v <- combos$ClinVar[i]; t <- combos$Time[i]
    sub <- master_sig_df_sp[master_sig_df_sp$ClinVar == v & master_sig_df_sp$Time == t, , drop = FALSE]
    if (nrow(sub) == 0) next
    sub_ord <- sub[order(abs(sub$lfc), decreasing = TRUE), ]
    top_by_variable_time_sp[[paste(v, t, sep = "_")]] <- head(sub_ord, 20)
  }
}
top_by_variable_time_df_sp <- if (length(top_by_variable_time_sp)) do.call(rbind, top_by_variable_time_sp) else data.frame()

# Save summaries to CSV
if (nrow(top_by_variable_df_sp))       write.csv(top_by_variable_df_sp,       "ANCOMBC2_top_by_variable_sp.csv", row.names = FALSE)
if (nrow(top_by_variable_time_df_sp)) write.csv(top_by_variable_time_df_sp, "ANCOMBC2_top_by_variable_time_sp.csv", row.names = FALSE)
```

### Detailed T0 vs. T12 comparison
```{r T0 vs T12 comparison, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
comparisons_list_sp <- list()
summary_counts_sp <- list()

# Define the variable to iterate over from the sputum master table
vars_sp <- sort(unique(master_sig_df_sp$ClinVar))

for (v in vars_sp) {
  sig_T0  <- master_sig_df_sp %>% dplyr::filter(ClinVar==v, Time=="T0")
  sig_T12 <- master_sig_df_sp %>% dplyr::filter(ClinVar==v, Time=="T12")

  sp_T0  <- unique(sig_T0$Species)
  sp_T12 <- unique(sig_T12$Species)

  if (length(sp_T0)==0 && length(sp_T12)==0) {
    summary_counts_sp[[v]] <- data.frame(
      ClinVar=v, n_T0_sig=0, n_T12_sig=0,
      n_common_same=0, n_common_flip=0,
      n_only_T0=0, n_only_T12=0
    )
    next
  }

  common <- dplyr::inner_join(
    dplyr::select(sig_T0,  Species, lfc_T0 = lfc,  q_T0 = q,  sign_T0 = sign),
    dplyr::select(sig_T12, Species, lfc_T12 = lfc, q_T12 = q, sign_T12 = sign),
    by = "Species"
  )

  common_same <- common %>% dplyr::filter(sign_T0 %in% c("+","-") & sign_T0 == sign_T12)
  common_flip <- common %>% dplyr::filter(sign_T0 %in% c("+","-") & sign_T12 %in% c("+","-") & sign_T0 != sign_T12)

  only_T0  <- setdiff(sp_T0,  sp_T12)
  only_T12 <- setdiff(sp_T12, sp_T0)

  # --- Creating the detailed comparison tables ---
  only_T0_tab <- sig_T0  %>% dplyr::filter(Species %in% only_T0) %>%
    dplyr::transmute(ClinVar=v, Category="only_T0", Species,
                     lfc_T0=lfc, q_T0=q, sign_T0=sign,
                     lfc_T12=NA_real_, q_T12=NA_real_, sign_T12=NA_character_)

  only_T12_tab <- sig_T12 %>% dplyr::filter(Species %in% only_T12) %>%
    dplyr::transmute(ClinVar=v, Category="only_T12", Species,
                     lfc_T0=NA_real_, q_T0=NA_real_, sign_T0=NA_character_,
                     lfc_T12=lfc, q_T12=q, sign_T12=sign)

  common_same_tab <- common_same %>% dplyr::mutate(ClinVar=v, Category="common_same") %>%
    dplyr::select(ClinVar, Category, Species, lfc_T0, q_T0, sign_T0, lfc_T12, q_T12, sign_T12)

  common_flip_tab <- common_flip %>% dplyr::mutate(ClinVar=v, Category="common_flip") %>%
    dplyr::select(ClinVar, Category, Species, lfc_T0, q_T0, sign_T0, lfc_T12, q_T12, sign_T12)

  comparisons_list_sp[[v]] <- dplyr::bind_rows(common_same_tab, common_flip_tab, only_T0_tab, only_T12_tab)

  summary_counts_sp[[v]] <- data.frame(
    ClinVar = v,
    n_T0_sig  = length(sp_T0),
    n_T12_sig = length(sp_T12),
    n_common_same = nrow(common_same),
    n_common_flip = nrow(common_flip),
    n_only_T0  = length(only_T0),
    n_only_T12 = length(only_T12)
  )
}

# Combine the lists into final data frames
comparisons_df_sp   <- if (length(comparisons_list_sp)) dplyr::bind_rows(comparisons_list_sp) else data.frame()
summary_counts_df_sp <- if (length(summary_counts_sp))   dplyr::bind_rows(summary_counts_sp)   else data.frame()

# Save the results to CSV files
if (nrow(comparisons_df_sp)) write.csv(comparisons_df_sp, "ANCOMBC2_T0vsT12_comparisons_sp.csv", row.names=FALSE)
if (nrow(summary_counts_df_sp)) write.csv(summary_counts_df_sp, "ANCOMBC2_T0vsT12_summary_counts_sp.csv", row.names=FALSE)
```

### Visualization
#### Forest plots
##### Prepare data for forest plots
```{r Prepare data, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
forest_plot_data_list_sp <- list()
alpha <- 0.05

for (it in items_sp) {
  if (!exists(it$obj)) next
  out <- get(it$obj)
  res <- out$res

  lfc_col  <- paste0("lfc_",  it$term)
  q_col    <- paste0("q_",    it$term)
  se_col   <- paste0("se_",   it$term)
  diff_col <- paste0("diff_", it$term)

  if (!all(c(lfc_col, q_col, se_col, diff_col) %in% names(res))) next

  temp_df <- data.frame(
    Taxon = res$taxon,
    Time = it$time,
    ClinVar = it$clin,
    lfc = res[[lfc_col]],
    se  = res[[se_col]],
    q   = res[[q_col]],
    diff = as.logical(res[[diff_col]])
  )
  
  forest_plot_data_list_sp[[it$obj]] <- temp_df
}

forest_plot_data_full_sp <- bind_rows(forest_plot_data_list_sp) %>%
  mutate(Species = sub(".*\\|s__", "", Taxon))
```

##### Generate and save Forest plots
```{r Generate and save Forest plots, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
dir.create("figures_svg_sp", showWarnings = FALSE)

for (variable_to_plot in unique(forest_plot_data_full_sp$ClinVar)) {

  top_species_for_var_sp <- forest_plot_data_full_sp %>%
    filter(ClinVar == variable_to_plot) %>%
    filter(!is.na(q) & q < alpha & diff == TRUE) %>%
    arrange(q) %>%
    distinct(Species) %>%
    slice_head(n = 20) %>%
    pull(Species)

  if (length(top_species_for_var_sp) == 0) {
    print(paste("No significant species found for", variable_to_plot, "- skipping plot."))
    next
  }
  
  plot_data_sp <- forest_plot_data_full_sp %>%
    filter(ClinVar == variable_to_plot, Species %in% top_species_for_var_sp) %>%
    mutate(
      lfc_lower = lfc - 1.96 * se,
      lfc_upper = lfc + 1.96 * se,
      is_significant = !is.na(q) & q < alpha & diff == TRUE,
      Species = factor(Species, levels = rev(unique(Species)))
    )

  p_sp <- ggplot(plot_data_sp, aes(y = Species, x = lfc, xmin = lfc_lower, xmax = lfc_upper, color = is_significant)) +
    geom_vline(xintercept = 0, linetype="dashed", color="grey70") +
    geom_errorbar(width = 0.2, linewidth = 0.8) +
    geom_point(size = 2.5) +
    facet_grid(. ~ Time) +
    scale_color_manual(
      name = "Significance", 
      values = c(`TRUE` = "#d95f02", `FALSE` = "grey70"),
      labels = c(`TRUE` = "Significative", `FALSE` = "Not significative")
    ) +
    labs(
      title = paste("Forest Plot:", variable_to_plot, "in Sputum"),
      x = "LFC (coefficient)", 
      y = NULL
    ) +
    theme_bw(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.y = element_text(face = "italic"),
      strip.background = element_rect(fill = "grey90"),
      legend.position = "right"
    )

  plot_height_sp <- max(4, 0.3 * length(top_species_for_var_sp))
  
  ggsave(
    filename = file.path("figures_svg_sp", paste0("Faceted_ForestPlot_", variable_to_plot, "_sp.svg")),
    plot = p_sp,
    device = svglite::svglite,
    width = 8,
    height = plot_height_sp
  )
}
```

