---
title: "cf_clinical_variables_thesis"
author: "Natalia Bastón-Paz"
date: '2026'
output: html_document
---

# Set up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require("knitr")
opts_knit$set(root.dir = "C:/RProjects/cf_clinical_variables_thesis")
```

# Clean environments
```{# r Clean environments, message=FALSE, warning=FALSE}
rm(list = ls())
```

## General introduction

Cystic fibrosis (CF) is a genetic disorder caused by pathogenic variants in the CFTR (Cystic Fibrosis Transmembrane Conductance Regulator) gene. CFTR dysfunction leads to impaired ion transport and the production of thick mucus, primarily affecting the lungs and gastrointestinal tract. People with CF (pwCF) commonly develop chronic airway infections and gastrointestinal complications.

In recent years, CFTR modulator therapies—particularly the triple combination elexacaftor-tezacaftor-ivacaftor (ETI) have markedly improved clinical outcomes, especially respiratory function and nutritional status.

In this script, we analyze clinical variables from 35 people with CF (6–66 years) assessed before treatment (T0) and after 12 months of ETI (T12).

*Main question:*

-   *Have clinical variables changed after one year of ETI treatment?*

## Libraries
```{r Libraries, message=FALSE, warning=FALSE}
library(here)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggdist)
library(purrr)
library(stringr)
library(patchwork)
library(ggpubr)
library(ggridges)
library(ppcor)
library(lme4)
library(lmerTest)
library(grid)
library(stringr)
library(forcats)
library(scales)
library(gtools)
library(emmeans)
library(tibble)
library(broom.mixed)
library(ggbreak)
library(cowplot)
library(lme4)
library(lmerTest)
library(rlang)
```

## Directories
```{r Create directory, message=FALSE, warning=FALSE}
dir.create(here("results"), showWarnings = FALSE)
```


# ------------------------------------------------------------------------------
# DEMOGRAPHICS
# ------------------------------------------------------------------------------

### Load in the data
```{r Load in data lung function, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_demo <- read_delim(here("data", "only_demographics_all.csv"))
```

### Funtion to export statistics
```{r Statistics, echo=TRUE, results='show', message=FALSE, warning=FALSE}
df_base <- df_demo %>%
  dplyr::mutate(Time = factor(Time, levels = c("T0","T12"))) %>%
  dplyr::group_by(PatientID) %>%
  dplyr::filter(dplyr::n_distinct(Time) == 2) %>%   
  dplyr::ungroup() %>%
  dplyr::filter(Time == "T0")                       

n_patients <- dplyr::n_distinct(df_base$PatientID)


fmt_n_pct <- function(x, n_total) {
  n <- sum(!is.na(x))
  if (n == 0) return(NA_character_)
  tb <- table(x, useNA = "no")
  out <- paste0(names(tb), ": ", as.integer(tb), " (", round(100*as.integer(tb)/n_total, 1), "%)")
  paste(out, collapse = "; ")
}

fmt_mean_sd <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  sprintf("%.1f ± %.1f", mean(x), stats::sd(x))
}

fmt_median_iqr <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  q <- stats::quantile(x, c(0.25, 0.5, 0.75))
  sprintf("%.1f [%.1f–%.1f]", q[2], q[1], q[3])
}

# Normality test (Age at T0; one value per patient)
age_vec <- df_base$Age
age_vec <- age_vec[!is.na(age_vec)]

if (length(age_vec) >= 3) {
  shapiro_age <- shapiro.test(age_vec)
  shapiro_p_age <- unname(shapiro_age$p.value)
  age_is_normal <- shapiro_p_age >= 0.05
} else {
  shapiro_age <- NULL
  shapiro_p_age <- NA_real_
  age_is_normal <- NA
}

cat("\n=== Normality test (Age, T0) ===\n")
cat("n =", length(age_vec), "\n")

if (!is.null(shapiro_age)) {
  print(shapiro_age)
  if (isTRUE(age_is_normal)) {
    cat("Age distribution: Normal (p =", round(shapiro_p_age, 4), ")\n")
  } else {
    cat("Age distribution: Not normal (p =", round(shapiro_p_age, 4), ")\n")
  }
} else {
  cat("Not enough values for Shapiro test (n < 3)\n")
}


baseline_stats <- tibble::tibble(
  variable = c(
    "N (paired patients)",
    "Age (years), Shapiro–Wilk p-value",
    "Age (years), mean ± SD",
    "Age (years), median [IQR]",
    "AgeGroup, n (%)",
    "Sex, n (%)",
    "MutationGroup, n (%)",
    "Mutation1, n (%)",
    "Mutation2, n (%)",
    "PI, n (%)",
    "CFLD, n (%)",
    "CFRD, n (%)",
    "coeliac_disease, n (%)",
    "colon_carcinoma, n (%)",
    "mCFTRGroup, n (%)"
  ),
  value = c(
    as.character(n_patients),
    ifelse(is.na(shapiro_p_age), NA_character_, signif(shapiro_p_age, 4)),
    fmt_mean_sd(df_base$Age),
    fmt_median_iqr(df_base$Age),
    fmt_n_pct(df_base$AgeGroup, n_patients),
    fmt_n_pct(df_base$Sex, n_patients),
    fmt_n_pct(df_base$MutationGroup, n_patients),
    fmt_n_pct(df_base$Mutation1, n_patients),
    fmt_n_pct(df_base$Mutation2, n_patients),
    fmt_n_pct(df_base$PI, n_patients),
    fmt_n_pct(df_base$CFLD, n_patients),
    fmt_n_pct(df_base$CFRD, n_patients),
    fmt_n_pct(df_base$coeliac_disease, n_patients),
    fmt_n_pct(df_base$colon_carcinoma, n_patients),
    fmt_n_pct(df_base$mCFTRGroup, n_patients)
  )
)

baseline_stats

if (!dir.exists(here::here("results"))) dir.create(here::here("results"))
readr::write_excel_csv(baseline_stats, here::here("results", "baseline_characteristics_T0.csv"))
```

# ------------------------------------------------------------------------------
# ADVERSE EVENTS
# ------------------------------------------------------------------------------

### Load in the data
```{r , echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_ae <- read_delim(here("data", "adv_events.csv"))
```

### Clean the table and prepare data
```{r Clean the table and prepare data, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Clean the table
df_ae_clean_ae <- df_ae %>%
  mutate(
    Adv_events = trimws(Adv_events),
    Which_AE = na_if(trimws(Which_AE), "NA"),
    Which_AE = na_if(Which_AE, ""),
    AgeGroup = factor(AgeGroup, levels = c("pediatric", "adult"))
  )

# Denominator: unique patients per age group
denom_ae <- df_ae_clean_ae %>%
  distinct(PatientID, AgeGroup) %>%
  count(AgeGroup, name = "n_total_ae")

# Numerator: unique patients with each side effect in each age group
summary_ae <- df_ae_clean_ae %>%
  filter(Adv_events == "Yes", !is.na(Which_AE)) %>%
  distinct(PatientID, AgeGroup, Which_AE) %>%
  count(AgeGroup, Which_AE, name = "n_patients_ae") %>%
  complete(AgeGroup, Which_AE, fill = list(n_patients_ae = 0)) %>%
  left_join(denom_ae, by = "AgeGroup") %>%
  mutate(percent_ae = 100 * n_patients_ae / n_total_ae)

# Order side effects by total frequency
order_ae <- summary_ae %>%
  group_by(Which_AE) %>%
  summarise(total_n_ae = sum(n_patients_ae), .groups = "drop") %>%
  arrange(desc(total_n_ae)) %>%
  pull(Which_AE)

summary_ae <- summary_ae %>%
  mutate(
    Which_AE = factor(Which_AE, levels = order_ae),
    Which_AE_label = str_replace_all(as.character(Which_AE), "_", " "),
    Which_AE_label = str_to_title(Which_AE_label)
  )
```

#### Export results
```{r Export results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
summary_ae_export <- summary_ae %>%
  dplyr::select(
    AgeGroup,
    Which_AE,
    Which_AE_label,
    n_patients_ae,
    n_total_ae,
    percent_ae
  ) %>%
  dplyr::arrange(Which_AE, AgeGroup)

readr::write_csv(
  summary_ae_export,
  here("results", "adv_events_summary_ae_export.csv")
)
```

### Plot
```{r Plot, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
p_ae <- ggplot(
  summary_ae,
  aes(
    x = fct_reorder(Which_AE_label, n_patients_ae, .fun = sum),
    y = percent_ae,
    fill = AgeGroup
  )
) +
  geom_col(
    position = position_dodge2(width = 0.8, preserve = "single"),
    width = 0.7,
    color = "black",
    linewidth = 0.25
  ) +
  geom_text(
    aes(label = ifelse(n_patients_ae == 0, "", paste0(n_patients_ae, "/", n_total_ae))),
    position = position_dodge2(width = 0.8, preserve = "single"),
    hjust = -0.15,
    size = 3
  ) +
  scale_fill_manual(
    values = c(
      "pediatric" = "#d3d4e0",  
      "adult"     = "#999ab1"   
    )
  ) +
  scale_y_continuous(
    limits = c(0, max(summary_ae$percent_ae) * 1.20),
    labels = label_number(suffix = "%"),
    expand = expansion(mult = c(0, 0.05))
  ) +
  coord_flip() +
  labs(
    title = "Adverse events reported after ETI initiation",
    x = "Adverse event",
    y = "Participants with adverse events (%)",
    fill = "Age group"
  ) +
  theme_classic() +
  theme(
    axis.text.y = element_text(size = 11),
    plot.title = element_text(face = "plain"),
    legend.title = element_text(face = "plain"),
    legend.text = element_text(face = "plain"),
    panel.grid.major = element_line(color = "grey85", linewidth = 0.4),
    panel.grid.minor = element_line(color = "grey92", linewidth = 0.25)
  )

p_ae

ggsave(here("results","adv_events.svg"), p_ae, width = 8, height = 5, bg = "white")
```


# ------------------------------------------------------------------------------
# ANTIBIOTICS
# ------------------------------------------------------------------------------

### Load in the data
```{r Load in data ATB, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_atb <- read_delim(here("data", "ATB.csv"))
```

### Prepare the data
```{r Prepare data, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_atb$Time <- factor(df_atb$Time, levels = c("T0", "T12"))

vars_atb <- c("ATB_inh_oral", "ATB_azi")

df_wide <- df_atb %>%
  dplyr::select(PatientID, Time, dplyr::all_of(vars_atb)) %>%
  tidyr::pivot_wider(
    id_cols = PatientID,
    names_from = Time,
    values_from = dplyr::all_of(vars_atb),
    names_sep = "."
  )
```

### Function exact McNemar
```{r Function exact McNemar, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
mcnemar_exact_result <- function(data, varname) {
  
  col_t0  <- paste0(varname, ".T0")
  col_t12 <- paste0(varname, ".T12")
  
  x_t0  <- data[[col_t0]]
  x_t12 <- data[[col_t12]]
  
  keep <- complete.cases(x_t0, x_t12)
  x_t0  <- x_t0[keep]
  x_t12 <- x_t12[keep]
  
  tab <- table(
    factor(x_t0, levels = c(0, 1)),
    factor(x_t12, levels = c(0, 1)),
    dnn = c("T0", "T12")
  )
  
  b <- unname(tab["1", "0"])   # yes in T0, no in T12
  c <- unname(tab["0", "1"])   # no in T0, yes in T12
  
  p_exact <- if ((b + c) == 0) 1 else binom.test(b, b + c, p = 0.5)$p.value
  
  data.frame(
    variable  = varname,
    n         = sum(keep),
    T0_yes    = sum(x_t0 == 1),
    T12_yes   = sum(x_t12 == 1),
    yes_yes   = unname(tab["1", "1"]),
    yes_no    = b,
    no_yes    = c,
    no_no     = unname(tab["0", "0"]),
    p_exact   = p_exact
  )
}
```

#### Run function exact McNemar
```{r Run function, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
results_atb <- dplyr::bind_rows(
  mcnemar_exact_result(df_wide, "ATB_inh_oral"),
  mcnemar_exact_result(df_wide, "ATB_azi")
)

results_atb$p_holm <- p.adjust(results_atb$p_exact, method = "holm")

results_atb <- results_atb %>%
  dplyr::mutate(
    variable_label = dplyr::recode(
      variable,
      "ATB_inh_oral" = "Inhaled/oral antibiotics",
      "ATB_azi"      = "Azithromycin"
    ),
    label_stats = paste0(
      "Holm p = ", signif(p_holm, 3),
      "\nYes→No = ", yes_no,
      "\nNo→Yes = ", no_yes
    )
  )

results_atb

results_atb_export <- results_atb %>%
  dplyr::mutate(
    T0_pct = round(100 * T0_yes / n, 1),
    T12_pct = round(100 * T12_yes / n, 1),
    T0_n_pct = paste0(T0_yes, "/", n, " (", T0_pct, "%)"),
    T12_n_pct = paste0(T12_yes, "/", n, " (", T12_pct, "%)")
  ) %>%
  dplyr::select(
    variable_label,
    n,
    T0_n_pct,
    T12_n_pct,
    yes_yes,
    yes_no,
    no_yes,
    no_no,
    p_exact,
    p_holm
  )

readr::write_csv(results_atb_export, here::here("results", "ATB_T0_vs_T12_stats_HOLM.csv"))
```

### Prepare data for plot
```{r Data for plot, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
plot_df_atb <- df_atb %>%
  dplyr::select(PatientID, Time, dplyr::all_of(vars_atb)) %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(vars_atb),
    names_to = "variable",
    values_to = "value"
  ) %>%
  dplyr::group_by(variable, Time) %>%
  dplyr::summarise(
    n_yes   = sum(value == 1, na.rm = TRUE),
    n_total = sum(!is.na(value)),
    prop    = n_yes / n_total,
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    variable_label = dplyr::recode(
      variable,
      "ATB_inh_oral" = "Inhaled/oral antibiotics",
      "ATB_azi"      = "Azithromycin"
    )
  )

ann_df_atb <- results_atb %>%
  dplyr::select(variable_label, label_stats) %>%
  dplyr::mutate(
    x = 1.5,
    y = 1.08
  )
```

#### Plot
```{r Plot, echo=TRUE, results='show', message=FALSE, warning=FALSE}
p_atb <- ggplot(plot_df_atb, aes(x = Time, y = prop, fill = Time)) +
  geom_col(width = 0.65, color = "black") +
  geom_text(
    aes(label = paste0(n_yes, "/", n_total)),
    vjust = -0.35,
    size = 4
  ) +
  geom_text(
    data = ann_df_atb,
    aes(x = x, y = y, label = label_stats),
    inherit.aes = FALSE,
    size = 3.5,
    lineheight = 1.05
  ) +
  facet_wrap(~ variable_label, nrow = 1) +
  scale_fill_manual(values = c("T0" = "#9EC1CF", "T12" = "#F9A1BC")) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1.18),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(
    x = NULL,
    y = "% participants exposed"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "none",
    strip.background = element_blank(),
    panel.grid.major = element_line(color = "grey80", linewidth = 0.5),
    panel.grid.minor = element_line(color = "grey90", linewidth = 0.3)
  )

p_atb

ggsave(filename = here("results", "ATB_T0_vs_T12_plot_separate.svg"),
       plot = p_atb, width = 8, height = 4.5, dpi = 300)


p_atb <- ggplot(plot_df_atb, aes(x = Time, y = prop, fill = Time)) +
  geom_col(width = 0.65, color = "black") +
  geom_text(
    aes(label = paste0(n_yes, "/", n_total)),
    vjust = -0.35,
    size = 4
  ) +
  geom_text(
    data = ann_df_atb,
    aes(x = x, y = y, label = label_stats),
    inherit.aes = FALSE,
    size = 3.5,
    lineheight = 1.05
  ) +
  facet_wrap(~ variable_label, nrow = 1) +
  scale_fill_manual(values = c("T0" = "#9EC1CF", "T12" = "#F9A1BC")) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1.18),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(
    x = NULL,
    y = "% participants exposed"
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "none",
    strip.background = element_blank(),
    panel.grid.major = element_line(color = "grey85", linewidth = 0.5),
    panel.grid.minor = element_line(color = "grey92", linewidth = 0.3)
  )

p_atb

ggsave(filename = here("results", "ATB_T0_vs_T12_plot_join.svg"),
       plot = p_atb, width = 8, height = 4.5, dpi = 300)
```

#### HOLM vs FDR
```{r Check Holm vs FDR, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
results_atb <- dplyr::bind_rows(
  mcnemar_exact_result(df_wide, "ATB_inh_oral"),
  mcnemar_exact_result(df_wide, "ATB_azi")
)

results_atb$p_fdr <- p.adjust(results_atb$p_exact, method = "fdr")

results_atb <- results_atb %>%
  dplyr::mutate(
    variable_label = dplyr::recode(
      variable,
      "ATB_inh_oral" = "Inhaled/oral antibiotics",
      "ATB_azi"      = "Azithromycin"
    ),
    label_stats = paste0(
      "FDR p = ", signif(p_fdr, 3),
      "\nYes→No = ", yes_no,
      "\nNo→Yes = ", no_yes
    )
  )

results_atb

results_atb_export <- results_atb %>%
  dplyr::mutate(
    T0_pct = round(100 * T0_yes / n, 1),
    T12_pct = round(100 * T12_yes / n, 1),
    T0_n_pct = paste0(T0_yes, "/", n, " (", T0_pct, "%)"),
    T12_n_pct = paste0(T12_yes, "/", n, " (", T12_pct, "%)")
  ) %>%
  dplyr::select(
    variable_label,
    n,
    T0_n_pct,
    T12_n_pct,
    yes_yes,
    yes_no,
    no_yes,
    no_no,
    p_exact,
    p_fdr
  )

readr::write_csv(results_atb_export, here::here("results", "ATB_T0_vs_T12_stats_FDR.csv"))
```

```{r Export results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
results_atb <- dplyr::bind_rows(
  mcnemar_exact_result(df_wide, "ATB_inh_oral"),
  mcnemar_exact_result(df_wide, "ATB_azi")
)

results_atb$p_holm <- p.adjust(results_atb$p_exact, method = "holm")
results_atb$p_fdr  <- p.adjust(results_atb$p_exact, method = "fdr")

results_atb <- results_atb %>%
  dplyr::mutate(
    variable_label = dplyr::recode(
      variable,
      "ATB_inh_oral" = "Inhaled/oral antibiotics",
      "ATB_azi"      = "Azithromycin"
    ),
    T0_pct = round(100 * T0_yes / n, 1),
    T12_pct = round(100 * T12_yes / n, 1),
    T0_n_pct = paste0(T0_yes, "/", n, " (", T0_pct, "%)"),
    T12_n_pct = paste0(T12_yes, "/", n, " (", T12_pct, "%)"),
    label_stats_holm = paste0(
      "Holm p = ", signif(p_holm, 3),
      "\nYes→No = ", yes_no,
      "\nNo→Yes = ", no_yes
    ),
    label_stats_fdr = paste0(
      "FDR p = ", signif(p_fdr, 3),
      "\nYes→No = ", yes_no,
      "\nNo→Yes = ", no_yes
    )
  )

results_atb_export <- results_atb %>%
  dplyr::select(
    variable,
    variable_label,
    n,
    T0_yes,
    T12_yes,
    T0_n_pct,
    T12_n_pct,
    yes_yes,
    yes_no,
    no_yes,
    no_no,
    p_exact,
    p_holm,
    p_fdr
  )

readr::write_csv(
  results_atb_export,
  here::here("results", "ATB_T0_vs_T12_stats_Holm_and_FDR.csv")
)

results_atb_export
```


# ------------------------------------------------------------------------------
# CULTURES
# ------------------------------------------------------------------------------
## Dynamics of P. aeruginosa and S. aureus sputum counts
# ------------------------------------------------------------------------------

### Load the data
```{r Load the data, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df <- read_delim(here("data", "cultures_simple.csv"))
```

### Prepare data
```{r Prepare data, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df <- df %>%
  mutate(across(where(is.character), ~ trimws(.))) %>%
  filter(
    !is.na(PatientID),
    !(PatientID %in% c("ONLY PAE", "ONLY SAE", "Pae + Sau a T0"))
  )

cols_counts <- c("Pae_T0","Pae_T12","Sau_T0","Sau_T12")
df <- df %>%
  mutate(across(
    any_of(cols_counts) & where(is.character),
    ~ suppressWarnings(as.numeric(.))
  ))

df <- df %>%
  mutate(
    Pae_any = (coalesce(Pae_T0, 0) > 0 | coalesce(Pae_T12, 0) > 0),
    Sau_any = (coalesce(Sau_T0, 0) > 0 | coalesce(Sau_T12, 0) > 0)
  )

# Create vectors
only_Pae     <- df %>% filter(Pae_any & !Sau_any) %>% pull(PatientID)
only_Sau     <- df %>% filter(Sau_any & !Pae_any) %>% pull(PatientID)
Pae_and_Sau  <- df %>% filter(Pae_any &  Sau_any) %>% pull(PatientID)

only_Pae; only_Sau; Pae_and_Sau
```

#### Only Pae
```{r only_Pae, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
pae_only <- df %>%
  dplyr::filter(PatientID %in% only_Pae) %>%
  dplyr::select(PatientID, Pae_T0, Pae_T12) %>%
  dplyr::filter(!is.na(Pae_T0), !is.na(Pae_T12)) %>%
  dplyr::mutate(change = dplyr::case_when(
    Pae_T12 > Pae_T0 ~ "up",
    Pae_T12 < Pae_T0 ~ "down",
    TRUE             ~ "no_change"
  ))

tab_pae <- pae_only %>%
  dplyr::count(change) %>%
  dplyr::mutate(n_total = sum(n),
                pct = round(100 * n / n_total, 1))

tab_pae

pae_up_ids   <- pae_only %>% dplyr::filter(change == "up") %>% dplyr::pull(PatientID)
pae_down_ids <- pae_only %>% dplyr::filter(change == "down") %>% dplyr::pull(PatientID)
pae_same_ids <- pae_only %>% dplyr::filter(change == "no_change") %>% dplyr::pull(PatientID)

pae_up_ids; pae_down_ids; pae_same_ids
```

#### Only Sau
```{r only_Sau, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
sau_only <- df %>%
  dplyr::filter(PatientID %in% only_Sau) %>%
  dplyr::select(PatientID, Sau_T0, Sau_T12) %>%
  dplyr::filter(!is.na(Sau_T0), !is.na(Sau_T12)) %>%
  dplyr::mutate(change = dplyr::case_when(
    Sau_T12 > Sau_T0 ~ "up",
    Sau_T12 < Sau_T0 ~ "down",
    TRUE             ~ "no_change"
  ))

tab_sau <- sau_only %>%
  dplyr::count(change) %>%
  dplyr::mutate(n_total = sum(n),
                pct = round(100 * n / n_total, 1))

tab_sau

sau_up_ids   <- sau_only %>% dplyr::filter(change == "up") %>% dplyr::pull(PatientID)
sau_down_ids <- sau_only %>% dplyr::filter(change == "down") %>% dplyr::pull(PatientID)
sau_same_ids <- sau_only %>% dplyr::filter(change == "no_change") %>% dplyr::pull(PatientID)

sau_up_ids; sau_down_ids; sau_same_ids
```

#### Pae and Sau
```{r Pae_and_Sau, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
both <- df %>%
  filter(PatientID %in% Pae_and_Sau)

# Classify changes by pathogen
pae_changes <- both %>%
  dplyr::mutate(change_pae = case_when(
    Pae_T12 > Pae_T0 ~ "UP Pae",
    Pae_T12 < Pae_T0 ~ "DOWN Pae",
    TRUE ~ "NOCHANGE Pae"
  )) %>%
  dplyr::select(PatientID, change_pae)

sau_changes <- both %>%
  dplyr::mutate(change_sau = case_when(
    Sau_T12 > Sau_T0 ~ "UP Sau",
    Sau_T12 < Sau_T0 ~ "DOWN Sau",
    TRUE ~ "NOCHANGE Sau"
  )) %>%
  dplyr::select(PatientID, change_sau)


combo <- pae_changes %>%
  left_join(sau_changes, by = "PatientID") %>%
  mutate(combo = paste(change_pae, "+", change_sau))


summary_combo <- combo %>%
  group_by(combo) %>%
  summarise(
    n = n(),
    pacientes = paste(PatientID, collapse = ", "),
    .groups = "drop"
  )

combo
summary_combo
```

### Normality and statistics
#### Only Pae
```{r Normality only_Pae, echo=TRUE, results='show', message=FALSE, warning=FALSE}
df_pae <- df %>%
  filter(PatientID %in% only_Pae) %>%
  transmute(PatientID, T0 = Pae_T0, T12 = Pae_T12) %>%
  filter(!is.na(T0), !is.na(T12)) %>%
  mutate(delta = T12 - T0,
         dir   = case_when(T12 > T0 ~ "up",
                           T12 < T0 ~ "down",
                           TRUE     ~ "no_change"))

sw_all <- shapiro.test(df_pae$delta)

if (sw_all$p.value > 0.05) {
  test_all <- t.test(df_pae$T12, df_pae$T0, paired = TRUE)
  method_all <- "paired t-test"
} else {
  test_all <- wilcox.test(df_pae$T12, df_pae$T0, paired = TRUE, exact = FALSE)
  method_all <- "Wilcoxon signed-rank"
}

df_pae_up   <- df_pae %>% filter(dir == "up")
df_pae_down <- df_pae %>% filter(dir == "down")

# UP
if (nrow(df_pae_up) >= 3) {
  sw_up <- shapiro.test(df_pae_up$delta)
  if (sw_up$p.value > 0.05) {
    test_up <- t.test(df_pae_up$T12, df_pae_up$T0, paired = TRUE)
    method_up <- "paired t-test"
  } else {
    test_up <- wilcox.test(df_pae_up$T12, df_pae_up$T0, paired = TRUE, exact = FALSE)
    method_up <- "Wilcoxon signed-rank"
  }
} else {
  sw_up <- list(p.value = NA)
  test_up <- list(statistic = NA, p.value = NA)
  method_up <- "insufficient pairs (<3)"
}

# DOWN
if (nrow(df_pae_down) >= 3) {
  sw_down <- shapiro.test(df_pae_down$delta)
  if (sw_down$p.value > 0.05) {
    test_down <- t.test(df_pae_down$T12, df_pae_down$T0, paired = TRUE)
    method_down <- "paired t-test"
  } else {
    test_down <- wilcox.test(df_pae_down$T12, df_pae_down$T0, paired = TRUE, exact = FALSE)
    method_down <- "Wilcoxon signed-rank"
  }
} else {
  sw_down <- list(p.value = NA)
  test_down <- list(statistic = NA, p.value = NA)
  method_down <- "insufficient pairs (<3)"
}

res_pae <- tibble::tibble(
  group = c("only_Pae (ALL)", "only_Pae (UP)", "only_Pae (DOWN)"),
  n     = c(nrow(df_pae), nrow(df_pae_up), nrow(df_pae_down)),
  shapiro_p = c(sw_all$p.value, sw_up$p.value, sw_down$p.value),
  test      = c(method_all, method_up, method_down),
  stat      = c(unname(test_all$statistic), unname(test_up$statistic), unname(test_down$statistic)),
  p_value   = c(test_all$p.value, test_up$p.value, test_down$p.value),
  mean_delta   = c(mean(df_pae$delta), mean(df_pae_up$delta), mean(df_pae_down$delta)),
  median_delta = c(median(df_pae$delta), median(df_pae_up$delta), median(df_pae_down$delta))
)

res_pae

readr::write_csv(res_pae, here::here("results", "stats_only_Pae.csv"))
```

#### Only Sau
```{r Normality only_Sau, echo=TRUE, results='show', message=FALSE, warning=FALSE}
df_sau <- df %>%
  filter(PatientID %in% only_Sau) %>%
  transmute(PatientID, T0 = Sau_T0, T12 = Sau_T12) %>%
  filter(!is.na(T0), !is.na(T12)) %>%
  mutate(delta = T12 - T0,
         dir   = case_when(T12 > T0 ~ "up",
                           T12 < T0 ~ "down",
                           TRUE     ~ "no_change"))

sw_all_sau <- shapiro.test(df_sau$delta)

if (sw_all_sau$p.value > 0.05) {
  test_all_sau <- t.test(df_sau$T12, df_sau$T0, paired = TRUE)
  method_all_sau <- "paired t-test"
} else {
  test_all_sau <- wilcox.test(df_sau$T12, df_sau$T0, paired = TRUE, exact = FALSE)
  method_all_sau <- "Wilcoxon signed-rank"
}

df_sau_up   <- df_sau %>% filter(dir == "up")
df_sau_down <- df_sau %>% filter(dir == "down")

# UP
if (nrow(df_sau_up) >= 3) {
  sw_up_sau <- shapiro.test(df_sau_up$delta)
  if (sw_up_sau$p.value > 0.05) {
    test_up_sau <- t.test(df_sau_up$T12, df_sau_up$T0, paired = TRUE)
    method_up_sau <- "paired t-test"
  } else {
    test_up_sau <- wilcox.test(df_sau_up$T12, df_sau_up$T0, paired = TRUE, exact = FALSE)
    method_up_sau <- "Wilcoxon signed-rank"
  }
} else {
  sw_up_sau <- list(p.value = NA)
  test_up_sau <- list(statistic = NA, p.value = NA)
  method_up_sau <- "insufficient pairs (<3)"
}

# DOWN
if (nrow(df_sau_down) >= 3) {
  sw_down_sau <- shapiro.test(df_sau_down$delta)
  if (sw_down_sau$p.value > 0.05) {
    test_down_sau <- t.test(df_sau_down$T12, df_sau_down$T0, paired = TRUE)
    method_down_sau <- "paired t-test"
  } else {
    test_down_sau <- wilcox.test(df_sau_down$T12, df_sau_down$T0, paired = TRUE, exact = FALSE)
    method_down_sau <- "Wilcoxon signed-rank"
  }
} else {
  sw_down_sau <- list(p.value = NA)
  test_down_sau <- list(statistic = NA, p.value = NA)
  method_down_sau <- "insufficient pairs (<3)"
}

res_sau <- tibble::tibble(
  group = c("only_Sau (ALL)", "only_Sau (UP)", "only_Sau (DOWN)"),
  n     = c(nrow(df_sau), nrow(df_sau_up), nrow(df_sau_down)),
  shapiro_p = c(sw_all_sau$p.value, sw_up_sau$p.value, sw_down_sau$p.value),
  test      = c(method_all_sau, method_up_sau, method_down_sau),
  stat      = c(unname(test_all_sau$statistic), unname(test_up_sau$statistic), unname(test_down_sau$statistic)),
  p_value   = c(test_all_sau$p.value, test_up_sau$p.value, test_down_sau$p.value),
  mean_delta   = c(mean(df_sau$delta), mean(df_sau_up$delta), mean(df_sau_down$delta)),
  median_delta = c(median(df_sau$delta), median(df_sau_up$delta), median(df_sau_down$delta))
)

res_sau

readr::write_csv(res_sau, here::here("results", "stats_only_Sau.csv"))
```

### Function for plots
```{r Function for plots, echo=TRUE, results='hide', message=FALSE, warning=FALSE}}
# 1) Data preparation: reshape counts and annotate subgroups
df_plot <- df %>%
  pivot_longer(
    cols = c(Pae_T0, Pae_T12, Sau_T0, Sau_T12),
    names_to = c("bug", "Time"), names_sep = "_", values_to = "count"
  ) %>%
  mutate(
    bug  = recode(bug, Pae = "Pae", Sau = "Sau"),
    Time = factor(Time, levels = c("T0","T12")),
    subgroup = case_when(
      PatientID %in% only_Pae    ~ "only_Pae",
      PatientID %in% only_Sau    ~ "only_Sau",
      PatientID %in% Pae_and_Sau ~ "Pae_and_Sau",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(subgroup))

pae_dir <- df %>%
  filter(PatientID %in% only_Pae) %>%
  transmute(PatientID,
            change = case_when(Pae_T12 > Pae_T0 ~ "up", Pae_T12 < Pae_T0 ~ "down", TRUE ~ "no_change"))

sau_dir <- df %>%
  filter(PatientID %in% only_Sau) %>%
  transmute(PatientID,
            change = case_when(Sau_T12 > Sau_T0 ~ "up", Sau_T12 < Sau_T0 ~ "down", TRUE ~ "no_change"))

long_only_pae_dir <- df_plot %>% filter(subgroup == "only_Pae", bug == "Pae") %>% left_join(pae_dir, by = "PatientID")
long_only_sau_dir <- df_plot %>% filter(subgroup == "only_Sau", bug == "Sau") %>% left_join(sau_dir, by = "PatientID")

# 2) Define aesthetics
time_colors <- c("T0" = "#9EC1CF", "T12" = "#F9A1BC")

# 3) Function
create_dot_plot <- function(data, p_value, y_label, title, shape = 21) {
  
  if (is.null(data) || nrow(data) < 2) {
    cat("Skipping plot for '", title, "' due to empty data.\n", sep = "")
    return(NULL)
  }
  
  d_paired <- data %>% 
    mutate(
      Time = factor(Time, levels = c("T0", "T12")),
      x    = as.numeric(Time),
      count_log10p = log10(count + 1)
    )
  
  y_max <- max(d_paired$count_log10p, na.rm = TRUE)
  y_limit <- y_max * 1.25
  
  p_label <- if (is.null(p_value) || is.na(p_value) || !is.finite(p_value)) {
    "p=N/A"
  } else if (p_value < 0.001) {
    "p<0.001"
  } else {
    paste0("p=", signif(p_value, 3))
  }
  
  p_y_pos <- y_max * 1.10
  
  p <- ggplot(d_paired, aes(x = x, y = count_log10p, group = PatientID)) +
    geom_line(color = "grey40", alpha = 0.5, linetype = "dashed", linewidth = 0.5) +
    geom_point(aes(fill = Time),
               shape = shape, size = 3, stroke = 0.5,
               color = "black", alpha = 0.8) +
    
    annotate("segment",
             x = 1, xend = 2,
             y = p_y_pos, yend = p_y_pos,
             linewidth = 0.5) +
    annotate("text",
             x = 1.5,
             y = p_y_pos + 0.05 * y_max,
             label = p_label, size = 4) +
    
    scale_x_continuous(
      breaks = c(1, 2),
      labels = c("T0", "T12"),
      limits = c(0.5, 2.5)
    ) +
    scale_y_continuous(
      breaks = pretty(c(0, y_max), n = 5),
      limits = c(0, y_limit)
    ) +
    scale_fill_manual(values = time_colors, name = "Time") +
    labs(x = NULL, y = y_label, title = title) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
  
  return(p)
}
```

##### Plots
```{r Plots, echo=TRUE, results='show', message=FALSE, warning=FALSE}
p_pae_all  <- res_pae$p_value[res_pae$group == "only_Pae (ALL)"]
p_pae_up   <- res_pae$p_value[res_pae$group == "only_Pae (UP)"]
p_pae_down <- res_pae$p_value[res_pae$group == "only_Pae (DOWN)"]
p_sau_all  <- res_sau$p_value[res_sau$group == "only_Sau (ALL)"]
p_sau_up   <- res_sau$p_value[res_sau$group == "only_Sau (UP)"]
p_sau_down <- res_sau$p_value[res_sau$group == "only_Sau (DOWN)"]

p_pae_all_dot  <- create_dot_plot(long_only_pae_dir, p_pae_all,
                                  "Log10 (Pae CFU/mL + 1)", "Only Pae - ALL")
p_pae_up_dot   <- create_dot_plot(
  filter(long_only_pae_dir, change == "up"),
  p_pae_up, "", "Only Pae - UP"
)
p_pae_down_dot <- create_dot_plot(
  filter(long_only_pae_dir, change == "down"),
  p_pae_down, "", "Only Pae - DOWN"
)

p_sau_all_dot  <- create_dot_plot(long_only_sau_dir, p_sau_all,
                                  "Log10 (Sau CFU/mL + 1)", "Only Sau - ALL",
                                  shape = 24)
p_sau_up_dot   <- create_dot_plot(
  filter(long_only_sau_dir, change == "up"),
  p_sau_up, "", "Only Sau - UP",
  shape = 24
)
p_sau_down_dot <- create_dot_plot(
  filter(long_only_sau_dir, change == "down"),
  p_sau_down, "", "Only Sau - DOWN",
  shape = 24
)


pae_dot_row <- p_pae_all_dot + p_pae_up_dot + p_pae_down_dot +
  plot_layout(nrow = 1, guides = "collect") &
  theme(legend.position = "bottom")

sau_dot_row <- p_sau_all_dot + p_sau_up_dot + p_sau_down_dot +
  plot_layout(nrow = 1, guides = "collect") &
  theme(legend.position = "bottom")

pae_sau_all_row <- p_pae_all_dot + p_sau_all_dot +
  plot_layout(nrow = 1, guides = "collect") &
  theme(legend.position = "bottom")


print(pae_dot_row)
print(sau_dot_row)
print(pae_sau_all_row)

ggsave(here::here("results", "Pae_only.svg"), pae_dot_row,
       device = "svg", width = 10, height = 6, dpi = 300)
ggsave(here::here("results", "Sau_only.svg"), sau_dot_row,
       device = "svg", width = 10, height = 6, dpi = 300)
ggsave(here::here("results", "Pae_Sau_all_only.svg"), pae_sau_all_row,
       device = "svg", width = 10, height = 6, dpi = 300)
```

##### Panel plot Pae and Sau
```{r Panel plot Pae and Sau, echo=TRUE, results='show', message=FALSE, warning=FALSE}
panel_pae_sau <- (pae_dot_row) / (sau_dot_row) +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

panel_pae_sau <- panel_pae_sau + plot_annotation(tag_levels = "A")

ggsave(here("results","panel_pae_sau.svg"), panel_pae_sau, width = 8, height = 8, bg = "white")
```


#### Co-colonization Pae + Sau
```{r Co-colonization Pae and Sau, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_both_plot <- df_plot %>%
  dplyr::filter(subgroup == "Pae_and_Sau") %>%
  dplyr::left_join(
    combo %>% dplyr::select(PatientID, combo),
    by = "PatientID"
  ) %>%
  dplyr::mutate(
    Time = factor(Time, levels = c("T0", "T12")),
    bug  = factor(bug,  levels = c("Pae", "Sau")),
    facet_lab = paste0(PatientID, "\n", combo)
  )

y_max_both <- max(df_both_plot$count, na.rm = TRUE)
```

##### Plots
```{r Plots co-colonization Pae and Sau, echo=TRUE, results='show', message=FALSE, warning=FALSE}
time_colors <- c(
  "T0"  = "#9EC1CF",
  "T12" = "#F9A1BC"
)

df_both_plot2 <- df_both_plot %>%
  mutate(
    count_log10p = log10(count + 1)
  )

y_max_both <- max(df_both_plot2$count_log10p, na.rm = TRUE)

p_both <- ggplot(
  df_both_plot2,
  aes(
    x      = Time,
    y      = count_log10p,
    group  = interaction(PatientID, bug),
    colour = Time,
    fill   = Time,
    shape  = bug
  )
) +

  geom_line(
    aes(group = interaction(PatientID, bug)),
    linetype    = "dashed",
    colour      = "grey60",
    linewidth   = 0.4,
    alpha       = 0.7,
    show.legend = FALSE
  ) +

  geom_point(
    size   = 3,
    stroke = 0.5
  ) +

  geom_point(
    aes(shape = bug),
    size        = 3,
    stroke      = 0.6,
    colour      = "black",
    fill        = NA,
    show.legend = FALSE
  ) +
  facet_wrap(~ facet_lab, ncol = 3) +
  
  scale_colour_manual(
    name   = "Time",
    values = time_colors,
    breaks = c("T0", "T12"),
    labels = c("T0", "T12")
  ) +
  scale_fill_manual(
    values = time_colors,
    guide  = "none"
  ) +
  scale_shape_manual(
    values = c(Pae = 21, Sau = 24),
    name   = "Pathogen"
  ) +
  scale_y_continuous(
    breaks = pretty(c(0, y_max_both), n = 5),
    limits = c(0, y_max_both * 1.3)
  ) +

  guides(
    colour = guide_legend(
      order = 1,
      override.aes = list(
        shape  = 21,
        size   = 3,
        stroke = 0.6,
        colour = "black",
        fill   = time_colors
      )
    ),
    shape = guide_legend(
      order = 2,
      override.aes = list(
        size   = 3,
        stroke = 0.6,
        colour = "black",
        fill   = "white"
      )
    )
  ) +
  
  labs(
    title = "Co-colonization by P. aeruginosa and S. aureus",
    x     = "Time",
    y     = "log10(P. aeruginosa + S. aureus CFU/mL + 1)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text.x     = element_text(face = "plain"),
    legend.position  = "right"
  )

print(p_both)

ggsave(
  here::here("results", "Pae_Sau_cocolonization.svg"),
  p_both,
  device = "svg",
  width  = 8,
  height = 8,
  dpi    = 300
)
```

# ---------------------------------------------------------------
## Association between ppFEV1 changes and P. aeruginosa dynamics
# ---------------------------------------------------------------

### Load in data
```{r Load in data, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
data_Pae <- read_delim(here("data", "only_Pae_ppFEV1.csv"))
```

### Create categorical variables
```{r Create categorical variables, echo=TRUE, results='show', message=FALSE, warning=FALSE}
analysis_data_Pae <- data_Pae %>%
  mutate(
    # Variable 1: P. aeruginosa change (Pae Increases vs. Does Not Increase)
    pae_change = factor(case_when(
      Pae_direction == "up"                   ~ "Pae Increases",
      Pae_direction %in% c("down", "neutral") ~ "Pae Does Not Increase",
      TRUE                                    ~ NA_character_
    )),
    
    # Variable 2: FEV1 change (Improves vs. Does Not Improve)
    fev1_improvement = factor(ifelse(ppFEV1_T12 > ppFEV1_T0, "Improves", "Does Not Improve"))
  )

# Create the contingency table
contingency_table_Pae <- table(analysis_data_Pae$pae_change, analysis_data_Pae$fev1_improvement)

print("Contingency Table:")
print(contingency_table_Pae)
```

### Perform Fisher's test
```{r Fisher test, echo=TRUE, results='show', message=FALSE, warning=FALSE}
fisher_test_result_Pae <- fisher.test(contingency_table_Pae)

# Print the results
print("Fisher's Exact Test Results:")
print(fisher_test_result_Pae)
```

### Normality test and statistics
```{r Shapiro-Wilk, echo=TRUE, results='show', message=FALSE, warning=FALSE}
# Filter for patients with a decrease in P. aeruginosa (ONLY 1 INCREASED)
patients_down_Pae <- data_Pae %>%
  filter(Pae_direction == "down")

# Calculate the differences between T12 and T0 (the normality test for a paired analysis is performed on these differences).
differences_Pae <- patients_down_Pae$ppFEV1_T12 - patients_down_Pae$ppFEV1_T0

# Perform the Shapiro-Wilk normality test on the differences
shapiro_result_Pae <- shapiro.test(differences_Pae)

cat("--- Shapiro-Wilk Normality Test Result ---\n")
print(shapiro_result_Pae)
cat("\n")

if (shapiro_result_Pae$p.value > 0.05) {
  
  cat("Data appear to be normally distributed (p > 0.05).\n")
  cat("--- Running Paired t-test ---\n")
  
  final_test_Pae <- t.test(patients_down_Pae$ppFEV1_T12, 
                         patients_down_Pae$ppFEV1_T0, 
                         paired = TRUE, 
                         alternative = "greater")
  
} else {
  
  cat("Data do not appear to be normally distributed (p <= 0.05).\n")
  cat("--- Running Paired Wilcoxon Signed-Rank Test ---\n")
  
  final_test_Pae <- wilcox.test(patients_down_Pae$ppFEV1_T12, 
                              patients_down_Pae$ppFEV1_T0, 
                              paired = TRUE, 
                              alternative = "greater")
}

cat("\n--- Final Statistical Test Result ---\n")
print(final_test_Pae)
```

### Save the results
```{r Save the results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
readr::write_csv(tibble::tibble(test = final_test_Pae$method,
                                statistic = unname(final_test_Pae$statistic),
                                p_value = final_test_Pae$p.value,
                                n = length(differences_Pae)),
                 here::here("results", "ppFEV1_Pae_down_stats.csv"))
```

# ---------------------------------------------------------------
## Association between ppFEV1 changes and S. aureus dynamics
# ---------------------------------------------------------------

#### Load in data
```{r Load in data, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
data_Sau <- read_delim(here("data", "only_Sau_ppFEV1.csv"))
```

### Create categorical variables
```{r Create categorical variables, echo=TRUE, results='show', message=FALSE, warning=FALSE}
analysis_data_Sau <- data_Sau %>%
  mutate(
    # Variable 1: S. aureus change (Aureus Increases vs. Does Not Increase)
    sau_change = factor(case_when(
      Sau_direction == "up"                   ~ "Aureus Increases",
      Sau_direction %in% c("down", "neutral") ~ "Aureus Does Not Increase",
      TRUE                                    ~ NA_character_
    )),
    
    # Variable 2: FEV1 change (Improves vs. Does Not Improve)
    fev1_improvement = factor(ifelse(ppFEV1_T12 > ppFEV1_T0, "Improves", "Does Not Improve"))
  )

# Create the contingency table
contingency_table_Sau <- table(analysis_data_Sau$sau_change, analysis_data_Sau$fev1_improvement)

# Print the table to verify it's correct
print("Contingency Table:")
print(contingency_table_Sau)
```

### Perform Fisher's test
```{r Fisher test, echo=TRUE, results='show', message=FALSE, warning=FALSE}
fisher_test_result_Sau <- fisher.test(contingency_table_Sau)

print("Fisher's Exact Test Results:")
print(fisher_test_result_Sau)
```

### Normality test and statistics
```{r Shapiro-Wilk, echo=TRUE, results='show', message=FALSE, warning=FALSE}
# Filter for patients with an increase in S. aureus
patients_up_Sau <- data_Sau %>%
  filter(Sau_direction == "up")

# Calculate the differences between T12 and T0 (the normality test for a paired analysis is performed on these differences).
differences_Sau <- patients_up_Sau$ppFEV1_T12 - patients_up_Sau$ppFEV1_T0

# Perform the Shapiro-Wilk normality test on the differences
shapiro_result_Sau <- shapiro.test(differences_Sau)

cat("--- Shapiro-Wilk Normality Test Result ---\n")
print(shapiro_result_Sau)
cat("\n")
if (shapiro_result_Sau$p.value > 0.05) {
  
  cat("Data appear to be normally distributed (p > 0.05).\n")
  cat("--- Running Paired t-test ---\n")
  
  final_test_Sau <- t.test(patients_up_Sau$ppFEV1_T12, 
                         patients_up_Sau$ppFEV1_T0, 
                         paired = TRUE, 
                         alternative = "greater")
  
} else {
  
  cat("Data do not appear to be normally distributed (p <= 0.05).\n")
  cat("--- Running Paired Wilcoxon Signed-Rank Test ---\n")
  
  final_test_Sau <- wilcox.test(patients_up_Sau$ppFEV1_T12, 
                              patients_up_Sau$ppFEV1_T0, 
                              paired = TRUE, 
                              alternative = "greater")
}

cat("\n--- Final Statistical Test Result ---\n")
print(final_test_Sau)
```

### Save the results
```{r Save the results, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
readr::write_csv(tibble::tibble(test = final_test_Sau$method,
                                statistic = unname(final_test_Sau$statistic),
                                p_value = final_test_Sau$p.value,
                                n = length(differences_Sau)),
                 here::here("results", "ppFEV1_Sau_up_stats.csv"))
```


# ------------------------------------------------------------------------------
# GLOBAL LUNG FUNCTION
# ------------------------------------------------------------------------------

### Load in the data
```{r Load in data lung function, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_lung <- read_delim(here("data", "lung_function.csv"))
```

### Change names of lung function group
```{r Change names of lung function groups, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_lung <- df_lung %>%
  dplyr::mutate(
    group_label = dplyr::case_when(
      group == 1 ~ "NM→NM",
      group == 2 ~ "MS→MS",
      group == 3 ~ "MS→NM",
      TRUE ~ NA_character_
    )
  )
```


### Filter data by time
```{r Filter data by time lung function, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
data_lung_T0  <- subset(df_lung, Time == "T0")
data_lung_T12 <- subset(df_lung, Time == "T12")
```

### Function for statistic test
```{r Function for statistic test based on shapiro results lung function, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
auto_paired_test_lung <- function(df,
                                  id_col   = "PatientID",
                                  time_col = "Time",
                                  value_col = "ppFEV1",
                                  levels   = c("T0","T12")) {
  stopifnot(all(c(id_col, time_col, value_col) %in% names(df)))
  
  dfw <- df[, c(id_col, time_col, value_col)]
  dfw <- tidyr::pivot_wider(
    dfw,
    names_from  = !!rlang::sym(time_col),
    values_from = !!rlang::sym(value_col)
  )
  if (!all(levels %in% names(dfw))) {
    stop("No columns found ", paste(levels, collapse = " and "), " after pivot_wider().")
  }
  dfw <- dfw[stats::complete.cases(dfw[, levels]), ]
  
  x0   <- dfw[[levels[1]]]
  x12  <- dfw[[levels[2]]]
  diff <- x12 - x0
  n    <- length(diff)
  if (n < 3) stop("Very few valid pairs (n < 3).")
  
  sh     <- shapiro.test(diff)
  normal <- sh$p.value >= 0.05
  
  if (normal) {
    tt <- t.test(x12, x0, paired = TRUE)
    
    mean_diff <- mean(diff, na.rm = TRUE)
    sd_diff   <- stats::sd(diff, na.rm = TRUE)
    d_paired  <- mean_diff / sd_diff
    
    mean_T0  <- mean(x0,  na.rm = TRUE); sd_T0  <- stats::sd(x0,  na.rm = TRUE)
    mean_T12 <- mean(x12, na.rm = TRUE); sd_T12 <- stats::sd(x12, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Paired t-test\n")
    cat("n =", n, "| t =", round(tt$statistic, 4), "| p =", signif(tt$p.value, 4), "\n")
    cat("T0:   mean ± SD = ", round(mean_T0, 3), " ± ", round(sd_T0, 3), "\n", sep = "")
    cat("T12:  mean ± SD = ", round(mean_T12, 3), " ± ", round(sd_T12, 3), "\n", sep = "")
    cat("Mean difference (T12−T0) =", round(mean_diff, 3),
    " | 95% CI [", round(tt$conf.int[1], 3), ", ", round(tt$conf.int[2], 3), "]\n", sep = "")
    cat("Effect size (Cohen's d paired) =", round(d_paired, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "paired_t",
      n = n,
      shapiro = sh,
      test = tt,
      group_summary = list(
        T0  = c(mean = mean_T0, sd = sd_T0),
        T12 = c(mean = mean_T12, sd = sd_T12)
      ),
      mean_diff = mean_diff,
      ci = tt$conf.int,
      effect_size = list(type = "cohen_d_paired", value = d_paired)
    ))
    
  } else {
    wt   <- wilcox.test(x0, x12, paired = TRUE, exact = FALSE, correct = FALSE)
    V    <- as.numeric(wt$statistic)
    r_rb <- 1 - (2 * V) / (n * (n + 1))
    
    med_T0   <- stats::median(x0,  na.rm = TRUE)
    q1_T0    <- stats::quantile(x0,  0.25, na.rm = TRUE)
    q3_T0    <- stats::quantile(x0,  0.75, na.rm = TRUE)
    med_T12  <- stats::median(x12, na.rm = TRUE)
    q1_T12   <- stats::quantile(x12, 0.25, na.rm = TRUE)
    q3_T12   <- stats::quantile(x12, 0.75, na.rm = TRUE)
    med_diff <- stats::median(diff, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Not normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Wilcoxon signed-rank (paired)\n")
    cat("n =", n, "| V =", V, "| p =", signif(wt$p.value, 4), "\n")
    cat("T0:   median [IQR] = ", round(med_T0, 3), " [",
        round(q1_T0, 3), "–", round(q3_T0, 3), "]\n", sep = "")
    cat("T12:  median [IQR] = ", round(med_T12, 3), " [",
        round(q1_T12, 3), "–", round(q3_T12, 3), "]\n", sep = "")
    cat("Median difference (T12−T0) =", round(med_diff, 3), "\n")
    cat("Effect size (rank-biserial r) =", round(r_rb, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "wilcoxon_paired",
      n = n,
      shapiro = sh,
      test = wt,
      group_summary = list(
        T0  = c(median = med_T0, Q1 = q1_T0, Q3 = q3_T0),
        T12 = c(median = med_T12, Q1 = q1_T12, Q3 = q3_T12)
      ),
      median_diff = med_diff,
      effect_size = list(type = "rank_biserial_r", value = r_rb, V = V)
    ))
  }
}
# Wrapper "safe": if n < 3, returns a valid res_obj object with p = NA instead of stopping
auto_paired_test_lung_safe <- function(df,
                                       id_col   = "PatientID",
                                       time_col = "Time",
                                       value_col = "ppFEV1",
                                       levels   = c("T0","T12")) {
  tryCatch(
    auto_paired_test_lung(df,
                          id_col = id_col,
                          time_col = time_col,
                          value_col = value_col,
                          levels = levels),
    error = function(e) {
      if (grepl("Very few valid pairs", e$message) || grepl("n < 3", e$message)) {
        list(
          variable = value_col,
          method   = "too_few_pairs",
          n        = NA_integer_,
          shapiro  = list(p.value = NA_real_),
          test     = list(statistic = NA_real_, p.value = NA_real_),
          effect_size = list(type = NA_character_, value = NA_real_)
        )
      } else {
        stop(e)
      }
    }
  )
}
```

### Perform the statistical test
#### For ppFEV1
##### Global
```{r Execute stats function lung function ppFEV1, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ppFEV1 <- auto_paired_test_lung(df_lung, value_col = "ppFEV1")
```
##### By Sex
###### Female
```{r Execute stats function lung function ppFEV1 by female, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ppFEV1_fem <- auto_paired_test_lung(
  df_lung %>% dplyr::filter(Sex == "female"),
  value_col = "ppFEV1"
)
```
###### Male
```{r Execute stats function lung function ppFEV1 by male, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ppFEV1_mal <- auto_paired_test_lung(
  df_lung %>% dplyr::filter(Sex == "male"),
  value_col = "ppFEV1"
)
```
##### By Age
###### Adult
```{r Execute stats function lung function ppFEV1 by adult, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ppFEV1_adu <- auto_paired_test_lung(
  df_lung %>% dplyr::filter(AgeGroup == "adult"),
  value_col = "ppFEV1"
)
```
###### Pediatric
```{r Execute stats function lung function ppFEV1 by pediatric, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ppFEV1_ped <- auto_paired_test_lung(
  df_lung %>% dplyr::filter(AgeGroup == "pediatric"),
  value_col = "ppFEV1"
)
```
##### By lung function group
###### NM-NM
```{r Execute stats function lung function ppFEV1 by NMNM, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ppFEV1_NMNM <- auto_paired_test_lung_safe(
  df_lung %>% dplyr::filter(group_label == "NM→NM"),
  value_col = "ppFEV1"
)
```
###### MS-MS
```{r Execute stats function lung function ppFEV1 by MSMS, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ppFEV1_MSMS <- auto_paired_test_lung_safe(
  df_lung %>% dplyr::filter(group_label == "MS→MS"),
  value_col = "ppFEV1"
)
```
###### MS-NM
```{r Execute stats function lung function ppFEV1 by MSNM, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ppFEV1_MSNM <- auto_paired_test_lung_safe(
  df_lung %>% dplyr::filter(group_label == "MS→NM"),
  value_col = "ppFEV1"
)
```

#### For ppFVC
##### Global
```{r Execute stats function lung function ppFVC, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ppFVC  <- auto_paired_test_lung(df_lung, value_col = "ppFVC")
```
##### By Sex
###### Female
```{r Execute stats function lung function ppFEV1 by female, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ppFVC_fem <- auto_paired_test_lung(
  df_lung %>% dplyr::filter(Sex == "female"),
  value_col = "ppFVC"
)
```
###### Male
```{r Execute stats function lung function ppFEV1 by male, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ppFVC_mal <- auto_paired_test_lung(
  df_lung %>% dplyr::filter(Sex == "male"),
  value_col = "ppFVC"
)
```
##### By Age
###### Adult
```{r Execute stats function lung function ppFEV1 by adult, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ppFVC_adu <- auto_paired_test_lung(
  df_lung %>% dplyr::filter(AgeGroup == "adult"),
  value_col = "ppFVC"
)
```
###### Pediatric
```{r Execute stats function lung function ppFEV1 by pediatric, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ppFVC_ped <- auto_paired_test_lung(
  df_lung %>% dplyr::filter(AgeGroup == "pediatric"),
  value_col = "ppFVC"
)
```
##### By lung function group
###### NM-NM
```{r Execute stats function lung function ppFVC by NMNM, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ppFVC_NMNM <- auto_paired_test_lung_safe(
  df_lung %>% dplyr::filter(group_label == "NM→NM"),
  value_col = "ppFVC"
)
```
###### MS-MS
```{r Execute stats function lung function ppFVC by MSMS, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ppFVC_MSMS <- auto_paired_test_lung_safe(
  df_lung %>% dplyr::filter(group_label == "MS→MS"),
  value_col = "ppFVC"
)
```
###### MS-NM
```{r Execute stats function lung function ppFVC by MSNM, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ppFVC_MSNM <- auto_paired_test_lung_safe(
  df_lung %>% dplyr::filter(group_label == "MS→NM"),
  value_col = "ppFVC"
)
```
#### For FEV1/FVC
##### Global
```{r Execute stats function lung function FEV1_FVC, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ratio  <- auto_paired_test_lung(df_lung, value_col = "FEV1_FVC")
```
##### By Sex
###### Female
```{r Execute stats function lung function ppFEV1 by female, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ratio_fem <- auto_paired_test_lung(
  df_lung %>% dplyr::filter(Sex == "female"),
  value_col = "FEV1_FVC"
)
```
###### Male
```{r Execute stats function lung function ppFEV1 by male, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ratio_mal <- auto_paired_test_lung(
  df_lung %>% dplyr::filter(Sex == "male"),
  value_col = "FEV1_FVC"
)
```
##### By Age
###### Adult
```{r Execute stats function lung function ppFEV1 by adult, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ratio_adu <- auto_paired_test_lung(
  df_lung %>% dplyr::filter(AgeGroup == "adult"),
  value_col = "FEV1_FVC"
)
```
###### Pediatric
```{r Execute stats function lung function ppFEV1 by pediatric, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ratio_ped <- auto_paired_test_lung(
  df_lung %>% dplyr::filter(AgeGroup == "pediatric"),
  value_col = "FEV1_FVC"
)
```
##### By lung function group
###### NM-NM
```{r Execute stats function lung function FEV1_FVC by NMNM, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ratio_NMNM <- auto_paired_test_lung_safe(
  df_lung %>% dplyr::filter(group_label == "NM→NM"),
  value_col = "FEV1_FVC"
)
```
###### MS-MS
```{r Execute stats function lung function FEV1_FVC by MSMS, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ratio_MSMS <- auto_paired_test_lung_safe(
  df_lung %>% dplyr::filter(group_label == "MS→MS"),
  value_col = "FEV1_FVC"
)
```
###### MS-NM
```{r Execute stats function lung function FEV1_FVC by MSNM, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ratio_MSNM <- auto_paired_test_lung_safe(
  df_lung %>% dplyr::filter(group_label == "MS→NM"),
  value_col = "FEV1_FVC"
)
```

### Females vs Males
```{r LMER fem vs mal, echo=TRUE, results='show', message=FALSE, warning=FALSE}
model_sex <- lmer(ppFEV1 ~ Time * Sex + (1 | PatientID), data = df_lung)
anova_sex <- as.data.frame(anova(model_sex))
anova_sex$term <- rownames(anova_sex)
anova_sex <- anova_sex[, c("term", setdiff(names(anova_sex), "term"))]

write.csv(anova_sex, "Results/lmer_ppFEV1_Time_Sex_ANOVA.csv", row.names = FALSE)
capture.output(summary(model_sex), file = "Results/lmer_ppFEV1_Time_Sex_summary.txt")
```

### Adults vs Pediatrics
```{r LMER adu vs ped, echo=TRUE, results='show', message=FALSE, warning=FALSE}
model_age <- lmer(ppFEV1 ~ Time * AgeGroup + (1 | PatientID), data = df_lung)
anova_age <- as.data.frame(anova(model_age))
anova_age$term <- rownames(anova_age)
anova_age <- anova_age[, c("term", setdiff(names(anova_age), "term"))]

write.csv(anova_age, "Results/lmer_ppFEV1_Time_AgeGroup_ANOVA.csv", row.names = FALSE)
capture.output(summary(model_age), file = "Results/lmer_ppFEV1_Time_AgeGroup_summary.txt")
```

### Function for plot
```{r Function plots, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
make_paired_plot <- function(df, value_var, res_obj, y_lab, outfile) {
  stopifnot(all(c("PatientID","Time", value_var) %in% names(df)))
  
  d <- df %>%
    mutate(Time = factor(Time, levels = c("T0","T12"))) %>%
    group_by(PatientID) %>% filter(n_distinct(Time) == 2) %>% ungroup() %>%
    mutate(x = as.numeric(Time))
  
  col_box   <- c("T0" = "#9EC1CF", "T12" = "#F9A1BC")
  col_point <- c("T0" = "#D6E9F5", "T12" = "#FCD6E0")
  
  d <- d %>%
    mutate(
      fill_box   = col_box[Time],
      fill_point = col_point[Time]
    )
  
  p_val   <- res_obj$test$p.value
  p_label <- ifelse(p_val < 0.001, "p < 0.001", paste0("p = ", signif(p_val, 3)))
  
  y_max    <- max(d[[value_var]], na.rm = TRUE) * 1.05
  p_values <- data.frame(x = 1, xend = 2, y = y_max, label = p_label)
  
  p <- ggplot() +
    stat_slab(
      data = d %>% filter(Time == "T0"),
      aes(x = x, y = .data[[value_var]], fill = fill_box),
      side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    stat_slab(
      data = d %>% filter(Time == "T12"),
      aes(x = x, y = .data[[value_var]], fill = fill_box),
      side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    geom_boxplot(
      data = d,
      aes(x = x, y = .data[[value_var]], fill = fill_box),
      width = 0.5, outlier.shape = NA,
      color = "black", alpha = 0.9, show.legend = TRUE
    ) +
    geom_point(
      data = d,
      aes(x = x, y = .data[[value_var]], fill = fill_point),
      shape = 21, size = 3, stroke = 0.3, color = "black",
      show.legend = FALSE
    ) +
    geom_line(
      data = d,
      aes(x = x, y = .data[[value_var]], group = PatientID),
      color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
    ) +
    geom_segment(
      data = p_values,
      aes(x = x, xend = xend, y = y, yend = y),
      color = "black", linewidth = 0.5
    ) +
    geom_text(
      data = p_values,
      aes(x = (x + xend) / 2, y = y * 1.02, label = label),
      size = 4
    ) +
    scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
    scale_fill_identity(guide = "legend", name = "Time",
                        breaks = col_box, labels = names(col_box)) +
    labs(x = NULL, y = y_lab) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "right",
          axis.text.x = element_text(size = 14))
  
  if (!dir.exists(here("results"))) dir.create(here("results"))
  ggsave(here("results", outfile), p, width = 8, height = 6, units = "in", dpi = 300)
  
  invisible(p)
}
```

### Function for polished panel plots
```{r Function for polish panel plots echo=TRUE, results='show', message=FALSE, warning=FALSE}
polish_panel_plot <- function(p) {
  p +
    theme(
      legend.position = "bottom",
      plot.margin = margin(5.5, 5.5, 5.5, 5.5),
      axis.title.y = element_text(margin = margin(r = 6))
    )
}
```


#### Panel plots
```{r Panel plots lung function, echo=TRUE, results='show', message=FALSE, warning=FALSE}
p_ppFEV1 <- polish_panel_plot(
  make_paired_plot(df_lung, "ppFEV1", res_ppFEV1, "ppFEV1 (%)", "lung_ppFEV1.svg")
)

p_ppFVC <- polish_panel_plot(
  make_paired_plot(df_lung, "ppFVC", res_ppFVC, "ppFVC (%)", "lung_ppFVC.svg")
)

p_ratio <- polish_panel_plot(
  make_paired_plot(df_lung, "FEV1_FVC", res_ratio, "FEV1/FVC", "lung_FEV1_FVC.svg")
)

panel_lung <- p_ppFEV1 | p_ppFVC | p_ratio

panel_lung <- panel_lung + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

ggsave(here("results","lung_function_panel.svg"),
       panel_lung, width = 12, height = 5, bg = "white")

# By Sex
## Female
p_ppFEV1_fem <- polish_panel_plot(
  make_paired_plot(df_lung, "ppFEV1", res_ppFEV1_fem, "ppFEV1 (%) - Females", "lung_ppFEV1_fem.svg")
)

p_ppFVC_fem <- polish_panel_plot(
  make_paired_plot(df_lung, "ppFVC", res_ppFVC_fem, "ppFVC (%) - Females", "lung_ppFVC_fem.svg")
)

p_ratio_fem <- polish_panel_plot(
  make_paired_plot(df_lung, "FEV1_FVC", res_ratio_fem, "FEV1/FVC - Females", "lung_FEV1_FVC_fem.svg")
)

panel_lung_fem <- p_ppFEV1_fem | p_ppFVC_fem | p_ratio_fem

panel_lung_fem <- panel_lung_fem + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

ggsave(here("results","lung_function_female_panel.svg"),
       panel_lung_fem, width = 12, height = 5, bg = "white")

## Male
p_ppFEV1_mal <- polish_panel_plot(
  make_paired_plot(df_lung, "ppFEV1", res_ppFEV1_mal, "ppFEV1 (%) - Males", "lung_ppFEV1_mal.svg")
)

p_ppFVC_mal <- polish_panel_plot(
  make_paired_plot(df_lung, "ppFVC", res_ppFVC_mal, "ppFVC (%) - Males", "lung_ppFVC_mal.svg")
)

p_ratio_mal <- polish_panel_plot(
  make_paired_plot(df_lung, "FEV1_FVC", res_ratio_mal, "FEV1/FVC - Males", "lung_FEV1_FVC_mal.svg")
)

panel_lung_mal <- p_ppFEV1_mal | p_ppFVC_mal | p_ratio_mal

panel_lung_mal <- panel_lung_mal + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

ggsave(here("results","lung_function_male_panel.svg"),
       panel_lung_mal, width = 12, height = 5, bg = "white")

# By Age
## Adult
p_ppFEV1_adu <- polish_panel_plot(
  make_paired_plot(df_lung, "ppFEV1", res_ppFEV1_adu, "ppFEV1 (%) - Adults", "lung_ppFEV1_adu.svg")
)

p_ppFVC_adu <- polish_panel_plot(
  make_paired_plot(df_lung, "ppFVC", res_ppFVC_adu, "ppFVC (%) - Adults", "lung_ppFVC_adu.svg")
)

p_ratio_adu <- polish_panel_plot(
  make_paired_plot(df_lung, "FEV1_FVC", res_ratio_adu, "FEV1/FVC - Adults", "lung_FEV1_FVC_adu.svg")
)

panel_lung_adu <- p_ppFEV1_adu | p_ppFVC_adu | p_ratio_adu

panel_lung_adu <- panel_lung_adu + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

ggsave(here("results","lung_function_adult_panel.svg"),
       panel_lung_adu, width = 12, height = 5, bg = "white")

## Pediatrics
p_ppFEV1_ped <- polish_panel_plot(
  make_paired_plot(df_lung, "ppFEV1", res_ppFEV1_ped, "ppFEV1 (%) - Pediatrics", "lung_ppFEV1_ped.svg")
)

p_ppFVC_ped <- polish_panel_plot(
  make_paired_plot(df_lung, "ppFVC", res_ppFVC_ped, "ppFVC (%) - Pediatrics", "lung_ppFVC_ped.svg")
)

p_ratio_ped <- polish_panel_plot(
  make_paired_plot(df_lung, "FEV1_FVC", res_ratio_ped, "FEV1/FVC - Pediatrics", "lung_FEV1_FVC_ped.svg")
)

panel_lung_ped <- p_ppFEV1_ped | p_ppFVC_ped | p_ratio_ped

panel_lung_ped <- panel_lung_ped + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

ggsave(here("results","lung_function_pediatric_panel.svg"),
       panel_lung_ped, width = 12, height = 5, bg = "white")

# By lung function group
## NM-NM
p_ppFEV1_NMNM <- polish_panel_plot(
  make_paired_plot(df_lung %>% dplyr::filter(group_label == "NM→NM"), "ppFEV1", res_ppFEV1_NMNM, "ppFEV1 (%) - NM→NM", "lung_ppFEV1_NMNM.svg")
)

p_ppFVC_NMNM <- polish_panel_plot(
  make_paired_plot(df_lung %>% dplyr::filter(group_label == "NM→NM"), "ppFVC", res_ppFVC_NMNM, "ppFVC (%) - NM→NM", "lung_ppFVC_NMNM.svg")
)

p_ratio_NMNM <- polish_panel_plot(
  make_paired_plot(df_lung %>% dplyr::filter(group_label == "NM→NM"), "FEV1_FVC", res_ratio_NMNM, "FEV1/FVC - NM→NM", "lung_FEV1_FVC_NMNM.svg")
)

panel_lung_NMNM <- p_ppFEV1_NMNM | p_ppFVC_NMNM | p_ratio_NMNM

panel_lung_NMNM <- panel_lung_NMNM + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

ggsave(here("results", "lung_function_panel_NMNM.svg"), panel_lung_NMNM, width = 12, height = 5, bg = "white")

## MS-MS
p_ppFEV1_MSMS <- polish_panel_plot(
  make_paired_plot(df_lung %>% dplyr::filter(group_label == "MS→MS"), "ppFEV1", res_ppFEV1_MSMS, "ppFEV1 (%) - MS→MS", "lung_ppFEV1_MSMS.svg")
)

p_ppFVC_MSMS <- polish_panel_plot(
  make_paired_plot(df_lung %>% dplyr::filter(group_label == "MS→MS"), "ppFVC", res_ppFVC_MSMS, "ppFVC (%) - MS→MS", "lung_ppFVC_MSMS.svg")
)

p_ratio_MSMS <- polish_panel_plot(
  make_paired_plot(df_lung %>% dplyr::filter(group_label == "MS→MS"), "FEV1_FVC", res_ratio_MSMS, "FEV1/FVC - MS→MS", "lung_FEV1_FVC_MSMS.svg")
)

panel_lung_MSMS <- p_ppFEV1_MSMS | p_ppFVC_MSMS | p_ratio_MSMS

panel_lung_MSMS <- panel_lung_MSMS + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

ggsave(here("results", "lung_function_panel_MSMS.svg"), panel_lung_MSMS, width = 12, height = 5, bg = "white")

## MS-NM
p_ppFEV1_MSNM <- polish_panel_plot(
  make_paired_plot(df_lung %>% dplyr::filter(group_label == "MS→NM"), "ppFEV1", res_ppFEV1_MSNM, "ppFEV1 (%) - MS→NM", "lung_ppFEV1_MSNM.svg")
)

p_ppFVC_MSNM <- polish_panel_plot(
  make_paired_plot(df_lung %>% dplyr::filter(group_label == "MS→NM"), "ppFVC", res_ppFVC_MSNM, "ppFVC (%) - MS→MS", "lung_ppFVC_MSNM.svg")
)

p_ratio_MSNM <- polish_panel_plot(
  make_paired_plot(df_lung %>% dplyr::filter(group_label == "MS→NM"), "FEV1_FVC", res_ratio_MSNM, "FEV1/FVC - MS→NM", "lung_FEV1_FVC_MSNM.svg")
)

panel_lung_MSNM <- p_ppFEV1_MSNM | p_ppFVC_MSNM | p_ratio_MSNM

panel_lung_MSNM <- panel_lung_MSNM + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

ggsave(here("results", "lung_function_panel_MSNM.svg"), panel_lung_MSNM, width = 12, height = 5, bg = "white")
```

### Function plot with shapes for each trajectory
```{r Plots lung function general colored by groups WITH SHAPES, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
make_paired_plot_transcolored <- function(df, value_var, res_obj, y_lab, outfile) {
  stopifnot(all(c("PatientID", "Time", value_var, "group_label") %in% names(df)))
  
  d <- df %>%
    dplyr::mutate(Time = factor(Time, levels = c("T0", "T12"))) %>%
    dplyr::group_by(PatientID) %>%
    dplyr::filter(dplyr::n_distinct(Time) == 2) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(x = as.numeric(Time))
  
  p_val   <- res_obj$test$p.value
  p_label <- ifelse(p_val < 0.001, "p < 0.001", paste0("p = ", signif(p_val, 3)))
  
  y_max    <- max(d[[value_var]], na.rm = TRUE) * 1.05
  p_values <- data.frame(x = 1, xend = 2, y = y_max, label = p_label)
  
  p <- ggplot() +
    ggdist::stat_slab(
      data = d %>% dplyr::filter(Time == "T0"),
      aes(x = x, y = .data[[value_var]], fill = Time),
      side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    ggdist::stat_slab(
      data = d %>% dplyr::filter(Time == "T12"),
      aes(x = x, y = .data[[value_var]], fill = Time),
      side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    geom_boxplot(
      data = d,
      aes(x = x, y = .data[[value_var]], fill = Time),
      width = 0.5, outlier.shape = NA,
      color = "black", alpha = 0.9
    ) +
    scale_fill_manual(values = c("T0" = "#9EC1CF", "T12" = "#F9A1BC"), name = "Time") +
    ggnewscale::new_scale_fill() +
    geom_point(
      data = d %>% dplyr::filter(!is.na(group_label)),
      aes(x = x, y = .data[[value_var]], fill = Time, shape = group_label),
      size = 3, stroke = 0.5, color = "black"
    ) +
    scale_fill_manual(values = c("T0" = "#D6E9F5", "T12" = "#FCD6E0"), guide = "none") +
    scale_shape_manual(
      values = c("NM→NM" = 21, "MS→MS" = 24, "MS→NM" = 22),
      name = "Trajectory"
    ) +
    geom_line(
      data = d,
      aes(x = x, y = .data[[value_var]], group = PatientID),
      color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
    ) +
    geom_segment(
      data = p_values,
      aes(x = x, xend = xend, y = y, yend = y),
      color = "black", linewidth = 0.5
    ) +
    geom_text(
      data = p_values,
      aes(x = (x + xend) / 2, y = y * 1.02, label = label),
      size = 4
    ) +
    scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
    labs(x = NULL, y = y_lab) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "right",
      axis.text.x = element_text(size = 14)
    )
  
  if (!dir.exists(here("results"))) dir.create(here("results"))
  ggsave(here("results", outfile), p, width = 8, height = 6, units = "in", dpi = 300)
  
  invisible(p)
}
```

#### Plot global trajectories
```{r Panel plots lung function trajectories, echo=TRUE, results='show', message=FALSE, warning=FALSE}
p_ppFEV1_trajec <- polish_panel_plot(
  make_paired_plot_transcolored(df_lung, "ppFEV1", res_ppFEV1, "ppFEV1 (%)", "lung_ppFEV1_groups_all_shapes.svg")
)

p_ppFVC_trajec <- polish_panel_plot(
  make_paired_plot_transcolored(df_lung, "ppFVC",    res_ppFVC,  "ppFVC (%)",  "lung_ppFVC_groups_all_shapes.svg")
)

p_ratio_trajec <- polish_panel_plot(
  make_paired_plot_transcolored(df_lung, "FEV1_FVC", res_ratio,  "FEV1/FVC",   "lung_FEV1FVC_groups_all_shapes.svg")
)
```

### Function to save stats lung 1
```{r Fuction to save stats results lung 1, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# === Export: lung function (ppFEV1, ppFVC, FEV1/FVC) ===
.build_row_lung <- function(df, value_col, res_obj, subgroup_label = "global",
                            fisher_tab = NULL, cutoff = NA_real_, p_fdr = NA_real_,
                            high_lbl = NULL, low_lbl = NULL) {
  stopifnot(all(c("PatientID","Time", value_col) %in% names(df)))
  is_t <- identical(res_obj$method, "paired_t")

  paired <- df %>%
    dplyr::select(PatientID, Time, !!rlang::sym(value_col)) %>%
    tidyr::pivot_wider(names_from = Time, values_from = !!rlang::sym(value_col)) %>%
    dplyr::filter(stats::complete.cases(T0, T12))

  mean_T0  <- mean(paired$T0)
  sd_T0    <- sd(paired$T0)
  mean_T12 <- mean(paired$T12)
  sd_T12   <- sd(paired$T12)

  med_T0 <- median(paired$T0)
  q1_T0  <- as.numeric(quantile(paired$T0, 0.25))
  q3_T0  <- as.numeric(quantile(paired$T0, 0.75))

  med_T12 <- median(paired$T12)
  q1_T12  <- as.numeric(quantile(paired$T12, 0.25))
  q3_T12  <- as.numeric(quantile(paired$T12, 0.75))

  fmt_group_T0  <- if (is_t) sprintf("%.3f ± %.3f", mean_T0, sd_T0) else sprintf("%.3f [%0.3f-%0.3f]", med_T0, q1_T0, q3_T0)
  fmt_group_T12 <- if (is_t) sprintf("%.3f ± %.3f", mean_T12, sd_T12) else sprintf("%.3f [%0.3f-%0.3f]", med_T12, q1_T12, q3_T12)

  if (!is.null(fisher_tab)) {
    fisher_res <- fisher.test(fisher_tab, conf.int = TRUE)

    tab_ha <- fisher_tab + 0.5
    a <- tab_ha[high_lbl, "T0"]
    b <- tab_ha[high_lbl, "T12"]
    c <- tab_ha[low_lbl, "T0"]
    d <- tab_ha[low_lbl, "T12"]

    or_ha   <- (a * d) / (b * c)
    se_log  <- sqrt(1/a + 1/b + 1/c + 1/d)
    ci_low  <- exp(log(or_ha) - 1.96 * se_log)
    ci_high <- exp(log(or_ha) + 1.96 * se_log)

    fisher_p       <- as.numeric(fisher_res$p.value)
    fisher_or      <- if (!is.null(fisher_res$estimate)) as.numeric(fisher_res$estimate) else NA_real_
    fisher_ci_low  <- if (!is.null(fisher_res$conf.int)) fisher_res$conf.int[1] else NA_real_
    fisher_ci_high <- if (!is.null(fisher_res$conf.int)) fisher_res$conf.int[2] else NA_real_
  } else {
    fisher_p <- fisher_or <- fisher_ci_low <- fisher_ci_high <- NA_real_
    or_ha <- ci_low <- ci_high <- NA_real_
  }

  tibble::tibble(
    subgroup      = subgroup_label,
    variable      = res_obj$variable,
    method        = res_obj$method,
    n_pairs       = res_obj$n,
    shapiro_p     = unname(res_obj$shapiro$p.value),
    stat          = unname(res_obj$test$statistic),
    p_value       = res_obj$test$p.value,
    group_T0      = fmt_group_T0,
    group_T12     = fmt_group_T12,
    mean_T0       = mean_T0,
    sd_T0         = sd_T0,
    median_T0     = med_T0,
    q1_T0         = q1_T0,
    q3_T0         = q3_T0,
    min_T0        = min(paired$T0),
    max_T0        = max(paired$T0),
    range_T0      = max(paired$T0) - min(paired$T0),
    mean_T12      = mean_T12,
    sd_T12        = sd_T12,
    median_T12    = med_T12,
    q1_T12        = q1_T12,
    q3_T12        = q3_T12,
    min_T12       = min(paired$T12),
    max_T12       = max(paired$T12),
    range_T12     = max(paired$T12) - min(paired$T12),
    diff_estimate = if (is_t) res_obj$mean_diff else res_obj$median_diff,
    ci_low        = if (is_t) res_obj$ci[1] else NA_real_,
    ci_high       = if (is_t) res_obj$ci[2] else NA_real_,
    effect_type   = res_obj$effect_size$type,
    effect_value  = res_obj$effect_size$value,
    fisher_p      = fisher_p,
    fisher_or     = fisher_or,
    fisher_ci_low = fisher_ci_low,
    fisher_ci_high= fisher_ci_high,
    or_ha         = or_ha,
    or_ha_ci_low  = if (!is.null(fisher_tab)) ci_low else NA_real_,
    or_ha_ci_high = if (!is.null(fisher_tab)) ci_high else NA_real_,
    cutoff_used   = cutoff
  )
}
```

### Function to save stats lung groups 2
```{r Fuction to save stats results lung groups 2, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
.build_row_lung_grp <- function(df, value_col, res_obj, subgroup_label) {
  stopifnot(all(c("PatientID", "Time", value_col) %in% names(df)))
  is_t <- identical(res_obj$method, "paired_t")

  paired <- df %>%
    dplyr::select(PatientID, Time, !!rlang::sym(value_col)) %>%
    tidyr::pivot_wider(names_from = Time, values_from = !!rlang::sym(value_col)) %>%
    dplyr::filter(stats::complete.cases(T0, T12))

  mean_T0  <- mean(paired$T0)
  sd_T0    <- sd(paired$T0)
  mean_T12 <- mean(paired$T12)
  sd_T12   <- sd(paired$T12)

  med_T0 <- median(paired$T0)
  q1_T0  <- as.numeric(quantile(paired$T0, 0.25))
  q3_T0  <- as.numeric(quantile(paired$T0, 0.75))

  med_T12 <- median(paired$T12)
  q1_T12  <- as.numeric(quantile(paired$T12, 0.25))
  q3_T12  <- as.numeric(quantile(paired$T12, 0.75))

  fmt_group_T0  <- if (is_t) sprintf("%.3f ± %.3f", mean_T0, sd_T0) else sprintf("%.3f [%0.3f-%0.3f]", med_T0, q1_T0, q3_T0)
  fmt_group_T12 <- if (is_t) sprintf("%.3f ± %.3f", mean_T12, sd_T12) else sprintf("%.3f [%0.3f-%0.3f]", med_T12, q1_T12, q3_T12)

  tibble::tibble(
    subgroup      = subgroup_label,
    variable      = res_obj$variable,
    method        = res_obj$method,
    n_pairs       = res_obj$n,
    shapiro_p     = unname(res_obj$shapiro$p.value),
    stat          = unname(res_obj$test$statistic),
    p_value       = res_obj$test$p.value,
    group_T0      = fmt_group_T0,
    group_T12     = fmt_group_T12,
    diff_estimate = if (is_t) res_obj$mean_diff else res_obj$median_diff,
    ci_low        = if (is_t) res_obj$ci[1] else NA_real_,
    ci_high       = if (is_t) res_obj$ci[2] else NA_real_,
    effect_type   = res_obj$effect_size$type,
    effect_value  = res_obj$effect_size$value
  )
}
```

### Save stats results ALL
```{r Save stats results lung function, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# === Save stats results ALL ===

row_ppFEV1 <- .build_row_lung(
  df_lung, "ppFEV1", res_ppFEV1,
  subgroup_label = "global",
  fisher_tab = NULL,
  cutoff     = NA_real_,
  high_lbl   = "normal-mild_ppFEV1",
  low_lbl    = "moderate-severe_ppFEV1"
)

row_ppFVC  <- .build_row_lung(df_lung, "ppFVC", res_ppFVC, subgroup_label = "global")
row_ratio  <- .build_row_lung(df_lung, "FEV1_FVC", res_ratio, subgroup_label = "global")

row_ppFEV1_fem <- .build_row_lung(df_lung %>% dplyr::filter(Sex == "female"), "ppFEV1", res_ppFEV1_fem, subgroup_label = "female")
row_ppFVC_fem  <- .build_row_lung(df_lung %>% dplyr::filter(Sex == "female"), "ppFVC", res_ppFVC_fem, subgroup_label = "female")
row_ratio_fem  <- .build_row_lung(df_lung %>% dplyr::filter(Sex == "female"), "FEV1_FVC", res_ratio_fem, subgroup_label = "female")

row_ppFEV1_mal <- .build_row_lung(df_lung %>% dplyr::filter(Sex == "male"), "ppFEV1", res_ppFEV1_mal, subgroup_label = "male")
row_ppFVC_mal  <- .build_row_lung(df_lung %>% dplyr::filter(Sex == "male"), "ppFVC", res_ppFVC_mal, subgroup_label = "male")
row_ratio_mal  <- .build_row_lung(df_lung %>% dplyr::filter(Sex == "male"), "FEV1_FVC", res_ratio_mal, subgroup_label = "male")

row_ppFEV1_adu <- .build_row_lung(df_lung %>% dplyr::filter(AgeGroup == "adult"), "ppFEV1", res_ppFEV1_adu, subgroup_label = "adult")
row_ppFVC_adu  <- .build_row_lung(df_lung %>% dplyr::filter(AgeGroup == "adult"), "ppFVC", res_ppFVC_adu, subgroup_label = "adult")
row_ratio_adu  <- .build_row_lung(df_lung %>% dplyr::filter(AgeGroup == "adult"), "FEV1_FVC", res_ratio_adu, subgroup_label = "adult")

row_ppFEV1_ped <- .build_row_lung(df_lung %>% dplyr::filter(AgeGroup == "pediatric"), "ppFEV1", res_ppFEV1_ped, subgroup_label = "pediatric")
row_ppFVC_ped  <- .build_row_lung(df_lung %>% dplyr::filter(AgeGroup == "pediatric"), "ppFVC", res_ppFVC_ped, subgroup_label = "pediatric")
row_ratio_ped  <- .build_row_lung(df_lung %>% dplyr::filter(AgeGroup == "pediatric"), "FEV1_FVC", res_ratio_ped, subgroup_label = "pediatric")

row_ppFEV1_NMNM <- .build_row_lung_grp(df_lung %>% dplyr::filter(group_label == "NM→NM"), "ppFEV1", res_ppFEV1_NMNM, "NM→NM")
row_ppFVC_NMNM  <- .build_row_lung_grp(df_lung %>% dplyr::filter(group_label == "NM→NM"), "ppFVC", res_ppFVC_NMNM, "NM→NM")
row_ratio_NMNM  <- .build_row_lung_grp(df_lung %>% dplyr::filter(group_label == "NM→NM"), "FEV1_FVC", res_ratio_NMNM, "NM→NM")

row_ppFEV1_MSMS <- .build_row_lung_grp(df_lung %>% dplyr::filter(group_label == "MS→MS"), "ppFEV1", res_ppFEV1_MSMS, "MS→MS")
row_ppFVC_MSMS  <- .build_row_lung_grp(df_lung %>% dplyr::filter(group_label == "MS→MS"), "ppFVC", res_ppFVC_MSMS, "MS→MS")
row_ratio_MSMS  <- .build_row_lung_grp(df_lung %>% dplyr::filter(group_label == "MS→MS"), "FEV1_FVC", res_ratio_MSMS, "MS→MS")

row_ppFEV1_MSNM <- .build_row_lung_grp(df_lung %>% dplyr::filter(group_label == "MS→NM"), "ppFEV1", res_ppFEV1_MSNM, "MS→NM")
row_ppFVC_MSNM  <- .build_row_lung_grp(df_lung %>% dplyr::filter(group_label == "MS→NM"), "ppFVC", res_ppFVC_MSNM, "MS→NM")
row_ratio_MSNM  <- .build_row_lung_grp(df_lung %>% dplyr::filter(group_label == "MS→NM"), "FEV1_FVC", res_ratio_MSNM, "MS→NM")

lung_stats <- dplyr::bind_rows(
  row_ppFEV1, row_ppFVC, row_ratio,
  row_ppFEV1_fem, row_ppFVC_fem, row_ratio_fem,
  row_ppFEV1_mal, row_ppFVC_mal, row_ratio_mal,
  row_ppFEV1_adu, row_ppFVC_adu, row_ratio_adu,
  row_ppFEV1_ped, row_ppFVC_ped, row_ratio_ped,
  row_ppFEV1_NMNM, row_ppFVC_NMNM, row_ratio_NMNM,
  row_ppFEV1_MSMS, row_ppFVC_MSMS, row_ratio_MSMS,
  row_ppFEV1_MSNM, row_ppFVC_MSNM, row_ratio_MSNM
)

lung_stats$p_fdr <- p.adjust(lung_stats$p_value, method = "BH")

if (!dir.exists(here("results"))) dir.create(here("results"))
readr::write_excel_csv(lung_stats, here("results", "lung_function_stats_complete_all.csv"))
```


### Save stats results of interest
```{r Save stats results lung function, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# === Export: lung function (ppFEV1, ppFVC, FEV1/FVC) ===

# === Save stats results of interest ===

row_ppFEV1 <- .build_row_lung(
  df_lung, "ppFEV1", res_ppFEV1,
  subgroup_label = "global",
  fisher_tab = NULL,
  cutoff     = NA_real_,
  high_lbl   = "normal-mild_ppFEV1",
  low_lbl    = "moderate-severe_ppFEV1"
)

row_ppFVC  <- .build_row_lung(df_lung, "ppFVC", res_ppFVC, subgroup_label = "global")
row_ratio  <- .build_row_lung(df_lung, "FEV1_FVC", res_ratio, subgroup_label = "global")

row_ppFEV1_fem <- .build_row_lung(df_lung %>% dplyr::filter(Sex == "female"), "ppFEV1", res_ppFEV1_fem, subgroup_label = "female")
row_ppFVC_fem  <- .build_row_lung(df_lung %>% dplyr::filter(Sex == "female"), "ppFVC", res_ppFVC_fem, subgroup_label = "female")
row_ratio_fem  <- .build_row_lung(df_lung %>% dplyr::filter(Sex == "female"), "FEV1_FVC", res_ratio_fem, subgroup_label = "female")

row_ppFEV1_mal <- .build_row_lung(df_lung %>% dplyr::filter(Sex == "male"), "ppFEV1", res_ppFEV1_mal, subgroup_label = "male")
row_ppFVC_mal  <- .build_row_lung(df_lung %>% dplyr::filter(Sex == "male"), "ppFVC", res_ppFVC_mal, subgroup_label = "male")
row_ratio_mal  <- .build_row_lung(df_lung %>% dplyr::filter(Sex == "male"), "FEV1_FVC", res_ratio_mal, subgroup_label = "male")

row_ppFEV1_adu <- .build_row_lung(df_lung %>% dplyr::filter(AgeGroup == "adult"), "ppFEV1", res_ppFEV1_adu, subgroup_label = "adult")
row_ppFVC_adu  <- .build_row_lung(df_lung %>% dplyr::filter(AgeGroup == "adult"), "ppFVC", res_ppFVC_adu, subgroup_label = "adult")
row_ratio_adu  <- .build_row_lung(df_lung %>% dplyr::filter(AgeGroup == "adult"), "FEV1_FVC", res_ratio_adu, subgroup_label = "adult")

row_ppFEV1_ped <- .build_row_lung(df_lung %>% dplyr::filter(AgeGroup == "pediatric"), "ppFEV1", res_ppFEV1_ped, subgroup_label = "pediatric")
row_ppFVC_ped  <- .build_row_lung(df_lung %>% dplyr::filter(AgeGroup == "pediatric"), "ppFVC", res_ppFVC_ped, subgroup_label = "pediatric")
row_ratio_ped  <- .build_row_lung(df_lung %>% dplyr::filter(AgeGroup == "pediatric"), "FEV1_FVC", res_ratio_ped, subgroup_label = "pediatric")

lung_stats <- dplyr::bind_rows(
  row_ppFEV1, row_ppFVC, row_ratio,
  row_ppFEV1_fem, row_ppFVC_fem, row_ratio_fem,
  row_ppFEV1_mal, row_ppFVC_mal, row_ratio_mal,
  row_ppFEV1_adu, row_ppFVC_adu, row_ratio_adu,
  row_ppFEV1_ped, row_ppFVC_ped, row_ratio_ped
)

lung_stats$p_fdr <- p.adjust(lung_stats$p_value, method = "BH")

if (!dir.exists(here("results"))) dir.create(here("results"))
readr::write_excel_csv(lung_stats, here("results", "lung_function_stats_complete_without_lunggroups.csv"))
```


# ------------------------------------------------------------------------------
# TRANSAMINASES
# ------------------------------------------------------------------------------

### Load in the data
```{r Load in data trans, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_trans <- read_delim(here("data", "transaminases.csv"))
```

### Descriptive statistics
```{r Descriptive stats, echo=TRUE, results='show', message=FALSE, warning=FALSE}
range(df_trans$ALT, na.rm = TRUE)
min_val <- min(df_trans$ALT, na.rm = TRUE)
max_val <- max(df_trans$ALT, na.rm = TRUE)

range(df_trans$AST, na.rm = TRUE)
min_val <- min(df_trans$AST, na.rm = TRUE)
max_val <- max(df_trans$AST, na.rm = TRUE)

range(df_trans$GGT, na.rm = TRUE)
min_val <- min(df_trans$GGT, na.rm = TRUE)
max_val <- max(df_trans$GGT, na.rm = TRUE)
```

### Filter data by time
```{r Filter data by time trans, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
data_trans_T0 <- subset(df_trans, Time == "T0")
data_trans_T12 <- subset(df_trans, Time == "T12")
```

### Function for statistic test
```{r Function for statistic test based on shapiro results trans, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
auto_paired_test_trans <- function(df,
                                   id_col   = "PatientID",
                                   time_col = "Time",
                                   value_col = "ALT",
                                   levels   = c("T0","T12")) {
  stopifnot(all(c(id_col, time_col, value_col) %in% names(df)))
  
  dfw <- df[, c(id_col, time_col, value_col)]
  dfw <- tidyr::pivot_wider(
    dfw,
    names_from  = !!rlang::sym(time_col),
    values_from = !!rlang::sym(value_col)
  )
  if (!all(levels %in% names(dfw))) {
    stop("No columns found ", paste(levels, collapse = " and "), " after pivot_wider().")
  }
  dfw <- dfw[stats::complete.cases(dfw[, levels]), ]
  
  x0   <- dfw[[levels[1]]]
  x12  <- dfw[[levels[2]]]
  diff <- x12 - x0
  n    <- length(diff)
  if (n < 3) stop("Very few valid pairs (n < 3).")
  
  sh     <- shapiro.test(diff)
  normal <- sh$p.value >= 0.05
  
  if (normal) {
    tt        <- t.test(x12, x0, paired = TRUE)
    mean_diff <- mean(diff, na.rm = TRUE)
    sd_diff   <- stats::sd(diff, na.rm = TRUE)
    d_paired  <- mean_diff / sd_diff
    
    mean_T0  <- mean(x0,  na.rm = TRUE); sd_T0  <- stats::sd(x0,  na.rm = TRUE)
    mean_T12 <- mean(x12, na.rm = TRUE); sd_T12 <- stats::sd(x12, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Paired t-test\n")
    cat("n =", n, "| t =", round(tt$statistic, 4), "| p =", signif(tt$p.value, 4), "\n")
    cat("T0:   mean ± SD = ", round(mean_T0, 3), " ± ", round(sd_T0, 3), "\n", sep = "")
    cat("T12:  mean ± SD = ", round(mean_T12, 3), " ± ", round(sd_T12, 3), "\n", sep = "")
    cat("Mean difference (T12−T0) =", round(mean_diff, 3),
        " | 95% CI [", round(tt$conf.int[1], 3), ", ", round(tt$conf.int[2], 3), "]\n", sep = "")
    cat("Effect size (Cohen's d paired) =", round(d_paired, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "paired_t",
      n = n,
      shapiro = sh,
      test = tt,
      group_summary = list(
        T0  = c(mean = mean_T0, sd = sd_T0),
        T12 = c(mean = mean_T12, sd = sd_T12)
      ),
      mean_diff = mean_diff,
      ci = tt$conf.int,
      effect_size = list(type = "cohen_d_paired", value = d_paired)
    ))
    
  } else {
    wt   <- wilcox.test(x0, x12, paired = TRUE, exact = FALSE, correct = FALSE)
    V    <- as.numeric(wt$statistic)
    r_rb <- 1 - (2 * V) / (n * (n + 1))
    
    med_T0   <- stats::median(x0,  na.rm = TRUE)
    q1_T0    <- stats::quantile(x0,  0.25, na.rm = TRUE)
    q3_T0    <- stats::quantile(x0,  0.75, na.rm = TRUE)
    med_T12  <- stats::median(x12, na.rm = TRUE)
    q1_T12   <- stats::quantile(x12, 0.25, na.rm = TRUE)
    q3_T12   <- stats::quantile(x12, 0.75, na.rm = TRUE)
    med_diff <- stats::median(diff, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Not normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Wilcoxon signed-rank (paired)\n")
    cat("n =", n, "| V =", V, "| p =", signif(wt$p.value, 4), "\n")
    cat("T0:   median [IQR] = ", round(med_T0, 3), " [",
        round(q1_T0, 3), "–", round(q3_T0, 3), "]\n", sep = "")
    cat("T12:  median [IQR] = ", round(med_T12, 3), " [",
        round(q1_T12, 3), "–", round(q3_T12, 3), "]\n", sep = "")
    cat("Median difference (T12−T0) =", round(med_diff, 3), "\n")
    cat("Effect size (rank-biserial r) =", round(r_rb, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "wilcoxon_paired",
      n = n,
      shapiro = sh,
      test = wt,
      group_summary = list(
        T0  = c(median = med_T0, Q1 = q1_T0, Q3 = q3_T0),
        T12 = c(median = med_T12, Q1 = q1_T12, Q3 = q3_T12)
      ),
      median_diff = med_diff,
      effect_size = list(type = "rank_biserial_r", value = r_rb, V = V)
    ))
  }
}
# Wrapper "safe": if n < 3, returns a valid res_obj object with p = NA instead of stopping
auto_paired_test_trans_safe <- function(df,
                                       id_col   = "PatientID",
                                       time_col = "Time",
                                       value_col = "ALT",
                                       levels   = c("T0","T12")) {
  tryCatch(
    auto_paired_test_trans(df,
                          id_col = id_col,
                          time_col = time_col,
                          value_col = value_col,
                          levels = levels),
    error = function(e) {
      if (grepl("Very few valid pairs", e$message) || grepl("n < 3", e$message)) {
        list(
          variable = value_col,
          method   = "too_few_pairs",
          n        = NA_integer_,
          shapiro  = list(p.value = NA_real_),
          test     = list(statistic = NA_real_, p.value = NA_real_),
          effect_size = list(type = NA_character_, value = NA_real_)
        )
      } else {
        stop(e)
      }
    }
  )
}
```

### Perform statistical test
#### ALT
```{r Execute stats function trans, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_ALT <- auto_paired_test_trans(df_trans, value_col = "ALT")
```
#### AST
```{r Execute stats function trans, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_AST <- auto_paired_test_trans(df_trans, value_col = "AST")
```
#### GGT
```{r Execute stats function trans, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_GGT <- auto_paired_test_trans(df_trans, value_col = "GGT")
```

#### Panel plots
```{r Panel plots trans, echo=TRUE, results='show', message=FALSE, warning=FALSE}
p_ALT <- polish_panel_plot(
  make_paired_plot(df_trans, "ALT", res_ALT, "ALT (U/L)", "trans_ALT.svg")
)

p_AST <- polish_panel_plot(
  make_paired_plot(df_trans, "AST", res_AST, "AST (U/L)", "trans_AST.svg")
)

p_GGT <- polish_panel_plot(
  make_paired_plot(df_trans, "GGT", res_GGT, "GGT (U/L)", "trans_GGT.svg")
)


panel_trans <- p_ALT | p_AST | p_GGT


panel_trans <- panel_trans + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

ggsave(here("results","transaminases_panel.svg"),
       panel_trans, width = 12, height = 5, bg = "white")
```


#### Manual transformed y-axis plot function with explicit raincloud
```{r Function break y axis trans, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
compress_y_axis <- function(y, break_at = 80, top_actual = 300,
                            top_plot = 120, gap = 7) {
  
  ifelse(
    y <= break_at,
    y,
    break_at + gap +
      ((y - break_at) / (top_actual - break_at)) *
      (top_plot - break_at - gap)
  )
}


make_half_violin_data <- function(d, y_var = "y_plot",
                                  cloud_width = 0.45,
                                  adjust = 0.6,
                                  n = 512) {
  
  out <- lapply(split(d, d$Time), function(dd) {
    
    vals <- dd[[y_var]]
    vals <- vals[is.finite(vals)]
    
    if (length(unique(vals)) < 2) {
      return(NULL)
    }
    
    dens <- stats::density(
      vals,
      adjust = adjust,
      from = min(vals, na.rm = TRUE),
      to = max(vals, na.rm = TRUE),
      n = n
    )
    
    dens_scaled <- dens$y / max(dens$y, na.rm = TRUE) * cloud_width
    
    time_value <- as.character(unique(dd$Time)[1])
    x0 <- unique(dd$x)[1]
    fill_box <- unique(dd$fill_box)[1]
    
    # T0 cloud goes to the left, T12 cloud goes to the right
    side <- ifelse(time_value == "T0", -1, 1)
    
    curve_x <- x0 + side * dens_scaled
    
    tibble::tibble(
      Time = time_value,
      fill_box = fill_box,
      x_poly = c(rep(x0, length(dens$x)), rev(curve_x)),
      y_poly = c(dens$x, rev(dens$x))
    )
  })
  
  dplyr::bind_rows(out)
}


make_paired_plot_transformed <- function(df, value_var, res_obj, y_lab,
                                         break_axis = FALSE,
                                         break_at = 80,
                                         top_actual = 300,
                                         top_plot = 120,
                                         gap = 7,
                                         y_breaks_actual = c(0, 20, 40, 60, 80),
                                         show_p = TRUE,
                                         show_legend = TRUE,
                                         show_break_marks = TRUE) {
  
  stopifnot(all(c("PatientID", "Time", value_var) %in% names(df)))
  
  d <- df %>%
    mutate(Time = factor(Time, levels = c("T0", "T12"))) %>%
    group_by(PatientID) %>%
    filter(n_distinct(Time) == 2) %>%
    ungroup() %>%
    mutate(x = as.numeric(Time))
  
  col_box   <- c("T0" = "#9EC1CF", "T12" = "#F9A1BC")
  col_point <- c("T0" = "#D6E9F5", "T12" = "#FCD6E0")
  
  if (break_axis) {
    d <- d %>%
      mutate(
        y_plot = compress_y_axis(
          .data[[value_var]],
          break_at = break_at,
          top_actual = top_actual,
          top_plot = top_plot,
          gap = gap
        )
      )
    
    y_breaks_plot <- compress_y_axis(
      y_breaks_actual,
      break_at = break_at,
      top_actual = top_actual,
      top_plot = top_plot,
      gap = gap
    )
    
    y_limit_plot <- top_plot
    
  } else {
    d <- d %>%
      mutate(y_plot = .data[[value_var]])
    
    y_breaks_plot <- y_breaks_actual
    y_limit_plot <- max(y_breaks_actual, na.rm = TRUE)
  }
  
  d <- d %>%
    mutate(
      fill_box   = col_box[Time],
      fill_point = col_point[Time]
    )
  
  cloud_data <- make_half_violin_data(
    d,
    y_var = "y_plot",
    cloud_width = 0.45,
    adjust = 0.6
  )
  
  p_val <- res_obj$test$p.value
  p_label <- ifelse(
    p_val < 0.001,
    "p < 0.001",
    paste0("p = ", signif(p_val, 3))
  )
  
  p <- ggplot() +
    geom_polygon(
      data = cloud_data,
      aes(x = x_poly, y = y_poly, group = Time, fill = fill_box),
      alpha = 0.5,
      color = NA,
      show.legend = FALSE
    ) +
    geom_boxplot(
      data = d,
      aes(x = x, y = y_plot, fill = fill_box),
      width = 0.5,
      outlier.shape = NA,
      color = "black",
      alpha = 0.9,
      show.legend = show_legend
    ) +
    geom_line(
      data = d,
      aes(x = x, y = y_plot, group = PatientID),
      color = "grey40",
      alpha = 0.4,
      linetype = "dashed",
      linewidth = 0.4
    ) +
    geom_point(
      data = d,
      aes(x = x, y = y_plot, fill = fill_point),
      shape = 21,
      size = 3,
      stroke = 0.3,
      color = "black",
      show.legend = FALSE
    ) +
    scale_x_continuous(
      breaks = c(1, 2),
      labels = c("T0", "T12"),
      limits = c(0.45, 2.55)
    ) +
    scale_y_continuous(
      limits = c(0, y_limit_plot),
      breaks = y_breaks_plot,
      labels = y_breaks_actual,
      expand = expansion(mult = c(0.02, 0.03))
    ) +
    scale_fill_identity(
      guide = "legend",
      name = "Time",
      breaks = unname(col_box),
      labels = names(col_box)
    ) +
    labs(x = NULL, y = y_lab) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = ifelse(show_legend, "bottom", "none"),
      axis.text.x = element_text(size = 14)
    )
  
  if (show_p) {
    p <- p +
      annotate(
        "segment",
        x = 1,
        xend = 2,
        y = 76,
        yend = 76,
        linewidth = 0.5
      ) +
      annotate(
        "text",
        x = 1.5,
        y = 78,
        label = p_label,
        size = 4
      )
  }
  
  if (break_axis && show_break_marks) {
    p <- p +
      annotate(
        "segment",
        x = 0.60,
        xend = 0.65,
        y = break_at + 1.5,
        yend = break_at + 4.5,
        linewidth = 0.6
      ) +
      annotate(
        "segment",
        x = 0.69,
        xend = 0.74,
        y = break_at + 1.5,
        yend = break_at + 4.5,
        linewidth = 0.6
      )
  }
  
  p
}
```

#### Transaminases transformed y-axis plots
```{r Plots break y axis trans, echo=TRUE, results='show', message=FALSE, warning=FALSE}
p_ALT_y <- make_paired_plot_transformed(
  df_trans,
  value_var = "ALT",
  res_obj = res_ALT,
  y_lab = "ALT (U/L)",
  break_axis = TRUE,
  break_at = 80,
  top_actual = 300,
  top_plot = 120,
  gap = 7,
  y_breaks_actual = c(0, 20, 40, 60, 80, 100, 200, 300),
  show_p = TRUE,
  show_legend = FALSE
)

p_AST_y <- make_paired_plot_transformed(
  df_trans,
  value_var = "AST",
  res_obj = res_AST,
  y_lab = "AST (U/L)",
  break_axis = FALSE,
  y_breaks_actual = c(0, 20, 40, 60, 80),
  show_p = TRUE,
  show_legend = FALSE
)

p_GGT_y <- make_paired_plot_transformed(
  df_trans,
  value_var = "GGT",
  res_obj = res_GGT,
  y_lab = "GGT (U/L)",
  break_axis = TRUE,
  break_at = 80,
  top_actual = 620,
  top_plot = 120,
  gap = 7,
  y_breaks_actual = c(0, 20, 40, 60, 80, 100, 300, 600),
  show_p = TRUE,
  show_legend = FALSE
)
```

#### Panel and legend
```{r Panel plots break y axis trans, echo=TRUE, results='show', message=FALSE, warning=FALSE}
legend_plot <- make_paired_plot_transformed(
  df_trans,
  value_var = "ALT",
  res_obj = res_ALT,
  y_lab = "ALT (U/L)",
  break_axis = TRUE,
  break_at = 80,
  top_actual = 300,
  top_plot = 120,
  gap = 7,
  y_breaks_actual = c(0, 20, 40, 60, 80, 100, 200, 300),
  show_p = FALSE,
  show_legend = TRUE
)

legend <- cowplot::get_legend(legend_plot)

panel_trans <- (p_ALT_y | p_AST_y | p_GGT_y) /
  cowplot::ggdraw(legend) +
  plot_layout(heights = c(1, 0.12))

ggsave(
  here("results", "transaminases_panel_transformed_axis_raincloud.svg"),
  plot = panel_trans, width = 12, height = 6, bg = "white")
```


### Save stats results
```{r Save stats results trans, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# === Export: ALT, AST, GGT ===
.build_row_trans <- function(df, value_col, res_obj, p_fdr = NA_real_) {
  stopifnot(all(c("PatientID","Time", value_col) %in% names(df)))
  is_t <- identical(res_obj$method, "paired_t")

  paired <- df %>%
    dplyr::select(PatientID, Time, !!rlang::sym(value_col)) %>%
    tidyr::pivot_wider(names_from = Time, values_from = !!rlang::sym(value_col)) %>%
    dplyr::filter(stats::complete.cases(T0, T12))

  mean_T0  <- mean(paired$T0);  sd_T0  <- sd(paired$T0)
  mean_T12 <- mean(paired$T12); sd_T12 <- sd(paired$T12)

  med_T0 <- median(paired$T0)
  q1_T0  <- as.numeric(quantile(paired$T0,  0.25))
  q3_T0  <- as.numeric(quantile(paired$T0,  0.75))

  med_T12 <- median(paired$T12)
  q1_T12  <- as.numeric(quantile(paired$T12, 0.25))
  q3_T12  <- as.numeric(quantile(paired$T12, 0.75))

  fmt_group_T0  <- if (is_t) sprintf("%.3f ± %.3f", mean_T0, sd_T0)  else sprintf("%.3f [%0.3f-%0.3f]", med_T0,  q1_T0,  q3_T0)
  fmt_group_T12 <- if (is_t) sprintf("%.3f ± %.3f", mean_T12, sd_T12) else sprintf("%.3f [%0.3f-%0.3f]", med_T12, q1_T12, q3_T12)

  tibble::tibble(
    variable      = res_obj$variable,
    method        = res_obj$method,
    n_pairs       = res_obj$n,
    shapiro_p     = unname(res_obj$shapiro$p.value),
    stat          = unname(res_obj$test$statistic),
    p_value       = res_obj$test$p.value,
    group_T0      = fmt_group_T0,
    group_T12     = fmt_group_T12,
    mean_T0  = mean_T0,
    sd_T0    = sd_T0,
    median_T0= med_T0,
    q1_T0    = q1_T0,
    q3_T0    = q3_T0,
    min_T0   = min(paired$T0),
    max_T0   = max(paired$T0),
    range_T0 = max(paired$T0) - min(paired$T0),
    mean_T12   = mean_T12,
    sd_T12     = sd_T12,
    median_T12 = med_T12,
    q1_T12     = q1_T12,
    q3_T12     = q3_T12,
    min_T12    = min(paired$T12),
    max_T12    = max(paired$T12),
    range_T12  = max(paired$T12) - min(paired$T12),
    diff_estimate = if (is_t) res_obj$mean_diff else res_obj$median_diff,
    ci_low        = if (is_t) res_obj$ci[1] else NA_real_,
    ci_high       = if (is_t) res_obj$ci[2] else NA_real_,
    effect_type   = res_obj$effect_size$type,
    effect_value  = res_obj$effect_size$value
  )
}

row_ALT <- .build_row_trans(df_trans, "ALT", res_ALT)
row_AST <- .build_row_trans(df_trans, "AST", res_AST)
row_GGT <- .build_row_trans(df_trans, "GGT", res_GGT)

trans_stats <- dplyr::bind_rows(row_ALT, row_AST, row_GGT)

trans_stats$p_fdr <- p.adjust(trans_stats$p_value, method = "BH")

if (!dir.exists(here("results"))) dir.create(here("results"))
readr::write_excel_csv(trans_stats, here("results", "transaminases_stats.csv"))
```


# ------------------------------------------------------------------------------
# CALPROTECTIN
# ------------------------------------------------------------------------------

### Load in the data
```{r Load in data calpro, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_calpro <- read_delim(here("data", "calprotectin.csv"))
```

### Filter data by time
```{r Filter data by time calpro, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
data_calpro_T0 <- subset(df_calpro, Time == "T0")
data_calpro_T12 <- subset(df_calpro, Time == "T12")
```

### Function for statistic test
```{r Function for statistic test based on shapiro results calpro, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
auto_paired_test_calpro <- function(df,
                                    id_col   = "PatientID",
                                    time_col = "Time",
                                    value_col = "Calprotectin",
                                    levels   = c("T0","T12")) {
  stopifnot(all(c(id_col, time_col, value_col) %in% names(df)))
  
  dfw <- df[, c(id_col, time_col, value_col)]
  dfw <- tidyr::pivot_wider(
    dfw,
    names_from  = !!rlang::sym(time_col),
    values_from = !!rlang::sym(value_col)
  )
  if (!all(levels %in% names(dfw))) {
    stop("No columns found ", paste(levels, collapse = " and "), " after pivot_wider().")
  }
  dfw <- dfw[stats::complete.cases(dfw[, levels]), ]
  
  x0   <- dfw[[levels[1]]]
  x12  <- dfw[[levels[2]]]
  diff <- x12 - x0
  n    <- length(diff)
  if (n < 3) stop("Very few valid pairs (n < 3).")
  
  sh     <- shapiro.test(diff)
  normal <- sh$p.value >= 0.05
  
  if (normal) {
    tt        <- t.test(x12, x0, paired = TRUE)
    mean_diff <- mean(diff, na.rm = TRUE)
    sd_diff   <- stats::sd(diff, na.rm = TRUE)
    d_paired  <- mean_diff / sd_diff
    
    mean_T0  <- mean(x0,  na.rm = TRUE); sd_T0  <- stats::sd(x0,  na.rm = TRUE)
    mean_T12 <- mean(x12, na.rm = TRUE); sd_T12 <- stats::sd(x12, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Paired t-test\n")
    cat("n =", n, "| t =", round(tt$statistic, 4), "| p =", signif(tt$p.value, 4), "\n")
    cat("T0:   mean ± SD = ", round(mean_T0, 3), " ± ", round(sd_T0, 3), "\n", sep = "")
    cat("T12:  mean ± SD = ", round(mean_T12, 3), " ± ", round(sd_T12, 3), "\n", sep = "")
    cat("Mean difference (T12−T0) =", round(mean_diff, 3),
        " | 95% CI [", round(tt$conf.int[1], 3), ", ", round(tt$conf.int[2], 3), "]\n", sep = "")
    cat("Effect size (Cohen's d paired) =", round(d_paired, 3), "\n")
    
    invisible(list(variable = value_col, method = "paired_t",
                   n = n, shapiro = sh, test = tt,
                   group_summary = list(
                     T0  = c(mean = mean_T0, sd = sd_T0),
                     T12 = c(mean = mean_T12, sd = sd_T12)
                   ),
                   mean_diff = mean_diff, ci = tt$conf.int,
                   effect_size = list(type = "cohen_d_paired", value = d_paired)))
    
  } else {
    wt   <- wilcox.test(x0, x12, paired = TRUE, exact = FALSE, correct = FALSE)
    V    <- as.numeric(wt$statistic)
    r_rb <- 1 - (2 * V) / (n * (n + 1))
    
    med_T0   <- stats::median(x0,  na.rm = TRUE)
    q1_T0    <- stats::quantile(x0,  0.25, na.rm = TRUE)
    q3_T0    <- stats::quantile(x0,  0.75, na.rm = TRUE)
    med_T12  <- stats::median(x12, na.rm = TRUE)
    q1_T12   <- stats::quantile(x12, 0.25, na.rm = TRUE)
    q3_T12   <- stats::quantile(x12, 0.75, na.rm = TRUE)
    med_diff <- stats::median(diff, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Not normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Wilcoxon signed-rank (paired)\n")
    cat("n =", n, "| V =", V, "| p =", signif(wt$p.value, 4), "\n")
    cat("T0:   median [IQR] = ", round(med_T0, 3), " [",
        round(q1_T0, 3), "–", round(q3_T0, 3), "]\n", sep = "")
    cat("T12:  median [IQR] = ", round(med_T12, 3), " [",
        round(q1_T12, 3), "–", round(q3_T12, 3), "]\n", sep = "")
    cat("Median difference (T12−T0) =", round(med_diff, 3), "\n")
    cat("Effect size (rank-biserial r) =", round(r_rb, 3), "\n")
    
    invisible(list(variable = value_col, method = "wilcoxon_paired",
                   n = n, shapiro = sh, test = wt,
                   group_summary = list(
                     T0  = c(median = med_T0, Q1 = q1_T0, Q3 = q3_T0),
                     T12 = c(median = med_T12, Q1 = q1_T12, Q3 = q3_T12)
                   ),
                   median_diff = med_diff,
                   effect_size = list(type = "rank_biserial_r", value = r_rb, V = V)))
  }
}
# Wrapper "safe": if n < 3, returns a valid res_obj object with p = NA instead of stopping
auto_paired_test_calpro_safe <- function(df,
                                       id_col   = "PatientID",
                                       time_col = "Time",
                                       value_col = "Calprotectin",
                                       levels   = c("T0","T12")) {
  tryCatch(
    auto_paired_test_calpro(df,
                          id_col = id_col,
                          time_col = time_col,
                          value_col = value_col,
                          levels = levels),
    error = function(e) {
      if (grepl("Very few valid pairs", e$message) || grepl("n < 3", e$message)) {
        list(
          variable = value_col,
          method   = "too_few_pairs",
          n        = NA_integer_,
          shapiro  = list(p.value = NA_real_),
          test     = list(statistic = NA_real_, p.value = NA_real_),
          effect_size = list(type = NA_character_, value = NA_real_)
        )
      } else {
        stop(e)
      }
    }
  )
}
```

### Perform statistical test
```{r Execute stats function calpro, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_calpro <- auto_paired_test_calpro(df_calpro, value_col = "Calprotectin")
```

### Run plots function
```{r Execute plot calpro , echo=TRUE, results='show', message=FALSE, warning=FALSE}
p_calpro <- polish_panel_plot(
  make_paired_plot(df_calpro, "Calprotectin", res_calpro,
                        "Fecal calprotectin (µg/g)", "calprotectin.svg")
)
```

### Save stats results
```{r Save stats results calpro , echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# === Export: calprotectin stats to CSV ===
is_t <- identical(res_calpro$method, "paired_t")

paired_calpro <- df_calpro %>%
  dplyr::select(PatientID, Time, Calprotectin) %>%
  tidyr::pivot_wider(names_from = Time, values_from = Calprotectin) %>%
  dplyr::filter(stats::complete.cases(T0, T12))

mean_T0 <- mean(paired_calpro$T0);  sd_T0 <- sd(paired_calpro$T0)
mean_T12<- mean(paired_calpro$T12); sd_T12<- sd(paired_calpro$T12)

med_T0 <- median(paired_calpro$T0);  q1_T0 <- as.numeric(quantile(paired_calpro$T0, 0.25))
q3_T0  <- as.numeric(quantile(paired_calpro$T0, 0.75))
med_T12<- median(paired_calpro$T12); q1_T12<- as.numeric(quantile(paired_calpro$T12,0.25))
q3_T12 <- as.numeric(quantile(paired_calpro$T12,0.75))

fmt_group_T0  <- if (is_t) {
  sprintf("%.3f ± %.3f", mean_T0, sd_T0)
} else {
  sprintf("%.3f [%0.3f-%0.3f]", med_T0, q1_T0, q3_T0)
}

fmt_group_T12 <- if (is_t) {
  sprintf("%.3f ± %.3f", mean_T12, sd_T12)
} else {
  sprintf("%.3f [%0.3f-%0.3f]", med_T12, q1_T12, q3_T12)
}

fisher_res <- fisher.test(table_calpro, conf.int = TRUE)

row_calpro <- tibble::tibble(
  variable   = res_calpro$variable,
  method     = res_calpro$method,
  n_pairs    = res_calpro$n,
  shapiro_p  = unname(res_calpro$shapiro$p.value),
  stat       = unname(res_calpro$test$statistic),
  p_value    = res_calpro$test$p.value,
  group_T0   = fmt_group_T0,
  group_T12  = fmt_group_T12,
  mean_T0    = mean_T0,
  sd_T0      = sd_T0,
  median_T0  = med_T0,
  q1_T0      = q1_T0,
  q3_T0      = q3_T0,
  min_T0     = min(paired_calpro$T0),
  max_T0     = max(paired_calpro$T0),
  range_T0   = max(paired_calpro$T0) - min(paired_calpro$T0),
  mean_T12   = mean_T12,
  sd_T12     = sd_T12,
  median_T12 = med_T12,
  q1_T12     = q1_T12,
  q3_T12     = q3_T12,
  min_T12    = min(paired_calpro$T12),
  max_T12    = max(paired_calpro$T12),
  range_T12  = max(paired_calpro$T12) - min(paired_calpro$T12),
  diff_estimate = if (is_t) res_calpro$mean_diff else res_calpro$median_diff,
  ci_low        = if (is_t) res_calpro$ci[1] else NA_real_,
  ci_high       = if (is_t) res_calpro$ci[2] else NA_real_,
  effect_type   = res_calpro$effect_size$type,
  effect_value  = res_calpro$effect_size$value,
  fisher_p      = as.numeric(fisher_res$p.value),
  fisher_or     = if (!is.null(fisher_res$estimate)) as.numeric(fisher_res$estimate) else NA_real_,
  fisher_ci_low = if (!is.null(fisher_res$conf.int)) fisher_res$conf.int[1] else NA_real_,
  fisher_ci_high= if (!is.null(fisher_res$conf.int)) fisher_res$conf.int[2] else NA_real_,
  or_ha         = or_ha,
  or_ha_ci_low  = ci_low,
  or_ha_ci_high = ci_high,
  cutoff_used   = cutoff_calpro
)

if (!dir.exists(here("results"))) dir.create(here("results"))

readr::write_excel_csv(row_calpro, here("results", "calprotectin_stats.csv"))
```


# ------------------------------------------------------------------------------
# ELASTASE
# ------------------------------------------------------------------------------

### Load in the data
```{r Load in data elas, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_elas <- read_delim(here("data", "elastase.csv"))
```

### Filter data by time
```{r Filter data by time elas, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
data_elas_T0 <- subset(df_elas, Time == "T0")
data_elas_T12 <- subset(df_elas, Time == "T12")
```

### Function for statistic test
```{r Function for statistic test based on shapiro results elas, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
auto_paired_test_elas <- function(df,
                                  id_col   = "PatientID",
                                  time_col = "Time",
                                  value_col = "Elastase",
                                  levels   = c("T0","T12")) {
  stopifnot(all(c(id_col, time_col, value_col) %in% names(df)))
  
  dfw <- df[, c(id_col, time_col, value_col)]
  dfw <- tidyr::pivot_wider(
    dfw,
    names_from  = !!rlang::sym(time_col),
    values_from = !!rlang::sym(value_col)
  )
  if (!all(levels %in% names(dfw))) {
    stop("No columns found ", paste(levels, collapse = " and "), " after pivot_wider().")
  }
  dfw <- dfw[stats::complete.cases(dfw[, levels]), ]
  
  x0   <- dfw[[levels[1]]]
  x12  <- dfw[[levels[2]]]
  diff <- x12 - x0
  n    <- length(diff)
  if (n < 3) stop("Very few valid pairs (n < 3).")
  
  sh     <- shapiro.test(diff)
  normal <- sh$p.value >= 0.05
  
  if (normal) {
    tt        <- t.test(x12, x0, paired = TRUE)
    mean_diff <- mean(diff, na.rm = TRUE)
    sd_diff   <- stats::sd(diff, na.rm = TRUE)
    d_paired  <- mean_diff / sd_diff
    
    mean_T0  <- mean(x0,  na.rm = TRUE); sd_T0  <- stats::sd(x0,  na.rm = TRUE)
    mean_T12 <- mean(x12, na.rm = TRUE); sd_T12 <- stats::sd(x12, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Paired t-test\n")
    cat("n =", n, "| t =", round(tt$statistic, 4), "| p =", signif(tt$p.value, 4), "\n")
    cat("T0:   mean ± SD = ", round(mean_T0, 3), " ± ", round(sd_T0, 3), "\n", sep = "")
    cat("T12:  mean ± SD = ", round(mean_T12, 3), " ± ", round(sd_T12, 3), "\n", sep = "")
    cat("Mean difference (T12−T0) =", round(mean_diff, 3),
        " | 95% CI [", round(tt$conf.int[1], 3), ", ", round(tt$conf.int[2], 3), "]\n", sep = "")
    cat("Effect size (Cohen's d paired) =", round(d_paired, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "paired_t",
      n = n,
      shapiro = sh,
      test = tt,
      group_summary = list(
        T0  = c(mean = mean_T0, sd = sd_T0),
        T12 = c(mean = mean_T12, sd = sd_T12)
      ),
      mean_diff = mean_diff,
      ci = tt$conf.int,
      effect_size = list(type = "cohen_d_paired", value = d_paired)
    ))
    
  } else {
    wt   <- wilcox.test(x0, x12, paired = TRUE, exact = FALSE, correct = FALSE)
    V    <- as.numeric(wt$statistic)
    r_rb <- 1 - (2 * V) / (n * (n + 1))
    
    med_T0   <- stats::median(x0,  na.rm = TRUE)
    q1_T0    <- stats::quantile(x0,  0.25, na.rm = TRUE)
    q3_T0    <- stats::quantile(x0,  0.75, na.rm = TRUE)
    med_T12  <- stats::median(x12, na.rm = TRUE)
    q1_T12   <- stats::quantile(x12, 0.25, na.rm = TRUE)
    q3_T12   <- stats::quantile(x12, 0.75, na.rm = TRUE)
    med_diff <- stats::median(diff, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Not normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Wilcoxon signed-rank (paired)\n")
    cat("n =", n, "| V =", V, "| p =", signif(wt$p.value, 4), "\n")
    cat("T0:   median [IQR] = ", round(med_T0, 3), " [",
        round(q1_T0, 3), "–", round(q3_T0, 3), "]\n", sep = "")
    cat("T12:  median [IQR] = ", round(med_T12, 3), " [",
        round(q1_T12, 3), "–", round(q3_T12, 3), "]\n", sep = "")
    cat("Median difference (T12−T0) =", round(med_diff, 3), "\n")
    cat("Effect size (rank-biserial r) =", round(r_rb, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "wilcoxon_paired",
      n = n,
      shapiro = sh,
      test = wt,
      group_summary = list(
        T0  = c(median = med_T0, Q1 = q1_T0, Q3 = q3_T0),
        T12 = c(median = med_T12, Q1 = q1_T12, Q3 = q3_T12)
      ),
      median_diff = med_diff,
      effect_size = list(type = "rank_biserial_r", value = r_rb, V = V)
    ))
  }
}
# Wrapper "safe": if n < 3, returns a valid res_obj object with p = NA instead of stopping
auto_paired_test_elas_safe <- function(df,
                                       id_col   = "PatientID",
                                       time_col = "Time",
                                       value_col = "Elastase",
                                       levels   = c("T0","T12")) {
  tryCatch(
    auto_paired_test_elas(df,
                          id_col = id_col,
                          time_col = time_col,
                          value_col = value_col,
                          levels = levels),
    error = function(e) {
      if (grepl("Very few valid pairs", e$message) || grepl("n < 3", e$message)) {
        list(
          variable = value_col,
          method   = "too_few_pairs",
          n        = NA_integer_,
          shapiro  = list(p.value = NA_real_),
          test     = list(statistic = NA_real_, p.value = NA_real_),
          effect_size = list(type = NA_character_, value = NA_real_)
        )
      } else {
        stop(e)
      }
    }
  )
}
```

### Perform statistical test
#### Elastase
```{r Execute stats function elas, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_elas <- auto_paired_test_elas(df_elas, value_col = "Elastase")
```

### Run plots function
```{r Execute plot elas , echo=TRUE, results='show', message=FALSE, warning=FALSE}
p_elas <- polish_panel_plot(
  make_paired_plot(df_elas, "Elastase", res_elas,
                        "Fecal elastase (µg/g)", "elastase.svg")
)
```

### Save stats results
```{r Save stats results elas , echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# === Export: elastase stats to CSV ===
is_t <- identical(res_elas$method, "paired_t")

paired_elas <- df_elas %>%
  dplyr::select(PatientID, Time, Elastase) %>%
  tidyr::pivot_wider(names_from = Time, values_from = Elastase) %>%
  dplyr::filter(stats::complete.cases(T0, T12))

mean_T0 <- mean(paired_elas$T0);  sd_T0 <- sd(paired_elas$T0)
mean_T12<- mean(paired_elas$T12); sd_T12<- sd(paired_elas$T12)

med_T0 <- median(paired_elas$T0);  q1_T0 <- as.numeric(quantile(paired_elas$T0, 0.25))
q3_T0  <- as.numeric(quantile(paired_elas$T0, 0.75))
med_T12<- median(paired_elas$T12); q1_T12<- as.numeric(quantile(paired_elas$T12,0.25))
q3_T12 <- as.numeric(quantile(paired_elas$T12,0.75))

fmt_group_T0  <- if (is_t) {
  sprintf("%.3f ± %.3f", mean_T0, sd_T0)
} else {
  sprintf("%.3f [%0.3f-%0.3f]", med_T0, q1_T0, q3_T0)
}

fmt_group_T12 <- if (is_t) {
  sprintf("%.3f ± %.3f", mean_T12, sd_T12)
} else {
  sprintf("%.3f [%0.3f-%0.3f]", med_T12, q1_T12, q3_T12)
}

fisher_res <- fisher.test(table_elas, conf.int = TRUE)

row_elas <- tibble::tibble(
  variable   = res_elas$variable,
  method     = res_elas$method,
  n_pairs    = res_elas$n,
  shapiro_p  = unname(res_elas$shapiro$p.value),
  stat       = unname(res_elas$test$statistic),
  p_value    = res_elas$test$p.value,
  group_T0   = fmt_group_T0,
  group_T12  = fmt_group_T12,
  mean_T0    = mean_T0,
  sd_T0      = sd_T0,
  median_T0  = med_T0,
  q1_T0      = q1_T0,
  q3_T0      = q3_T0,
  min_T0     = min(paired_elas$T0),
  max_T0     = max(paired_elas$T0),
  range_T0   = max(paired_elas$T0) - min(paired_elas$T0),
  mean_T12   = mean_T12,
  sd_T12     = sd_T12,
  median_T12 = med_T12,
  q1_T12     = q1_T12,
  q3_T12     = q3_T12,
  min_T12    = min(paired_elas$T12),
  max_T12    = max(paired_elas$T12),
  range_T12  = max(paired_elas$T12) - min(paired_elas$T12),
  diff_estimate = if (is_t) res_elas$mean_diff else res_elas$median_diff,
  ci_low        = if (is_t) res_elas$ci[1] else NA_real_,
  ci_high       = if (is_t) res_elas$ci[2] else NA_real_,
  effect_type   = res_elas$effect_size$type,
  effect_value  = res_elas$effect_size$value,
  fisher_p      = as.numeric(fisher_res$p.value),
  fisher_or     = if (!is.null(fisher_res$estimate)) as.numeric(fisher_res$estimate) else NA_real_,
  fisher_ci_low = if (!is.null(fisher_res$conf.int)) fisher_res$conf.int[1] else NA_real_,
  fisher_ci_high= if (!is.null(fisher_res$conf.int)) fisher_res$conf.int[2] else NA_real_,
  or_ha         = or_ha,
  or_ha_ci_low  = ci_low,
  or_ha_ci_high = ci_high,
  cutoff_used   = cutoff_elas
)

if (!dir.exists(here("results"))) dir.create(here("results"))

readr::write_excel_csv(row_elas, here("results", "elastase_stats.csv"))
```


# ------------------------------------------------------------------------------
# ANTHROPOMETRY ADULTS
# ------------------------------------------------------------------------------

### Load in the data
```{r Load in data Z-scores, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_bmi <- read_delim(here("data", "BMI_adults.csv"))
```

### Filter data by time
```{r Filter data by time Z-scores, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
data_bmi_T0  <- subset(df_bmi, Time == "T0")
data_bmi_T12 <- subset(df_bmi, Time == "T12")
```

#### Function for statistic test
```{r Function for statistic test based on shapiro results Z-scores, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
auto_paired_test_bmi <- function(df,
                               id_col   = "PatientID",
                               time_col = "Time",
                               value_col = "BMI_kgm2",
                               levels   = c("T0","T12")) {
  stopifnot(all(c(id_col, time_col, value_col) %in% names(df)))
  
  dfw <- df[, c(id_col, time_col, value_col)]
  dfw <- tidyr::pivot_wider(
    dfw,
    names_from  = !!rlang::sym(time_col),
    values_from = !!rlang::sym(value_col)
  )
  if (!all(levels %in% names(dfw))) {
    stop("No columns found ", paste(levels, collapse = " and "), " after pivot_wider().")
  }
  dfw <- dfw[stats::complete.cases(dfw[, levels]), ]
  
  x0   <- dfw[[levels[1]]]
  x12  <- dfw[[levels[2]]]
  diff <- x12 - x0
  n    <- length(diff)
  if (n < 3) stop("Very few valid pairs (n < 3).")
  
  sh     <- shapiro.test(diff)
  normal <- sh$p.value >= 0.05
  
  if (normal) {
    tt        <- t.test(x12, x0, paired = TRUE)
    mean_diff <- mean(diff, na.rm = TRUE)
    sd_diff   <- stats::sd(diff, na.rm = TRUE)
    d_paired  <- mean_diff / sd_diff
    
    mean_T0  <- mean(x0,  na.rm = TRUE); sd_T0  <- stats::sd(x0,  na.rm = TRUE)
    mean_T12 <- mean(x12, na.rm = TRUE); sd_T12 <- stats::sd(x12, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Paired t-test\n")
    cat("n =", n, "| t =", round(tt$statistic, 4), "| p =", signif(tt$p.value, 4), "\n")
    cat("T0:   mean ± SD = ", round(mean_T0, 3), " ± ", round(sd_T0, 3), "\n", sep = "")
    cat("T12:  mean ± SD = ", round(mean_T12, 3), " ± ", round(sd_T12, 3), "\n", sep = "")
    cat("Mean difference (T12−T0) =", round(mean_diff, 3),
        " | 95% CI [", round(tt$conf.int[1], 3), ", ", round(tt$conf.int[2], 3), "]\n", sep = "")
    cat("Effect size (Cohen's d paired) =", round(d_paired, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "paired_t",
      n = n,
      shapiro = sh,
      test = tt,
      group_summary = list(
        T0  = c(mean = mean_T0, sd = sd_T0),
        T12 = c(mean = mean_T12, sd = sd_T12)
      ),
      mean_diff = mean_diff,
      ci = tt$conf.int,
      effect_size = list(type = "cohen_d_paired", value = d_paired)
    ))
    
  } else {
    wt   <- wilcox.test(x0, x12, paired = TRUE, exact = FALSE, correct = FALSE)
    V    <- as.numeric(wt$statistic)
    r_rb <- 1 - (2 * V) / (n * (n + 1))
    
    med_T0   <- stats::median(x0,  na.rm = TRUE)
    q1_T0    <- stats::quantile(x0,  0.25, na.rm = TRUE)
    q3_T0    <- stats::quantile(x0,  0.75, na.rm = TRUE)
    med_T12  <- stats::median(x12, na.rm = TRUE)
    q1_T12   <- stats::quantile(x12, 0.25, na.rm = TRUE)
    q3_T12   <- stats::quantile(x12, 0.75, na.rm = TRUE)
    med_diff <- stats::median(diff, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Not normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Wilcoxon signed-rank (paired)\n")
    cat("n =", n, "| V =", V, "| p =", signif(wt$p.value, 4), "\n")
    cat("T0:   median [IQR] = ", round(med_T0, 3), " [",
        round(q1_T0, 3), "–", round(q3_T0, 3), "]\n", sep = "")
    cat("T12:  median [IQR] = ", round(med_T12, 3), " [",
        round(q1_T12, 3), "–", round(q3_T12, 3), "]\n", sep = "")
    cat("Median difference (T12−T0) =", round(med_diff, 3), "\n")
    cat("Effect size (rank-biserial r) =", round(r_rb, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "wilcoxon_paired",
      n = n,
      shapiro = sh,
      test = wt,
      group_summary = list(
        T0  = c(median = med_T0, Q1 = q1_T0, Q3 = q3_T0),
        T12 = c(median = med_T12, Q1 = q1_T12, Q3 = q3_T12)
      ),
      median_diff = med_diff,
      effect_size = list(type = "rank_biserial_r", value = r_rb, V = V)
    ))
  }
}
# Wrapper "safe": if n < 3, returns a valid res_obj object with p = NA instead of stopping
auto_paired_test_bmi_safe <- function(df,
                                       id_col   = "PatientID",
                                       time_col = "Time",
                                       value_col = "BMI_kgm2",
                                       levels   = c("T0","T12")) {
  tryCatch(
    auto_paired_test_bmi(df,
                          id_col = id_col,
                          time_col = time_col,
                          value_col = value_col,
                          levels = levels),
    error = function(e) {
      if (grepl("Very few valid pairs", e$message) || grepl("n < 3", e$message)) {
        list(
          variable = value_col,
          method   = "too_few_pairs",
          n        = NA_integer_,
          shapiro  = list(p.value = NA_real_),
          test     = list(statistic = NA_real_, p.value = NA_real_),
          effect_size = list(type = NA_character_, value = NA_real_)
        )
      } else {
        stop(e)
      }
    }
  )
}
```

### Perform the statistical test
#### Weight
```{r Execute stats function Z-scores, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_bmi_wt  <- auto_paired_test_bmi(df_bmi, value_col = "Weight_kg")
```
#### Height
```{r Execute stats function Z-scores, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_bmi_ht  <- auto_paired_test_bmi(df_bmi, value_col = "Height_cm")
```
#### BMI
```{r Execute stats function Z-scores, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_bmi_bmi <- auto_paired_test_bmi(df_bmi, value_col = "BMI_kgm2")
```

#### Panel plots
```{r Panel plots Z-scores, echo=TRUE, results='show', message=FALSE, warning=FALSE}
p_bmi_wt <- polish_panel_plot(
  make_paired_plot(df_bmi, "Weight_kg",
                   res_bmi_wt, "Weight (kg)", "weight.svg")
)

p_bmi_ht <- polish_panel_plot(
  make_paired_plot(df_bmi, "Height_cm",
                   res_bmi_ht, "Height (cm)", "height.svg")
)

p_bmi_bmi <- polish_panel_plot(
  make_paired_plot(df_bmi, "BMI_kgm2",
                   res_bmi_bmi, "BMI (kg/m2)", "BMI.svg")
)

panel_bmi <- p_bmi_wt | p_bmi_ht | p_bmi_bmi

panel_bmi <- panel_bmi + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

ggsave(here("results","zscore_adults_panel.svg"),
       panel_bmi, width = 12, height = 5, bg = "white")
```

### Save stats results
```{r Save stats results z_score, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# === Export: anthropometry (weight, height, BMI) ===

.build_row_bmi <- function(df, value_col, res_obj, p_fdr = NA_real_) {
  stopifnot(all(c("PatientID","Time", value_col) %in% names(df)))
  is_t <- identical(res_obj$method, "paired_t")

  paired <- df %>%
    dplyr::select(PatientID, Time, !!rlang::sym(value_col)) %>%
    tidyr::pivot_wider(names_from = Time, values_from = !!rlang::sym(value_col)) %>%
    dplyr::filter(stats::complete.cases(T0, T12))

  mean_T0  <- mean(paired$T0);  sd_T0  <- sd(paired$T0)
  mean_T12 <- mean(paired$T12); sd_T12 <- sd(paired$T12)

  med_T0 <- median(paired$T0)
  q1_T0  <- as.numeric(quantile(paired$T0,  0.25))
  q3_T0  <- as.numeric(quantile(paired$T0,  0.75))

  med_T12 <- median(paired$T12)
  q1_T12  <- as.numeric(quantile(paired$T12, 0.25))
  q3_T12  <- as.numeric(quantile(paired$T12, 0.75))

  fmt_group_T0  <- if (is_t) sprintf("%.3f ± %.3f", mean_T0, sd_T0)  else sprintf("%.3f [%0.3f-%0.3f]", med_T0,  q1_T0,  q3_T0)
  fmt_group_T12 <- if (is_t) sprintf("%.3f ± %.3f", mean_T12, sd_T12) else sprintf("%.3f [%0.3f-%0.3f]", med_T12, q1_T12, q3_T12)

  tibble::tibble(
    variable      = res_obj$variable,
    method        = res_obj$method,
    n_pairs       = res_obj$n,
    shapiro_p     = unname(res_obj$shapiro$p.value),
    stat          = unname(res_obj$test$statistic),
    p_value       = res_obj$test$p.value,
    group_T0      = fmt_group_T0,
    group_T12     = fmt_group_T12,
    mean_T0  = mean_T0,
    sd_T0    = sd_T0,
    median_T0= med_T0,
    q1_T0    = q1_T0,
    q3_T0    = q3_T0,
    min_T0   = min(paired$T0),
    max_T0   = max(paired$T0),
    range_T0 = max(paired$T0) - min(paired$T0),
    mean_T12   = mean_T12,
    sd_T12     = sd_T12,
    median_T12 = med_T12,
    q1_T12     = q1_T12,
    q3_T12     = q3_T12,
    min_T12    = min(paired$T12),
    max_T12    = max(paired$T12),
    range_T12  = max(paired$T12) - min(paired$T12),
    diff_estimate = if (is_t) res_obj$mean_diff else res_obj$median_diff,
    ci_low        = if (is_t) res_obj$ci[1] else NA_real_,
    ci_high       = if (is_t) res_obj$ci[2] else NA_real_,
    effect_type   = res_obj$effect_size$type,
    effect_value  = res_obj$effect_size$value
  )
}

row_bmi_wt <- .build_row_bmi(df_bmi, "Weight_kg", res_bmi_wt)
row_bmi_ht <- .build_row_bmi(df_bmi, "Height_cm", res_bmi_ht)
row_bmi_bmi <- .build_row_bmi(df_bmi, "BMI_kgm2", res_bmi_bmi)

bmi_stats <- dplyr::bind_rows(row_bmi_wt, row_bmi_ht, row_bmi_bmi)

bmi_stats$p_fdr <- p.adjust(bmi_stats$p_value, method = "BH")

if (!dir.exists(here("results"))) dir.create(here("results"))
readr::write_excel_csv(bmi_stats, here("results", "bmi_stats.csv"))
```


# ------------------------------------------------------------------------------
# ANTHROPOMETRY PEDIATRICS
# ------------------------------------------------------------------------------

### Load in the data
```{r Load in data Z-scores, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_z <- read_delim(here("data", "zscore.csv"))
```

### Filter data by time
```{r Filter data by time Z-scores, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
data_z_T0  <- subset(df_z, Time == "T0")
data_z_T12 <- subset(df_z, Time == "T12")
```

#### Function for statistic test
```{r Function for statistic test based on shapiro results Z-scores, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
auto_paired_test_z <- function(df,
                               id_col   = "PatientID",
                               time_col = "Time",
                               value_col = "SD_z_score_BMI_2010",
                               levels   = c("T0","T12")) {
  stopifnot(all(c(id_col, time_col, value_col) %in% names(df)))
  
  dfw <- df[, c(id_col, time_col, value_col)]
  dfw <- tidyr::pivot_wider(
    dfw,
    names_from  = !!rlang::sym(time_col),
    values_from = !!rlang::sym(value_col)
  )
  if (!all(levels %in% names(dfw))) {
    stop("No columns found ", paste(levels, collapse = " and "), " after pivot_wider().")
  }
  dfw <- dfw[stats::complete.cases(dfw[, levels]), ]
  
  x0   <- dfw[[levels[1]]]
  x12  <- dfw[[levels[2]]]
  diff <- x12 - x0
  n    <- length(diff)
  if (n < 3) stop("Very few valid pairs (n < 3).")
  
  sh     <- shapiro.test(diff)
  normal <- sh$p.value >= 0.05
  
  if (normal) {
    tt        <- t.test(x12, x0, paired = TRUE)
    mean_diff <- mean(diff, na.rm = TRUE)
    sd_diff   <- stats::sd(diff, na.rm = TRUE)
    d_paired  <- mean_diff / sd_diff
    
    mean_T0  <- mean(x0,  na.rm = TRUE); sd_T0  <- stats::sd(x0,  na.rm = TRUE)
    mean_T12 <- mean(x12, na.rm = TRUE); sd_T12 <- stats::sd(x12, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Paired t-test\n")
    cat("n =", n, "| t =", round(tt$statistic, 4), "| p =", signif(tt$p.value, 4), "\n")
    cat("T0:   mean ± SD = ", round(mean_T0, 3), " ± ", round(sd_T0, 3), "\n", sep = "")
    cat("T12:  mean ± SD = ", round(mean_T12, 3), " ± ", round(sd_T12, 3), "\n", sep = "")
    cat("Mean difference (T12−T0) =", round(mean_diff, 3),
        " | 95% CI [", round(tt$conf.int[1], 3), ", ", round(tt$conf.int[2], 3), "]\n", sep = "")
    cat("Effect size (Cohen's d paired) =", round(d_paired, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "paired_t",
      n = n,
      shapiro = sh,
      test = tt,
      group_summary = list(
        T0  = c(mean = mean_T0, sd = sd_T0),
        T12 = c(mean = mean_T12, sd = sd_T12)
      ),
      mean_diff = mean_diff,
      ci = tt$conf.int,
      effect_size = list(type = "cohen_d_paired", value = d_paired)
    ))
    
  } else {
    wt   <- wilcox.test(x0, x12, paired = TRUE, exact = FALSE, correct = FALSE)
    V    <- as.numeric(wt$statistic)
    r_rb <- 1 - (2 * V) / (n * (n + 1))
    
    med_T0   <- stats::median(x0,  na.rm = TRUE)
    q1_T0    <- stats::quantile(x0,  0.25, na.rm = TRUE)
    q3_T0    <- stats::quantile(x0,  0.75, na.rm = TRUE)
    med_T12  <- stats::median(x12, na.rm = TRUE)
    q1_T12   <- stats::quantile(x12, 0.25, na.rm = TRUE)
    q3_T12   <- stats::quantile(x12, 0.75, na.rm = TRUE)
    med_diff <- stats::median(diff, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Not normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Wilcoxon signed-rank (paired)\n")
    cat("n =", n, "| V =", V, "| p =", signif(wt$p.value, 4), "\n")
    cat("T0:   median [IQR] = ", round(med_T0, 3), " [",
        round(q1_T0, 3), "–", round(q3_T0, 3), "]\n", sep = "")
    cat("T12:  median [IQR] = ", round(med_T12, 3), " [",
        round(q1_T12, 3), "–", round(q3_T12, 3), "]\n", sep = "")
    cat("Median difference (T12−T0) =", round(med_diff, 3), "\n")
    cat("Effect size (rank-biserial r) =", round(r_rb, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "wilcoxon_paired",
      n = n,
      shapiro = sh,
      test = wt,
      group_summary = list(
        T0  = c(median = med_T0, Q1 = q1_T0, Q3 = q3_T0),
        T12 = c(median = med_T12, Q1 = q1_T12, Q3 = q3_T12)
      ),
      median_diff = med_diff,
      effect_size = list(type = "rank_biserial_r", value = r_rb, V = V)
    ))
  }
}
# Wrapper "safe": if n < 3, returns a valid res_obj object with p = NA instead of stopping
auto_paired_test_z_safe <- function(df,
                                       id_col   = "PatientID",
                                       time_col = "Time",
                                       value_col = "SD_z_score_BMI_2010",
                                       levels   = c("T0","T12")) {
  tryCatch(
    auto_paired_test_z(df,
                          id_col = id_col,
                          time_col = time_col,
                          value_col = value_col,
                          levels = levels),
    error = function(e) {
      if (grepl("Very few valid pairs", e$message) || grepl("n < 3", e$message)) {
        list(
          variable = value_col,
          method   = "too_few_pairs",
          n        = NA_integer_,
          shapiro  = list(p.value = NA_real_),
          test     = list(statistic = NA_real_, p.value = NA_real_),
          effect_size = list(type = NA_character_, value = NA_real_)
        )
      } else {
        stop(e)
      }
    }
  )
}
```

### Perform the statistical test
#### Weight
```{r Execute stats function Z-scores, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_z_wt  <- auto_paired_test_z(df_z, value_col = "SD_z_score_weight_2010")
```
#### Height
```{r Execute stats function Z-scores, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_z_ht  <- auto_paired_test_z(df_z, value_col = "SD_z_score_height_2010")
```
#### BMI z-score
```{r Execute stats function Z-scores, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_z_bmi <- auto_paired_test_z(df_z, value_col = "SD_z_score_BMI_2010")
```

#### OPTIONAL: BMI increased/decreased
```{r BMI increased or decreased Z-scores, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Paired BMI z-score per patient
paired_bmiz <- df_z %>%
  dplyr::select(PatientID, Time, SD_z_score_BMI_2010) %>%
  tidyr::pivot_wider(names_from = Time, values_from = SD_z_score_BMI_2010) %>%
  dplyr::filter(stats::complete.cases(T0, T12)) %>%
  dplyr::mutate(delta = T12 - T0)

n_increase <- sum(paired_bmiz$delta > 0)
n_decrease <- sum(paired_bmiz$delta < 0)
n_nochange <- sum(paired_bmiz$delta == 0)
n_total    <- nrow(paired_bmiz)

cat("\n=== BMI z-score change (T12 - T0) ===\n")
cat("Paired patients:", n_total, "\n")
cat("Increased:", n_increase, "\n")
cat("Decreased:", n_decrease, "\n")
cat("No change:", n_nochange, "\n")

# Paired BMI per patient
paired_bmi <- df_z %>%
  dplyr::select(PatientID, Time, SD_z_score_BMI_2010) %>%                 
  tidyr::pivot_wider(names_from = Time, values_from = SD_z_score_BMI_2010) %>%
  dplyr::filter(stats::complete.cases(T0, T12)) %>%
  dplyr::mutate(delta = T12 - T0)

n_increase <- sum(paired_bmi$delta > 0)
n_decrease <- sum(paired_bmi$delta < 0)
n_nochange <- sum(paired_bmi$delta == 0)
n_total    <- nrow(paired_bmi)

cat("\n=== BMI change (kg/m2): T12 - T0 ===\n")
cat("Paired patients:", n_total, "\n")
cat("Increased:", n_increase, "\n")
cat("Decreased:", n_decrease, "\n")
cat("No change:", n_nochange, "\n")

# --- Create BMIz change groups (inc/dec) per patient ---
paired_bmiz <- df_z %>%
  dplyr::select(PatientID, Time, SD_z_score_BMI_2010) %>%
  tidyr::pivot_wider(names_from = Time, values_from = SD_z_score_BMI_2010) %>%
  dplyr::filter(stats::complete.cases(T0, T12)) %>%
  dplyr::mutate(delta_bmiz = T12 - T0)

bmiz_group <- paired_bmiz %>%
  dplyr::mutate(
    BMIz_group = dplyr::case_when(
      delta_bmiz > 0 ~ "BMIz_inc",
      delta_bmiz < 0 ~ "BMIz_dec",
      TRUE           ~ "BMIz_nochange"
    )
  ) %>%
  dplyr::select(PatientID, BMIz_group)

print(table(bmiz_group$BMIz_group))

# PatientID
bmiz_ids_compact <- bmiz_group %>%
  dplyr::arrange(BMIz_group, PatientID) %>%
  dplyr::group_by(BMIz_group) %>%
  dplyr::summarise(PatientIDs = paste(PatientID, collapse = ", "), .groups = "drop")

print(bmiz_ids_compact, n = Inf)
```

#### Panel plots
```{r Panel plots Z-scores, echo=TRUE, results='show', message=FALSE, warning=FALSE}
p_z_wt <- polish_panel_plot(
  make_paired_plot(df_z, "SD_z_score_weight_2010",
                   res_z_wt, "Weight z-score (2010)", "zscore_weight.svg")
)

p_z_ht <- polish_panel_plot(
  make_paired_plot(df_z, "SD_z_score_height_2010",
                   res_z_ht, "Height z-score (2010)", "zscore_height.svg")
)

p_z_bmi <- polish_panel_plot(
  make_paired_plot(df_z, "SD_z_score_BMI_2010",
                   res_z_bmi, "BMI z-score (Carrascosa 2010)", "zscore_BMI.svg")
)

panel_z <- p_z_wt | p_z_ht | p_z_bmi

panel_z <- panel_z + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

ggsave(here("results","zscore_panel.svg"),
       panel_z, width = 12, height = 5, bg = "white")
```

#### Alternative plot
```{r Alternative plot Z-scores, echo=TRUE, results='show', message=FALSE, warning=FALSE}
# Set colors
col_box   <- c("T0" = "#9EC1CF", "T12" = "#F9A1BC")
col_point <- c("T0" = "#D6E9F5", "T12" = "#FCD6E0")

# Long format
z_long <- df_z %>%
  dplyr::select(
    PatientID, Time,
    z_weight = SD_z_score_weight_2010,
    z_height = SD_z_score_height_2010,
    z_bmi    = SD_z_score_BMI_2010
  ) %>%
  tidyr::pivot_longer(cols = starts_with("z_"),
                      names_to = "marker", values_to = "zscore") %>%
  mutate(
    marker = recode(marker,
                    z_weight = "Weight z-score",
                    z_height = "Height z-score",
                    z_bmi    = "BMI z-score"),
    marker = factor(marker, levels = c("Weight z-score","Height z-score","BMI z-score")),
    Time   = factor(Time, levels = c("T0","T12"))
  )

# Explicit x positions (so paired lines work cleanly)
offset <- 0.18
z_long <- z_long %>%
  mutate(
    x0 = as.numeric(marker),
    x  = x0 + ifelse(Time == "T0", -offset, +offset)
  )

# Median + IQR per timepoint (for the summary bars + diamond)
summ_df <- z_long %>%
  group_by(marker, Time, x) %>%
  summarise(
    med = median(zscore, na.rm = TRUE),
    q1  = quantile(zscore, 0.25, na.rm = TRUE),
    q3  = quantile(zscore, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

# P-values
pvals_df <- tibble::tibble(
  marker = factor(c("Weight z-score","Height z-score","BMI z-score"),
                  levels = levels(z_long$marker)),
  p = c(res_z_wt$test$p.value, res_z_ht$test$p.value, res_z_bmi$test$p.value)
)

anno_df <- z_long %>%
  group_by(marker) %>%
  summarise(y = max(zscore, na.rm = TRUE) + 0.25, .groups = "drop") %>%
  left_join(pvals_df, by = "marker") %>%
  mutate(
    x0 = as.numeric(marker),
    x1 = x0 - offset,
    x2 = x0 + offset,
    p_label = paste0("p = ", format.pval(p, digits = 2, eps = 0.001))
  )

p_zscores_all <- ggplot() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major.x = element_line(color = "grey85", linewidth = 0.8),
    panel.grid.major.y = element_line(color = "grey85", linewidth = 0.8),
    panel.grid.minor.x = element_line(color = "grey92", linewidth = 0.6),
    panel.grid.minor.y = element_blank(),
    legend.position = "bottom"
    ) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.9, color = "grey55") +
  geom_linerange(
    data = summ_df,
    aes(x = x, ymin = q1, ymax = q3, color = Time),
    linewidth = 1.2, alpha = 0.95
  ) +
  geom_vline(
    xintercept = c(0.5, 1.5, 2.5, 3.5),
    color = "grey90",
    linewidth = 0.8
    ) +
  geom_segment(
    data = summ_df,
    aes(x = x - 0.04, xend = x + 0.04, y = q1, yend = q1, color = Time),
    linewidth = 1.5
  ) +
  geom_segment(
    data = summ_df,
    aes(x = x - 0.04, xend = x + 0.04, y = q3, yend = q3, color = Time),
    linewidth = 1.5
  ) +
  geom_line(
    data = z_long,
    aes(x = x, y = zscore, group = interaction(PatientID, marker)),
    color = "grey70", linetype = "dashed", linewidth = 0.8
  ) +
  geom_point(
    data = z_long,
    aes(x = x, y = zscore, fill = Time),
    shape = 21, size = 3.0, color = "black", stroke = 0.6, alpha = 0.95
  ) +
  geom_point(
    data = summ_df,
    aes(x = x, y = med, color = Time),
    shape = 23, size = 4.0, fill = "white", stroke = 1.0
  ) +
  geom_segment(
    data = anno_df,
    aes(x = x1, xend = x2, y = y, yend = y),
    linewidth = 0.8, color = "black"
  ) +
  geom_text(
    data = anno_df,
    aes(x = x0, y = y + 0.08, label = p_label),
    size = 4.2, color = "black"
  ) +

  scale_color_manual(values = col_box, guide = "none") +
  scale_fill_manual(values = col_point, name = "Time") +
  scale_x_continuous(
    limits = c(0.5, 3.5),
    breaks = 1:length(levels(z_long$marker)),
    labels = levels(z_long$marker),
    minor_breaks = sort(c(
      1 - offset, 1 + offset,
      2 - offset, 2 + offset,
      3 - offset, 3 + offset,
      1.5, 2.5))
    ) +
  labs(x = NULL, y = "z-score (SD units)")

p_zscores_all

ggplot2::ggsave(
  here::here("results", "anthropometry_zscores_T0_T12_combined.svg"),
  p_zscores_all, width = 9, height = 4.5, bg = "white"
)
```


### Save stats results
```{r Save stats results z_score, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# === Export: z-scores (weight, height, BMI) ===

.build_row_z <- function(df, value_col, res_obj, fisher_tab = NULL, cutoff = NA_real_, p_fdr = NA_real_) {
  stopifnot(all(c("PatientID","Time", value_col) %in% names(df)))
  is_t <- identical(res_obj$method, "paired_t")

  paired <- df %>%
    dplyr::select(PatientID, Time, !!rlang::sym(value_col)) %>%
    tidyr::pivot_wider(names_from = Time, values_from = !!rlang::sym(value_col)) %>%
    dplyr::filter(stats::complete.cases(T0, T12))

  mean_T0  <- mean(paired$T0);  sd_T0  <- sd(paired$T0)
  mean_T12 <- mean(paired$T12); sd_T12 <- sd(paired$T12)

  med_T0 <- median(paired$T0)
  q1_T0  <- as.numeric(quantile(paired$T0,  0.25))
  q3_T0  <- as.numeric(quantile(paired$T0,  0.75))

  med_T12 <- median(paired$T12)
  q1_T12  <- as.numeric(quantile(paired$T12, 0.25))
  q3_T12  <- as.numeric(quantile(paired$T12, 0.75))

  fmt_group_T0  <- if (is_t) sprintf("%.3f ± %.3f", mean_T0, sd_T0)  else sprintf("%.3f [%0.3f-%0.3f]", med_T0,  q1_T0,  q3_T0)
  fmt_group_T12 <- if (is_t) sprintf("%.3f ± %.3f", mean_T12, sd_T12) else sprintf("%.3f [%0.3f-%0.3f]", med_T12, q1_T12, q3_T12)

  if (!is.null(fisher_tab)) {
    fisher_res <- fisher.test(fisher_tab, conf.int = TRUE)

    tab_ha <- fisher_tab + 0.5

    a <- tab_ha["high_BMIz","T0"];  b <- tab_ha["high_BMIz","T12"]
    c <- tab_ha["low_BMIz","T0"];   d <- tab_ha["low_BMIz","T12"]

    or_ha   <- (a*d) / (b*c)
    se_log  <- sqrt(1/a + 1/b + 1/c + 1/d)
    ci_low  <- exp(log(or_ha) - 1.96 * se_log)
    ci_high <- exp(log(or_ha) + 1.96 * se_log)

    fisher_p      <- as.numeric(fisher_res$p.value)
    fisher_or     <- if (!is.null(fisher_res$estimate)) as.numeric(fisher_res$estimate) else NA_real_
    fisher_ci_low <- if (!is.null(fisher_res$conf.int)) fisher_res$conf.int[1] else NA_real_
    fisher_ci_high<- if (!is.null(fisher_res$conf.int)) fisher_res$conf.int[2] else NA_real_
  } else {
    fisher_p <- fisher_or <- fisher_ci_low <- fisher_ci_high <- NA_real_
    or_ha <- ci_low <- ci_high <- NA_real_
  }

  tibble::tibble(
    variable      = res_obj$variable,
    method        = res_obj$method,
    n_pairs       = res_obj$n,
    shapiro_p     = unname(res_obj$shapiro$p.value),
    stat          = unname(res_obj$test$statistic),
    p_value       = res_obj$test$p.value,
    group_T0      = fmt_group_T0,
    group_T12     = fmt_group_T12,
    mean_T0    = mean_T0,
    sd_T0      = sd_T0,
    median_T0  = med_T0,
    q1_T0      = q1_T0,
    q3_T0      = q3_T0,
    min_T0     = min(paired$T0),
    max_T0     = max(paired$T0),
    range_T0   = max(paired$T0) - min(paired$T0),
    mean_T12   = mean_T12,
    sd_T12     = sd_T12,
    median_T12 = med_T12,
    q1_T12     = q1_T12,
    q3_T12     = q3_T12,
    min_T12    = min(paired$T12),
    max_T12    = max(paired$T12),
    range_T12  = max(paired$T12) - min(paired$T12),
    diff_estimate = if (is_t) res_obj$mean_diff else res_obj$median_diff,
    ci_low        = if (is_t) res_obj$ci[1] else NA_real_,
    ci_high       = if (is_t) res_obj$ci[2] else NA_real_,
    effect_type   = res_obj$effect_size$type,
    effect_value  = res_obj$effect_size$value,
    fisher_p      = fisher_p,
    fisher_or     = fisher_or,
    fisher_ci_low = fisher_ci_low,
    fisher_ci_high= fisher_ci_high,
    or_ha         = or_ha,
    or_ha_ci_low  = if (!is.null(fisher_tab)) ci_low else NA_real_,
    or_ha_ci_high = if (!is.null(fisher_tab)) ci_high else NA_real_,
    cutoff_used   = cutoff
  )
}

row_z_wt  <- .build_row_z(df_z, "SD_z_score_weight_2010",             res_z_wt)
row_z_ht  <- .build_row_z(df_z, "SD_z_score_height_2010",             res_z_ht)
row_z_bmi <- .build_row_z(df_z, "SD_z_score_BMI_2010",      res_z_bmi,
                          fisher_tab = table_bmiz, cutoff = cutoff_z)

z_scores_stats <- dplyr::bind_rows(row_z_wt, row_z_ht, row_z_bmi)

z_scores_stats$p_fdr <- p.adjust(z_scores_stats$p_value, method = "BH")

if (!dir.exists(here("results"))) dir.create(here("results"))
readr::write_excel_csv(z_scores_stats, here("results", "Zscores_pediatrics_stats.csv"))
```





# -----------------------------------------------------------------------------
# -----------------------------------------------------------------------------
# GLOBAL FDR
# -----------------------------------------------------------------------------
# -----------------------------------------------------------------------------

### Including ONLY of interest
```{r Stats results FDR, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# 1) Put all your test objects in a single nested list (system -> variables -> res_object)
clinical_tests <- list(
  `Lung function` = list(
    ppFEV1       = res_ppFEV1,
    ppFVC        = res_ppFVC,
    FEV1_FVC     = res_ratio,
    ppFEV1_fem   = res_ppFEV1_fem,
    ppFVC_fem    = res_ppFVC_fem,
    FEV1_FVC_fem = res_ratio_fem,
    ppFEV1_mal   = res_ppFEV1_mal,
    ppFVC_mal    = res_ppFVC_mal,
    FEV1_FVC_mal = res_ratio_mal,
    ppFEV1_adu   = res_ppFEV1_adu,
    ppFVC_adu    = res_ppFVC_adu,
    FEV1_FVC_adu = res_ratio_adu,
    ppFEV1_ped   = res_ppFEV1_ped,
    ppFVC_ped    = res_ppFVC_ped,
    FEV1_FVC_ped = res_ratio_ped
  ),
  `Transaminases` = list(
    ALT = res_ALT,
    AST = res_AST,
    GGT = res_GGT
  ),
  `Intestinal inflammation` = list(
    Calprotectin = res_calpro
  ),
  `Absorption markers` = list(
    Elastase = res_elas
  ),
  `Anthropometry_BMI` = list(
    Weight_kg = res_bmi_wt,
    Height_cm = res_bmi_ht,
    BMI_kgm2 = res_bmi_bmi
  ),
  `Anthropometry_zscore` = list(
    SD_z_score_weight_2010 = res_z_wt,
    SD_z_score_height_2010 = res_z_ht,
    SD_z_score_BMI_2010 = res_z_bmi
  )
)

# 2) Helper to extract p-value (adapt if your object structure differs)
get_p <- function(x) x$test$p.value

# 3) Build one table with all clinical p-values
clinical_pvals <- purrr::imap_dfr(clinical_tests, \(tests, sys) {
  tibble::tibble(
    system   = sys,
    variable = names(tests),
    p_value  = purrr::map_dbl(tests, get_p)
  )
})

# 4) Global BH-FDR across ALL clinical variables together
clinical_pvals_globalFDR <- clinical_pvals %>%
  dplyr::mutate(
    p_fdr_global = p.adjust(p_value, method = "BH"),
    signif_fdr   = p_fdr_global < 0.05
  ) %>%
  dplyr::arrange(p_fdr_global)

print(clinical_pvals_globalFDR, n = Inf)

# 5) Save
if (!dir.exists(here::here("results"))) dir.create(here::here("results"))
readr::write_excel_csv(clinical_pvals_globalFDR, here::here("results", "clinical_globalFDR_pvalues.csv"))
```

## Export table
```{r Export final table, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# 0) Collect per-system result tables
results_by_system <- list(
  "Lung function"           = lung_stats,
  "Transaminases"           = trans_stats,
  "Intestinal inflammation" = row_calpro,
  "Absorption markers"      = row_elas,
  "Anthropometry_BMI"       = bmi_stats,
  "Anthropometry_zscore"    = z_scores_stats
)

# 1) Build mega table (do NOT compute BH here)
mega_tbl <- dplyr::bind_rows(results_by_system, .id = "system") %>%
  dplyr::mutate(
    system   = stringr::str_trim(system),
    variable = stringr::str_trim(variable)
  )

# 2) Load the GLOBAL FDR table created in your "ONLY of interest" code
global_fdr_tbl <- readr::read_csv(
  here::here("results", "clinical_globalFDR_pvalues.csv"),
  show_col_types = FALSE
) %>%
  dplyr::transmute(
    system   = stringr::str_trim(system),
    variable = stringr::str_trim(variable),
    p_fdr_global
  )

# 3) Keep ONLY rows that exist in the global list + join global FDR + reorder column
mega_tbl_final <- mega_tbl %>%
  dplyr::semi_join(global_fdr_tbl, by = c("system", "variable")) %>%
  dplyr::select(-dplyr::any_of(c("p_fdr", "p_fdr_global"))) %>%
  dplyr::left_join(global_fdr_tbl, by = c("system", "variable")) %>%
  dplyr::relocate(p_fdr_global, .after = p_value)

# 4) Export (Excel-friendly UTF-8 CSV)
readr::write_excel_csv(
  mega_tbl_final, here::here("results", "clinical_results_mega_table_globalFDR_FINAL.csv"))
```


# -----------------------------------------------------------------------------
#### PANELS PLOTS (run again the FIRST function plot)
# -----------------------------------------------------------------------------
##### Lung function
```{r Lung panel, echo=TRUE, results='show', message=FALSE, warning=FALSE}
# Panel 3
panel_lung_all <- (p_ppFEV1 | p_ppFVC | p_ratio) /
  (p_ppFEV1_fem | p_ppFEV1_mal | plot_spacer()) / 
  (p_ppFEV1_adu | p_ppFEV1_ped | plot_spacer()) +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

panel_lung_all <- panel_lung_all + plot_annotation(tag_levels = "A")

ggsave(here("results","lung_all_panel.svg"), panel_lung_all, width = 10, height = 9, bg = "white")
```

##### Lung function with trajectories
```{r Lung panel, echo=TRUE, results='show', message=FALSE, warning=FALSE}
# Panel 3
panel_lung_trajec <- (p_ppFEV1 | p_ppFEV1_trajec) /
  (p_ppFEV1_fem | p_ppFEV1_mal) / 
  (p_ppFEV1_adu | p_ppFEV1_ped) +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

panel_lung_trajec <- panel_lung_trajec + plot_annotation(tag_levels = "A")

ggsave(here("results","lung_trajec_panel.svg"), panel_lung_trajec, width = 8, height = 11, bg = "white")
```

##### Transaminases
```{r Trans panel, echo=TRUE, results='show', message=FALSE, warning=FALSE}
# Panel 3
panel_trans <- (p_ALT | p_AST  | p_GGT) +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

panel_trans <- panel_trans + plot_annotation(tag_levels = "A")

ggsave(here("results","trans_panel.svg"), panel_trans, width = 10, height = 5, bg = "white")
```

#### Transaminases with Elastase and calpro
```{r Transas Elas and Calpro panel, echo=TRUE, results='show', message=FALSE, warning=FALSE}
# Panel 3x"3"
panel_trans_calpro_elas <- (p_ALT | p_AST | p_GGT) /
  (p_calpro | p_elas | plot_spacer()) +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

panel_trans_calpro_elas <- panel_trans_calpro_elas + plot_annotation(tag_levels = "A")

ggsave(here("results","trans_calpro_elas_panel.svg"), panel_trans_calpro_elas, width = 10, height = 7, bg = "white")
```

#### Transaminases with Elastase and calpro version 2
```{r Transas Elas and Calpro version 2  panel, echo=TRUE, results='show', message=FALSE, warning=FALSE}
top_panel <- p_ALT_y | p_AST_y | p_GGT_y
bottom_panel <- p_calpro | p_elas | plot_spacer()

panel_trans_calpro_elas_final <- top_panel / bottom_panel +
  plot_layout(
    guides = "collect",
    heights = c(1.4, 1)
  ) &
  theme(legend.position = "bottom")

#panel_trans_calpro_elas_final <- panel_trans_calpro_elas_final +
#  plot_annotation(tag_levels = "A")

ggsave(
  here("results", "trans_calpro_elas_panel_final.svg"),
  plot = panel_trans_calpro_elas_final, width = 10, height = 8, bg = "white")
```

##### Elastase and calpro
```{r Elas and Calpro panel, echo=TRUE, results='show', message=FALSE, warning=FALSE}
# Panel 2x2
panel_calpro_elas <- (p_calpro | p_elas) +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

panel_calpro_elas <- panel_calpro_elas + plot_annotation(tag_levels = "A")

ggsave(here("results","calpro_elas_panel.svg"), panel_calpro_elas, width = 8, height = 4, bg = "white")
```

##### BMI and Z-scores
```{r BMI and Zscores panel, echo=TRUE, results='show', message=FALSE, warning=FALSE}
# Panel 3x3
panel_bmi_z <- (p_bmi_wt | p_bmi_ht | p_bmi_bmi) /
  (p_z_wt | p_z_ht | p_z_bmi) +
  
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

panel_bmi_z <- panel_bmi_z + plot_annotation(tag_levels = "A")

ggsave(here("results","bmi_zscores_panel.svg"), panel_bmi_z, width = 10, height = 7, bg = "white")


panel_bmi_z_all <- (p_bmi_wt | p_bmi_ht | p_bmi_bmi) /
  (p_zscores_all) +
  
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

panel_bmi_z_all <- panel_bmi_z_all + plot_annotation(tag_levels = "A")

ggsave(here("results","bmi_zscores_all_panel.svg"), panel_bmi_z_all, width = 10, height = 7, bg = "white")
```



