---
title: "cf_metagenomics_thesis_chapter_I"
author: "Natalia Bastón-Paz"
date: "2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Original: knitr::opts_chunk$set(echo = TRUE)
```

```{#r Clean environments, include=FALSE}
rm(list = ls())
```

## General introduction

Cystic Fibrosis (CF) is a genetic disorder that primarily affects the respiratory, digestive, and reproductive systems. It is caused by mutations in the CFTR (Cystic Fibrosis Transmembrane Conductance Regulator) gene, which encodes a protein responsible for regulating chloride and bicarbonate transport across epithelial cell membranes. These mutations result in abnormal ion transport, leading to thick, sticky mucus accumulation in the lungs, pancreas, and other organs. CF is one of the most common autosomal recessive disorders in Caucasian populations, with an incidence of approximately 1 in 2,500–3,500 live births. People with CF (pwCF) typically present with chronic lung infections due to impaired mucociliary clearance, ancreatic insufficiency, leading to malabsorption and nutritional deficiencies, liver disease, diabetes, and other systemic complications. The CF lung microbiome is characterized by chronic colonization and infection with pathogens such as *Pseudomonas aeruginosa*, and *Staphylococcus aureus*, among others. These infections contribute to progressive lung damage, inflammation, and exacerbations. Significant advances in CF management, including CFTR modulators, improved airway clearance techniques, and antibiotic therapies, have increased life expectancy and quality of life for pwCF. The triple therapy elexacaftor-tezacaftor-ivacaftor (ETI) was approved in the USA in 2019, but its effects over the microbiota remains unclear.

*Questions to be answered:*

-   *Have the gut and lung microbiota changed after one year of treatment with ETI?*

## Libraries

```{r Libraries, message=FALSE, warning=FALSE}
library(mia)     
library(miaViz)
library(tidyverse)         
library(castor)
library(readr)
library(ggpubr)
library(utils)
library(ggplot2)          
library(patchwork)
library(plotrix)
library(ComplexUpset)
library(vegan)
library(stats)
library(rstatix)
library(data.table)
library(Maaslin2)
library(MicrobiomeStat)
library(ANCOMBC)
library(ALDEx2)
library(MADAV)
library(SummarizedExperiment)
library(tibble)
library(TreeSummarizedExperiment)
library(stringr)
library(here)
library(svglite)
library(aplot)
library(patchwork)
library(tidyr)
library(ggsci)
library(purrr)
library(openxlsx)
library(broom)
library(ggthemes)
library(RColorBrewer)
library(gghalves)
library(scales)
library(ggdist)
library(forcats)
library(dplyr)

# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")

# BiocManager::install("package_name")
```

## Load the data

#### Optional: If we want to know which samples have 0 counts:

```{r Samples with 0 counts, echo=TRUE, results='show'}
df <- read.table(here("data", "cf_metaphlan_counts.txt"),
                 header = TRUE, sep = "\t", 
                 row.names = 1, stringsAsFactors = FALSE)

print(names(which(colSums(df) == 100000000)))
```

#### Sample Metadata file

Load in sample metadata file --\> colData_Metaphlan object:

```{r Load in Sample Metadata file}
Metaphlan_data_cf <- read_delim(here("data", "cf_metadata_profiles.txt"), 
                                "\t", 
                                escape_double = FALSE, 
                                trim_ws = TRUE)
# Fix column names
colnames(Metaphlan_data_cf)[1] <- 'SampleID'
```

Remove the samples that have 0 counts and others of interest:

```{r Remove samples of choice}
samples_to_remove <- c("A12E0", "A12E12", "A20E0", "A20E12", "A26E0", "A26E12", "P16E0", "P16E12", "P21E0", "P21E12", "P3H0", "P3H12")  # Replace with actual SampleIDs
Metaphlan_data_cf <- Metaphlan_data_cf[!Metaphlan_data_cf$SampleID %in% samples_to_remove, ]
# view(Metaphlan_data_cf)
```

Convert the column SampleID into row names:

```{r Convert col SampleID into row names}
Metaphlan_data_cf <- Metaphlan_data_cf %>%
  column_to_rownames(var = 'SampleID') %>%                                          
  data.frame()
# view(Metaphlan_data_cf)
```

#### Abundance file

Load in abundance file (change file name and FeatureID_column) --> assay_Metaphlan object:

```{r Load in abundance file}
M_assay_cf <- read_delim(here("data", "cf_metaphlan_counts.txt"), "\t", escape_double = FALSE, trim_ws = TRUE)
colnames(M_assay_cf)[1] <- 'FeatureID'
```

Get the taxonomy data from the FeatureID column of the assay_Metaphlan object --> rowData_Metaphlan object:

```{r Get the taxonomy data from FeatureID}
M_rowData_cf <- M_assay_cf[,'FeatureID'] %>% 
  mutate(split = FeatureID) %>%  
  separate(col = split, into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species",'Type'), sep = "\\|.__") %>%
  mutate(Kingdom = str_remove_all(Kingdom,'^k__')) %>%  
  column_to_rownames(var = "FeatureID") %>%        
  data.frame()
```

Filter the assay_Metaphlan object (M_assay_cf) using the rowData_Metaphlan (M_rowData_cf) & colData_Metaphlan (Metaphlan_data_cf) objects and create the TSE:

```{r Filter and create the TSE}
M_assay_cf <- M_assay_cf[,c('FeatureID',rownames(Metaphlan_data_cf))] %>% # The assay data columns are reorder to match the colData
  column_to_rownames(var = 'FeatureID') %>%        
  as.matrix()
M_TSE_cf_unfiltered <- TreeSummarizedExperiment(assays = M_assay_cf, # To create TreeSummarizedExperiment object                                   
                                                    rowData = M_rowData_cf,
                                                     colData = Metaphlan_data_cf)
assayNames(M_TSE_cf_unfiltered) <-'counts'
```

Rename NA's in 'UNCLASSIFIED' row of rowData(TSE_Metaphlan) --\> Replaces NA values (and any other values) with "UNCLASSIFIED" in the UNCLASSIFIED row only:

```{r Rename UNCLASSIFIED}
rowData(M_TSE_cf_unfiltered)[rowData(M_TSE_cf_unfiltered)$Kingdom == 'UNCLASSIFIED',1:ncol(rowData(M_TSE_cf_unfiltered))] <- 'UNCLASSIFIED' 
```

Filter to 'Species' level and retain 'UNCLASSIFIED':

```{r Filter to Species and retain UNCLASSIFIED}
M_TSE_cf_unfiltered <- M_TSE_cf_unfiltered[
  (!is.na(rowData(M_TSE_cf_unfiltered)$Species) & is.na(rowData(M_TSE_cf_unfiltered)$Type)) | 
    (rowData(M_TSE_cf_unfiltered)$Kingdom == 'UNCLASSIFIED'),
]
```

Optional filtering steps: Remove all non-bacteria taxa (check first whether Bacteria are in the 'Kingdom' or 'Domain' class). Here we also keep unclassified Bacteria and UNCLASSIFIED

```{r Remove non-bacteria taxa}
M_TSE_cf_bacteria <- M_TSE_cf_unfiltered[rowData(M_TSE_cf_unfiltered)$Kingdom %in% c('UNCLASSIFIED','Bacteria'), ]
```

Remove all singleton features:

```{r Remove singletons}
M_TSE_cf_singletons <- M_TSE_cf_bacteria[apply(assay(M_TSE_cf_bacteria), 1, function(x) { sum(x > 0) > 1 }), ]
```

Remove 'UNCLASSIFIED' feature:

```{r Remove UNCLASSIFIED}
M_TSE_cf_UN <- M_TSE_cf_singletons
M_TSE_cf_no_UN <- M_TSE_cf_singletons[rownames(M_TSE_cf_singletons) != 'UNCLASSIFIED',]
```

*REMEMBER:*

-   *For Relative Abundance use: M_TSE_cf_UN*

-   *For alpha and beta diversity, for differential abundance and for the relative abundance of the differential abundance analysis use: M_TSE_cf_UN*

## Create subsets with different variables

#### Subsets by Time

```{r Subset by Time}
subset_T0 <- M_TSE_cf_UN[, M_TSE_cf_UN$Time %in% 'T0']
subset_T12 <- M_TSE_cf_UN[, M_TSE_cf_UN$Time %in% 'T12']
subset_T0_no_UN <- M_TSE_cf_no_UN[, M_TSE_cf_no_UN$Time %in% 'T0']
subset_T12_no_UN <- M_TSE_cf_no_UN[, M_TSE_cf_no_UN$Time %in% 'T12']
```

#### Subsets by SampleType

```{r Subset by SampleType}
subset_feces <- M_TSE_cf_UN[, M_TSE_cf_UN$SampleType %in% 'feces']
subset_sputum <- M_TSE_cf_UN[, M_TSE_cf_UN$SampleType %in% 'sputum']
subset_feces_no_UN <- M_TSE_cf_no_UN[, M_TSE_cf_no_UN$SampleType %in% 'feces']
subset_sputum_no_UN <- M_TSE_cf_no_UN[, M_TSE_cf_no_UN$SampleType %in% 'sputum']
```

# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------
# RELATIVE ABUNDANCE (RA)
# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------

```{r RA CSV at T0 and T12, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
# Phylum
rank_level <- "Phylum"

RA_agg <- mia::transformAssay(M_TSE_cf_UN, method = "relabundance") %>%
  mia::agglomerateByRank(rank_level, onRankOnly = TRUE, na.rm = TRUE)

ra_mat <- SummarizedExperiment::assay(RA_agg, "relabundance")

ra_wide <- as.data.frame(ra_mat) %>%
  tibble::rownames_to_column(var = "FeatureID")

readr::write_csv(ra_wide, here::here("results", paste0("relabundance_", rank_level, "_wide.csv")))

# Species
rank_level <- "Species"  # "Genus", etc.

RA_agg <- mia::transformAssay(M_TSE_cf_UN, method = "relabundance") %>%
  mia::agglomerateByRank(rank_level, onRankOnly = TRUE, na.rm = TRUE)

ra_mat <- SummarizedExperiment::assay(RA_agg, "relabundance")

ra_wide <- as.data.frame(ra_mat) %>%
  tibble::rownames_to_column(var = "FeatureID")

readr::write_csv(ra_wide, here::here("results", paste0("relabundance_", rank_level, "_wide.csv")))
```

#------------------
## SAMPLE TYPE
#------------------

#### Feces at T0 and T12

##### Phylum

```{r RA of Feces at T0 and T12 phylum, echo=TRUE, fig.show='show', message=FALSE, warning=FALSE}
RA_bar_cf_fc_ph <- RA_bar(tse = subset_feces,
                    var = c('Time'),
                    var_levels = c('T0', 'T12'),
                    level = 'Phylum',
                    top_features = 5,
                    feature_order = 'abund',
                    sample_order = 'SampleType',
                    sample_order_levels = NULL,
                    feature_bar = NULL,
                    feature_bar_levels = 'SampleType',
                    group_average = FALSE,
                    plot_title = 'Relative Abundance of Feces',
                    cutoff_features = 'None',
                    show_counts = FALSE)
# Save as SVG
ggsave(filename = here("results", "RA_bar_cf_fc_phylum.svg"), plot = RA_bar_cf_fc_ph$RA_bar_plot, width = 12, height = 8, dpi = 300, device = "svg") # Use RA_bar_plot when feature_bar = NULL
```

##### Species

```{r RA of Feces at T0 and T12 species, echo=TRUE, fig.show='show', message=FALSE, warning=FALSE}
RA_bar_cf_fc_sps <- RA_bar(tse = subset_feces,
                    var = c('Time'),
                    var_levels = c('T0', 'T12'),
                    level = 'Species',
                    top_features = 10,
                    feature_order = 'abund',
                    sample_order = 'SampleType',
                    sample_order_levels = NULL,
                    feature_bar = NULL,
                    feature_bar_levels = 'SampleType',
                    group_average = FALSE,
                    plot_title = 'Relative Abundance of Feces',
                    cutoff_features = 'None',
                    show_counts = FALSE)
# Save as SVG
ggsave(filename = here("results", "RA_bar_cf_fc_species.svg"), plot = RA_bar_cf_fc_sps$RA_bar_plot, width = 12, height = 8, dpi = 300, device = "svg") # Use RA_bar_plot when feature_bar = NULL
```

#### Sputum at T0 and T12

##### Phylum

```{r RA of Sputum at T0 and T12 phylum, echo=TRUE, fig.show='show', message=FALSE, warning=FALSE}
RA_bar_cf_sp_ph <- RA_bar(tse = subset_sputum,
                    var = c('Time'),
                    var_levels = c('T0', 'T12'),
                    level = 'Phylum',
                    top_features = 5,
                    feature_order = 'abund',
                    sample_order = 'SampleType',
                    sample_order_levels = NULL,
                    feature_bar = NULL,
                    feature_bar_levels = 'SampleType',
                    group_average = FALSE,
                    plot_title = 'Relative Abundance of Sputum',
                    cutoff_features = 'None',
                    show_counts = FALSE)
# Save as SVG
ggsave(filename = here("results", "RA_bar_cf_sp_phylum.svg"), plot = RA_bar_cf_sp_ph$RA_bar_plot, width = 12, height = 8, dpi = 300, device = "svg") # Use RA_bar_plot when feature_bar = NULL
```

##### Species

```{r RA of Sputum at T0 and T12 species, echo=TRUE, fig.show='show', message=FALSE, warning=FALSE}
RA_bar_cf_sp_sps <- RA_bar(tse = subset_sputum,
                    var = c('Time'),
                    var_levels = c('T0', 'T12'),
                    level = 'Species',
                    top_features = 10,
                    feature_order = 'abund',
                    sample_order = 'SampleType',
                    sample_order_levels = NULL,
                    feature_bar = NULL,
                    feature_bar_levels = 'SampleType',
                    group_average = FALSE,
                    plot_title = 'Relative Abundance of Sputum',
                    cutoff_features = 'None',
                    show_counts = FALSE)
# Save as SVG
ggsave(filename = here("results", "RA_bar_cf_sp_species.svg"), plot = RA_bar_cf_sp_sps$RA_bar_plot, width = 12, height = 8, dpi = 300, device = "svg") # Use RA_bar_plot when feature_bar = NULL
```

#-------------------
### FECES AND SPUTUM
#-------------------

# Function to calculate relative abundance
```{r Function to calculate RA, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# RELATIVE ABUNDANCE TABLES BY SAMPLE TYPE AND TAXONOMIC LEVEL
# Phylum in feces, Phylum in sputum, Species in feces, Species in sputum
# ------------------------------------------------------------------------------

calc_relative_abundance <- function(tse, rank_level, output_prefix, group_vars = c("Time")) {
  
  # 0. Standardize phylum names before relative abundance calculation
  # This is only applied when calculating relative abundance at Phylum level
  if (rank_level == "Phylum") {
    
    phylum_map <- c(
      "Firmicutes"      = "Bacillota",
      "Bacteroidetes"   = "Bacteroidota",
      "Proteobacteria"  = "Pseudomonadota",
      "Actinobacteria"  = "Actinomycetota",
      "Verrucomicrobia" = "Verrucomicrobiota",
      "Fusobacteria"    = "Fusobacteriota",
      "Spirochaetes"    = "Spirochaetota",
      "Tenericutes"     = "Mycoplasmatota"
    )
    
    phylum <- as.character(SummarizedExperiment::rowData(tse)$Phylum)
    
    phylum[phylum %in% names(phylum_map)] <- phylum_map[
      phylum[phylum %in% names(phylum_map)]
    ]
    
    SummarizedExperiment::rowData(tse)$Phylum <- phylum
  }
  
  # 1. Calculate relative abundance per sample and agglomerate at selected rank
  RA_agg <- mia::transformAssay(tse, method = "relabundance") %>%
    mia::agglomerateByRank(rank_level, onRankOnly = TRUE, na.rm = TRUE)
  
  # 2. Extract relative abundance matrix and convert to percentage
  ra_mat <- SummarizedExperiment::assay(RA_agg, "relabundance") * 100
  
  # 3. Taxonomy table
  tax_table <- as.data.frame(SummarizedExperiment::rowData(RA_agg)) %>%
    tibble::rownames_to_column(var = "FeatureID") %>%
    dplyr::mutate(
      Rank = rank_level,
      Taxon = .data[[rank_level]]
    ) %>%
    dplyr::select(FeatureID, Rank, Taxon)
  
  # 4. Metadata table
  meta_table <- as.data.frame(SummarizedExperiment::colData(RA_agg)) %>%
    tibble::rownames_to_column(var = "SampleID")
  
  # 5. Wide table: taxa as rows, samples as columns
  ra_wide <- as.data.frame(ra_mat) %>%
    tibble::rownames_to_column(var = "FeatureID") %>%
    dplyr::left_join(tax_table, by = "FeatureID") %>%
    dplyr::select(FeatureID, Rank, Taxon, dplyr::everything())
  
  # 6. Long table: one row per taxon and sample
  ra_long <- as.data.frame(ra_mat) %>%
    tibble::rownames_to_column(var = "FeatureID") %>%
    tidyr::pivot_longer(
      cols = -FeatureID,
      names_to = "SampleID",
      values_to = "Relative_abundance_percent"
    ) %>%
    dplyr::left_join(tax_table, by = "FeatureID") %>%
    dplyr::left_join(meta_table, by = "SampleID") %>%
    dplyr::select(
      SampleID,
      SampleType,
      dplyr::all_of(group_vars),
      Rank,
      Taxon,
      Relative_abundance_percent,
      dplyr::everything()
    )
  
  # 7. Summary table by group, usually T0 and T12
  ra_summary <- ra_long %>%
    dplyr::group_by(
      dplyr::across(dplyr::all_of(c("SampleType", group_vars, "Rank", "Taxon")))
    ) %>%
    dplyr::summarise(
      n_samples = dplyr::n(),
      mean_RA_percent = mean(Relative_abundance_percent, na.rm = TRUE),
      sd_RA_percent = sd(Relative_abundance_percent, na.rm = TRUE),
      median_RA_percent = median(Relative_abundance_percent, na.rm = TRUE),
      Q1_RA_percent = quantile(Relative_abundance_percent, 0.25, na.rm = TRUE),
      Q3_RA_percent = quantile(Relative_abundance_percent, 0.75, na.rm = TRUE),
      min_RA_percent = min(Relative_abundance_percent, na.rm = TRUE),
      max_RA_percent = max(Relative_abundance_percent, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::arrange(SampleType, dplyr::across(dplyr::all_of(group_vars)), desc(mean_RA_percent))
  
  # 8. Export CSV files
  readr::write_csv(
    ra_wide,
    here::here("results", paste0(output_prefix, "_wide_by_sample.csv"))
  )
  
  readr::write_csv(
    ra_long,
    here::here("results", paste0(output_prefix, "_long_by_sample.csv"))
  )
  
  readr::write_csv(
    ra_summary,
    here::here("results", paste0(output_prefix, "_summary_by_", paste(group_vars, collapse = "_"), ".csv"))
  )
  
  # 9. Return objects to R environment
  return(list(
    RA_agg = RA_agg,
    wide = ra_wide,
    long = ra_long,
    summary = ra_summary
  ))
}
```

## Run to calculate RA
```{r Run function to calculate RA, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
# Phylum in feces
RA_phylum_feces <- calc_relative_abundance(
  tse = subset_feces,
  rank_level = "Phylum",
  output_prefix = "relabundance_phylum_feces"
)

# Phylum in sputum
RA_phylum_sputum <- calc_relative_abundance(
  tse = subset_sputum,
  rank_level = "Phylum",
  output_prefix = "relabundance_phylum_sputum"
)

# Species in feces
RA_species_feces <- calc_relative_abundance(
  tse = subset_feces,
  rank_level = "Species",
  output_prefix = "relabundance_species_feces"
)

# Species in sputum
RA_species_sputum <- calc_relative_abundance(
  tse = subset_sputum,
  rank_level = "Species",
  output_prefix = "relabundance_species_sputum"
)
```

### Function to check distribution
```{r Function to check distribution, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# CHECK DISTRIBUTION OF RELATIVE ABUNDANCE VALUES
# Safe version: avoids Shapiro-Wilk errors when all values are identical
# ------------------------------------------------------------------------------

safe_shapiro_p <- function(x) {
  
  x <- x[!is.na(x)]
  
  if (length(x) < 3) {
    return(NA_real_)
  }
  
  if (length(x) > 5000) {
    return(NA_real_)
  }
  
  if (length(unique(x)) < 3) {
    return(NA_real_)
  }
  
  return(shapiro.test(x)$p.value)
}

check_RA_distribution <- function(ra_long) {
  
  distribution_table <- ra_long %>%
    dplyr::group_by(SampleType, Time, Rank, Taxon) %>%
    dplyr::summarise(
      n_samples = sum(!is.na(Relative_abundance_percent)),
      n_nonzero = sum(Relative_abundance_percent > 0, na.rm = TRUE),
      n_unique_values = length(unique(Relative_abundance_percent[!is.na(Relative_abundance_percent)])),
      zero_percent = mean(Relative_abundance_percent == 0, na.rm = TRUE) * 100,
      mean_RA_percent = mean(Relative_abundance_percent, na.rm = TRUE),
      sd_RA_percent = sd(Relative_abundance_percent, na.rm = TRUE),
      median_RA_percent = median(Relative_abundance_percent, na.rm = TRUE),
      Q1_RA_percent = quantile(Relative_abundance_percent, 0.25, na.rm = TRUE),
      Q3_RA_percent = quantile(Relative_abundance_percent, 0.75, na.rm = TRUE),
      min_RA_percent = min(Relative_abundance_percent, na.rm = TRUE),
      max_RA_percent = max(Relative_abundance_percent, na.rm = TRUE),
      shapiro_p = safe_shapiro_p(Relative_abundance_percent),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      distribution = dplyr::case_when(
        n_unique_values < 3 ~ "Not assessed: identical or nearly identical values",
        is.na(shapiro_p) ~ "Not assessed",
        shapiro_p < 0.05 ~ "Non-normal",
        shapiro_p >= 0.05 ~ "Compatible with normality"
      ),
      recommended_summary = dplyr::case_when(
        n_unique_values < 3 ~ "Median [IQR]",
        zero_percent > 20 ~ "Median [IQR]",
        is.na(shapiro_p) ~ "Median [IQR]",
        shapiro_p < 0.05 ~ "Median [IQR]",
        shapiro_p >= 0.05 ~ "Mean +/- SD"
      )
    ) %>%
    dplyr::arrange(SampleType, Time, Rank, desc(mean_RA_percent))
  
  return(distribution_table)
}
```

### Run distribution
```{r Run distribution, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
dist_phylum_feces <- check_RA_distribution(RA_phylum_feces$long)
dist_phylum_sputum <- check_RA_distribution(RA_phylum_sputum$long)

dist_species_feces <- check_RA_distribution(RA_species_feces$long)
dist_species_sputum <- check_RA_distribution(RA_species_sputum$long)
```

### Export distribution
```{r Export distribution, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
readr::write_csv(dist_phylum_feces, here::here("results", "distribution_phylum_feces.csv"))
readr::write_csv(dist_phylum_sputum, here::here("results", "distribution_phylum_sputum.csv"))
readr::write_csv(dist_species_feces, here::here("results", "distribution_species_feces.csv"))
readr::write_csv(dist_species_sputum, here::here("results", "distribution_species_sputum.csv"))
```

## Patient by patient FECES

### Taxonomy harmonization
```{r Taxonomy harmonization, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# TAXONOMIC NAME STANDARDIZATION
# ------------------------------------------------------------------------------

standardize_taxon_names_in_ra_long <- function(ra_long) {
  
  phylum_map <- c(
    "Firmicutes"      = "Bacillota",
    "Bacteroidetes"   = "Bacteroidota",
    "Proteobacteria"  = "Pseudomonadota",
    "Actinobacteria"  = "Actinomycetota",
    "Verrucomicrobia" = "Verrucomicrobiota",
    "Fusobacteria"    = "Fusobacteriota",
    "Spirochaetes"    = "Spirochaetota",
    "Tenericutes"     = "Mycoplasmatota"
  )
  
  ra_long_std <- ra_long %>%
    dplyr::mutate(
      Taxon = dplyr::case_when(
        Rank == "Phylum" & Taxon %in% names(phylum_map) ~ phylum_map[Taxon],
        TRUE ~ Taxon
      )
    ) %>%
    dplyr::group_by(
      SampleID,
      SampleType,
      Time,
      Rank,
      Taxon,
      dplyr::across(dplyr::any_of(c("PatientID", "AgeGroup", "Sex")))
    ) %>%
    dplyr::summarise(
      Relative_abundance_percent = sum(Relative_abundance_percent, na.rm = TRUE),
      .groups = "drop"
    )
  
  return(ra_long_std)
}
```

### Functions to paired patient-level RA changes
```{r Function to paired patient RA changes, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# PAIRED PATIENT-LEVEL RELATIVE ABUNDANCE CHANGES
# For taxa shown in the fecal relative abundance figure
# ------------------------------------------------------------------------------

# Taxa shown in Figure 8A
feces_phyla_fig8 <- c(
  "Bacillota",
  "Actinomycetota",
  "Verrucomicrobiota",
  "Pseudomonadota",
  "Bacteroidota"
)

# Taxa shown in Figure 8B
feces_species_fig8 <- c(
  "Bifidobacterium_longum",
  "Ruminococcus_bromii",
  "Blautia_wexlerae",
  "Akkermansia_muciniphila",
  "Eubacterium_rectale",
  "Ruminococcus_gnavus",
  "Blautia_massiliensis",
  "Dialister_invisus",
  "Anaerostipes_hadrus"
)

# ------------------------------------------------------------------------------
# Function 1: create patient-level paired table
# ------------------------------------------------------------------------------

get_paired_RA_changes <- function(ra_long,
                                  taxa_keep,
                                  detection_threshold = 0,
                                  change_threshold = 0) {
  
  # Standardize phylum names and merge duplicated taxa after renaming
  ra_long <- standardize_taxon_names_in_ra_long(ra_long)
  
  # Also standardize taxa_keep, in case old phylum names are used
  phylum_map <- c(
    "Firmicutes"      = "Bacillota",
    "Bacteroidetes"   = "Bacteroidota",
    "Proteobacteria"  = "Pseudomonadota",
    "Actinobacteria"  = "Actinomycetota",
    "Verrucomicrobia" = "Verrucomicrobiota",
    "Fusobacteria"    = "Fusobacteriota",
    "Spirochaetes"    = "Spirochaetota",
    "Tenericutes"     = "Mycoplasmatota"
  )
  
  taxa_keep[taxa_keep %in% names(phylum_map)] <- phylum_map[
    taxa_keep[taxa_keep %in% names(phylum_map)]
  ]
  
  df <- ra_long %>%
    dplyr::filter(
      Taxon %in% taxa_keep,
      Time %in% c("T0", "T12"),
      !Taxon %in% c("UNCLASSIFIED", "Unclassified", "Other")
    )
  
  # If PatientID is not present, derive it from SampleID
  if (!"PatientID" %in% colnames(df)) {
    df <- df %>%
      dplyr::mutate(
        PatientID = stringr::str_remove(SampleID, "(H|E)(0|12)$")
      )
  }
  
  paired_df <- df %>%
    dplyr::select(
      PatientID,
      SampleID,
      SampleType,
      Time,
      Rank,
      Taxon,
      Relative_abundance_percent
    ) %>%
    dplyr::group_by(PatientID, SampleType, Rank, Taxon, Time) %>%
    dplyr::summarise(
      SampleID = paste(unique(SampleID), collapse = ";"),
      RA_percent = sum(Relative_abundance_percent, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    tidyr::pivot_wider(
      names_from = Time,
      values_from = c(SampleID, RA_percent),
      names_sep = "_"
    ) %>%
    dplyr::filter(
      !is.na(RA_percent_T0),
      !is.na(RA_percent_T12)
    ) %>%
    dplyr::mutate(
      detected_T0 = RA_percent_T0 > detection_threshold,
      detected_T12 = RA_percent_T12 > detection_threshold,
      delta_T12_minus_T0 = RA_percent_T12 - RA_percent_T0,
      
      detection_status = dplyr::case_when(
        detected_T0 & detected_T12 ~ "detected_T0_and_T12",
        detected_T0 & !detected_T12 ~ "detected_T0_only",
        !detected_T0 & detected_T12 ~ "detected_T12_only",
        !detected_T0 & !detected_T12 ~ "not_detected"
      ),
      
      change_direction = dplyr::case_when(
        delta_T12_minus_T0 > change_threshold ~ "increase",
        delta_T12_minus_T0 < -change_threshold ~ "decrease",
        TRUE ~ "stable"
      )
    ) %>%
    dplyr::arrange(Taxon, PatientID)
  
  return(paired_df)
}

# ------------------------------------------------------------------------------
# Function 2: summarise paired changes by taxon
# ------------------------------------------------------------------------------

summarise_paired_RA_changes <- function(paired_df) {
  
  summary_df <- paired_df %>%
    dplyr::group_by(SampleType, Rank, Taxon) %>%
    dplyr::summarise(
      n_paired = dplyr::n(),
      
      detected_T0_n = sum(detected_T0, na.rm = TRUE),
      detected_T12_n = sum(detected_T12, na.rm = TRUE),
      
      detected_T0_percent = detected_T0_n / n_paired * 100,
      detected_T12_percent = detected_T12_n / n_paired * 100,
      
      detected_T0_only_n = sum(detection_status == "detected_T0_only", na.rm = TRUE),
      detected_T12_only_n = sum(detection_status == "detected_T12_only", na.rm = TRUE),
      detected_both_n = sum(detection_status == "detected_T0_and_T12", na.rm = TRUE),
      not_detected_n = sum(detection_status == "not_detected", na.rm = TRUE),
      
      increased_n = sum(change_direction == "increase", na.rm = TRUE),
      decreased_n = sum(change_direction == "decrease", na.rm = TRUE),
      stable_n = sum(change_direction == "stable", na.rm = TRUE),
      
      median_T0_all = median(RA_percent_T0, na.rm = TRUE),
      Q1_T0_all = quantile(RA_percent_T0, 0.25, na.rm = TRUE),
      Q3_T0_all = quantile(RA_percent_T0, 0.75, na.rm = TRUE),
      
      median_T12_all = median(RA_percent_T12, na.rm = TRUE),
      Q1_T12_all = quantile(RA_percent_T12, 0.25, na.rm = TRUE),
      Q3_T12_all = quantile(RA_percent_T12, 0.75, na.rm = TRUE),
      
      median_delta_all = median(delta_T12_minus_T0, na.rm = TRUE),
      Q1_delta_all = quantile(delta_T12_minus_T0, 0.25, na.rm = TRUE),
      Q3_delta_all = quantile(delta_T12_minus_T0, 0.75, na.rm = TRUE),
      
      median_T0_among_T0_detected = if (sum(detected_T0, na.rm = TRUE) > 0) {
        median(RA_percent_T0[detected_T0], na.rm = TRUE)
      } else {
        NA_real_
      },
      
      median_T12_among_T0_detected = if (sum(detected_T0, na.rm = TRUE) > 0) {
        median(RA_percent_T12[detected_T0], na.rm = TRUE)
      } else {
        NA_real_
      },
      
      median_delta_among_T0_detected = if (sum(detected_T0, na.rm = TRUE) > 0) {
        median(delta_T12_minus_T0[detected_T0], na.rm = TRUE)
      } else {
        NA_real_
      },
      
      patients_detected_T0_only = paste(PatientID[detection_status == "detected_T0_only"], collapse = "; "),
      patients_detected_T12_only = paste(PatientID[detection_status == "detected_T12_only"], collapse = "; "),
      patients_with_decrease = paste(PatientID[change_direction == "decrease"], collapse = "; "),
      patients_with_increase = paste(PatientID[change_direction == "increase"], collapse = "; "),
      
      .groups = "drop"
    ) %>%
    dplyr::arrange(Rank, desc(detected_T0_n), desc(median_T0_all))
  
  return(summary_df)
}

# ------------------------------------------------------------------------------
# Function 3: extract the individual samples with highest RA per taxon and timepoint
# ------------------------------------------------------------------------------

get_top_RA_samples <- function(ra_long,
                               taxa_keep,
                               n_top = 5,
                               detection_threshold = 0) {
  
  # Standardize phylum names and merge duplicated taxa after renaming
  ra_long <- standardize_taxon_names_in_ra_long(ra_long)
  
  # Also standardize taxa_keep
  phylum_map <- c(
    "Firmicutes"      = "Bacillota",
    "Bacteroidetes"   = "Bacteroidota",
    "Proteobacteria"  = "Pseudomonadota",
    "Actinobacteria"  = "Actinomycetota",
    "Verrucomicrobia" = "Verrucomicrobiota",
    "Fusobacteria"    = "Fusobacteriota",
    "Spirochaetes"    = "Spirochaetota",
    "Tenericutes"     = "Mycoplasmatota"
  )
  
  taxa_keep[taxa_keep %in% names(phylum_map)] <- phylum_map[
    taxa_keep[taxa_keep %in% names(phylum_map)]
  ]
  
  df <- ra_long %>%
    dplyr::filter(
      Taxon %in% taxa_keep,
      Time %in% c("T0", "T12"),
      !Taxon %in% c("UNCLASSIFIED", "Unclassified", "Other")
    )
  
  if (!"PatientID" %in% colnames(df)) {
    df <- df %>%
      dplyr::mutate(
        PatientID = stringr::str_remove(SampleID, "(H|E)(0|12)$")
      )
  }
  
  top_df <- df %>%
    dplyr::filter(Relative_abundance_percent > detection_threshold) %>%
    dplyr::group_by(SampleType, Rank, Taxon, Time) %>%
    dplyr::arrange(desc(Relative_abundance_percent), .by_group = TRUE) %>%
    dplyr::mutate(rank_within_taxon_time = dplyr::row_number()) %>%
    dplyr::filter(rank_within_taxon_time <= n_top) %>%
    dplyr::ungroup() %>%
    dplyr::select(
      SampleType,
      Rank,
      Taxon,
      Time,
      rank_within_taxon_time,
      PatientID,
      SampleID,
      Relative_abundance_percent
    )
  
  return(top_df)
}
```

### Run functions
```{r Run functions, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# Apply to fecal phyla
# ------------------------------------------------------------------------------

feces_phylum_patient_changes <- get_paired_RA_changes(
  ra_long = RA_phylum_feces$long,
  taxa_keep = feces_phyla_fig8,
  detection_threshold = 0,
  change_threshold = 0
)

feces_phylum_summary_changes <- summarise_paired_RA_changes(
  feces_phylum_patient_changes
)

feces_phylum_top_samples <- get_top_RA_samples(
  ra_long = RA_phylum_feces$long,
  taxa_keep = feces_phyla_fig8,
  n_top = 5,
  detection_threshold = 0
)

# ------------------------------------------------------------------------------
# Apply to fecal species
# ------------------------------------------------------------------------------

feces_species_patient_changes <- get_paired_RA_changes(
  ra_long = RA_species_feces$long,
  taxa_keep = feces_species_fig8,
  detection_threshold = 0,
  change_threshold = 0
)

feces_species_summary_changes <- summarise_paired_RA_changes(
  feces_species_patient_changes
)

feces_species_top_samples <- get_top_RA_samples(
  ra_long = RA_species_feces$long,
  taxa_keep = feces_species_fig8,
  n_top = 5,
  detection_threshold = 0
)
```

### Export results
```{r Export results, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# EXPORT FECAL PATIENT-LEVEL TABLES
# ------------------------------------------------------------------------------
# Phylum-level fecal tables
readr::write_csv(
  feces_phylum_patient_changes,
  here::here("results", "feces_phylum_patient_level_T0_T12_changes.csv")
)

readr::write_csv(
  feces_phylum_summary_changes,
  here::here("results", "feces_phylum_summary_T0_T12_changes.csv")
)

readr::write_csv(
  feces_phylum_top_samples,
  here::here("results", "feces_phylum_top_samples_by_taxon_time.csv")
)

# Species-level fecal tables
readr::write_csv(
  feces_species_patient_changes,
  here::here("results", "feces_species_patient_level_T0_T12_changes.csv")
)

readr::write_csv(
  feces_species_summary_changes,
  here::here("results", "feces_species_summary_T0_T12_changes.csv")
)

readr::write_csv(
  feces_species_top_samples,
  here::here("results", "feces_species_top_samples_by_taxon_time.csv")
)
```

## Patient by Patient SPUTUM

### Taxonomy harmonization

```{r Taxonomy harmoniztion, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# SPUTUM: PATIENT-LEVEL RELATIVE ABUNDANCE CHANGES
# Figure 15A and Figure 15B
# ------------------------------------------------------------------------------
# Phyla shown in Figure 15A
sputum_phyla_fig15 <- c(
  "Actinomycetota",
  "Bacillota",
  "Bacteroidota",
  "Pseudomonadota",
  "Verrucomicrobiota"
)

# Species shown in Figure 15B
sputum_species_fig15 <- c(
  "Rothia_mucilaginosa",
  "Streptococcus_mitis",
  "Porphyromonas_bobii",
  "Pseudomonas_aeruginosa",
  "Staphylococcus_aureus",
  "Moraxella_catarrhalis",
  "Achromobacter_xylosoxidans",
  "Stenotrophomonas_maltophilia",
  "Burkholderia_contaminans"
)
```

### Run functions
```{r Run functions, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# Paired patient-level changes: phylum level
# ------------------------------------------------------------------------------

sputum_phylum_patient_changes <- get_paired_RA_changes(
  ra_long = RA_phylum_sputum$long,
  taxa_keep = sputum_phyla_fig15,
  detection_threshold = 0,
  change_threshold = 0
)

sputum_phylum_summary_changes <- summarise_paired_RA_changes(
  sputum_phylum_patient_changes
)

sputum_phylum_top_samples <- get_top_RA_samples(
  ra_long = RA_phylum_sputum$long,
  taxa_keep = sputum_phyla_fig15,
  n_top = 5,
  detection_threshold = 0
)

# ------------------------------------------------------------------------------
# Paired patient-level changes: species level
# ------------------------------------------------------------------------------

sputum_species_patient_changes <- get_paired_RA_changes(
  ra_long = RA_species_sputum$long,
  taxa_keep = sputum_species_fig15,
  detection_threshold = 0,
  change_threshold = 0
)

sputum_species_summary_changes <- summarise_paired_RA_changes(
  sputum_species_patient_changes
)

sputum_species_top_samples <- get_top_RA_samples(
  ra_long = RA_species_sputum$long,
  taxa_keep = sputum_species_fig15,
  n_top = 5,
  detection_threshold = 0
)
```

### Export results
```{r Export results, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# Export sputum tables
# ------------------------------------------------------------------------------
# Phylum-level sputum tables
readr::write_csv(
  sputum_phylum_patient_changes,
  here::here("results", "sputum_phylum_patient_level_T0_T12_changes.csv")
)

readr::write_csv(
  sputum_phylum_summary_changes,
  here::here("results", "sputum_phylum_summary_T0_T12_changes.csv")
)

readr::write_csv(
  sputum_phylum_top_samples,
  here::here("results", "sputum_phylum_top_samples_by_taxon_time.csv")
)

# Species-level sputum tables
readr::write_csv(
  sputum_species_patient_changes,
  here::here("results", "sputum_species_patient_level_T0_T12_changes.csv")
)

readr::write_csv(
  sputum_species_summary_changes,
  here::here("results", "sputum_species_summary_T0_T12_changes.csv")
)

readr::write_csv(
  sputum_species_top_samples,
  here::here("results", "sputum_species_top_samples_by_taxon_time.csv")
)
```


#-----------------------
### Statistics by SampleType
#### Function
```{r Function statistics, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
get_ra_stats_paired <- function(tse_obj, rank_level = "Phylum") {
  
  # 1) Agglomerate + relative abundance
  tse_ra <- tse_obj %>%
    mia::agglomerateByRank(rank = rank_level, onRankOnly = TRUE, na.rm = TRUE) %>%
    mia::transformAssay(method = "relabundance", pseudocount = 0, name = "relabundance")
  
  # 2) Extract matrix
  ra_mat <- SummarizedExperiment::assay(tse_ra, "relabundance")
  
  # 3) Long format + metadata
  ra_long <- as.data.frame(ra_mat) %>%
    tibble::rownames_to_column(var = "FeatureID") %>%
    tidyr::pivot_longer(
      cols = -FeatureID,
      names_to = "SampleID",
      values_to = "abundance"
    ) %>%
    dplyr::left_join(
      as.data.frame(SummarizedExperiment::colData(tse_ra)) %>%
        tibble::rownames_to_column(var = "SampleID"),
      by = "SampleID"
    ) %>%
    dplyr::filter(Time %in% c("T0", "T12")) %>%
    dplyr::mutate(Time = factor(Time, levels = c("T0", "T12")))
  
  # 4) Basic stats by taxon and time
  stats_basic <- ra_long %>%
    dplyr::group_by(FeatureID, Time) %>%
    dplyr::summarise(
      n = sum(!is.na(abundance)),
      prevalence = 100 * mean(abundance > 0, na.rm = TRUE),
      mean = mean(abundance, na.rm = TRUE),
      sd = sd(abundance, na.rm = TRUE),
      median = median(abundance, na.rm = TRUE),
      iqr = IQR(abundance, na.rm = TRUE),
      sum_abundance = sum(abundance, na.rm = TRUE),
      .groups = "drop"
    )
  
  # 5) Paired test per taxon: T0 vs T12
  paired_results <- ra_long %>%
    dplyr::select(FeatureID, PatientID, Time, abundance) %>%
    tidyr::pivot_wider(names_from = Time, values_from = abundance) %>%
    dplyr::group_by(FeatureID) %>%
    dplyr::group_modify(~{
      dat <- .x
      
      pair_ok <- stats::complete.cases(dat$T0, dat$T12)
      x0 <- dat$T0[pair_ok]
      x12 <- dat$T12[pair_ok]
      delta <- x12 - x0
      
      n_pairs <- length(delta)
      
      prev_T0 <- if (n_pairs > 0) 100 * mean(x0 > 0, na.rm = TRUE) else NA_real_
      prev_T12 <- if (n_pairs > 0) 100 * mean(x12 > 0, na.rm = TRUE) else NA_real_
      
      sum_T0 <- sum(x0, na.rm = TRUE)
      sum_T12 <- sum(x12, na.rm = TRUE)
      
      n_unique_delta <- dplyr::n_distinct(delta)
      all_zero_delta <- if (n_pairs > 0) all(delta == 0, na.rm = TRUE) else NA
      
      shapiro_p <- if (n_pairs >= 3 && n_unique_delta >= 2 && !isTRUE(all_zero_delta)) {
        stats::shapiro.test(delta)$p.value
      } else {
        NA_real_
      }
      
      method <- if (n_pairs < 3) {
        NA_character_
      } else if (isTRUE(all_zero_delta)) {
        "no_change_constant"
      } else if (!is.na(shapiro_p) && shapiro_p >= 0.05) {
        "paired_t"
      } else {
        "wilcoxon_paired"
      }
      
      p_value <- if (n_pairs < 3) {
        NA_real_
      } else if (isTRUE(all_zero_delta)) {
        1
      } else if (!is.na(shapiro_p) && shapiro_p >= 0.05) {
        stats::t.test(x12, x0, paired = TRUE)$p.value
      } else {
        stats::wilcox.test(x12, x0, paired = TRUE, exact = FALSE)$p.value
      }
      
      tibble::tibble(
        n_pairs = n_pairs,
        prev_T0 = prev_T0,
        prev_T12 = prev_T12,
        sum_T0 = sum_T0,
        sum_T12 = sum_T12,
        shapiro_p = shapiro_p,
        method = method,
        p_value = p_value,
        mean_delta = if (n_pairs > 0) mean(delta, na.rm = TRUE) else NA_real_,
        median_delta = if (n_pairs > 0) median(delta, na.rm = TRUE) else NA_real_
      )
    }) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      pooled_RA_T0 = 100 * sum_T0 / sum(sum_T0, na.rm = TRUE),
      pooled_RA_T12 = 100 * sum_T12 / sum(sum_T12, na.rm = TRUE),
      p_fdr = stats::p.adjust(p_value, method = "fdr")
    ) %>%
    dplyr::select(
      FeatureID,
      n_pairs,
      prev_T0, prev_T12,
      pooled_RA_T0, pooled_RA_T12,
      mean_delta, median_delta,
      shapiro_p, method, p_value, p_fdr
    ) %>%
    dplyr::arrange(p_fdr)
  
  list(
    ra_long = ra_long,
    stats_basic = stats_basic,
    paired_results = paired_results
  )
}
```

##### Feces
```{r Run function, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
# Phylum
ra_stats_fc_ph <- get_ra_stats_paired(
  tse_obj = subset_feces,
  rank_level = "Phylum"
)

readr::write_excel_csv(
  ra_stats_fc_ph$stats_basic,
  here::here("results", "phyla_basic_stats_by_time_feces.csv")
)

readr::write_excel_csv(
  ra_stats_fc_ph$paired_results,
  here::here("results", "phyla_paired_tests_T0_vs_T12_feces.csv")
)

# Species
ra_stats_fc_sps <- get_ra_stats_paired(
  tse_obj = subset_feces,
  rank_level = "Species"
)

readr::write_excel_csv(
  ra_stats_fc_sps$stats_basic,
  here::here("results", "species_basic_stats_by_time_feces.csv")
)

readr::write_excel_csv(
  ra_stats_fc_sps$paired_results,
  here::here("results", "species_paired_tests_T0_vs_T12_feces.csv")
)
```

##### Sputum
```{r Run function, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
# Phylum
ra_stats_sp_ph <- get_ra_stats_paired(
  tse_obj = subset_sputum,
  rank_level = "Phylum"
)

readr::write_excel_csv(
  ra_stats_sp_ph$stats_basic,
  here::here("results", "phyla_basic_stats_by_time_sputum.csv")
)

readr::write_excel_csv(
  ra_stats_sp_ph$paired_results,
  here::here("results", "phyla_paired_tests_T0_vs_T12_sputum.csv")
)

# Species
ra_stats_sp_sps <- get_ra_stats_paired(
  tse_obj = subset_sputum,
  rank_level = "Species"
)

readr::write_excel_csv(
  ra_stats_sp_sps$stats_basic,
  here::here("results", "species_basic_stats_by_time_sputum.csv")
)

readr::write_excel_csv(
  ra_stats_sp_sps$paired_results,
  here::here("results", "species_paired_tests_T0_vs_T12_sputum.csv")
)
```


# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------
# ALPHA DIVERSITY (AD)
# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------

**Goal:** Measure within-sample diversity changes over time, age, and sex.

Indices for Alpha Diversity:

-   Chao1: estimated species richness

-   Shannon: species richness and evenness

-   Inverse Simpson: dominance (evenness)

## Modified function of AD_plot {MADAV}: AD_plot_paired_csv

```{r AD_plot MODIFIED for paired results and CSV, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
# Drop-in replacement of AD_plot:
# - Adds `paired` and `export_csv` parameters
# - Exports alpha-diversity values to CSV if requested
# - Uses paired comparisons in geom_pwc() when paired = TRUE
AD_plot_paired_csv <- function (tse,
                     assay_type = "counts",
                     var,
                     var_levels = NULL,
                     level = NULL,
                     index = "inverse_simpson",
                     plot_title = "Alpha-diversity",
                     hide.ns = FALSE,
                     paired = FALSE,            # new argument
                     export_csv = NULL) {       # new argument

  `%>%` <- magrittr::`%>%`
  AD <- tse

  if (is.null(level)) {
    level <- mia::taxonomyRanks(tse)[length(mia::taxonomyRanks(tse))]
    warning(paste("No level was specified, alpha-diversity analysis was performed on",
                  level, "level"))
  }

  if (level %in% colnames(SummarizedExperiment::rowData(AD)) &
      !(level %in% mia::taxonomyRanks(AD))) {
    if ("domain" %in% colnames(SummarizedExperiment::rowData(AD))) {
      colnames(SummarizedExperiment::rowData(AD))[colnames(SummarizedExperiment::rowData(AD)) ==
        "domain"] <- "true_domain_column"
    }
    colnames(SummarizedExperiment::rowData(AD))[colnames(SummarizedExperiment::rowData(AD)) ==
      level] <- "domain"
    rank <- "domain"
  } else if (level %in% mia::taxonomyRanks(AD)) {
    rank <- level
  }

  AD_agg <- mia::agglomerateByRank(AD, rank, onRankOnly = TRUE, na.rm = TRUE)

  if (!is.null(var_levels)) {
    SummarizedExperiment::colData(AD_agg)[, var] <- factor(
      SummarizedExperiment::colData(AD_agg)[, var], levels = var_levels
    )
  } else {
    SummarizedExperiment::colData(AD_agg)[, var] <- factor(
      SummarizedExperiment::colData(AD_agg)[, var]
    )
  }

  set_levels <- levels(SummarizedExperiment::colData(AD_agg)[, var])

  if (index %in% c("coverage", "fisher", "gini_simpson", "inverse_simpson",
                   "log_modulo_skewness", "shannon")) {
    AD_index <- mia::estimateDiversity(AD_agg, index = index, assay.type = assay_type)
  } else if (index %in% c("ace", "chao1", "hill", "observed")) {
    AD_index <- mia::estimateRichness(AD_agg, index = index, assay.type = assay_type)
  }

  # NEW: build a tidy alpha-diversity table and optionally export it
  alpha_df <- SummarizedExperiment::colData(AD_index) %>%
    as.data.frame() %>%
    dplyr::transmute(
      SampleID  = rownames(.),
      PatientID = .data[["PatientID"]],
      Time      = .data[[var]],
      AlphaValue = .data[[index]]
  )
  if (!is.null(export_csv)) {
    readr::write_csv(alpha_df, export_csv)
    message("Alpha-diversity exported to: ", export_csv)
  }

  y_label <- stringr::str_replace_all(index, "_", " ") %>%
    stringr::str_to_title() %>% paste("Index", sep = " ")

  plot_palette <- c("#1F77B4FF", "#AEC7E8FF", "#FF7F0EFF",
                    "#FFBB78FF", "#2CA02CFF", "#98DF8AFF", "#D62728FF", "#FF9896FF",
                    "#9467BDFF", "#C5B0D5FF", "#8C564BFF", "#C49C94FF", "#E377C2FF",
                    "#F7B6D2FF", "#7F7F7FFF", "#C7C7C7FF", "#BCBD22FF", "#DBDB8DFF",
                    "#17BECFFF", "#9EDAE5FF")

  if (length(set_levels) <= 10) {
    plot_colors <- plot_palette[seq(1, 19, 2)][1:length(set_levels)]
    color_scale <- ggplot2::scale_color_manual(values = plot_colors)
  } else if (length(set_levels) <= 20) {
    plot_colors <- plot_palette[sort(c(seq(2, 20, 2)[1:(length(set_levels) - 5)],
                                       seq(1, 19, 2)))]
    color_scale <- ggplot2::scale_color_manual(values = plot_colors)
  } else {
    color_scale <- ggraph::scale_color_viridis(discrete = TRUE)
  }

  AD_plot_final <- ggplot2::ggplot(
    SummarizedExperiment::colData(AD_index),
    ggplot2::aes(SummarizedExperiment::colData(AD_index)[, var],
                 SummarizedExperiment::colData(AD_index)[, index])
  ) +
    ggplot2::geom_boxplot(lwd = 0.65, show.legend = FALSE, outlier.shape = NA) +
    ggplot2::geom_jitter(
      ggplot2::aes(color = SummarizedExperiment::colData(AD_index)[, var]),
      position = ggplot2::position_jitterdodge(0.2),
      alpha = 0.75, show.legend = FALSE
    ) +
    color_scale +
    ggplot2::theme_bw(base_size = 16, base_line_size = 1, base_rect_size = 1) +
    ggplot2::theme(
      axis.title.x = ggplot2::element_blank(),
      plot.title   = ggplot2::element_text(size = 17),
      panel.grid   = ggplot2::element_blank(),
      axis.text.x  = ggplot2::element_text(color = "Grey1", size = 12, angle = 45, hjust = 1, face = "plain"),
      axis.text.y  = ggplot2::element_text(color = "Grey1", size = 12, face = "plain"),
      axis.title.y = ggplot2::element_text(color = "Grey1", size = 14, face = "plain")
    ) +
    ggpubr::geom_pwc(
      label = "p.adj.signif",
      p.adjust.method = "BH",
      hide.ns = hide.ns,
      method.args = list(paired = paired)  # NEW: pass paired to tests
    ) +
    ggplot2::ggtitle(plot_title) +
    ggplot2::ylab(y_label)

  print(AD_plot_final)
  return(invisible(AD_plot_final))
}
```

#----------------------
## SAMPLETYPE AND TIME
#----------------------

### Feces at T0 and T12

```{r AD of Feces at T0 and T12, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
AD_plot_fc_ch_csv <- AD_plot_paired_csv(tse = subset_feces_no_UN,
                                var = 'Time',
                                var_levels = c('T0','T12'),
                                level = 'Species',
                                index = 'chao1',  
                                plot_title = 'Alpha diversity of Feces',
                                hide.ns = FALSE,
                                paired = TRUE,
                                export_csv = here::here("results", "alpha_feces_species_chao1.csv"))
ggsave(filename = here("results", "AD_plot_fc_ch_csv.svg"), plot = AD_plot_fc_ch_csv, width = 8, height = 6, dpi = 300, device = "svg")

AD_plot_fc_sh_csv <- AD_plot_paired_csv(tse = subset_feces_no_UN,
                                var = 'Time',
                                var_levels = c('T0','T12'),
                                level = 'Species',
                                index = 'shannon',  
                                plot_title = 'Alpha diversity of Feces',
                                hide.ns = FALSE,
                                paired = TRUE,
                                export_csv = here::here("results", "alpha_feces_species_shannon.csv"))

ggsave(filename = here("results", "AD_plot_fc_sh_csv.svg"), plot = AD_plot_fc_sh_csv, width = 8, height = 6, dpi = 300, device = "svg")

AD_plot_fc_is_csv <- AD_plot_paired_csv(tse = subset_feces_no_UN,
                                var = 'Time',
                                var_levels = c('T0','T12'),
                                level = 'Species',
                                index = 'inverse_simpson',
                                plot_title = 'Alpha diversity of Feces',
                                hide.ns = FALSE,
                                paired = TRUE,
                                export_csv = here::here("results", "alpha_feces_species_inverse_simpson.csv"))

ggsave(filename = here("results", "AD_plot_fc_is_csv.svg"), plot = AD_plot_fc_is_csv, width = 8, height = 6, dpi = 300, device = "svg")
```

#### Plots

##### Chao1

```{r Plots alpha diversity Chao1 FECES, echo=TRUE, fig.show='show', message=FALSE, warning=FALSE}
# 1) Load data
alpha_df_fc <- read_csv(here::here("results", "alpha_feces_species_chao1.csv")) %>%
  mutate(Time = factor(Time, levels = c("T0","T12"))) %>%
  group_by(PatientID) %>% filter(n_distinct(Time) == 2) %>% ungroup()

# 2) Compute paired Wilcoxon test
wilc_fc <- alpha_df_fc %>%
  wilcox_test(AlphaValue ~ Time, paired = TRUE) %>%
  adjust_pvalue(method = "BH")

# 3) Format p-value for label
p_value_label_fc <- ifelse(wilc_fc$p.adj < 0.001, "p.adj < 0.001",
                           paste0("p.adj = ", signif(wilc_fc$p.adj, 3)))

# 4) Define colors for boxplots and points
col_box   <- c("T0" = "#F4E1A1", "T12" = "#A1E1E1")
col_point <- c("T0" = "#FAF3DC", "T12" = "#D6F5F5")

# 5) Add color columns and numeric x-axis values for plotting
alpha_df_fc <- alpha_df_fc %>%
  mutate(
    fill_box   = col_box[Time],
    fill_point = col_point[Time],
    x          = as.numeric(Time)  # for continuous x-axis
  )

# 6) Data for p-value line and label
p_values_fc <- data.frame(
  x = 1, xend = 2,
  y = max(alpha_df_fc$AlphaValue) * 1.05,
  label = p_value_label_fc
)

# 7) Raincloud-style plot with connecting lines
p_chao1_fc <- ggplot() +
  # Raincloud T0
  stat_slab(
    data = alpha_df_fc %>% filter(Time == "T0"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Raincloud T12
  stat_slab(
    data = alpha_df_fc %>% filter(Time == "T12"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Boxplots
  geom_boxplot(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, fill = fill_box),
    width = 0.5, outlier.shape = NA,
    color = "black", alpha = 0.9, show.legend = TRUE
  ) +
  # Points
  geom_point(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, fill = fill_point),
    shape = 21, size = 3, stroke = 0.3, color = "black",
    show.legend = FALSE
  ) +
  # Connecting lines
  geom_line(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, group = PatientID),
    color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
  ) +
  # p-value line and text
  geom_segment(
    data = p_values_fc,
    aes(x = x, xend = xend, y = y, yend = y),
    color = "black", linewidth = 0.5
  ) +
  geom_text(
    data = p_values_fc,
    aes(x = (x + xend) / 2, y = y + 0.02 * y, label = label),
    size = 4
  ) +
  scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
  scale_fill_identity(guide = "legend", name = "Time",
                      breaks = col_box, labels = names(col_box)) +
  labs(x = NULL, y = "Chao1 index - Feces") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 14)
  )

# 8) Save plot
ggsave(filename = here("results", "alpha_diversity_chao1_feces.svg"), plot = p_chao1_fc, width = 8, height = 6, dpi = 300, device = "svg")
```

##### Shannon

```{r Plots alpha diversity Shanon FECES, echo=TRUE, fig.show='show', message=FALSE, warning=FALSE}
# 1) Load data
alpha_df_fc <- read_csv(here::here("results", "alpha_feces_species_shannon.csv")) %>%
  mutate(Time = factor(Time, levels = c("T0","T12"))) %>%
  group_by(PatientID) %>% filter(n_distinct(Time) == 2) %>% ungroup()

# 2) Compute paired Wilcoxon test
wilc_fc <- alpha_df_fc %>%
  wilcox_test(AlphaValue ~ Time, paired = TRUE) %>%
  adjust_pvalue(method = "BH")

# 3) Format p-value for label
p_value_label_fc <- ifelse(wilc_fc$p.adj < 0.001, "p.adj < 0.001",
                           paste0("p.adj = ", signif(wilc_fc$p.adj, 3)))

# 4) Define colors for boxplots and points
col_box   <- c("T0" = "#F4E1A1", "T12" = "#A1E1E1")
col_point <- c("T0" = "#FAF3DC", "T12" = "#D6F5F5")

# 5) Add color columns and numeric x-axis values for plotting
alpha_df_fc <- alpha_df_fc %>%
  mutate(
    fill_box   = col_box[Time],
    fill_point = col_point[Time],
    x          = as.numeric(Time)  # for continuous x-axis
  )

# 6) Data for p-value line and label
p_values_fc <- data.frame(
  x = 1, xend = 2,
  y = max(alpha_df_fc$AlphaValue) * 1.05,
  label = p_value_label_fc
)

# 7) Raincloud-style plot with connecting lines
p_shannon_fc <- ggplot() +
  # Raincloud T0
  stat_slab(
    data = alpha_df_fc %>% filter(Time == "T0"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Raincloud T12
  stat_slab(
    data = alpha_df_fc %>% filter(Time == "T12"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Boxplots
  geom_boxplot(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, fill = fill_box),
    width = 0.5, outlier.shape = NA,
    color = "black", alpha = 0.9, show.legend = TRUE
  ) +
  # Points
  geom_point(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, fill = fill_point),
    shape = 21, size = 3, stroke = 0.3, color = "black",
    show.legend = FALSE
  ) +
  # Connecting lines
  geom_line(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, group = PatientID),
    color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
  ) +
  # p-value line and text
  geom_segment(
    data = p_values_fc,
    aes(x = x, xend = xend, y = y, yend = y),
    color = "black", linewidth = 0.5
  ) +
  geom_text(
    data = p_values_fc,
    aes(x = (x + xend) / 2, y = y + 0.02 * y, label = label),
    size = 4
  ) +
  scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
  scale_fill_identity(guide = "legend", name = "Time",
                      breaks = col_box, labels = names(col_box)) +
  labs(x = NULL, y = "Shannon index - Feces") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 14)
  )

# 8) Save plot
ggsave(filename = here("results", "alpha_diversity_shannon_feces.svg"), plot = p_shannon_fc, width = 8, height = 6, dpi = 300, device = "svg")
```


##### Inverse Simpson

```{r Plots alpha diversity Inverse Simpson FECES, echo=TRUE, fig.show='show', message=FALSE, warning=FALSE}
# 1) Load data
alpha_df_fc <- read_csv(here::here("results", "alpha_feces_species_inverse_simpson.csv")) %>%
  mutate(Time = factor(Time, levels = c("T0","T12"))) %>%
  group_by(PatientID) %>% filter(n_distinct(Time) == 2) %>% ungroup()

# 2) Compute paired Wilcoxon test
wilc_fc <- alpha_df_fc %>%
  wilcox_test(AlphaValue ~ Time, paired = TRUE) %>%
  adjust_pvalue(method = "BH")

# 3) Format p-value for label
p_value_label_fc <- ifelse(wilc_fc$p.adj < 0.001, "p.adj < 0.001",
                           paste0("p.adj = ", signif(wilc_fc$p.adj, 3)))

# 4) Define colors for boxplots and points
col_box   <- c("T0" = "#F4E1A1", "T12" = "#A1E1E1")
col_point <- c("T0" = "#FAF3DC", "T12" = "#D6F5F5")

# 5) Add color columns and numeric x-axis values for plotting
alpha_df_fc <- alpha_df_fc %>%
  mutate(
    fill_box   = col_box[Time],
    fill_point = col_point[Time],
    x          = as.numeric(Time)  # for continuous x-axis
  )

# 6) Raincloud-style plot with connecting lines
p_values_fc <- data.frame(
  x = 1, xend = 2,
  y = max(alpha_df_fc$AlphaValue) * 1.05,
  label = p_value_label_fc
)

# 7) Raincloud-style plot with connecting lines
p_invsimp_fc <- ggplot() +
  # Raincloud T0
  stat_slab(
    data = alpha_df_fc %>% filter(Time == "T0"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Raincloud T12
  stat_slab(
    data = alpha_df_fc %>% filter(Time == "T12"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Boxplots
  geom_boxplot(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, fill = fill_box),
    width = 0.5, outlier.shape = NA,
    color = "black", alpha = 0.9, show.legend = TRUE
  ) +
  # Points
  geom_point(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, fill = fill_point),
    shape = 21, size = 3, stroke = 0.3, color = "black",
    show.legend = FALSE
  ) +
  # Connecting lines
  geom_line(
    data = alpha_df_fc,
    aes(x = x, y = AlphaValue, group = PatientID),
    color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
  ) +
  # p-value line and text
  geom_segment(
    data = p_values_fc,
    aes(x = x, xend = xend, y = y, yend = y),
    color = "black", linewidth = 0.5
  ) +
  geom_text(
    data = p_values_fc,
    aes(x = (x + xend) / 2, y = y + 0.02 * y, label = label),
    size = 4
  ) +
  scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
  scale_fill_identity(guide = "legend", name = "Time",
                      breaks = col_box, labels = names(col_box)) +
  labs(x = NULL, y = "Inverse Simpson index - Feces") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 14)
  )

# 8) Save plot
ggsave(filename = here("results", "alpha_diversity_invsimp_feces.svg"), plot = p_invsimp_fc, width = 8, height = 6, dpi = 300, device = "svg")
```


### Sputum at T0 and T12

```{r AD of Sputum at T0 and T12, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
AD_plot_sp_ch_csv <- AD_plot_paired_csv(tse = subset_sputum_no_UN,
                                var = 'Time',
                                var_levels = c('T0','T12'),
                                level = 'Species',
                                index = 'chao1',
                                plot_title = 'Alpha diversity of Sputum',
                                hide.ns = FALSE,
                                paired = TRUE,
                                export_csv = here::here("results", "alpha_sputum_species_chao1.csv"))
ggsave(filename = here("results", "AD_plot_sp_ch_csv.svg"), plot = AD_plot_sp_ch_csv, width = 8, height = 6, dpi = 300, device = "svg")

AD_plot_sp_sh_csv <- AD_plot_paired_csv(tse = subset_sputum_no_UN,
                                var = 'Time',
                                var_levels = c('T0','T12'),
                                level = 'Species',
                                index = 'shannon',
                                plot_title = 'Alpha diversity of Sputum',
                                hide.ns = FALSE,
                                paired = TRUE,
                                export_csv = here::here("results", "alpha_sputum_species_shannon.csv"))
ggsave(filename = here("results", "AD_plot_sp_sh_csv.svg"), plot = AD_plot_sp_sh_csv, width = 8, height = 6, dpi = 300, device = "svg")

AD_plot_sp_is_csv <- AD_plot_paired_csv(tse = subset_sputum_no_UN,
                                var = 'Time',
                                var_levels = c('T0','T12'),
                                level = 'Species',
                                index = 'inverse_simpson',
                                plot_title = 'Alpha diversity of Sputum',
                                hide.ns = FALSE,
                                paired = TRUE,
                                export_csv = here::here("results", "alpha_sputum_species_inverse_simpson.csv"))

ggsave(filename = here("results", "AD_plot_sp_is_csv.svg"), plot = AD_plot_sp_is_csv, width = 8, height = 6, dpi = 300, device = "svg")
```


#### Plots

##### Chao1

```{r Plots alpha diversity Chao1 SPUTUM, echo=TRUE, fig.show='show', message=FALSE, warning=FALSE}
# 1) Load data
alpha_df_sp <- read_csv(here::here("results", "alpha_sputum_species_chao1.csv")) %>%
  mutate(Time = factor(Time, levels = c("T0","T12"))) %>%
  group_by(PatientID) %>% filter(n_distinct(Time) == 2) %>% ungroup()

# 2) Compute paired Wilcoxon test
wilc_sp <- alpha_df_sp %>%
  wilcox_test(AlphaValue ~ Time, paired = TRUE) %>%
  adjust_pvalue(method = "BH")

# 3) Format p-value for label
p_value_label_sp <- ifelse(wilc_sp$p.adj < 0.001, "p.adj < 0.001",
                           paste0("p.adj = ", signif(wilc_sp$p.adj, 3)))

# 4) Define colors for boxplots and points
col_box   <- c("T0" = "#D9C2E9", "T12" = "#B3E1B6")
col_point <- c("T0" = "#EDE3F4", "T12" = "#E2F5E3")

# 5) Add color columns and numeric x-axis values for plotting
alpha_df_sp <- alpha_df_sp %>%
  mutate(
    fill_box   = col_box[Time],
    fill_point = col_point[Time],
    x          = as.numeric(Time) # for continuous x-axis
  )

# 6) Data for p-value line and label
p_values_sp <- data.frame(
  x = 1, xend = 2,
  y = max(alpha_df_sp$AlphaValue) * 1.05,
  label = p_value_label_sp
)

# 7) Raincloud-style plot with connecting lines
p_chao1_sp <- ggplot() +
  # Raincloud T0
  stat_slab(
    data = alpha_df_sp %>% filter(Time == "T0"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Raincloud T12
  stat_slab(
    data = alpha_df_sp %>% filter(Time == "T12"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Boxplots
  geom_boxplot(
    data = alpha_df_sp,
    aes(x = x, y = AlphaValue, fill = fill_box),
    width = 0.5, outlier.shape = NA,
    color = "black", alpha = 0.9, show.legend = TRUE
  ) +
  # Points
  geom_point(
    data = alpha_df_sp,
    aes(x = x, y = AlphaValue, fill = fill_point),
    shape = 21, size = 3, stroke = 0.3, color = "black",
    show.legend = FALSE
  ) +
  # Connecting lines
  geom_line(
    data = alpha_df_sp,
    aes(x = x, y = AlphaValue, group = PatientID),
    color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
  ) +
  # p-value line and text
  geom_segment(
    data = p_values_sp,
    aes(x = x, xend = xend, y = y, yend = y),
    color = "black", linewidth = 0.5
  ) +
  geom_text(
    data = p_values_sp,
    aes(x = (x + xend) / 2, y = y + 0.02 * y, label = label),
    size = 4
  ) +
  scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
  scale_fill_identity(guide = "legend", name = "Time",
                      breaks = col_box, labels = names(col_box)) +
  labs(x = NULL, y = "Chao1 index - Sputum") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 14)
  )

# 8) Save plot
ggsave(filename = here("results", "alpha_diversity_chao1_sputum.svg"), plot = p_chao1_sp, width = 8, height = 6, dpi = 300, device = "svg")
```

##### Shannon

```{r Plots alpha diversity Shannon SPUTUM, echo=TRUE, fig.show='show', message=FALSE, warning=FALSE}
# 1) Load data
alpha_df_sp <- read_csv(here::here("results", "alpha_sputum_species_shannon.csv")) %>%
  mutate(Time = factor(Time, levels = c("T0","T12"))) %>%
  group_by(PatientID) %>% filter(n_distinct(Time) == 2) %>% ungroup()

# 2) Compute paired Wilcoxon test
wilc_sp <- alpha_df_sp %>%
  wilcox_test(AlphaValue ~ Time, paired = TRUE) %>%
  adjust_pvalue(method = "BH")

# 3) Format p-value for label
p_value_label_sp <- ifelse(wilc_sp$p.adj < 0.001, "p.adj < 0.001",
                           paste0("p.adj = ", signif(wilc_sp$p.adj, 3)))

# 4) Define colors for boxplots and points
col_box   <- c("T0" = "#D9C2E9", "T12" = "#B3E1B6")
col_point <- c("T0" = "#EDE3F4", "T12" = "#E2F5E3")

# 5) Add color columns and numeric x-axis values for plotting
alpha_df_sp <- alpha_df_sp %>%
  mutate(
    fill_box   = col_box[Time],
    fill_point = col_point[Time],
    x          = as.numeric(Time)  # for continuous x-axis
  )

# 6) Data for p-value line and label
p_values_sp <- data.frame(
  x = 1, xend = 2,
  y = max(alpha_df_sp$AlphaValue) * 1.05,
  label = p_value_label_sp
)

# 7) Raincloud-style plot with connecting lines
p_shannon_sp <- ggplot() +
  # Raincloud T0
  stat_slab(
    data = alpha_df_sp %>% filter(Time == "T0"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Raincloud T12
  stat_slab(
    data = alpha_df_sp %>% filter(Time == "T12"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Boxplots
  geom_boxplot(
    data = alpha_df_sp,
    aes(x = x, y = AlphaValue, fill = fill_box),
    width = 0.5, outlier.shape = NA,
    color = "black", alpha = 0.9, show.legend = TRUE
  ) +
  # Points
  geom_point(
    data = alpha_df_sp,
    aes(x = x, y = AlphaValue, fill = fill_point),
    shape = 21, size = 3, stroke = 0.3, color = "black",
    show.legend = FALSE
  ) +
  # Connecting lines
  geom_line(
    data = alpha_df_sp,
    aes(x = x, y = AlphaValue, group = PatientID),
    color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
  ) +
  # p-value line and text
  geom_segment(
    data = p_values_sp,
    aes(x = x, xend = xend, y = y, yend = y),
    color = "black", linewidth = 0.5
  ) +
  geom_text(
    data = p_values_sp,
    aes(x = (x + xend) / 2, y = y + 0.02 * y, label = label),
    size = 4
  ) +
  scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
  scale_fill_identity(guide = "legend", name = "Time",
                      breaks = col_box, labels = names(col_box)) +
  labs(x = NULL, y = "Shannon index - Sputum") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 14)
  )

# 8) Save plot
ggsave(filename = here("results", "alpha_diversity_shannon_sputum.svg"), plot = p_shannon_sp, width = 8, height = 6, dpi = 300, device = "svg")
```

##### Inverse Simpson

```{r Plots alpha diversity Inverse Simpson SPUTUM, echo=TRUE, fig.show='show', message=FALSE, warning=FALSE}
# 1) Load data
alpha_df_sp <- read_csv(here::here("results", "alpha_sputum_species_inverse_simpson.csv")) %>%
  mutate(Time = factor(Time, levels = c("T0","T12"))) %>%
  group_by(PatientID) %>% filter(n_distinct(Time) == 2) %>% ungroup()

# 2) Compute paired Wilcoxon test
wilc_sp <- alpha_df_sp %>%
  wilcox_test(AlphaValue ~ Time, paired = TRUE) %>%
  adjust_pvalue(method = "BH")

# 3) Format p-value for label
p_value_label_sp <- ifelse(wilc_sp$p.adj < 0.001, "p.adj < 0.001",
                           paste0("p.adj = ", signif(wilc_sp$p.adj, 3)))

# 4) Define colors for boxplots and points
col_box   <- c("T0" = "#D9C2E9", "T12" = "#B3E1B6")
col_point <- c("T0" = "#EDE3F4", "T12" = "#E2F5E3")

# 5) Add color columns and numeric x-axis values for plotting
alpha_df_sp <- alpha_df_sp %>%
  mutate(
    fill_box   = col_box[Time],
    fill_point = col_point[Time],
    x          = as.numeric(Time)  # for continuous x-axis
  )

# 6) Data for p-value line and label
p_values_sp <- data.frame(
  x = 1, xend = 2,
  y = max(alpha_df_sp$AlphaValue) * 1.05,
  label = p_value_label_sp
)

# 7) Raincloud-style plot with connecting lines
p_invsimp_sp <- ggplot() +
  # Raincloud T0
  stat_slab(
    data = alpha_df_sp %>% filter(Time == "T0"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Raincloud T12
  stat_slab(
    data = alpha_df_sp %>% filter(Time == "T12"),
    aes(x = x, y = AlphaValue, fill = fill_box),
    side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
  ) +
  # Boxplots
  geom_boxplot(
    data = alpha_df_sp,
    aes(x = x, y = AlphaValue, fill = fill_box),
    width = 0.5, outlier.shape = NA,
    color = "black", alpha = 0.9, show.legend = TRUE
  ) +
  # Points
  geom_point(
    data = alpha_df_sp,
    aes(x = x, y = AlphaValue, fill = fill_point),
    shape = 21, size = 3, stroke = 0.3, color = "black",
    show.legend = FALSE
  ) +
  # Connecting lines
  geom_line(
    data = alpha_df_sp,
    aes(x = x, y = AlphaValue, group = PatientID),
    color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
  ) +
  # p-value line and text
  geom_segment(
    data = p_values_sp,
    aes(x = x, xend = xend, y = y, yend = y),
    color = "black", linewidth = 0.5
  ) +
  geom_text(
    data = p_values_sp,
    aes(x = (x + xend) / 2, y = y + 0.02 * y, label = label),
    size = 4
  ) +
  scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
  scale_fill_identity(guide = "legend", name = "Time",
                      breaks = col_box, labels = names(col_box)) +
  labs(x = NULL, y = "Inverse Simpson index - Sputum") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 14)
  )

# 8) Save plot
ggsave(filename = here("results", "alpha_diversity_invsimp_sputum.svg"), plot = p_invsimp_sp, width = 8, height = 6, dpi = 300, device = "svg")
```


# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------
# BETA DIVERSITY (AD)
# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------

**Goal:** Measure between-sample diversity differences over time, age, and sex.

Dissimilarity Metrics:

-   Bray-Curtis: species abundance dissimilarity

-   Weighted UniFrac: phylogenetic structure and abundance


### Modified function of BD_plot {MADAV}: BD_plot_paired_csv

```{r BD_plot MODIFIED paired csv, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
BD_plot_paired_csv <- function(tse, assay_type = "counts", var, var_levels = NULL, 
    level = NULL, method = "bray", weighted_UniFrac = TRUE, ordination = "PCoA", 
    constraining = var, plot_title = "Beta Diversity", dissimilarity_plot = FALSE, 
    hide.ns = FALSE, export_prefix = NULL) {

  set.seed(1234)
  `%>%` <- magrittr::`%>%`
  BD <- tse

  if (is.null(level)) {
    level <- mia::taxonomyRanks(tse)[length(mia::taxonomyRanks(tse))]
    warning(paste("No level was specified, beta-diversity analysis was performed on", level, "level"))
  }

  if (level %in% colnames(SummarizedExperiment::rowData(BD)) & !(level %in% mia::taxonomyRanks(BD))) {
    if ("domain" %in% colnames(SummarizedExperiment::rowData(BD))) {
      colnames(SummarizedExperiment::rowData(BD))[colnames(SummarizedExperiment::rowData(BD)) == "domain"] <- "true_domain_column"
    }
    colnames(SummarizedExperiment::rowData(BD))[colnames(SummarizedExperiment::rowData(BD)) == level] <- "domain"
    rank <- "domain"
  } else if (level %in% mia::taxonomyRanks(BD)) {
    rank <- level
  }

  BD_agg <- mia::agglomerateByRank(BD, rank, onRankOnly = TRUE, agglomerateTree = TRUE, na.rm = TRUE)

  if (length(var) == 1) var_levels <- list(var_levels)
  for (cov in 1:length(var)) {
    if (!is.null(var_levels[[cov]])) {
      SummarizedExperiment::colData(BD_agg)[, var[cov]] <- factor(SummarizedExperiment::colData(BD_agg)[, var[cov]], levels = var_levels[[cov]])
    } else {
      SummarizedExperiment::colData(BD_agg)[, var[cov]] <- factor(SummarizedExperiment::colData(BD_agg)[, var[cov]])
    }
  }

  set_levels <- lapply(var, function(x) levels(SummarizedExperiment::colData(BD_agg)[, x]))

  if (method %in% c("manhattan","euclidean","canberra","clark","bray","kulczynski","gower","altGower",
                    "morisita","horn","mountford","raup","binomial","chao","cao","mahalanobis","chisq",
                    "chord","hellinger","aitchison","robust.aitchison")) {
    BD_dist <- vegan::vegdist(as.matrix(t(SummarizedExperiment::assays(BD_agg)[[assay_type]])), method = method, binary = FALSE)
  } else if (method == "jaccard") {
    BD_dist <- vegan::vegdist(as.matrix(t(SummarizedExperiment::assays(BD_agg)[[assay_type]])), method = method, binary = TRUE)
  } else if (method == "UniFrac") {
    BD_dist <- calculateUnifrac(BD_agg, assay.type = assay_type, weighted = weighted_UniFrac)
  } else if (method == "JSD") {
    BD_dist <- calculateJSD(BD_agg, assay.type = assay_type)
  }

  # PERMANOVA (pareado por PatientID)
  permanova <- vegan::adonis2(
    BD_dist ~ SummarizedExperiment::colData(BD_agg)[, var[1]],
    strata = SummarizedExperiment::colData(BD_agg)$PatientID
  ) %>% data.frame()
  rownames(permanova)[1] <- var[1]
  print(permanova)

  permanova_R2 <- signif(permanova[1, 3], 2)
  permanova_P  <- signif(permanova[1, 5], 2)
  permanova_plot_text <- bquote(atop(R^2:~.(permanova_R2), paste("     Pr(>F): ", .(permanova_P))))

  # Dispersion and permutest
  BD_dispr  <- vegan::betadisper(BD_dist, SummarizedExperiment::colData(BD_agg)[, var[1]])
  permutest <- vegan::permutest(BD_dispr, pairwise = TRUE)
  print(permutest)

  # Ordenation and dataframe for plot
  pct1 <- pct2 <- NA_real_

  if (ordination == "PCoA") {
    BD_ord <- stats::cmdscale(BD_dist, eig = TRUE)
    BD_ord_df <- data.frame(BD_ord$points[,1], BD_ord$points[,2], SummarizedExperiment::colData(BD_agg)[, var])
    pct1 <- round(100 * BD_ord$eig[1] / sum(BD_ord$eig), 2)
    pct2 <- round(100 * BD_ord$eig[2] / sum(BD_ord$eig), 2)
    axis_title_x <- paste0("PCoA1 (", pct1, "%)")
    axis_title_y <- paste0("PCoA2 (", pct2, "%)")
  } else if (ordination == "NMDS") {
    BD_ord <- vegan::metaMDS(BD_dist, k = 2)
    BD_ord_df <- data.frame(BD_ord$points, SummarizedExperiment::colData(BD_agg)[, var])
    axis_title_x <- "NMDS1"; axis_title_y <- "NMDS2"
  } else if (ordination == "RDA") {
    BD_ord <- mia::calculateRDA(BD_agg, stats::formula(paste("data ~", paste(constraining, collapse = " + "))),
                                method = method, assay.type = assay_type, homogeneity.test = "permanova")
    BD_ord_df <- data.frame(BD_ord[,1:2], SummarizedExperiment::colData(BD_agg)[, var])
    axis_title_x <- paste0("dbRDA1 (", round(100*attr(BD_ord,"rda")$CCA$eig[1]/attr(BD_ord,"rda")$CCA$tot.chi, 2), "%)")
    axis_title_y <- paste0("dbRDA2 (", round(100*attr(BD_ord,"rda")$CCA$eig[2]/attr(BD_ord,"rda")$CCA$tot.chi, 2), "%)")
  } else if (ordination == "CCA") {
    BD_ord <- mia::calculateCCA(BD_agg, stats::formula(paste("data ~", paste(constraining, collapse = " + "))),
                                method = method, assay.type = assay_type, homogeneity.test = "permanova")
    BD_ord_df <- data.frame(BD_ord[,1:2], SummarizedExperiment::colData(BD_agg)[, var])
    axis_title_x <- paste0("CCA1 (", round(100*attr(BD_ord,"cca")$CCA$eig[1]/attr(BD_ord,"cca")$CCA$tot.chi, 2), "%)")
    axis_title_y <- paste0("CCA2 (", round(100*attr(BD_ord,"cca")$CCA$eig[2]/attr(BD_ord,"cca")$CCA$tot.chi, 2), "%)")
  } else if (ordination == "DPCoA") {
    BD_ord <- mia::calculateDPCoA(BD_agg)
    BD_ord_df <- data.frame(BD_ord[,1:2], SummarizedExperiment::colData(BD_agg)[, var])
    axis_title_x <- paste0("DPCoA1 (", round(100*attr(BD_ord,"eig")[1]/sum(attr(BD_ord,"eig")),2), "%)")
    axis_title_y <- paste0("DPCoA2 (", round(100*attr(BD_ord,"eig")[2]/sum(attr(BD_ord,"eig")),2), "%)")
  }

  if (length(var) > 1) { shape <- BD_ord_df[,4]; shape_lab <- var[2] } else { shape <- NULL; shape_lab <- NULL }

  plot_palette <- c("#1F77B4FF","#AEC7E8FF","#FF7F0EFF","#FFBB78FF","#2CA02CFF","#98DF8AFF",
                    "#D62728FF","#FF9896FF","#9467BDFF","#C5B0D5FF","#8C564BFF","#C49C94FF",
                    "#E377C2FF","#F7B6D2FF","#7F7F7FFF","#C7C7C7FF","#BCBD22FF","#DBDB8DFF",
                    "#17BECFFF","#9EDAE5FF")

  if (length(set_levels[[1]]) <= 5) {
    plot_colors <- plot_palette[seq(1, 9, 2)][1:length(set_levels[[1]])]
    color_scale <- ggplot2::scale_color_manual(values = plot_colors)
  } else if (length(set_levels[[1]]) <= 10) {
    plot_colors <- plot_palette[sort(c(seq(2, 10, 2)[1:(length(set_levels[[1]]) - 5)], seq(1, 9, 2)))]
    color_scale <- ggplot2::scale_color_manual(values = plot_colors)
  } else if (length(set_levels[[1]]) <= 20) {
    plot_colors <- plot_palette[sort(c(seq(2, 20, 2)[1:(length(set_levels[[1]]) - 10)], seq(1, 19, 2)))]
    color_scale <- ggplot2::scale_color_manual(values = plot_colors)
  } else {
    color_scale <- ggraph::scale_color_viridis(discrete = TRUE)
  }

  BD_ord_plot <- ggplot2::ggplot(BD_ord_df, ggplot2::aes(BD_ord_df[,1], BD_ord_df[,2], color = BD_ord_df[,3])) +
    ggplot2::geom_point(ggplot2::aes(shape = shape), size = 2) +
    ggplot2::ggtitle(plot_title) +
    color_scale +
    ggplot2::stat_ellipse(ggplot2::aes(group = BD_ord_df[,3]), linetype = 1, linewidth = 1) +
    ggplot2::labs(caption = permanova_plot_text, x = axis_title_x, y = axis_title_y, shape = shape_lab) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      plot.title = ggplot2::element_text(size = 17),
      legend.text = ggplot2::element_text(size = 12),
      legend.title = ggplot2::element_text(size = 13),
      panel.grid = ggplot2::element_blank(),
      axis.text.x = ggplot2::element_text(color = "Black", size = 12),
      axis.text.y = ggplot2::element_text(color = "Black", size = 12),
      axis.title.x = ggplot2::element_text(color = "Black", size = 14),
      axis.title.y = ggplot2::element_text(color = "Black", size = 14),
      plot.caption.position = "plot",
      plot.caption = ggplot2::element_text(hjust = 0.08, size = 11)
    ) +
    ggplot2::guides(color = ggplot2::guide_legend(order = 1, title = var[1]),
                    shape = ggplot2::guide_legend(order = 2, title = var[2]))

  print(BD_ord_plot)

  # ---------- EXPORTS ----------
  if (!is.null(export_prefix)) {
    # 1) PERMANOVA
    readr::write_csv(permanova,
      file = here::here("results", paste0(export_prefix, "_permanova.csv"))
    )
    # 2) PERMUTEST (table)
    readr::write_csv(as.data.frame(permutest$tab),
      file = here::here("results", paste0(export_prefix, "_permutest.csv"))
    )
    # 3) COORDINATES
    sample_ids <- rownames(if (ordination %in% c("PCoA","NMDS")) BD_ord$points else as.matrix(BD_ord_df[,1:2]))
    if (is.null(sample_ids)) sample_ids <- colnames(SummarizedExperiment::assays(BD_agg)[[assay_type]])
    coords_df <- data.frame(
      SampleID = sample_ids,
      Axis1    = BD_ord_df[[1]],
      Axis2    = BD_ord_df[[2]],
      Group    = BD_ord_df[[3]]
    )
    readr::write_csv(coords_df,
      file = here::here("results", paste0(export_prefix, "_coords.csv"))
    )
    # 4) % of axis (only PCoA; if not, it is omitted)
    if (ordination == "PCoA" && is.finite(pct1) && is.finite(pct2)) {
      readr::write_csv(
        tibble::tibble(Axis = c("PCoA1","PCoA2"), Percent = c(pct1, pct2)),
        here::here("results", paste0(export_prefix, "_axes.csv"))
      )
    }
  }
  # -----------------------------

  BD_output <- list(BD_ord_plot, permanova, permutest)
  names(BD_output) <- c("BD_plot", "PERMANOVA_results", "Permutest_results")
  invisible(BD_output)
}
```


#----------------------
## SAMPLETYPE AND TIME
#----------------------

### Feces at T0 and T12

```{r Beta diversity paired CSV Feces at T0 and T12 no error message, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE, error=TRUE}
BD_plot_cf_fc_csv <- BD_plot_paired_csv(
  tse = subset_feces_no_UN,
  var = 'Time',
  var_levels = c('T0','T12'),
  level = 'Species',
  method = 'bray',
  weighted_UniFrac = TRUE,
  ordination = 'PCoA',
  constraining = var,
  plot_title = 'Beta diversity of Feces',
  dissimilarity_plot = FALSE,
  hide.ns = FALSE,
  export_prefix = "beta_feces_species_bray"
)

ggsave(filename = here("results", "BD_plot_cf_fc_csv.svg"), plot = BD_plot_cf_fc_csv$BD_plot_paired_csv, width = 10, height = 7, dpi = 300, device = "svg")
```

#### Plot

```{r Plot BD feces, echo=TRUE, fig.show='show', message=FALSE, warning=FALSE}
# ==== files prefixes ====
prefix <- "beta_feces_species_bray"
# =============================

# 1) Load data
coords <- read_csv(here("results", paste0(prefix, "_coords.csv"))) %>%
  mutate(Group = factor(Group, levels = c("T0","T12")))

perm <- read_csv(here("results", paste0(prefix, "_permanova.csv")))

# Helpers to extract numbers or text
get_num1 <- function(x) {
  if (is.numeric(x)) x[1] else readr::parse_number(as.character(x[1]))
}

# R2 and p
R2_val <- signif(get_num1(perm$R2), 3)

p_col  <- names(perm)[grepl("Pr\\(>F\\)|Pr|p", names(perm), ignore.case = TRUE)][1]
p_num  <- get_num1(perm[[p_col]])
p_txt  <- if (is.na(p_num)) "NA" else if (p_num < 1e-3) "< 0.001" else signif(p_num, 3)

cap_text <- paste0("R\u00B2 = ", R2_val, "   •   p = ", p_txt)

# % of axis
axes_file <- here("results", paste0(prefix, "_axes.csv"))
if (file.exists(axes_file)) {
  ax <- read_csv(axes_file)
  ax1 <- ax$Percent[ax$Axis %in% c("PCoA1","PC1")][1]
  ax2 <- ax$Percent[ax$Axis %in% c("PCoA2","PC2")][1]
  x_lab <- paste0("PCoA1 (", ax1, "%)")
  y_lab <- paste0("PCoA2 (", ax2, "%)")
} else {
  x_lab <- "PCoA1"; y_lab <- "PCoA2"
}

# Colours
col_ellipse <- c("T0" = "#F4E1A1", "T12" = "#A1E1E1")
col_point   <- c("T0" = "#FAF3DC", "T12" = "#D6F5F5")

# 3) Plot
p_beta_fc <- ggplot(coords, aes(Axis1, Axis2)) +
  stat_ellipse(aes(fill = Group, color = Group),
               geom = "polygon", level = 0.95,
               alpha = 0.35, linewidth = 0.6, show.legend = TRUE) +
  geom_point(
    data = coords %>% filter(Group == "T0"),
    shape = 21, size = 3, stroke = 0.3, color = "black",
    fill  = col_point["T0"]
  ) +
  geom_point(
    data = coords %>% filter(Group == "T12"),
    shape = 21, size = 3, stroke = 0.3, color = "black",
    fill  = col_point["T12"]
  ) +
  scale_fill_manual(values = col_ellipse, name = "Time") +
  scale_color_manual(values = col_ellipse, guide = "none") +
  guides(fill = guide_legend(override.aes = list(alpha = 0.35, linewidth = 0.6, color = col_ellipse))) +
  labs(
    x = x_lab, y = y_lab,
    title = "Beta diversity (PCoA) - Feces",
    caption = cap_text
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor      = element_blank(),
    legend.position       = "right",
    plot.title            = element_text(size = 14),
    plot.caption.position = "plot",
    plot.caption          = element_text(hjust = 0, size = 11)
  )

p_beta_fc

# 4) Save SVG
ggsave(here::here("results", "beta_feces_pcoa.svg"),
       p_beta_fc, width = 8, height = 6, units = "in")
```

### Sputum at T0 and T12

```{r Beta diversity paired CSV Sputum at T0 and T12 no error message, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE, error=TRUE}
BD_plot_cf_sp_csv <- BD_plot_paired_csv(
  tse = subset_sputum_no_UN,
  var = 'Time',
  var_levels = c('T0','T12'),
  level = 'Species',
  method = 'bray',
  weighted_UniFrac = TRUE,
  ordination = 'PCoA',
  constraining = var,
  plot_title = 'Beta diversity of Sputum',
  dissimilarity_plot = FALSE,
  hide.ns = FALSE,
  export_prefix = "beta_sputum_species_bray"
)

ggsave(filename = here("results", "BD_plot_cf_sp_csv.svg"), plot = BD_plot_cf_sp_csv$BD_plot_paired_csv, width = 10, height = 7, dpi = 300, device = "svg")
```

#### Plot

```{r Plot BD sputum, echo=TRUE, fig.show='show', message=FALSE, warning=FALSE}
# ==== files prefixes ====
prefix <- "beta_sputum_species_bray"
# =============================

# 1) Load data
coords <- read_csv(here("results", paste0(prefix, "_coords.csv"))) %>%
  mutate(Group = factor(Group, levels = c("T0","T12")))

perm <- read_csv(here("results", paste0(prefix, "_permanova.csv")))

# Helpers to extract numbers or text
get_num1 <- function(x) {
  if (is.numeric(x)) x[1] else readr::parse_number(as.character(x[1]))
}

# R2 and p
R2_val <- signif(get_num1(perm$R2), 3)

p_col  <- names(perm)[grepl("Pr\\(>F\\)|Pr|p", names(perm), ignore.case = TRUE)][1]
p_num  <- get_num1(perm[[p_col]])
p_txt  <- if (is.na(p_num)) "NA" else if (p_num < 1e-3) "< 0.001" else signif(p_num, 3)

cap_text <- paste0("R\u00B2 = ", R2_val, "   •   p = ", p_txt)

# % of axis
axes_file <- here("results", paste0(prefix, "_axes.csv"))
if (file.exists(axes_file)) {
  ax <- read_csv(axes_file)
  ax1 <- ax$Percent[ax$Axis %in% c("PCoA1","PC1")][1]
  ax2 <- ax$Percent[ax$Axis %in% c("PCoA2","PC2")][1]
  x_lab <- paste0("PCoA1 (", ax1, "%)")
  y_lab <- paste0("PCoA2 (", ax2, "%)")
} else {
  x_lab <- "PCoA1"; y_lab <- "PCoA2"
}

# Colours
col_ellipse <- c("T0" = "#D9C2E9", "T12" = "#B3E1B6")
col_point   <- c("T0" = "#EDE3F4", "T12" = "#E2F5E3")

# 3) Plot
p_beta_sp <- ggplot(coords, aes(Axis1, Axis2)) +
  stat_ellipse(aes(fill = Group, color = Group),
               geom = "polygon", level = 0.95,
               alpha = 0.35, linewidth = 0.6, show.legend = TRUE) +
  geom_point(
    data = coords %>% filter(Group == "T0"),
    shape = 21, size = 3, stroke = 0.3, color = "black",
    fill  = col_point["T0"]
  ) +
  geom_point(
    data = coords %>% filter(Group == "T12"),
    shape = 21, size = 3, stroke = 0.3, color = "black",
    fill  = col_point["T12"]
  ) +
  scale_fill_manual(values = col_ellipse, name = "Time") +
  scale_color_manual(values = col_ellipse, guide = "none") +
  guides(fill = guide_legend(override.aes = list(alpha = 0.35, linewidth = 0.6, color = col_ellipse))) +
  labs(
    x = x_lab, y = y_lab,
    title = "Beta diversity (PCoA) - Sputum",
    caption = cap_text
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor      = element_blank(),
    legend.position       = "right",
    plot.title            = element_text(size = 14),
    plot.caption.position = "plot",
    plot.caption          = element_text(hjust = 0, size = 11)
  )

p_beta_sp

# 4) Save SVG
ggsave(here::here("results", "beta_sputum_pcoa.svg"),
       p_beta_sp, width = 8, height = 6, units = "in")
```


# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------
# DIFFERENTIAL ABUNDANCE ANALYSIS (DAA)
# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------

**Goal:** Identify differentially abundant species between groups.

## Modified function of RA_plot {MADAV}: RA_plot_csv
A modification for the RA_plot() function from MADAV package was used in order to export the results.

```{r RA_plot MODIFIED for CSV, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
RA_plot_csv <- function (features, tse, var, var_levels = NULL, level = NULL, 
                         total_tools = 0, plot_title = "Relative Abundance", 
                         hide.ns = FALSE, panel_dim = c(4, 4),
                         export_csv = NULL) {  # NEW: optional CSV export path
  
  `%>%` <- magrittr::`%>%`
  RAP <- tse
  
  if (is.null(level)) {
    level <- mia::taxonomyRanks(tse)[length(mia::taxonomyRanks(tse))]
    warning(paste("No level was specified, DA analysis was performed on", level, "level"))
  }
  
  if (level %in% colnames(SummarizedExperiment::rowData(RAP)) & !(level %in% mia::taxonomyRanks(RAP))) {
    if ("domain" %in% colnames(SummarizedExperiment::rowData(RAP))) {
      colnames(SummarizedExperiment::rowData(RAP))[colnames(SummarizedExperiment::rowData(RAP)) == "domain"] <- "true_domain_column"
    }
    colnames(SummarizedExperiment::rowData(RAP))[colnames(SummarizedExperiment::rowData(RAP)) == level] <- "domain"
    rank <- "domain"
  } else if (level %in% mia::taxonomyRanks(RAP)) {
    rank <- level
  }
  
  RAP_agg <- mia::transformAssay(RAP, method = "relabundance") %>% 
    mia::agglomerateByRank(rank, onRankOnly = TRUE, na.rm = TRUE)
  
  if (!is.null(attr(features, "var"))) {
    feature_frame <- features$comparison[features$comparison$total_tools >= total_tools, c("featureID", "var")]
    if (nrow(feature_frame) != 0) {
      feature_frame <- feature_frame[!(duplicated(feature_frame)), ] %>% 
        dplyr::mutate(var = paste("sign", var, sep = "|"))
      feature_frame <- tidyr::pivot_wider(feature_frame, names_from = "var", values_from = "var")
      feature_frame[, 2:length(feature_frame)] <- apply(feature_frame[, 2:length(feature_frame)], 2, function(x) ifelse(is.na(x), FALSE, TRUE))
    } else {
      stop("No features were detected by the selected number of total_tools")
    }
  } else {
    feature_frame <- data.frame(matrix(ncol = length(var) + 1, nrow = length(features))) %>% 
      stats::setNames(c("featureID", paste("sign", var, sep = "|")))
    feature_frame$featureID <- features
    feature_frame[, paste("sign", var, sep = "|")] <- TRUE
  }
  
  if (all(feature_frame$featureID %in% rownames(RAP_agg))) {
    RAP_sel <- RAP_agg[rownames(RAP_agg) %in% feature_frame$featureID, ]
  } else {
    stop("Not all features are present in the tse object at the selected level")
  }
  
  RAP_meta <- tibble::rownames_to_column(data.frame(SummarizedExperiment::colData(RAP_sel)), var = "sampleID") %>% 
    dplyr::select(dplyr::all_of(c("sampleID", var)))
  
  RAP_melt <- as.data.frame(SummarizedExperiment::assays(RAP_sel)[["relabundance"]]) %>% 
    tibble::rownames_to_column(var = "featureID") %>% 
    reshape2::melt(id = "featureID") %>% 
    stats::setNames(c("featureID", "sampleID", "abundance")) %>% 
    dplyr::right_join(RAP_meta, by = "sampleID") %>% 
    dplyr::right_join(feature_frame, by = "featureID")
  
  # NEW: export relative abundance table to CSV (long format)
  if (!is.null(export_csv)) {
    dir.create(dirname(export_csv), recursive = TRUE, showWarnings = FALSE)
    
    rap_export <- as.data.frame(RAP_melt)
    first_cols <- c("featureID", "sampleID", "abundance", var)
    first_cols <- first_cols[first_cols %in% colnames(rap_export)]
    rap_export <- rap_export[, c(first_cols, setdiff(colnames(rap_export), first_cols)), drop = FALSE]
    
    names(rap_export)[names(rap_export) == "featureID"] <- "FeatureID"
    names(rap_export)[names(rap_export) == "sampleID"]  <- "SampleID"
    names(rap_export)[names(rap_export) == "abundance"] <- "RelAbundance"
    
    readr::write_csv(rap_export, export_csv)
    message("Relative abundance exported to: ", export_csv)
  }
  
  feature_order <- data.frame(SummarizedExperiment::assays(RAP_sel)[["relabundance"]]) %>% 
    rowSums() %>% sort(decreasing = TRUE) %>% names()
  RAP_melt$featureID <- factor(RAP_melt$featureID, levels = feature_order)
  
  RAP_out_list <- list()
  RAP_mean <- data.frame(feature_order) %>% stats::setNames("featureID")
  
  if (length(var) == 1) {
    var_levels <- list(var_levels)
  }
  
  for (cov in rev(var)) {
    if (paste("sign", cov, sep = "|") %in% colnames(RAP_melt)) {
      
      RAP_melt_sel <- RAP_melt[RAP_melt[, paste("sign", cov, sep = "|")] == TRUE, ] %>% 
        dplyr::select(all_of(c("featureID", "sampleID", "abundance", cov))) %>% 
        stats::setNames(c("featureID", "sampleID", "abundance", "variable"))
      
      RAP_cov_list <- list()
      
      if (is.character(RAP_melt_sel$variable) | is.factor(RAP_melt_sel$variable)) {
        RAP_melt_cov <- RAP_melt_sel
        
        if (!is.null(var_levels[[which(var == cov)]])) {
          RAP_melt_cov$variable <- factor(RAP_melt_cov$variable, levels = var_levels[[which(var == cov)]])
        } else {
          RAP_melt_cov$variable <- factor(RAP_melt_cov$variable)
        }
        
        plot_palette <- c("#1F77B4FF", "#AEC7E8FF", "#FF7F0EFF", 
                          "#FFBB78FF", "#2CA02CFF", "#98DF8AFF", "#D62728FF", 
                          "#FF9896FF", "#9467BDFF", "#C5B0D5FF", "#8C564BFF", 
                          "#C49C94FF", "#E377C2FF", "#F7B6D2FF", "#7F7F7FFF", 
                          "#C7C7C7FF", "#BCBD22FF", "#DBDB8DFF", "#17BECFFF", 
                          "#9EDAE5FF")
        
        if (length(levels(RAP_melt_cov$variable)) <= 10) {
          plot_colors <- plot_palette[seq(1, 19, 2)][1:length(levels(RAP_melt_cov$variable))]
          color_scale <- ggplot2::scale_color_manual(values = plot_colors)
        } else if (length(levels(RAP_melt_cov$variable)) <= 20) {
          # NOTE: kept logic; uses levels(variable) instead of undefined set_levels
          plot_colors <- plot_palette[sort(c(seq(2, 20, 2)[1:(length(levels(RAP_melt_cov$variable)) - 10)], seq(1, 19, 2)))]
          color_scale <- ggplot2::scale_color_manual(values = plot_colors)
        } else {
          color_scale <- ggraph::scale_color_viridis(discrete = TRUE)
        }
        
      } else {
        RAP_melt_cov <- suppressMessages(
          RAP_melt_sel %>% dplyr::group_by(featureID, variable) %>% 
            dplyr::summarise(mean = mean(abundance), std = plotrix::std.error(abundance))
        )
        
        lm_list <- lapply(unique(RAP_melt_cov$featureID), function(x) {
          summary(stats::lm(as.data.frame(RAP_melt_cov[RAP_melt_cov$featureID == x, ])$mean ~ 
                              as.data.frame(RAP_melt_cov[RAP_melt_cov$featureID == x, ])$variable))
        })
        
        RAP_lm_stats <- data.frame(
          as.character(unique(RAP_melt_cov$featureID)),
          unlist(lapply(lm_list, function(x) paste("R2:", signif(x$adj.r.squared, digits = 2)))),
          unlist(lapply(lm_list, function(x) paste("Pr(>|t|):", signif(x$coefficients[2, 4], digits = 2))))
        ) %>% stats::setNames(c("featureID", "R2", "Pr(>|t|)"))
      }
      
      for (part in (ceiling(length(unique(RAP_melt_sel$featureID)) / prod(panel_dim))):1) {
        feature_range <- (prod(panel_dim) * part - prod(panel_dim) + 1):(prod(panel_dim) * part)
        feature_sel <- unique(RAP_melt_sel$featureID)[feature_range]
        RAP_melt_panel <- RAP_melt_cov[RAP_melt_cov$featureID %in% feature_sel, ]
        
        if (is.numeric(RAP_melt_sel$variable) | is.double(RAP_melt_sel$variable)) {
          RAP_lm_stats_sel <- RAP_lm_stats[RAP_lm_stats$featureID %in% feature_sel, ]
        }
        
        if (length(unique(RAP_melt_sel$featureID)) > prod(panel_dim)) {
          plot_title_panel <- paste(plot_title, ": ", cov, " (", part, ")", sep = "")
        } else {
          plot_title_panel <- paste(plot_title, ": ", cov, sep = "")
        }
        
        if (is.factor(RAP_melt_cov$variable)) {
          RAP_out_panel <- ggplot2::ggplot(RAP_melt_panel, ggplot2::aes(x = variable, y = abundance)) +
            ggplot2::geom_boxplot(lwd = 0.5, show.legend = FALSE, outlier.shape = NA) +
            ggplot2::geom_jitter(ggplot2::aes(color = variable), size = 0.65,
                                 position = ggplot2::position_jitterdodge(0.2),
                                 alpha = 0.75, show.legend = FALSE) +
            ggpubr::geom_pwc(label = "p.adj.signif", p.adjust.method = "BH", hide.ns = hide.ns) +
            color_scale
        } else {
          RAP_out_panel <- ggplot2::ggplot(RAP_melt_panel, ggplot2::aes(x = variable, y = mean)) +
            ggplot2::geom_point(size = 2, color = "#26456E") +
            ggplot2::geom_line(linetype = 2, color = "#26456E") +
            ggplot2::geom_errorbar(ggplot2::aes(ymin = mean - std, ymax = mean + std), width = 0.2, color = "grey") +
            ggplot2::geom_text(data = RAP_lm_stats_sel, ggplot2::aes(x = -Inf, y = Inf, label = R2),
                               size = 2.5, vjust = 1.2, hjust = -0.1) +
            ggplot2::geom_text(data = RAP_lm_stats_sel, ggplot2::aes(x = -Inf, y = Inf, label = `Pr(>|t|)`),
                               size = 2.5, vjust = 2.4, hjust = -0.07)
        }
        
        RAP_out_panel <- RAP_out_panel +
          ggplot2::labs(x = "", y = "Relative abundance", title = plot_title_panel, fill = cov) +
          ggplot2::theme_bw() +
          ggplot2::theme(plot.title = ggplot2::element_text(size = 20, face = "plain"),
                         axis.title.y = ggplot2::element_text(size = 15, face = "plain"),
                         strip.text.x = ggplot2::element_text(size = 9, face = "italic"),
                         panel.grid = ggplot2::element_blank()) +
          ggplot2::facet_wrap(~featureID, scales = "free", ncol = panel_dim[2])
        
        RAP_cov_list <- append(RAP_cov_list, list(RAP_out_panel))
        names(RAP_cov_list)[length(RAP_cov_list)] <- paste(cov, part, sep = "_")
      }
      
      print(RAP_cov_list)
      
      for (feat in unique(RAP_melt_sel$featureID)) {
        RAP_melt_feat <- RAP_melt_cov[RAP_melt_cov$featureID == feat, ]
        
        if (is.numeric(RAP_melt_sel$variable) | is.double(RAP_melt_sel$variable)) {
          RAP_lm_stats_feat <- RAP_lm_stats[RAP_lm_stats$featureID == feat, ]
        }
        
        if (is.factor(RAP_melt_cov$variable)) {
          RAP_out_feat <- ggplot2::ggplot(RAP_melt_feat, ggplot2::aes(x = variable, y = abundance)) +
            ggplot2::geom_boxplot(lwd = 0.65, show.legend = FALSE, outlier.shape = NA) +
            ggplot2::geom_jitter(ggplot2::aes(color = variable),
                                 position = ggplot2::position_jitterdodge(0.2),
                                 alpha = 0.75, show.legend = FALSE) +
            ggpubr::geom_pwc(label = "p.adj.signif", p.adjust.method = "BH",
                             hide.ns = hide.ns, label.size = 2.8) +
            color_scale
        } else {
          RAP_out_feat <- ggplot2::ggplot(RAP_melt_feat, ggplot2::aes(x = variable, y = mean)) +
            ggplot2::geom_point(size = 2, color = "#26456E") +
            ggplot2::geom_line(linetype = 2, color = "#26456E") +
            ggplot2::geom_errorbar(ggplot2::aes(ymin = mean - std, ymax = mean + std), width = 0.2, color = "grey") +
            ggplot2::geom_text(data = RAP_lm_stats_feat, ggplot2::aes(x = -Inf, y = Inf, label = R2),
                               size = 4, vjust = 1.2, hjust = -0.1) +
            ggplot2::geom_text(data = RAP_lm_stats_feat, ggplot2::aes(x = -Inf, y = Inf, label = `Pr(>|t|)`),
                               size = 4, vjust = 2.4, hjust = -0.07)
        }
        
        RAP_out_feat <- RAP_out_feat +
          ggplot2::labs(x = "", y = "", fill = cov, title = feat) +
          ggplot2::theme_bw() +
          ggplot2::theme(plot.title = ggplot2::element_text(size = 13, face = "italic"),
                         axis.title.y = ggplot2::element_text(size = 12, face = "plain"),
                         strip.text.x = ggplot2::element_text(size = 9, face = "italic"),
                         panel.grid = ggplot2::element_blank())
        
        RAP_cov_list$features <- append(RAP_cov_list$features, list(RAP_out_feat))
        names(RAP_cov_list$features)[length(RAP_cov_list$features)] <- feat
      }
      
      RAP_out_list <- append(RAP_out_list, list(RAP_cov_list))
      names(RAP_out_list)[length(RAP_out_list)] <- cov
      
      RAP_mean_add <- stats::aggregate(stats::as.formula(paste("abundance ~ featureID +", cov)), RAP_melt, mean) %>% 
        tidyr::pivot_wider(names_from = dplyr::all_of(cov), values_from = abundance)
      colnames(RAP_mean_add) <- c("featureID", paste(cov, colnames(RAP_mean_add)[-1], sep = ""))
      RAP_mean <- dplyr::right_join(RAP_mean, RAP_mean_add, by = "featureID")
      
    } else {
      warning(paste("No significant comparisons were identified in the DA_analysis object for selected conditions for variable:", cov))
    }
  }
  
  RAP_out_list <- append(RAP_out_list, list(RAP_mean))
  names(RAP_out_list)[length(RAP_out_list)] <- "mean_relabundance"
  
  return(invisible(RAP_out_list))
}
```

## Function to export p- values and p-adjusted values from RA

```{r Function to export p values and p adj values from RA, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
RA_padj_from_RA <- function(ra_long,
                            feature_col  = "FeatureID",
                            value_col    = "RelAbundance",
                            group_col    = "Time",
                            group_levels = c("T0","T12"),
                            paired       = FALSE,
                            id_col       = "PatientID",
                            padj_method  = "BH") {

  df <- ra_long %>%
    dplyr::mutate(
      .feature = .data[[feature_col]],
      .value   = .data[[value_col]],
      .group   = factor(.data[[group_col]], levels = group_levels)
    ) %>%
    dplyr::filter(!is.na(.group), !is.na(.value))

  if (paired) {
    if (!(id_col %in% colnames(df))) {
      stop(sprintf("paired=TRUE: the column '%s' does not exist in ra_long", id_col))
    }

    df <- df %>%
      dplyr::group_by(.feature, .data[[id_col]], .group) %>%
      dplyr::summarise(.value = stats::median(.value, na.rm = TRUE), .groups = "drop") %>%
      dplyr::group_by(.feature, .data[[id_col]]) %>%
      dplyr::filter(all(group_levels %in% .group)) %>%
      dplyr::ungroup()

    out <- df %>%
      tidyr::pivot_wider(names_from = .group, values_from = .value) %>%
      dplyr::group_by(.feature) %>%
      dplyr::summarise(
        n_pairs = sum(!is.na(.data[[group_levels[1]]]) & !is.na(.data[[group_levels[2]]])),
        n_T0  = n_pairs,
        n_T12 = n_pairs,
        p = tryCatch(
          stats::wilcox.test(
            .data[[group_levels[1]]],
            .data[[group_levels[2]]],
            paired = TRUE,
            exact = FALSE
          )$p.value,
          error = function(e) NA_real_
        ),
        .groups = "drop"
      )

  } else {

    out <- df %>%
      dplyr::group_by(.feature) %>%
      dplyr::summarise(
        n_T0  = sum(.group == group_levels[1]),
        n_T12 = sum(.group == group_levels[2]),
        p = tryCatch(
          stats::wilcox.test(.value ~ .group, paired = FALSE, exact = FALSE)$p.value,
          error = function(e) NA_real_
        ),
        .groups = "drop"
      )
  }

  out %>%
    dplyr::mutate(
      padj = stats::p.adjust(p, method = padj_method),
      padj_signif = dplyr::case_when(
        is.na(padj)      ~ NA_character_,
        padj <= 0.0001   ~ "****",
        padj <= 0.001    ~ "***",
        padj <= 0.01     ~ "**",
        padj <= 0.05     ~ "*",
        TRUE             ~ "ns"
      )
    ) %>%
    dplyr::rename(FeatureID = .feature)
}
```

## Function: percent zeros (prevalence of zeros) by Feature and Time

```{r Zeros percent, echo=TRUE, fig.show='hide', message=FALSE, warning=FALSE}
# ---- function: percent zeros (prevalence of zeros) by Feature and Time ----
zeros_by_feature_time <- function(ra_long,
                                  feature_col  = "FeatureID",
                                  value_col    = "RelAbundance",
                                  group_col    = "Time",
                                  group_levels = c("T0","T12"),
                                  zero_eps     = 0) {

  ra_long %>%
    mutate(
      .feature = .data[[feature_col]],
      .value   = .data[[value_col]],
      .group   = factor(.data[[group_col]], levels = group_levels)
    ) %>%
    filter(!is.na(.feature), !is.na(.group), !is.na(.value)) %>%
    group_by(.feature, .group) %>%
    summarise(
      n = n(),
      n_zero = sum(.value <= zero_eps),
      pct_zero = 100 * n_zero / n,
      .groups = "drop"
    ) %>%
    pivot_wider(
      names_from  = .group,
      values_from = c(n, n_zero, pct_zero),
      names_sep   = "_"
    ) %>%
    rename(FeatureID = .feature)
}
```

#----------------------
## SAMPLETYPE AND TIME
#----------------------

### Feces at T0 and T12

```{r DAA ANCOMBC2 of Feces at T0 and T12 min_prevalence 0.10, results='hide', echo=TRUE, warning=FALSE, message=FALSE}
DAA_Metaphlan_cf_fc_abc <- DA_analysis(tse = subset_feces_no_UN,
                               assay_type = 'counts',
                               var = 'Time',
                               main_var = NULL,
                               var_levels = c('T0','T12'),
                               pairwise = T,
                               level = 'Species',
                               min_prevalence = 0.10,
                               min_abundance = 0,
                               padj_method = 'BH',
                               padj_m = 'default',
                               rand_formula = NULL,
                               tools = 'ANCOM_BC2',
                               sign_cut = 0.05,
                               effect_cut = 0,
                               n_cores = 1,
                               verbose = T,
                               struc_zero = F,
                               neg_lb = F,
                               pseudo_sens = T,
                               s0_perc = 0.05)
```

#### Print significant results

```{r Print significant features of DAA_Metaphlan_cf_fc}
print(DAA_Metaphlan_cf_fc_abc$sign_features)
```

#### Save significant results

```{r Save results feces T0 and T12}
write.csv(DAA_Metaphlan_cf_fc_abc$sign_features$ANCOM_BC2,
          file = here("results", "DAA_ANCOMBC2_sign_species_feces.csv"),
          row.names = FALSE)
```

#### Plot significant results

```{r Plot significat results feces}
sign_df_fc <- DAA_Metaphlan_cf_fc_abc$sign_features$ANCOM_BC2

DAA_Metaphlan_cf_fc_abc_bar <- ggplot(sign_df_fc, 
    aes(x = reorder(featureID, effect), y = effect, fill = effect)) +
  geom_bar(stat = "identity", color = "black", size = 0.2) +
  coord_flip() +
  scale_fill_gradient2(low = "#F4E1A1", mid = "white", high = "#A1E1E1", midpoint = 0) +
  labs(title = "Differentially abundant species at T12 vs T0",
       x = "",
       y = "Effect size (log fold change)") +
  theme_minimal()
ggsave(filename = here("results", "DAA_Metaphlan_cf_fc_abc_bar.svg"), plot = DAA_Metaphlan_cf_fc_abc_bar, width = 10, height = 7, dpi = 300, device = "svg")
```

##### Relative abundance of DAA of Feces at T0 and T12
###### Export CSV

```{r RA CSV of DAA of Feces at T0 and T12, echo=TRUE, fig.show='hide', warning=FALSE, message=FALSE}
RAP_out_fc <- RA_plot_csv(
  features = DAA_Metaphlan_cf_fc_abc,
  tse = subset_feces_no_UN,
  var = c("Time"),
  var_levels = c("T0","T12"),
  level = "Species",
  plot_title = "Relative Abundance",
  export_csv = here::here("results", "RA_DAA_plot_relabundance_long_fc.csv")
)
```

###### Run function to export p-values and p-adjusted values from RA PAIRED

```{r Run function to export pval and p adj values from RA PAIRED, echo=TRUE, warning=FALSE, message=FALSE}
ra_df <- read_csv(here::here("results","RA_DAA_plot_relabundance_long_fc.csv"), show_col_types = FALSE) %>%
  mutate(
    PatientID = str_extract(SampleID, "^[A-Za-z]+\\d+")
  )

ra_padj <- RA_padj_from_RA(
  ra_long = ra_df,
  paired = TRUE
)

write_csv(ra_padj, here::here("results","RA_DAA_padj_from_RA_Time_feces.csv"))
```

###### Plot

```{r Plot of RA of DAA of Feces at T0 and T12, echo=TRUE, fig.show='hide', warning=FALSE, message=FALSE}
# Load ANCOM-BC2 significant features
ancom_tab <- read_csv(here::here("results", "DAA_ANCOMBC2_sign_species_feces.csv"),
                      show_col_types = FALSE) %>%
  rename(FeatureID = featureID)

sig_feats <- ancom_tab %>%
  filter(padj < 0.05) %>%
  pull(FeatureID) %>%
  unique()

# Load RA long CSV
ra_df <- read_csv(here::here("results", "RA_DAA_plot_relabundance_long_fc.csv"),
                  show_col_types = FALSE) %>%
  mutate(
    PatientID = str_extract(SampleID, "^[A-Za-z]+\\d+"),
    Time = factor(Time, levels = c("T0", "T12"))
  )

# Keep only sig features + paired patients (T0 and T12)
ra_sig <- ra_df %>%
  filter(FeatureID %in% sig_feats) %>%
  group_by(FeatureID, PatientID) %>%
  filter(n_distinct(Time) == 2) %>%
  ungroup()

# ---- Compute Wilcoxon paired p-values on Relative Abundance (for labels) ----
ra_pvals <- ra_sig %>%
  group_by(FeatureID, PatientID, Time) %>%
  summarise(RelAbundance = median(RelAbundance, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Time, values_from = RelAbundance) %>%
  group_by(FeatureID) %>%
  summarise(
    p_RA = tryCatch(
      wilcox.test(T0, T12, paired = TRUE, exact = FALSE)$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  )

# Colors
col_box   <- c("T0" = "#F4E1A1", "T12" = "#A1E1E1")
col_point <- c("T0" = "#FAF3DC", "T12" = "#D6F5F5")

# Paired plot for ONE feature
plot_ra_feature <- function(feat) {

  df <- ra_sig %>%
    filter(FeatureID == feat) %>%
    mutate(
      fill_box   = col_box[Time],
      fill_point = col_point[Time],
      x          = as.numeric(Time)
    )

  # ---- Label using RA Wilcoxon paired p-value ----
  p_val <- ra_pvals %>%
    filter(FeatureID == feat) %>%
    slice(1) %>%
    pull(p_RA)

  lab <- ifelse(is.na(p_val), "p = NA",
                ifelse(p_val < 0.001, "p < 0.001",
                       paste0("p = ", signif(p_val, 3))))

  y_max <- max(df$RelAbundance, na.rm = TRUE)
  if (!is.finite(y_max) || y_max == 0) y_max <- 1e-6

  y_bracket <- y_max * 1.15
  y_text    <- y_max * 1.22
  y_lim     <- y_max * 1.30

  p_values <- data.frame(x = 1, xend = 2, y = y_bracket, label = lab)

  ggplot() +
    stat_slab(
      data = df %>% filter(Time == "T0"),
      aes(x = x, y = RelAbundance, fill = fill_box),
      side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    stat_slab(
      data = df %>% filter(Time == "T12"),
      aes(x = x, y = RelAbundance, fill = fill_box),
      side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    geom_boxplot(
      data = df,
      aes(x = x, y = RelAbundance, fill = fill_box),
      width = 0.5, outlier.shape = NA,
      color = "black", alpha = 0.9, show.legend = TRUE
    ) +
    geom_point(
      data = df,
      aes(x = x, y = RelAbundance, fill = fill_point),
      shape = 21, size = 3, stroke = 0.3, color = "black",
      show.legend = FALSE
    ) +
    geom_line(
      data = df,
      aes(x = x, y = RelAbundance, group = PatientID),
      color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
    ) +
    geom_segment(
      data = p_values,
      aes(x = x, xend = xend, y = y, yend = y),
      color = "black", linewidth = 0.5
    ) +
    geom_text(
      data = p_values,
      aes(x = 1.5, y = y_text, label = label),
      size = 4
    ) +
    scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
    scale_fill_identity(
      guide = "legend", name = "Time",
      breaks = col_box, labels = names(col_box)
    ) +
    coord_cartesian(ylim = c(0, y_lim)) +
    labs(x = NULL, y = "Relative abundance", title = feat) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "right",
      axis.text.x = element_text(size = 14)
    )
}

# Save ONLY the ANCOM-BC2 significant features into results/
for (feat in sig_feats) {
  p <- plot_ra_feature(feat)

  out_file <- here::here(
    "results",
    paste0("RA_DAA_fc_", gsub("[^A-Za-z0-9]+", "_", feat), ".svg")
  )

  ggsave(out_file, plot = p, width = 8, height = 6, dpi = 300, device = "svg")
  message("Saved: ", out_file)
}
```

####### Panel plot

```{r Panel plot of RA of DAA of Feces at T0 and T12, echo=TRUE, fig.show='show', warning=FALSE, message=FALSE}
# Create all plots first
plot_list <- lapply(sig_feats, plot_ra_feature)
names(plot_list) <- sig_feats

# Combine into one panel
panel_plot_fc <- wrap_plots(
  plotlist = plot_list,
  ncol = 4,
  guides = "collect"
) & theme(
  legend.position = "right"
)

# Save panel
n_rows_panel <- ceiling(length(plot_list) / 3)

ggsave(filename = here::here("results", "RA_DAA_panel_fc.svg"),
  plot = panel_plot_fc, width = 14, height = 2.5 * n_rows_panel, dpi = 300,
  device = "svg")
```

###### Prevalence of zeros (%) by taxon and by time

```{r Prevalence of zeros, echo=TRUE, fig.show='show', warning=FALSE, message=FALSE}
# ---- load your long RA table ----
ra_df <- read_csv(
  here::here("results", "RA_DAA_plot_relabundance_long_fc.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    PatientID = str_extract(SampleID, "^[A-Za-z]+\\d+"),
    Time = factor(Time, levels = c("T0","T12"))
  )

group_levels <- c("T0","T12")
zero_eps <- 0

# ---- 1) Prevalence of zeros by feature and time ----
zero_df <- ra_df %>%
  filter(!is.na(FeatureID), !is.na(Time), !is.na(RelAbundance)) %>%
  group_by(FeatureID, Time) %>%
  summarise(
    n = n(),
    n_zero = sum(RelAbundance <= zero_eps),
    pct_zero = 100 * n_zero / n,
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from  = Time,
    values_from = c(n, n_zero, pct_zero),
    names_sep   = "_"
  )

# ---- 2) Paired wilcoxon T0 vs T12 by feature (p-value) ----
p_df <- ra_df %>%
  filter(!is.na(FeatureID), !is.na(Time), !is.na(RelAbundance), !is.na(PatientID)) %>%
  group_by(FeatureID, PatientID, Time) %>%
  summarise(RelAbundance = median(RelAbundance, na.rm = TRUE), .groups = "drop") %>%
  group_by(FeatureID, PatientID) %>%
  filter(all(group_levels %in% Time)) %>%
  ungroup() %>%
  pivot_wider(names_from = Time, values_from = RelAbundance) %>%
  group_by(FeatureID) %>%
  summarise(
    n_pairs = sum(!is.na(.data[[group_levels[1]]]) & !is.na(.data[[group_levels[2]]])),
    p = tryCatch(
      wilcox.test(
        .data[[group_levels[1]]],
        .data[[group_levels[2]]],
        paired = TRUE,
        exact  = FALSE
      )$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) %>%
  mutate(
    padj = p.adjust(p, method = "BH"),
    p_signif = case_when(
      is.na(p) ~ NA_character_,
      p <= 0.0001 ~ "****",
      p <= 0.001  ~ "***",
      p <= 0.01   ~ "**",
      p <= 0.05   ~ "*",
      TRUE ~ "ns"
    ),
    padj_signif = case_when(
      is.na(padj) ~ NA_character_,
      padj <= 0.0001 ~ "****",
      padj <= 0.001  ~ "***",
      padj <= 0.01   ~ "**",
      padj <= 0.05   ~ "*",
      TRUE ~ "ns"
    )
  )

# ---- 3) Merge ----
ra_zero_p <- p_df %>%
  left_join(zero_df, by = "FeatureID") %>%
  arrange(p)

# ---- Export ----
write_csv(ra_zero_p, here::here("results", "RA_DAA_pvalues_and_percent_zeros_by_Time_feces.csv"))
```


### Sputum at T0 and T12

```{r DAA ANCOMBC2 of Sputum at T0 and T12 min_prevalence 0.10, results='hide', echo=TRUE, warning=FALSE, message=FALSE}
DAA_Metaphlan_cf_sp_abc <- DA_analysis(tse = subset_sputum_no_UN,
                               assay_type = 'counts',
                               var = 'Time',
                               main_var = NULL,
                               var_levels = c('T0','T12'),
                               pairwise = T,
                               level = 'Species',
                               min_prevalence = 0.10,
                               min_abundance = 0,
                               padj_method = 'BH',
                               padj_m = 'default',
                               rand_formula = NULL,
                               tools = 'ANCOM_BC2',
                               sign_cut = 0.05,
                               effect_cut = 0,
                               n_cores = 1,
                               verbose = T,
                               struc_zero = F,
                               neg_lb = F,
                               pseudo_sens = T,
                               s0_perc = 0.05)
```

#### Print significant results

```{r Print significant features of DAA_Metaphlan_cf_sp}
print(DAA_Metaphlan_cf_sp_abc$sign_features)
```

#### Save significant results

```{r Save results sputum T0 and T12}
write.csv(DAA_Metaphlan_cf_sp_abc$sign_features$ANCOM_BC2,
          file = here("results", "DAA_ANCOMBC2_sign_species_sputum.csv"),
          row.names = FALSE)
```

#### Plot significant results

```{r Plot significat results sputum T0 T12}
sign_df_sp <- DAA_Metaphlan_cf_sp_abc$sign_features$ANCOM_BC2

DAA_Metaphlan_cf_sp_abc_bar <- ggplot(sign_df_sp, 
    aes(x = reorder(featureID, effect), y = effect, fill = effect)) +
  geom_bar(stat = "identity", color = "black", size = 0.2) +
  coord_flip() +
  scale_fill_gradient2(low = "#D9C2E9", mid = "white", high = "#B3E1B6", midpoint = 0) +
  labs(title = "Differentially abundant species at T12 vs T0 in sputum",
       x = "",
       y = "Effect size (log fold change)") +
  theme_minimal()
ggsave(filename = here("results", "DAA_Metaphlan_cf_sp_abc_bar.svg"), plot = DAA_Metaphlan_cf_sp_abc_bar, width = 10, height = 7, dpi = 300, device = "svg")
```

##### Relative abundance of DAA of Sputum at T0 and T12
###### Export CSV

```{r RA CSV of DAA of Sputum at T0 and T12, echo=TRUE, fig.show='hide', warning=FALSE, message=FALSE}
RAP_out_sp <- RA_plot_csv(
  features = DAA_Metaphlan_cf_sp_abc,
  tse = subset_sputum_no_UN,
  var = c("Time"),
  var_levels = c("T0","T12"),
  level = "Species",
  plot_title = "Relative Abundance",
  export_csv = here::here("results", "RA_DAA_plot_relabundance_long_sp.csv")
)
```

###### Run function to export p-values and p-adjusted values from RA PAIRED

```{r Run function to export pval and p adj values from RA PAIRED, echo=TRUE, warning=FALSE, message=FALSE}
ra_df <- read_csv(here::here("results","RA_DAA_plot_relabundance_long_sp.csv"), show_col_types = FALSE) %>%
  mutate(
    PatientID = str_extract(SampleID, "^[A-Za-z]+\\d+")
  )

ra_padj <- RA_padj_from_RA(
  ra_long = ra_df,
  paired = TRUE
)

write_csv(ra_padj, here::here("results","RA_DAA_padj_from_RA_Time_sputum.csv"))
```

###### Plot

```{r Plot of RA of DAA of Feces at T0 and T12, echo=TRUE, fig.show='hide', warning=FALSE, message=FALSE}
# Load ANCOM-BC2 significant features
ancom_tab <- read_csv(here::here("results", "DAA_ANCOMBC2_sign_species_sputum.csv"),
                      show_col_types = FALSE) %>%
  rename(FeatureID = featureID)

sig_feats <- ancom_tab %>%
  filter(padj < 0.05) %>%
  pull(FeatureID) %>%
  unique()

# Load RA long CSV
ra_df <- read_csv(here::here("results", "RA_DAA_plot_relabundance_long_sp.csv"),
                  show_col_types = FALSE) %>%
  mutate(
    PatientID = str_extract(SampleID, "^[A-Za-z]+\\d+"),
    Time = factor(Time, levels = c("T0", "T12"))
  )

# Keep only sig features + paired patients (T0 and T12)
ra_sig <- ra_df %>%
  filter(FeatureID %in% sig_feats) %>%
  group_by(FeatureID, PatientID) %>%
  filter(n_distinct(Time) == 2) %>%
  ungroup()

# ---- Compute Wilcoxon paired p-values on Relative Abundance (for labels) ----
ra_pvals <- ra_sig %>%
  group_by(FeatureID, PatientID, Time) %>%
  summarise(RelAbundance = median(RelAbundance, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Time, values_from = RelAbundance) %>%
  group_by(FeatureID) %>%
  summarise(
    p_RA = tryCatch(
      wilcox.test(T0, T12, paired = TRUE, exact = FALSE)$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  )

# Colors
col_box   <- c("T0" = "#D9C2E9", "T12" = "#B3E1B6")
col_point <- c("T0" = "#EDE3F4", "T12" = "#E2F5E3")

# Paired plot for ONE feature
plot_ra_feature <- function(feat) {

  df <- ra_sig %>%
    filter(FeatureID == feat) %>%
    mutate(
      fill_box   = col_box[Time],
      fill_point = col_point[Time],
      x          = as.numeric(Time)
    )

  # ---- Label using RA Wilcoxon paired p-value ----
  p_val <- ra_pvals %>%
    filter(FeatureID == feat) %>%
    slice(1) %>%
    pull(p_RA)

  lab <- ifelse(is.na(p_val), "p = NA",
                ifelse(p_val < 0.001, "p < 0.001",
                       paste0("p = ", signif(p_val, 3))))

  y_max <- max(df$RelAbundance, na.rm = TRUE)
  if (!is.finite(y_max) || y_max == 0) y_max <- 1e-6

  y_bracket <- y_max * 1.15
  y_text    <- y_max * 1.22
  y_lim     <- y_max * 1.30

  p_values <- data.frame(x = 1, xend = 2, y = y_bracket, label = lab)

  ggplot() +
    stat_slab(
      data = df %>% filter(Time == "T0"),
      aes(x = x, y = RelAbundance, fill = fill_box),
      side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    stat_slab(
      data = df %>% filter(Time == "T12"),
      aes(x = x, y = RelAbundance, fill = fill_box),
      side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    geom_boxplot(
      data = df,
      aes(x = x, y = RelAbundance, fill = fill_box),
      width = 0.5, outlier.shape = NA,
      color = "black", alpha = 0.9, show.legend = TRUE
    ) +
    geom_point(
      data = df,
      aes(x = x, y = RelAbundance, fill = fill_point),
      shape = 21, size = 3, stroke = 0.3, color = "black",
      show.legend = FALSE
    ) +
    geom_line(
      data = df,
      aes(x = x, y = RelAbundance, group = PatientID),
      color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
    ) +
    geom_segment(
      data = p_values,
      aes(x = x, xend = xend, y = y, yend = y),
      color = "black", linewidth = 0.5
    ) +
    geom_text(
      data = p_values,
      aes(x = 1.5, y = y_text, label = label),
      size = 4
    ) +
    scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
    scale_fill_identity(
      guide = "legend", name = "Time",
      breaks = col_box, labels = names(col_box)
    ) +
    coord_cartesian(ylim = c(0, y_lim)) +
    labs(x = NULL, y = "Relative abundance", title = feat) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "right",
      axis.text.x = element_text(size = 14)
    )
}

# Save ONLY the ANCOM-BC2 significant features into results/
for (feat in sig_feats) {
  p <- plot_ra_feature(feat)

  out_file <- here::here(
    "results",
    paste0("RA_DAA_sp_", gsub("[^A-Za-z0-9]+", "_", feat), ".svg")
  )

  ggsave(out_file, plot = p, width = 8, height = 6, dpi = 300, device = "svg")
  message("Saved: ", out_file)
}
```

###### Panel plot

```{r Panel plot of RA of DAA of Sputum at T0 and T12, echo=TRUE, fig.show='show', warning=FALSE, message=FALSE}
# Create all plots first
plot_list <- lapply(sig_feats, plot_ra_feature)
names(plot_list) <- sig_feats

# Combine into one panel
panel_plot_sp <- wrap_plots(
  plotlist = plot_list,
  ncol = 2,
  guides = "collect"
) & theme(
  legend.position = "right"
)

# Save panel
n_rows_panel <- ceiling(length(plot_list) / 3)

ggsave(filename = here::here("results", "RA_DAA_panel_sp.svg"),
  plot = panel_plot_sp, width = 8, height = 3 * n_rows_panel, dpi = 300,
  device = "svg")
```

###### Prevalence of zeros (%) by taxon and by time

```{r Prevalence of zeros, echo=TRUE, fig.show='show', warning=FALSE, message=FALSE}
# ---- load your long RA table ----
ra_df <- read_csv(
  here::here("results", "RA_DAA_plot_relabundance_long_sp.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    PatientID = str_extract(SampleID, "^[A-Za-z]+\\d+"),
    Time = factor(Time, levels = c("T0","T12"))
  )

group_levels <- c("T0","T12")
zero_eps <- 0

# ---- 1) Prevalence of zeros by feature and time ----
zero_df <- ra_df %>%
  filter(!is.na(FeatureID), !is.na(Time), !is.na(RelAbundance)) %>%
  group_by(FeatureID, Time) %>%
  summarise(
    n = n(),
    n_zero = sum(RelAbundance <= zero_eps),
    pct_zero = 100 * n_zero / n,
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from  = Time,
    values_from = c(n, n_zero, pct_zero),
    names_sep   = "_"
  )

# ---- 2) Paired wilcoxon T0 vs T12 by feature (p-value) ----
p_df <- ra_df %>%
  filter(!is.na(FeatureID), !is.na(Time), !is.na(RelAbundance), !is.na(PatientID)) %>%
  group_by(FeatureID, PatientID, Time) %>%
  summarise(RelAbundance = median(RelAbundance, na.rm = TRUE), .groups = "drop") %>%
  group_by(FeatureID, PatientID) %>%
  filter(all(group_levels %in% Time)) %>%
  ungroup() %>%
  pivot_wider(names_from = Time, values_from = RelAbundance) %>%
  group_by(FeatureID) %>%
  summarise(
    n_pairs = sum(!is.na(.data[[group_levels[1]]]) & !is.na(.data[[group_levels[2]]])),
    p = tryCatch(
      wilcox.test(
        .data[[group_levels[1]]],
        .data[[group_levels[2]]],
        paired = TRUE,
        exact  = FALSE
      )$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) %>%
  mutate(
    padj = p.adjust(p, method = "BH"),
    p_signif = case_when(
      is.na(p) ~ NA_character_,
      p <= 0.0001 ~ "****",
      p <= 0.001  ~ "***",
      p <= 0.01   ~ "**",
      p <= 0.05   ~ "*",
      TRUE ~ "ns"
    ),
    padj_signif = case_when(
      is.na(padj) ~ NA_character_,
      padj <= 0.0001 ~ "****",
      padj <= 0.001  ~ "***",
      padj <= 0.01   ~ "**",
      padj <= 0.05   ~ "*",
      TRUE ~ "ns"
    )
  )

# ---- 3) Merge ----
ra_zero_p <- p_df %>%
  left_join(zero_df, by = "FeatureID") %>%
  arrange(p)

# ---- Export ----
write_csv(ra_zero_p, here::here("results", "RA_DAA_pvalues_and_percent_zeros_by_Time_sputum.csv"))
```

