---
title: "cf_SCFA_thesis_chapter_I"
author: "Natalia Bastón-Paz"
date: '2026'
output: html_document
---

# Set up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require("knitr")
opts_knit$set(root.dir = "C:/RProjects/cf_SCFA_thesis/cf_SCFA_thesis_chapter_I")
```

# Clean environments
```{# r Clean environments, message=FALSE, warning=FALSE}
rm(list = ls())
```

## General introduction

Cystic fibrosis (CF) is a genetic disorder caused by pathogenic variants in the CFTR (Cystic Fibrosis Transmembrane Conductance Regulator) gene. CFTR dysfunction leads to impaired ion transport and the production of thick mucus, primarily affecting the lungs and gastrointestinal tract. People with CF (pwCF) commonly develop chronic airway infections and gastrointestinal complications.

In recent years, CFTR modulator therapies, particularly the triple combination elexacaftor/tezacaftor/ivacaftor (ETI), have markedly improved clinical outcomes, especially respiratory function and nutritional status. Short-chain fatty acids (SCFAs) are produced through the fermentation of dietary fiber by colonic bacteria. Among them, butyrate has been associated with improved gut function. SCFAs can be quantified using gas chromatography–mass spectrometry (GC–MS).

In this script, we analyze SCFA concentrations in 35 people with CF aged 6–66 years, assessed before ETI treatment (T0) and after 12 months of treatment (T12).

*Main question:*

-   *Do fecal and sputum SCFA concentrations change after one year of ETI treatment?*

## Libraries
```{r Libraries, message=FALSE, warning=FALSE}
library(here)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggdist)
library(purrr)
library(stringr)
library(patchwork)
library(ggpubr)
library(ggridges)
library(ppcor)
library(lme4)
library(lmerTest)
library(grid)
library(stringr)
library(forcats)
library(scales)
library(gtools)
library(emmeans)
library(tibble)
library(broom.mixed)
library(ggbreak)
library(cowplot)
library(lme4)
library(lmerTest)
```

## Directories
```{r Create directory, message=FALSE, warning=FALSE}
dir.create(here("results"), showWarnings = FALSE)
```


# ------------------------------------------------------------------------------
# SCFA FECES
# ------------------------------------------------------------------------------

### Load in the data
```{r Load in data SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_scfa_fc <- read_delim(here("data", "SCFA_feces.csv"))
```

#### Descriptive stats
```{r Descriptive stats SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
range(df_scfa_fc$acetate, na.rm = TRUE)
min_val <- min(df_scfa_fc$acetate, na.rm = TRUE)
max_val <- max(df_scfa_fc$acetate, na.rm = TRUE)

range(df_scfa_fc$propionate, na.rm = TRUE)
min_val <- min(df_scfa_fc$propionate, na.rm = TRUE)
max_val <- max(df_scfa_fc$propionate, na.rm = TRUE)

range(df_scfa_fc$butyrate, na.rm = TRUE)
min_val <- min(df_scfa_fc$butyrate, na.rm = TRUE)
max_val <- max(df_scfa_fc$butyrate, na.rm = TRUE)
```

### Filter data by time
```{r Filter data by time SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
data_scfa_fc_T0 <- subset(df_scfa_fc, Time == "T0")
data_scfa_fc_T12 <- subset(df_scfa_fc, Time == "T12")
```

### Function for statistic test
```{r Function for statistic test based on shapiro results SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
auto_paired_test_scfa <- function(df,
                                   id_col   = "PatientID",
                                   time_col = "Time",
                                   value_col = "acetate",
                                   levels   = c("T0","T12")) {
  stopifnot(all(c(id_col, time_col, value_col) %in% names(df)))
  
  dfw <- df[, c(id_col, time_col, value_col)]
  dfw <- tidyr::pivot_wider(
    dfw,
    names_from  = !!rlang::sym(time_col),
    values_from = !!rlang::sym(value_col)
  )
  if (!all(levels %in% names(dfw))) {
    stop("No columns found ", paste(levels, collapse = " and "), " after pivot_wider().")
  }
  dfw <- dfw[stats::complete.cases(dfw[, levels]), ]
  
  x0   <- dfw[[levels[1]]]
  x12  <- dfw[[levels[2]]]
  diff <- x12 - x0
  n    <- length(diff)
  if (n < 3) stop("Very few valid pairs (n < 3).")
  
  sh     <- shapiro.test(diff)
  normal <- sh$p.value >= 0.05
  
  if (normal) {
    tt        <- t.test(x12, x0, paired = TRUE)
    mean_diff <- mean(diff, na.rm = TRUE)
    sd_diff   <- stats::sd(diff, na.rm = TRUE)
    d_paired  <- mean_diff / sd_diff
    
    mean_T0  <- mean(x0,  na.rm = TRUE); sd_T0  <- stats::sd(x0,  na.rm = TRUE)
    mean_T12 <- mean(x12, na.rm = TRUE); sd_T12 <- stats::sd(x12, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Paired t-test\n")
    cat("n =", n, "| t =", round(tt$statistic, 4), "| p =", signif(tt$p.value, 4), "\n")
    cat("T0:   mean ± SD = ", round(mean_T0, 3), " ± ", round(sd_T0, 3), "\n", sep = "")
    cat("T12:  mean ± SD = ", round(mean_T12, 3), " ± ", round(sd_T12, 3), "\n", sep = "")
    cat("Mean difference (T12−T0) =", round(mean_diff, 3),
        " | 95% CI [", round(tt$conf.int[1], 3), ", ", round(tt$conf.int[2], 3), "]\n", sep = "")
    cat("Effect size (Cohen's d paired) =", round(d_paired, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "paired_t",
      n = n,
      shapiro = sh,
      test = tt,
      group_summary = list(
        T0  = c(mean = mean_T0, sd = sd_T0),
        T12 = c(mean = mean_T12, sd = sd_T12)
      ),
      mean_diff = mean_diff,
      ci = tt$conf.int,
      effect_size = list(type = "cohen_d_paired", value = d_paired)
    ))
    
  } else {
    wt   <- wilcox.test(x12, x0, paired = TRUE, exact = FALSE, correct = FALSE)
    V    <- as.numeric(wt$statistic)
    r_rb <- 1 - (2 * V) / (n * (n + 1))
    
    med_T0   <- stats::median(x0,  na.rm = TRUE)
    q1_T0    <- stats::quantile(x0,  0.25, na.rm = TRUE)
    q3_T0    <- stats::quantile(x0,  0.75, na.rm = TRUE)
    med_T12  <- stats::median(x12, na.rm = TRUE)
    q1_T12   <- stats::quantile(x12, 0.25, na.rm = TRUE)
    q3_T12   <- stats::quantile(x12, 0.75, na.rm = TRUE)
    med_diff <- stats::median(diff, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Not normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Wilcoxon signed-rank (paired)\n")
    cat("n =", n, "| V =", V, "| p =", signif(wt$p.value, 4), "\n")
    cat("T0:   median [IQR] = ", round(med_T0, 3), " [",
        round(q1_T0, 3), "–", round(q3_T0, 3), "]\n", sep = "")
    cat("T12:  median [IQR] = ", round(med_T12, 3), " [",
        round(q1_T12, 3), "–", round(q3_T12, 3), "]\n", sep = "")
    cat("Median difference (T12−T0) =", round(med_diff, 3), "\n")
    cat("Effect size (rank-biserial r) =", round(r_rb, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "wilcoxon_paired",
      n = n,
      shapiro = sh,
      test = wt,
      group_summary = list(
        T0  = c(median = med_T0, Q1 = q1_T0, Q3 = q3_T0),
        T12 = c(median = med_T12, Q1 = q1_T12, Q3 = q3_T12)
      ),
      median_diff = med_diff,
      effect_size = list(type = "rank_biserial_r", value = r_rb, V = V)
    ))
  }
}
# Wrapper "safe": if n < 3, returns a valid res_obj object with p = NA instead of stopping
auto_paired_test_scfa_safe <- function(df,
                                       id_col   = "PatientID",
                                       time_col = "Time",
                                       value_col = "acetate",
                                       levels   = c("T0","T12")) {
  tryCatch(
    auto_paired_test_scfa(df,
                          id_col = id_col,
                          time_col = time_col,
                          value_col = value_col,
                          levels = levels),
    error = function(e) {
      if (grepl("Very few valid pairs", e$message) || grepl("n < 3", e$message)) {
        list(
          variable = value_col,
          method   = "too_few_pairs",
          n        = NA_integer_,
          shapiro  = list(p.value = NA_real_),
          test     = list(statistic = NA_real_, p.value = NA_real_),
          effect_size = list(type = NA_character_, value = NA_real_)
        )
      } else {
        stop(e)
      }
    }
  )
}
```

### Perform  the statistical test
#### Acetate
```{r Execute stats function SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_acetate_fc <- auto_paired_test_scfa(df_scfa_fc, value_col = "acetate")
```

#### Propionate
```{r Execute stats function SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_propionate_fc <- auto_paired_test_scfa(df_scfa_fc, value_col = "propionate")
```

#### Butyrate
```{r Execute stats function SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_butyrate_fc <- auto_paired_test_scfa(df_scfa_fc, value_col = "butyrate")
```

### Function for plots
```{r Function for plot SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
make_paired_plot_fc <- function(df, value_var, res_obj, y_lab, outfile) {
  stopifnot(all(c("PatientID","Time", value_var) %in% names(df)))
  
  d <- df %>%
    mutate(Time = factor(Time, levels = c("T0","T12"))) %>%
    group_by(PatientID) %>% filter(n_distinct(Time) == 2) %>% ungroup() %>%
    mutate(x = as.numeric(Time))
  
  col_box   <- c("T0" = "#F4E1A1", "T12" = "#A1E1E1")
  col_point <- c("T0" = "#FAF3DC", "T12" = "#D6F5F5")
  
  d <- d %>%
    mutate(
      fill_box   = col_box[Time],
      fill_point = col_point[Time]
    )
  
  p_val   <- res_obj$test$p.value
  p_label <- ifelse(p_val < 0.001, "p < 0.001", paste0("p = ", signif(p_val, 3)))
  
  y_max    <- max(d[[value_var]], na.rm = TRUE) * 1.05
  p_values <- data.frame(x = 1, xend = 2, y = y_max, label = p_label)
  
  p <- ggplot() +
    stat_slab(
      data = d %>% filter(Time == "T0"),
      aes(x = x, y = .data[[value_var]], fill = fill_box),
      side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    stat_slab(
      data = d %>% filter(Time == "T12"),
      aes(x = x, y = .data[[value_var]], fill = fill_box),
      side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    geom_boxplot(
      data = d,
      aes(x = x, y = .data[[value_var]], fill = fill_box),
      width = 0.5, outlier.shape = NA,
      color = "black", alpha = 0.9, show.legend = TRUE
    ) +
    geom_point(
      data = d,
      aes(x = x, y = .data[[value_var]], fill = fill_point),
      shape = 21, size = 3, stroke = 0.3, color = "black",
      show.legend = FALSE
    ) +
    geom_line(
      data = d,
      aes(x = x, y = .data[[value_var]], group = PatientID),
      color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
    ) +
    geom_segment(
      data = p_values,
      aes(x = x, xend = xend, y = y, yend = y),
      color = "black", linewidth = 0.5
    ) +
    geom_text(
      data = p_values,
      aes(x = (x + xend) / 2, y = y * 1.02, label = label),
      size = 4
    ) +
    scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
    scale_fill_identity(guide = "legend", name = "Time",
                        breaks = col_box, labels = names(col_box)) +
    labs(x = NULL, y = y_lab) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "right",
          axis.text.x = element_text(size = 14))
  
  if (!dir.exists(here("results"))) dir.create(here("results"))
  ggsave(here("results", outfile), p, width = 8, height = 6, units = "in", dpi = 300)
  
  invisible(p)
}
```


#### Panel plots
```{r Panel plots lung function, echo=TRUE, results='show', message=FALSE, warning=FALSE}
polish_panel_plot <- function(p) {
  p +
    theme(
      legend.position = "bottom",
      plot.margin = margin(5.5, 5.5, 5.5, 5.5),
      axis.title.y = element_text(margin = margin(r = 6))
    )
}

p_ace_fc <- polish_panel_plot(
  make_paired_plot_fc(df_scfa_fc, "acetate", res_acetate_fc, "Acetate (mM)", "SCFA_acetate_feces.svg")
)

p_pro_fc <- polish_panel_plot(
  make_paired_plot_fc(df_scfa_fc, "propionate", res_propionate_fc, "Propionate (mM)", "SCFA_propionate_feces.svg")
)

p_but_fc <- polish_panel_plot(
  make_paired_plot_fc(df_scfa_fc, "butyrate", res_butyrate_fc, "Butyrate (mM)", "SCFA_butyrate_feces.svg")
)

p_ace_pro_fc <- polish_panel_plot(
  make_paired_plot_fc(df_scfa_fc, "AcetatePropionate", res_ace_pro_fc, "Acetate/Propionate ratio", "acetate_propionate_ratio_fc.svg")
)

p_ace_but_fc <- polish_panel_plot(
  make_paired_plot_fc(df_scfa_fc, "AcetateButyrate", res_ace_but_fc, "Acetate/Butyrate ratio", "acetate_butyrate_ratio_fc.svg")
  )

p_pro_but_fc <- polish_panel_plot(
  make_paired_plot_fc(df_scfa_fc, "PropionateButyrate", res_pro_but_fc, "Propionate/Butyrate ratio", "propionate_butyrate_ratio_fc.svg")
  )


panel_scfa_fc <- (p_ace_fc | p_pro_fc | p_but_fc) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

panel_scfa_fc <- panel_scfa_fc + plot_annotation(tag_levels = "A")

ggsave(here("results","SCFA_feces_panel.svg"),
       panel_scfa_fc, width = 10, height = 4, bg = "white")
```


### Save stats results
```{r Save stats results SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# === Export: acetate, propionate, butyrate ===

.build_row_scfa <- function(df, value_col, res_obj, p_fdr = NA_real_) {
  stopifnot(all(c("PatientID","Time", value_col) %in% names(df)))
  is_t <- identical(res_obj$method, "paired_t")

  paired <- df %>%
    dplyr::select(PatientID, Time, !!rlang::sym(value_col)) %>%
    tidyr::pivot_wider(names_from = Time, values_from = !!rlang::sym(value_col)) %>%
    dplyr::filter(stats::complete.cases(T0, T12))

  mean_T0  <- mean(paired$T0);  sd_T0  <- sd(paired$T0)
  mean_T12 <- mean(paired$T12); sd_T12 <- sd(paired$T12)

  med_T0 <- median(paired$T0)
  q1_T0  <- as.numeric(quantile(paired$T0,  0.25))
  q3_T0  <- as.numeric(quantile(paired$T0,  0.75))

  med_T12 <- median(paired$T12)
  q1_T12  <- as.numeric(quantile(paired$T12, 0.25))
  q3_T12  <- as.numeric(quantile(paired$T12, 0.75))

  fmt_group_T0  <- if (is_t) sprintf("%.3f ± %.3f", mean_T0, sd_T0)  else sprintf("%.3f [%0.3f-%0.3f]", med_T0,  q1_T0,  q3_T0)
  fmt_group_T12 <- if (is_t) sprintf("%.3f ± %.3f", mean_T12, sd_T12) else sprintf("%.3f [%0.3f-%0.3f]", med_T12, q1_T12, q3_T12)

  tibble::tibble(
    variable      = res_obj$variable,
    method        = res_obj$method,
    n_pairs       = res_obj$n,
    shapiro_p     = unname(res_obj$shapiro$p.value),
    stat          = unname(res_obj$test$statistic),
    p_value       = res_obj$test$p.value,
    mean_T0  = mean_T0,
    sd_T0    = sd_T0,
    median_T0= med_T0,
    q1_T0    = q1_T0,
    q3_T0    = q3_T0,
    min_T0   = min(paired$T0),
    max_T0   = max(paired$T0),
    range_T0 = max(paired$T0) - min(paired$T0),
    mean_T12   = mean_T12,
    sd_T12     = sd_T12,
    median_T12 = med_T12,
    q1_T12     = q1_T12,
    q3_T12     = q3_T12,
    min_T12    = min(paired$T12),
    max_T12    = max(paired$T12),
    range_T12  = max(paired$T12) - min(paired$T12),
    group_T0      = fmt_group_T0,
    group_T12     = fmt_group_T12,
    diff_estimate = if (is_t) res_obj$mean_diff else res_obj$median_diff,
    ci_low        = if (is_t) res_obj$ci[1] else NA_real_,
    ci_high       = if (is_t) res_obj$ci[2] else NA_real_,
    effect_type   = res_obj$effect_size$type,
    effect_value  = res_obj$effect_size$value
  )
}

row_acetate_fc <- .build_row_scfa(df_scfa_fc, "acetate", res_acetate_fc)
row_propionate_fc <- .build_row_scfa(df_scfa_fc, "propionate", res_propionate_fc)
row_butyrate_fc <- .build_row_scfa(df_scfa_fc, "butyrate", res_butyrate_fc)

scfa_stats_fc <- dplyr::bind_rows(row_acetate_fc, row_propionate_fc, row_butyrate_fc)

scfa_stats_fc$p_fdr <- p.adjust(scfa_stats_fc$p_value, method = "BH")

if (!dir.exists(here("results"))) dir.create(here("results"))
readr::write_excel_csv(scfa_stats_fc, here("results", "SCFA_ratios_feces_stats_without_ratios.csv"))
```


# ------------------------------------------------------------------------------
# SCFA SPUTUM
# ------------------------------------------------------------------------------

### Load in the data
```{r Load in data SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
df_scfa_sp <- read_delim(here("data", "SCFA_sputum.csv"))
```

#### Descriptive stats
```{r Descriptive stats SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
range(df_scfa_sp$acetate, na.rm = TRUE)
min_val <- min(df_scfa_sp$acetate, na.rm = TRUE)
max_val <- max(df_scfa_sp$acetate, na.rm = TRUE)

range(df_scfa_sp$propionate, na.rm = TRUE)
min_val <- min(df_scfa_sp$propionate, na.rm = TRUE)
max_val <- max(df_scfa_sp$propionate, na.rm = TRUE)

range(df_scfa_sp$butyrate, na.rm = TRUE)
min_val <- min(df_scfa_sp$butyrate, na.rm = TRUE)
max_val <- max(df_scfa_sp$butyrate, na.rm = TRUE)
```

### Filter data by time
```{r Filter data by time SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
data_scfa_sp_T0 <- subset(df_scfa_sp, Time == "T0")
data_scfa_sp_T12 <- subset(df_scfa_sp, Time == "T12")
```

### Function for statistic test
```{r Function for statistic test based on shapiro results SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
auto_paired_test_scfa <- function(df,
                                   id_col   = "PatientID",
                                   time_col = "Time",
                                   value_col = "acetate",
                                   levels   = c("T0","T12")) {
  stopifnot(all(c(id_col, time_col, value_col) %in% names(df)))
  
  dfw <- df[, c(id_col, time_col, value_col)]
  dfw <- tidyr::pivot_wider(
    dfw,
    names_from  = !!rlang::sym(time_col),
    values_from = !!rlang::sym(value_col)
  )
  if (!all(levels %in% names(dfw))) {
    stop("No columns found ", paste(levels, collapse = " and "), " after pivot_wider().")
  }
  dfw <- dfw[stats::complete.cases(dfw[, levels]), ]
  
  x0   <- dfw[[levels[1]]]
  x12  <- dfw[[levels[2]]]
  diff <- x12 - x0
  n    <- length(diff)
  if (n < 3) stop("Very few valid pairs (n < 3).")
  
  sh     <- shapiro.test(diff)
  normal <- sh$p.value >= 0.05
  
  if (normal) {
    tt        <- t.test(x12, x0, paired = TRUE)
    mean_diff <- mean(diff, na.rm = TRUE)
    sd_diff   <- stats::sd(diff, na.rm = TRUE)
    d_paired  <- mean_diff / sd_diff
    
    mean_T0  <- mean(x0,  na.rm = TRUE); sd_T0  <- stats::sd(x0,  na.rm = TRUE)
    mean_T12 <- mean(x12, na.rm = TRUE); sd_T12 <- stats::sd(x12, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Paired t-test\n")
    cat("n =", n, "| t =", round(tt$statistic, 4), "| p =", signif(tt$p.value, 4), "\n")
    cat("T0:   mean ± SD = ", round(mean_T0, 3), " ± ", round(sd_T0, 3), "\n", sep = "")
    cat("T12:  mean ± SD = ", round(mean_T12, 3), " ± ", round(sd_T12, 3), "\n", sep = "")
    cat("Mean difference (T12−T0) =", round(mean_diff, 3),
        " | 95% CI [", round(tt$conf.int[1], 3), ", ", round(tt$conf.int[2], 3), "]\n", sep = "")
    cat("Effect size (Cohen's d paired) =", round(d_paired, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "paired_t",
      n = n,
      shapiro = sh,
      test = tt,
      group_summary = list(
        T0  = c(mean = mean_T0, sd = sd_T0),
        T12 = c(mean = mean_T12, sd = sd_T12)
      ),
      mean_diff = mean_diff,
      ci = tt$conf.int,
      effect_size = list(type = "cohen_d_paired", value = d_paired)
    ))
    
  } else {
    wt   <- wilcox.test(x12, x0, paired = TRUE, exact = FALSE, correct = FALSE)
    V    <- as.numeric(wt$statistic)
    r_rb <- 1 - (2 * V) / (n * (n + 1))
    
    med_T0   <- stats::median(x0,  na.rm = TRUE)
    q1_T0    <- stats::quantile(x0,  0.25, na.rm = TRUE)
    q3_T0    <- stats::quantile(x0,  0.75, na.rm = TRUE)
    med_T12  <- stats::median(x12, na.rm = TRUE)
    q1_T12   <- stats::quantile(x12, 0.25, na.rm = TRUE)
    q3_T12   <- stats::quantile(x12, 0.75, na.rm = TRUE)
    med_diff <- stats::median(diff, na.rm = TRUE)
    
    cat("\n=== Auto test (", value_col, ") ===\n", sep = "")
    cat("Normality of differences: Not normal (Shapiro p =", round(sh$p.value, 4), ")\n")
    cat("Chosen test: Wilcoxon signed-rank (paired)\n")
    cat("n =", n, "| V =", V, "| p =", signif(wt$p.value, 4), "\n")
    cat("T0:   median [IQR] = ", round(med_T0, 3), " [",
        round(q1_T0, 3), "–", round(q3_T0, 3), "]\n", sep = "")
    cat("T12:  median [IQR] = ", round(med_T12, 3), " [",
        round(q1_T12, 3), "–", round(q3_T12, 3), "]\n", sep = "")
    cat("Median difference (T12−T0) =", round(med_diff, 3), "\n")
    cat("Effect size (rank-biserial r) =", round(r_rb, 3), "\n")
    
    invisible(list(
      variable = value_col,
      method = "wilcoxon_paired",
      n = n,
      shapiro = sh,
      test = wt,
      group_summary = list(
        T0  = c(median = med_T0, Q1 = q1_T0, Q3 = q3_T0),
        T12 = c(median = med_T12, Q1 = q1_T12, Q3 = q3_T12)
      ),
      median_diff = med_diff,
      effect_size = list(type = "rank_biserial_r", value = r_rb, V = V)
    ))
  }
}
# Wrapper "safe": if n < 3, returns a valid res_obj object with p = NA instead of stopping
auto_paired_test_scfa_safe <- function(df,
                                       id_col   = "PatientID",
                                       time_col = "Time",
                                       value_col = "acetate",
                                       levels   = c("T0","T12")) {
  tryCatch(
    auto_paired_test_scfa(df,
                          id_col = id_col,
                          time_col = time_col,
                          value_col = value_col,
                          levels = levels),
    error = function(e) {
      if (grepl("Very few valid pairs", e$message) || grepl("n < 3", e$message)) {
        list(
          variable = value_col,
          method   = "too_few_pairs",
          n        = NA_integer_,
          shapiro  = list(p.value = NA_real_),
          test     = list(statistic = NA_real_, p.value = NA_real_),
          effect_size = list(type = NA_character_, value = NA_real_)
        )
      } else {
        stop(e)
      }
    }
  )
}
```

### Perform  the statistical test
#### Acetate
```{r Execute stats function SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_acetate_sp <- auto_paired_test_scfa(df_scfa_sp, value_col = "acetate")
```

#### Propionate
```{r Execute stats function SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_propionate_sp <- auto_paired_test_scfa(df_scfa_sp, value_col = "propionate")
```

#### Butyrate
```{r Execute stats function SCFA, echo=TRUE, results='show', message=FALSE, warning=FALSE}
res_butyrate_sp <- auto_paired_test_scfa(df_scfa_sp, value_col = "butyrate")
```

### Function for plots
```{r Function for plot SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
make_paired_plot_sp <- function(df, value_var, res_obj, y_lab, outfile) {
  stopifnot(all(c("PatientID","Time", value_var) %in% names(df)))
  
  d <- df %>%
    mutate(Time = factor(Time, levels = c("T0","T12"))) %>%
    group_by(PatientID) %>% filter(n_distinct(Time) == 2) %>% ungroup() %>%
    mutate(x = as.numeric(Time))
  
  col_box   <- c("T0" = "#D9C2E9", "T12" = "#B3E1B6")
  col_point <- c("T0" = "#EDE3F4", "T12" = "#E2F5E3")
  
  d <- d %>%
    mutate(
      fill_box   = col_box[Time],
      fill_point = col_point[Time]
    )
  
  p_val   <- res_obj$test$p.value
  p_label <- ifelse(p_val < 0.001, "p < 0.001", paste0("p = ", signif(p_val, 3)))
  
  y_max    <- max(d[[value_var]], na.rm = TRUE) * 1.05
  p_values <- data.frame(x = 1, xend = 2, y = y_max, label = p_label)
  
  p <- ggplot() +
    stat_slab(
      data = d %>% filter(Time == "T0"),
      aes(x = x, y = .data[[value_var]], fill = fill_box),
      side = "left", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    stat_slab(
      data = d %>% filter(Time == "T12"),
      aes(x = x, y = .data[[value_var]], fill = fill_box),
      side = "right", alpha = 0.5, adjust = 0.6, show.legend = FALSE
    ) +
    geom_boxplot(
      data = d,
      aes(x = x, y = .data[[value_var]], fill = fill_box),
      width = 0.5, outlier.shape = NA,
      color = "black", alpha = 0.9, show.legend = TRUE
    ) +
    geom_point(
      data = d,
      aes(x = x, y = .data[[value_var]], fill = fill_point),
      shape = 21, size = 3, stroke = 0.3, color = "black",
      show.legend = FALSE
    ) +
    geom_line(
      data = d,
      aes(x = x, y = .data[[value_var]], group = PatientID),
      color = "grey40", alpha = 0.4, linetype = "dashed", linewidth = 0.4
    ) +
    geom_segment(
      data = p_values,
      aes(x = x, xend = xend, y = y, yend = y),
      color = "black", linewidth = 0.5
    ) +
    geom_text(
      data = p_values,
      aes(x = (x + xend) / 2, y = y * 1.02, label = label),
      size = 4
    ) +
    scale_x_continuous(breaks = c(1, 2), labels = c("T0", "T12")) +
    scale_fill_identity(guide = "legend", name = "Time",
                        breaks = col_box, labels = names(col_box)) +
    labs(x = NULL, y = y_lab) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "right",
          axis.text.x = element_text(size = 14))
  
  if (!dir.exists(here("results"))) dir.create(here("results"))
  ggsave(here("results", outfile), p, width = 8, height = 6, units = "in", dpi = 300)
  
  invisible(p)
}
```


#### Panel plots
```{r Panel plots lung function, echo=TRUE, results='show', message=FALSE, warning=FALSE}
polish_panel_plot <- function(p) {
  p +
    theme(
      legend.position = "bottom",
      plot.margin = margin(5.5, 5.5, 5.5, 5.5),
      axis.title.y = element_text(margin = margin(r = 6))
    )
}

p_ace_sp <- polish_panel_plot(
  make_paired_plot_sp(df_scfa_sp, "acetate", res_acetate_sp, "Acetate (mM)", "SCFA_acetate_sputum.svg")
)

p_pro_sp <- polish_panel_plot(
  make_paired_plot_sp(df_scfa_sp, "propionate", res_propionate_sp, "Propionate (mM)", "SCFA_propionate_sputum.svg")
)

p_but_sp <- polish_panel_plot(
  make_paired_plot_sp(df_scfa_sp, "butyrate", res_butyrate_sp, "Butyrate (mM)", "SCFA_butyrate_sputum.svg")
)

p_ace_pro_sp <- polish_panel_plot(
  make_paired_plot_sp(df_scfa_sp, "AcetatePropionate", res_ace_pro_sp, "Acetate/Propionate ratio", "acetate_propionate_ratio_sp.svg")
)

p_ace_but_sp <- polish_panel_plot(
  make_paired_plot_sp(df_scfa_sp, "AcetateButyrate", res_ace_but_sp, "Acetate/Butyrate ratio", "acetate_butyrate_ratio_sp.svg")
  )

p_pro_but_sp <- polish_panel_plot(
  make_paired_plot_sp(df_scfa_sp, "PropionateButyrate", res_pro_but_sp, "Propionate/Butyrate ratio", "propionate_butyrate_ratio_sp.svg")
  )

panel_scfa_sp <- (p_ace_sp | p_pro_sp | p_but_sp) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

panel_scfa_sp <- panel_scfa_sp + plot_annotation(tag_levels = "A")

ggsave(here("results","SCFA_sputum_panel.svg"),
       panel_scfa_sp, width = 10, height = 4, bg = "white")
```


### Save stats results
```{r Save stats results SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# === Export: acetate, propionate, butyrate ===

.build_row_scfa <- function(df, value_col, res_obj, p_fdr = NA_real_) {
  stopifnot(all(c("PatientID","Time", value_col) %in% names(df)))
  is_t <- identical(res_obj$method, "paired_t")

  paired <- df %>%
    dplyr::select(PatientID, Time, !!rlang::sym(value_col)) %>%
    tidyr::pivot_wider(names_from = Time, values_from = !!rlang::sym(value_col)) %>%
    dplyr::filter(stats::complete.cases(T0, T12))

  mean_T0  <- mean(paired$T0);  sd_T0  <- sd(paired$T0)
  mean_T12 <- mean(paired$T12); sd_T12 <- sd(paired$T12)

  med_T0 <- median(paired$T0)
  q1_T0  <- as.numeric(quantile(paired$T0,  0.25))
  q3_T0  <- as.numeric(quantile(paired$T0,  0.75))

  med_T12 <- median(paired$T12)
  q1_T12  <- as.numeric(quantile(paired$T12, 0.25))
  q3_T12  <- as.numeric(quantile(paired$T12, 0.75))

  fmt_group_T0  <- if (is_t) sprintf("%.3f ± %.3f", mean_T0, sd_T0)  else sprintf("%.3f [%0.3f-%0.3f]", med_T0,  q1_T0,  q3_T0)
  fmt_group_T12 <- if (is_t) sprintf("%.3f ± %.3f", mean_T12, sd_T12) else sprintf("%.3f [%0.3f-%0.3f]", med_T12, q1_T12, q3_T12)

  tibble::tibble(
    variable      = res_obj$variable,
    method        = res_obj$method,
    n_pairs       = res_obj$n,
    shapiro_p     = unname(res_obj$shapiro$p.value),
    stat          = unname(res_obj$test$statistic),
    p_value       = res_obj$test$p.value,
    mean_T0  = mean_T0,
    sd_T0    = sd_T0,
    median_T0= med_T0,
    q1_T0    = q1_T0,
    q3_T0    = q3_T0,
    min_T0   = min(paired$T0),
    max_T0   = max(paired$T0),
    range_T0 = max(paired$T0) - min(paired$T0),
    mean_T12   = mean_T12,
    sd_T12     = sd_T12,
    median_T12 = med_T12,
    q1_T12     = q1_T12,
    q3_T12     = q3_T12,
    min_T12    = min(paired$T12),
    max_T12    = max(paired$T12),
    range_T12  = max(paired$T12) - min(paired$T12),
    group_T0      = fmt_group_T0,
    group_T12     = fmt_group_T12,
    diff_estimate = if (is_t) res_obj$mean_diff else res_obj$median_diff,
    ci_low        = if (is_t) res_obj$ci[1] else NA_real_,
    ci_high       = if (is_t) res_obj$ci[2] else NA_real_,
    effect_type   = res_obj$effect_size$type,
    effect_value  = res_obj$effect_size$value
  )
}

row_acetate_sp <- .build_row_scfa(df_scfa_sp, "acetate", res_acetate_sp)
row_propionate_sp <- .build_row_scfa(df_scfa_sp, "propionate", res_propionate_sp)
row_butyrate_sp <- .build_row_scfa(df_scfa_sp, "butyrate", res_butyrate_sp)

scfa_stats_sp <- dplyr::bind_rows(row_acetate_sp, row_propionate_sp, row_butyrate_sp)

scfa_stats_sp$p_fdr <- p.adjust(scfa_stats_sp$p_value, method = "BH")

if (!dir.exists(here("results"))) dir.create(here("results"))
readr::write_excel_csv(scfa_stats_sp, here("results", "SCFA_ratios_sputum_stats_without_ratios.csv"))
```


# -----------------------------------------------------------------------------
# GLOBAL FDR
# -----------------------------------------------------------------------------

### Including ONLY of interest
```{r Stats results FDR, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# 1) Put all your test objects in a single nested list (system -> variables -> res_object)
clinical_tests <- list(
  `Fecal SCFAs` = list(
    acetate = res_acetate_fc,
    propionate = res_propionate_fc,
    butyrate = res_butyrate_fc
  ),
  `Sputum SCFAs` = list(
    acetate = res_acetate_sp,
    propionate = res_propionate_sp,
    butyrate = res_butyrate_sp
  )
)

# 2) Helper to extract p-value (adapt if your object structure differs)
get_p <- function(x) x$test$p.value

# 3) Build one table with all clinical p-values
clinical_pvals <- purrr::imap_dfr(clinical_tests, \(tests, sys) {
  tibble::tibble(
    system   = sys,
    variable = names(tests),
    p_value  = purrr::map_dbl(tests, get_p)
  )
})

# 4) Global BH-FDR across ALL clinical variables together
clinical_pvals_globalFDR <- clinical_pvals %>%
  dplyr::mutate(
    p_fdr_global = p.adjust(p_value, method = "BH"),
    signif_fdr   = p_fdr_global < 0.05
  ) %>%
  dplyr::arrange(p_fdr_global)

print(clinical_pvals_globalFDR, n = Inf)

# 5) Save
if (!dir.exists(here::here("results"))) dir.create(here::here("results"))
readr::write_excel_csv(clinical_pvals_globalFDR, here::here("results", "clinical_globalFDR_pvalues_SCFAs_without_ratios.csv"))
```


#-------------------------------------------------------------------------------
#-------------------------------------------------------------------------------
# OPTIONAL ANALYSIS (NOT INCLUDED IN THE THESIS) -------------------------------
#-------------------------------------------------------------------------------
#-------------------------------------------------------------------------------

#-------------------------------------------------------------------------------
# MICROBIOME CORRELATIONS
#-------------------------------------------------------------------------------
#--------
## 0.1. Function: run LMM z(outcome) ~ z(taxon) + Time + (1|PatientID) or z(outcome) ~ z(taxon) + Time + PatientID if singular fit
```{r Function LMM, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
run_lmm_taxa_vs_outcomes <- function(dat_long,
                                    taxa_keep,
                                    outcomes,
                                    taxon_col = "taxon_val",
                                    id_col = "PatientID",
                                    time_col = "Time",
                                    tol_singular = 1e-4) {

  # safety check
  missing_out <- setdiff(outcomes, colnames(dat_long))
  if (length(missing_out) > 0) {
    stop("These outcomes are missing in dat_long: ", paste(missing_out, collapse = ", "))
  }

  purrr::map_dfr(outcomes, function(out) {

    out <- as.character(out)[1]

    purrr::map_dfr(taxa_keep, function(tx) {

      tx <- as.character(tx)[1]

      dat_tx <- dat_long %>%
        dplyr::filter(FeatureID == tx) %>%
        dplyr::select(FeatureID, dplyr::all_of(c(taxon_col, out, time_col, id_col))) %>%
        dplyr::filter(stats::complete.cases(.)) %>%
        dplyr::mutate(
          !!time_col := factor(.data[[time_col]], levels = c("T0","T12")),
          !!id_col   := factor(.data[[id_col]])
        )

      # skip if too few points or no variance
      if (nrow(dat_tx) < 6 ||
          sd(dat_tx[[taxon_col]], na.rm = TRUE) == 0 ||
          sd(dat_tx[[out]],       na.rm = TRUE) == 0) {
        return(tibble::tibble(
          Outcome = out, FeatureID = tx, n = nrow(dat_tx),
          beta_SD = NA_real_, conf.low = NA_real_, conf.high = NA_real_, p = NA_real_,
          singular_fit = NA, model_used = NA_character_
        ))
      }

      # standardize within this model dataset
      dat_tx <- dat_tx %>%
        dplyr::mutate(
          taxon_z = as.numeric(scale(.data[[taxon_col]])),
          y_z     = as.numeric(scale(.data[[out]]))
        )

      # 1) Fit LMM first
      fml_lmm <- stats::as.formula(paste0("y_z ~ taxon_z + ", time_col, " + (1|", id_col, ")"))
      fit_lmm <- lmerTest::lmer(fml_lmm, data = dat_tx)

      sing <- lme4::isSingular(fit_lmm, tol = tol_singular)

      # 2) If singular, refit with PatientID as fixed effect
      if (isTRUE(sing)) {

        fml_fe <- stats::as.formula(paste0("y_z ~ taxon_z + ", time_col, " + ", id_col))
        fit_fe <- stats::lm(fml_fe, data = dat_tx)

        co <- summary(fit_fe)$coefficients
        ci <- suppressWarnings(stats::confint(fit_fe, parm = "taxon_z"))

        return(tibble::tibble(
          Outcome   = out,
          FeatureID = tx,
          n         = nrow(dat_tx),
          beta_SD   = unname(co["taxon_z","Estimate"]),
          conf.low  = ci[1],
          conf.high = ci[2],
          p         = unname(co["taxon_z","Pr(>|t|)"]),
          singular_fit = TRUE,
          model_used   = "FE_PatientID"
        ))
      }

      # 3) Otherwise keep LMM results
      co <- summary(fit_lmm)$coefficients
      ci <- suppressWarnings(confint(fit_lmm, parm = "taxon_z", method = "Wald"))

      tibble::tibble(
        Outcome   = out,
        FeatureID = tx,
        n         = nrow(dat_tx),
        beta_SD   = unname(co["taxon_z","Estimate"]),
        conf.low  = ci[1],
        conf.high = ci[2],
        p         = unname(co["taxon_z","Pr(>|t|)"]),
        singular_fit = FALSE,
        model_used   = "LMM"
      )
    })

  }) %>%
    dplyr::group_by(Outcome) %>%
    dplyr::mutate(p_fdr_outcome = p.adjust(p, method = "BH")) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(p_fdr_global = p.adjust(p, method = "BH"))
}
```

## 0.2. Helper: build forest dataframe for a given group
```{r Helper forest, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
prep_forest_df <- function(res_table, outcomes_vec, labels_vec, taxa_order) {

  res_table %>%
    dplyr::filter(Outcome %in% outcomes_vec, FeatureID %in% taxa_order) %>%
    dplyr::mutate(
      Outcome_label = dplyr::recode(Outcome, !!!as.list(labels_vec)),
      Outcome_label = factor(Outcome_label,
                             levels = rev(unname(labels_vec[outcomes_vec]))),
      FeatureID = factor(FeatureID, levels = taxa_order),
      sig = !is.na(.data[[FDR_COL]]) & .data[[FDR_COL]] < 0.05
    ) %>%
    dplyr::filter(!is.na(beta_SD), !is.na(conf.low), !is.na(conf.high))
}
```

#----------------------------------------------------------------------------
# FECAL MICROBIOTA vs FECAL SCFAs
#----------------------------------------------------------------------------
- microbiota file: RA_DAA_plot_relabundance_long_fc.csv
- SCFA file: df_scfa_fc

## 0.3. Taxa deltas for fecal microbiota
```{r Taxa deltas, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Load the long RA table
ra <- readr::read_csv(
  here::here("data", "RA_DAA_plot_relabundance_long_fc.csv"),
  show_col_types = FALSE
)

colnames(ra)
sort(unique(ra$FeatureID))

# Keep ALL taxa
taxa_keep <- sort(unique(ra$FeatureID))

ra2 <- ra %>%
  dplyr::mutate(
    PatientID = sub("H\\d+$", "", SampleID),
    Time = dplyr::case_when(
      Time %in% c("T0", "time0", "Time0", "0") ~ "T0",
      Time %in% c("T12", "time12", "Time12", "12") ~ "T12",
      TRUE ~ as.character(Time)
    ),
    RelAbundance = as.numeric(RelAbundance)
  ) %>%
  dplyr::filter(FeatureID %in% taxa_keep)

# Check
dplyr::count(ra2, Time)
length(unique(ra2$PatientID))
length(unique(ra2$FeatureID))


# TAXA DELTAS (Δ = T12 - T0)
# log1p used for taxa as in your original code
taxa_delta <- ra2 %>%
  dplyr::mutate(val = log1p(RelAbundance)) %>%
  dplyr::select(PatientID, FeatureID, Time, val) %>%
  tidyr::pivot_wider(names_from = Time, values_from = val) %>%
  dplyr::filter(stats::complete.cases(T0, T12)) %>%
  dplyr::mutate(delta = T12 - T0) %>%
  dplyr::select(PatientID, FeatureID, delta) %>%
  tidyr::pivot_wider(
    names_from = FeatureID,
    values_from = delta,
    names_glue = "{FeatureID}_delta"
  )

colnames(taxa_delta)
```

#----------------

# 2. Fecal microbiome composition and SCFAs
## Deltas
```{r Deltas SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Fecal SCFA deltas (Δ = T12 - T0) from df_scfa_fc
scfa_vars <- c("acetate", "propionate", "butyrate")#, "AcetatePropionate", "AcetateButyrate", "PropionateButyrate")

fix_time <- function(x) {
  x <- as.character(x)
  dplyr::case_when(
    x %in% c("T0", "0", "Time0", "time0", "baseline") ~ "T0",
    x %in% c("T12", "12", "Time12", "time12", "followup") ~ "T12",
    TRUE ~ x
  )
}

scfa_delta_fc <- purrr::map(scfa_vars, function(v) {
  df_scfa_fc %>%
    dplyr::mutate(Time = fix_time(Time)) %>%
    dplyr::select(PatientID, Time, dplyr::all_of(v)) %>%
    tidyr::pivot_wider(names_from = Time, values_from = dplyr::all_of(v)) %>%
    dplyr::filter(stats::complete.cases(T0, T12)) %>%
    dplyr::mutate(delta = T12 - T0) %>%
    dplyr::select(PatientID, delta) %>%
    dplyr::rename(!!paste0(v, "_delta") := delta)
}) %>%
  purrr::reduce(dplyr::left_join, by = "PatientID")

colnames(scfa_delta_fc)

# Join taxa deltas + SCFA deltas
delta_all_taxa_fc <- taxa_delta %>%
  dplyr::left_join(scfa_delta_fc, by = "PatientID")

colnames(delta_all_taxa_fc)
```

## Spearman
```{r Spearman, echo=TRUE, results='show', message=FALSE, warning=FALSE}
taxa_y <- paste0(taxa_keep, "_delta")

vars_x <- c(
  "acetate_delta",
  "propionate_delta",
  "butyrate_delta"#,
  #"AcetatePropionate_delta",
  #"AcetateButyrate_delta",
  #"PropionateButyrate_delta"
)

min_n <- 6

res_spear_taxa_scfa_fc <- purrr::map_dfr(taxa_y, function(ty) {
  purrr::map_dfr(vars_x, function(vx) {

    tmp <- delta_all_taxa_fc %>%
      dplyr::select(dplyr::all_of(c(ty, vx))) %>%
      dplyr::filter(stats::complete.cases(.))

    if (nrow(tmp) < min_n) {
      return(tibble::tibble(
        taxon = ty,
        var = vx,
        n = nrow(tmp),
        rho = NA_real_,
        p = NA_real_
      ))
    }

    ct <- suppressWarnings(
      stats::cor.test(tmp[[1]], tmp[[2]], method = "spearman", exact = FALSE)
    )

    tibble::tibble(
      taxon = ty,
      var = vx,
      n = nrow(tmp),
      rho = unname(ct$estimate),
      p = ct$p.value
    )
  })
}) %>%
  dplyr::mutate(
    p_fdr = p.adjust(p, method = "fdr")
  ) %>%
  dplyr::arrange(p_fdr)

print(res_spear_taxa_scfa_fc, n = Inf)

readr::write_csv(res_spear_taxa_scfa_fc, here::here("results", "res_spearman_taxa_scfa_ALL_taxa_fc_without_ratios.csv")
)
```

### Heatmap
```{r Heatmap, echo=TRUE, results='show', message=FALSE, warning=FALSE}
scfa_labels <- c(
  acetate_delta            = "Acetate",
  propionate_delta         = "Propionate",
  butyrate_delta           = "Butyrate"#,
  #AcetatePropionate_delta  = "Acetate/Propionate",
  #AcetateButyrate_delta    = "Acetate/Butyrate",
  #PropionateButyrate_delta = "Propionate/Butyrate"
)

plot_df_fc <- res_spear_taxa_scfa_fc %>%
  dplyr::filter(!is.na(rho)) %>%
  dplyr::mutate(
    taxon = sub("_delta$", "", taxon),
    var_label = dplyr::recode(var, !!!scfa_labels),
    sig = dplyr::case_when(
      is.na(p_fdr)  ~ "",
      p_fdr < 0.001 ~ "***",
      p_fdr < 0.01  ~ "**",
      p_fdr < 0.05  ~ "*",
      TRUE ~ ""
    ),
    label = paste0(sprintf("%.2f", rho), sig)
  )

# Order taxa by best FDR
taxa_order <- plot_df_fc %>%
  dplyr::group_by(taxon) %>%
  dplyr::summarise(best_fdr = min(p_fdr, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(best_fdr, taxon) %>%
  dplyr::pull(taxon)

plot_df_fc <- plot_df_fc %>%
  dplyr::mutate(
    taxon = factor(taxon, levels = rev(taxa_order)),
    var_label = factor(
      var_label,
      levels = c(
        "Acetate",
        "Propionate",
        "Butyrate"#,
        #"Acetate/Propionate",
        #"Acetate/Butyrate",
        #"Propionate/Butyrate"
      )
    )
  )

p_heat_taxa_scfa_fc <- ggplot(plot_df_fc, aes(x = var_label, y = taxon, fill = rho)) +
  geom_tile() +
  geom_text(aes(label = sig), size = 4) +
  scale_fill_gradient2(
    low = "#2C7BB6",
    mid = "grey95",
    high = "#D7191C",
    midpoint = 0,
    limits = c(-1, 1),
    name = "Spearman rho"
  ) +
  labs(
    title = "Spearman correlations of changes (ΔT12-T0): all taxa vs fecal SCFAs",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

p_heat_taxa_scfa_fc

ggsave(
  filename = here::here("results", "heatmap_spearman_deltas_FDR_ALL_taxa_SCFAs_fc.svg"),
  plot = p_heat_taxa_scfa_fc,
  device = "svg",
  width = 8,
  height = max(6, 0.35 * length(unique(plot_df_fc$taxon))),
  units = "in"
)
```

## LMER
### Function
```{r LMER function, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# FUNCTION: LMM z(outcome) ~ z(taxon) + Time + (1|PatientID)
# fallback to FE model if singular
run_lmm_taxa_vs_outcomes <- function(dat_long,
                                     taxa_keep,
                                     outcomes,
                                     taxon_col = "RelAbundance",
                                     id_col = "PatientID",
                                     time_col = "Time",
                                     tol_singular = 1e-4) {

  missing_out <- setdiff(outcomes, colnames(dat_long))
  if (length(missing_out) > 0) {
    stop("These outcomes are missing in dat_long: ", paste(missing_out, collapse = ", "))
  }

  purrr::map_dfr(outcomes, function(out) {

    out <- as.character(out)[1]

    purrr::map_dfr(taxa_keep, function(tx) {

      tx <- as.character(tx)[1]

      dat_tx <- dat_long %>%
        dplyr::filter(FeatureID == tx) %>%
        dplyr::select(FeatureID, dplyr::all_of(c(taxon_col, out, time_col, id_col))) %>%
        dplyr::filter(stats::complete.cases(.)) %>%
        dplyr::mutate(
          !!time_col := factor(.data[[time_col]], levels = c("T0", "T12")),
          !!id_col   := factor(.data[[id_col]])
        )

      if (nrow(dat_tx) < 6 ||
          sd(dat_tx[[taxon_col]], na.rm = TRUE) == 0 ||
          sd(dat_tx[[out]], na.rm = TRUE) == 0) {
        return(tibble::tibble(
          Outcome = out,
          FeatureID = tx,
          n = nrow(dat_tx),
          beta_SD = NA_real_,
          conf.low = NA_real_,
          conf.high = NA_real_,
          p = NA_real_,
          singular_fit = NA,
          model_used = NA_character_
        ))
      }

      dat_tx <- dat_tx %>%
        dplyr::mutate(
          taxon_z = as.numeric(scale(.data[[taxon_col]])),
          y_z     = as.numeric(scale(.data[[out]]))
        )

      # 1) LMM
      fml_lmm <- stats::as.formula(
        paste0("y_z ~ taxon_z + ", time_col, " + (1|", id_col, ")")
      )
      fit_lmm <- lmerTest::lmer(fml_lmm, data = dat_tx)

      sing <- lme4::isSingular(fit_lmm, tol = tol_singular)

      # 2) FE fallback if singular
      if (isTRUE(sing)) {
        fml_fe <- stats::as.formula(
          paste0("y_z ~ taxon_z + ", time_col, " + ", id_col)
        )
        fit_fe <- stats::lm(fml_fe, data = dat_tx)

        co <- summary(fit_fe)$coefficients
        ci <- suppressWarnings(stats::confint(fit_fe, parm = "taxon_z"))

        return(tibble::tibble(
          Outcome = out,
          FeatureID = tx,
          n = nrow(dat_tx),
          beta_SD = unname(co["taxon_z", "Estimate"]),
          conf.low = ci[1],
          conf.high = ci[2],
          p = unname(co["taxon_z", "Pr(>|t|)"]),
          singular_fit = TRUE,
          model_used = "FE_PatientID"
        ))
      }

      # 3) Keep LMM
      co <- summary(fit_lmm)$coefficients
      ci <- suppressWarnings(confint(fit_lmm, parm = "taxon_z", method = "Wald"))

      tibble::tibble(
        Outcome = out,
        FeatureID = tx,
        n = nrow(dat_tx),
        beta_SD = unname(co["taxon_z", "Estimate"]),
        conf.low = ci[1],
        conf.high = ci[2],
        p = unname(co["taxon_z", "Pr(>|t|)"]),
        singular_fit = FALSE,
        model_used = "LMM"
      )
    })
  }) %>%
    dplyr::group_by(Outcome) %>%
    dplyr::mutate(p_fdr_outcome = p.adjust(p, method = "BH")) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(p_fdr_global = p.adjust(p, method = "BH"))
}
```

### Data
```{r Data, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# One row per PatientID-Time for fecal SCFAs
meta_scfa_fc <- df_scfa_fc %>%
  dplyr::mutate(Time = fix_time(Time)) %>%
  dplyr::select(PatientID, Time, dplyr::all_of(scfa_vars)) %>%
  dplyr::group_by(PatientID, Time) %>%
  dplyr::summarise(
    dplyr::across(
      dplyr::all_of(scfa_vars),
      ~ if (all(is.na(.x))) NA_real_ else mean(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  )

# Long microbiome table
dat_long <- ra2 %>%
  dplyr::select(PatientID, Time, FeatureID, RelAbundance)

# Join SCFAs into microbiome long table
dat_long_scfa_full_fc <- dat_long %>%
  dplyr::mutate(
    Time = fix_time(Time),
    PatientID = as.character(PatientID)
  ) %>%
  dplyr::left_join(meta_scfa_fc, by = c("PatientID", "Time")) %>%
  dplyr::mutate(
    Time = factor(Time, levels = c("T0", "T12")),
    PatientID = factor(PatientID)
  )

# Check
colnames(dat_long_scfa_full_fc)
```


### Run LMER
```{r Run LMER, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
res_lmm_scfa_fc <- run_lmm_taxa_vs_outcomes(
  dat_long  = dat_long_scfa_full_fc,
  taxa_keep = taxa_keep,
  outcomes  = scfa_vars,
  taxon_col = "RelAbundance"
)

print(res_lmm_scfa_fc, n = Inf)

readr::write_csv(
  res_lmm_scfa_fc,
  here::here("results", "res_lmm_taxa_vs_SCFAs_LMM_SD_ALL_taxa_fc.csv")
)
```

### Forest plot
```{r Forest plot, echo=TRUE, results='show', message=FALSE, warning=FALSE}
# =========================================================
# Forest plot: ONLY acetate, propionate, butyrate
# taxa on y-axis, SCFAs on top
# =========================================================

FDR_COL <- "p_fdr_global"   # or "p_fdr_outcome"

scfa_labels_basic <- c(
  acetate    = "Acetate",
  propionate = "Propionate",
  butyrate   = "Butyrate"
)

scfa_basic <- names(scfa_labels_basic)

# Keep only the 3 SCFAs
df_scfa_forest_basic_fc <- res_lmm_scfa_fc %>%
  dplyr::filter(
    Outcome %in% scfa_basic,
    !is.na(beta_SD),
    !is.na(conf.low),
    !is.na(conf.high)
  ) %>%
  dplyr::mutate(
    Outcome_label = dplyr::recode(Outcome, !!!as.list(scfa_labels_basic)),
    Outcome_label = factor(
      Outcome_label,
      levels = c("Acetate", "Propionate", "Butyrate")
    ),
    sig = !is.na(.data[[FDR_COL]]) & .data[[FDR_COL]] < 0.05
  )

# Optional: order taxa by their best FDR across the 3 SCFAs
taxa_order_forest_fc <- df_scfa_forest_basic_fc %>%
  dplyr::group_by(FeatureID) %>%
  dplyr::summarise(best_fdr = min(.data[[FDR_COL]], na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(best_fdr, FeatureID) %>%
  dplyr::pull(FeatureID)

df_scfa_forest_basic_fc <- df_scfa_forest_basic_fc %>%
  dplyr::mutate(
    FeatureID = factor(FeatureID, levels = rev(taxa_order_forest_fc))
  )

p_taxa_scfa_basic_fc <- ggplot(df_scfa_forest_basic_fc, aes(x = beta_SD, y = FeatureID)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    width = 0.2,
    linewidth = 0.8,
    orientation = "y"
  ) +
  geom_point(aes(fill = sig), shape = 21, size = 3, color = "black") +
  facet_grid(. ~ Outcome_label) +
  scale_fill_manual(
    name = "BH-FDR (q < 0.05)",
    values = c(`TRUE` = "black", `FALSE` = "white"),
    breaks = c(TRUE, FALSE),
    labels = c("Significant", "Not significant")
  ) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE)) +
  labs(
    title = "Association between fecal taxa and fecal SCFAs (standardized)",
    subtitle = "Only acetate, propionate and butyrate; models adjusted for Time",
    x = "Standardized effect (SD outcome per 1 SD taxon)",
    y = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "grey85", linewidth = 0.6),
    panel.grid.minor = element_line(color = "grey92", linewidth = 0.4),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 9),
    strip.text  = element_text(face = "plain"),
    legend.position = "bottom"
  )

p_taxa_scfa_basic_fc

ggsave(
  filename = here::here("results", "forest_SCFAs_basic_taxa_on_y_fc.svg"),
  plot = p_taxa_scfa_basic_fc,
  device = "svg",
  width = 10,
  height = max(6, 0.28 * length(unique(df_scfa_forest_basic_fc$FeatureID))),
  units = "in"
)
```


#----------------------------------------------------------------------------
# LUNG MICROBIOTA vs SPUTUM SCFAs
#----------------------------------------------------------------------------
- microbiota file: RA_DAA_plot_relabundance_long_sp.csv
- SCFA file: df_scfa_sp

## 0.3. Taxa deltas for fecal microbiota
```{r Taxa deltas, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Load the long RA table
ra <- readr::read_csv(
  here::here("data", "RA_DAA_plot_relabundance_long_sp.csv"),
  show_col_types = FALSE
)

colnames(ra)
sort(unique(ra$FeatureID))

# Keep ALL taxa
taxa_keep <- sort(unique(ra$FeatureID))

ra2 <- ra %>%
  dplyr::mutate(
    PatientID = sub("E\\d+$", "", SampleID),
    Time = dplyr::case_when(
      Time %in% c("T0", "time0", "Time0", "0") ~ "T0",
      Time %in% c("T12", "time12", "Time12", "12") ~ "T12",
      TRUE ~ as.character(Time)
    ),
    RelAbundance = as.numeric(RelAbundance)
  ) %>%
  dplyr::filter(FeatureID %in% taxa_keep)

# Check
dplyr::count(ra2, Time)
length(unique(ra2$PatientID))
length(unique(ra2$FeatureID))


# TAXA DELTAS (Δ = T12 - T0)
# log1p used for taxa as in your original code
taxa_delta <- ra2 %>%
  dplyr::mutate(val = log1p(RelAbundance)) %>%
  dplyr::select(PatientID, FeatureID, Time, val) %>%
  tidyr::pivot_wider(names_from = Time, values_from = val) %>%
  dplyr::filter(stats::complete.cases(T0, T12)) %>%
  dplyr::mutate(delta = T12 - T0) %>%
  dplyr::select(PatientID, FeatureID, delta) %>%
  tidyr::pivot_wider(
    names_from = FeatureID,
    values_from = delta,
    names_glue = "{FeatureID}_delta"
  )

colnames(taxa_delta)
```

#----------------

# 2. Lung microbiome composition and SCFAs
## Deltas
```{r Deltas SCFA, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Sputum SCFA deltas (Δ = T12 - T0) from df_scfa_sp
scfa_vars <- c("acetate", "propionate", "butyrate")#, "AcetatePropionate", "AcetateButyrate", "PropionateButyrate")

fix_time <- function(x) {
  x <- as.character(x)
  dplyr::case_when(
    x %in% c("T0", "0", "Time0", "time0", "baseline") ~ "T0",
    x %in% c("T12", "12", "Time12", "time12", "followup") ~ "T12",
    TRUE ~ x
  )
}

scfa_delta_sp <- purrr::map(scfa_vars, function(v) {
  df_scfa_sp %>%
    dplyr::mutate(Time = fix_time(Time)) %>%
    dplyr::select(PatientID, Time, dplyr::all_of(v)) %>%
    tidyr::pivot_wider(names_from = Time, values_from = dplyr::all_of(v)) %>%
    dplyr::filter(stats::complete.cases(T0, T12)) %>%
    dplyr::mutate(delta = T12 - T0) %>%
    dplyr::select(PatientID, delta) %>%
    dplyr::rename(!!paste0(v, "_delta") := delta)
}) %>%
  purrr::reduce(dplyr::left_join, by = "PatientID")

colnames(scfa_delta_sp)

# Join taxa deltas + SCFA deltas
delta_all_taxa_sp <- taxa_delta %>%
  dplyr::left_join(scfa_delta_sp, by = "PatientID")

colnames(delta_all_taxa_sp)
```

## Spearman
```{r Spearman, echo=TRUE, results='show', message=FALSE, warning=FALSE}
taxa_y <- paste0(taxa_keep, "_delta")

vars_x <- c(
  "acetate_delta",
  "propionate_delta",
  "butyrate_delta"#,
  #"AcetatePropionate_delta",
  #"AcetateButyrate_delta",
  #"PropionateButyrate_delta"
)

min_n <- 6

res_spear_taxa_scfa_sp <- purrr::map_dfr(taxa_y, function(ty) {
  purrr::map_dfr(vars_x, function(vx) {

    tmp <- delta_all_taxa_sp %>%
      dplyr::select(dplyr::all_of(c(ty, vx))) %>%
      dplyr::filter(stats::complete.cases(.))

    if (nrow(tmp) < min_n) {
      return(tibble::tibble(
        taxon = ty,
        var = vx,
        n = nrow(tmp),
        rho = NA_real_,
        p = NA_real_
      ))
    }

    ct <- suppressWarnings(
      stats::cor.test(tmp[[1]], tmp[[2]], method = "spearman", exact = FALSE)
    )

    tibble::tibble(
      taxon = ty,
      var = vx,
      n = nrow(tmp),
      rho = unname(ct$estimate),
      p = ct$p.value
    )
  })
}) %>%
  dplyr::mutate(
    p_fdr = p.adjust(p, method = "fdr")
  ) %>%
  dplyr::arrange(p_fdr)

print(res_spear_taxa_scfa_sp, n = Inf)

readr::write_csv(res_spear_taxa_scfa_sp, here::here("results", "res_spearman_taxa_scfa_ALL_taxa_sp_without_ratios.csv")
)
```

### Heatmap
```{r Heatmap, echo=TRUE, results='show', message=FALSE, warning=FALSE}
scfa_labels <- c(
  acetate_delta            = "Acetate",
  propionate_delta         = "Propionate",
  butyrate_delta           = "Butyrate"#,
  #AcetatePropionate_delta  = "Acetate/Propionate",
  #AcetateButyrate_delta    = "Acetate/Butyrate",
  #PropionateButyrate_delta = "Propionate/Butyrate"
)

plot_df_sp <- res_spear_taxa_scfa_sp %>%
  dplyr::filter(!is.na(rho)) %>%
  dplyr::mutate(
    taxon = sub("_delta$", "", taxon),
    var_label = dplyr::recode(var, !!!scfa_labels),
    sig = dplyr::case_when(
      is.na(p_fdr)  ~ "",
      p_fdr < 0.001 ~ "***",
      p_fdr < 0.01  ~ "**",
      p_fdr < 0.05  ~ "*",
      TRUE ~ ""
    ),
    label = paste0(sprintf("%.2f", rho), sig)
  )

# Order taxa by best FDR
taxa_order <- plot_df_sp %>%
  dplyr::group_by(taxon) %>%
  dplyr::summarise(best_fdr = min(p_fdr, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(best_fdr, taxon) %>%
  dplyr::pull(taxon)

plot_df_sp <- plot_df_sp %>%
  dplyr::mutate(
    taxon = factor(taxon, levels = rev(taxa_order)),
    var_label = factor(
      var_label,
      levels = c(
        "Acetate",
        "Propionate",
        "Butyrate"#,
        #"Acetate/Propionate",
        #"Acetate/Butyrate",
        #"Propionate/Butyrate"
      )
    )
  )

p_heat_taxa_scfa_sp <- ggplot(plot_df_sp, aes(x = var_label, y = taxon, fill = rho)) +
  geom_tile() +
  geom_text(aes(label = sig), size = 4) +
  scale_fill_gradient2(
    low = "#2C7BB6",
    mid = "grey95",
    high = "#D7191C",
    midpoint = 0,
    limits = c(-1, 1),
    name = "Spearman rho"
  ) +
  labs(
    title = "Spearman correlations of changes (ΔT12-T0): all taxa vs sputum SCFAs",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

p_heat_taxa_scfa_sp

ggsave(
  filename = here::here("results", "heatmap_spearman_deltas_FDR_ALL_taxa_SCFAs_sp.svg"),
  plot = p_heat_taxa_scfa_sp,
  device = "svg",
  width = 8,
  height = max(6, 0.35 * length(unique(plot_df_sp$taxon))),
  units = "in"
)
```

## LMER
### Function
```{r LMER function, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# FUNCTION: LMM z(outcome) ~ z(taxon) + Time + (1|PatientID)
# fallback to FE model if singular
run_lmm_taxa_vs_outcomes <- function(dat_long,
                                     taxa_keep,
                                     outcomes,
                                     taxon_col = "RelAbundance",
                                     id_col = "PatientID",
                                     time_col = "Time",
                                     tol_singular = 1e-4) {

  missing_out <- setdiff(outcomes, colnames(dat_long))
  if (length(missing_out) > 0) {
    stop("These outcomes are missing in dat_long: ", paste(missing_out, collapse = ", "))
  }

  purrr::map_dfr(outcomes, function(out) {

    out <- as.character(out)[1]

    purrr::map_dfr(taxa_keep, function(tx) {

      tx <- as.character(tx)[1]

      dat_tx <- dat_long %>%
        dplyr::filter(FeatureID == tx) %>%
        dplyr::select(FeatureID, dplyr::all_of(c(taxon_col, out, time_col, id_col))) %>%
        dplyr::filter(stats::complete.cases(.)) %>%
        dplyr::mutate(
          !!time_col := factor(.data[[time_col]], levels = c("T0", "T12")),
          !!id_col   := factor(.data[[id_col]])
        )

      if (nrow(dat_tx) < 6 ||
          sd(dat_tx[[taxon_col]], na.rm = TRUE) == 0 ||
          sd(dat_tx[[out]], na.rm = TRUE) == 0) {
        return(tibble::tibble(
          Outcome = out,
          FeatureID = tx,
          n = nrow(dat_tx),
          beta_SD = NA_real_,
          conf.low = NA_real_,
          conf.high = NA_real_,
          p = NA_real_,
          singular_fit = NA,
          model_used = NA_character_
        ))
      }

      dat_tx <- dat_tx %>%
        dplyr::mutate(
          taxon_z = as.numeric(scale(.data[[taxon_col]])),
          y_z     = as.numeric(scale(.data[[out]]))
        )

      # 1) LMM
      fml_lmm <- stats::as.formula(
        paste0("y_z ~ taxon_z + ", time_col, " + (1|", id_col, ")")
      )
      fit_lmm <- lmerTest::lmer(fml_lmm, data = dat_tx)

      sing <- lme4::isSingular(fit_lmm, tol = tol_singular)

      # 2) FE fallback if singular
      if (isTRUE(sing)) {
        fml_fe <- stats::as.formula(
          paste0("y_z ~ taxon_z + ", time_col, " + ", id_col)
        )
        fit_fe <- stats::lm(fml_fe, data = dat_tx)

        co <- summary(fit_fe)$coefficients
        ci <- suppressWarnings(stats::confint(fit_fe, parm = "taxon_z"))

        return(tibble::tibble(
          Outcome = out,
          FeatureID = tx,
          n = nrow(dat_tx),
          beta_SD = unname(co["taxon_z", "Estimate"]),
          conf.low = ci[1],
          conf.high = ci[2],
          p = unname(co["taxon_z", "Pr(>|t|)"]),
          singular_fit = TRUE,
          model_used = "FE_PatientID"
        ))
      }

      # 3) Keep LMM
      co <- summary(fit_lmm)$coefficients
      ci <- suppressWarnings(confint(fit_lmm, parm = "taxon_z", method = "Wald"))

      tibble::tibble(
        Outcome = out,
        FeatureID = tx,
        n = nrow(dat_tx),
        beta_SD = unname(co["taxon_z", "Estimate"]),
        conf.low = ci[1],
        conf.high = ci[2],
        p = unname(co["taxon_z", "Pr(>|t|)"]),
        singular_fit = FALSE,
        model_used = "LMM"
      )
    })
  }) %>%
    dplyr::group_by(Outcome) %>%
    dplyr::mutate(p_fdr_outcome = p.adjust(p, method = "BH")) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(p_fdr_global = p.adjust(p, method = "BH"))
}
```

### Data
```{r Data, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# One row per PatientID-Time for sputum SCFAs
meta_scfa_sp <- df_scfa_sp %>%
  dplyr::mutate(Time = fix_time(Time)) %>%
  dplyr::select(PatientID, Time, dplyr::all_of(scfa_vars)) %>%
  dplyr::group_by(PatientID, Time) %>%
  dplyr::summarise(
    dplyr::across(
      dplyr::all_of(scfa_vars),
      ~ if (all(is.na(.x))) NA_real_ else mean(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  )

# Long microbiome table
dat_long <- ra2 %>%
  dplyr::select(PatientID, Time, FeatureID, RelAbundance)

# Join SCFAs into microbiome long table
dat_long_scfa_full_sp <- dat_long %>%
  dplyr::mutate(
    Time = fix_time(Time),
    PatientID = as.character(PatientID)
  ) %>%
  dplyr::left_join(meta_scfa_sp, by = c("PatientID", "Time")) %>%
  dplyr::mutate(
    Time = factor(Time, levels = c("T0", "T12")),
    PatientID = factor(PatientID)
  )

# Check
colnames(dat_long_scfa_full_sp)
```


### Run LMER
```{r Run LMER, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
res_lmm_scfa_sp <- run_lmm_taxa_vs_outcomes(
  dat_long  = dat_long_scfa_full_sp,
  taxa_keep = taxa_keep,
  outcomes  = scfa_vars,
  taxon_col = "RelAbundance"
)

print(res_lmm_scfa_sp, n = Inf)

readr::write_csv(
  res_lmm_scfa_sp,
  here::here("results", "res_lmm_taxa_vs_SCFAs_LMM_SD_ALL_taxa_sp.csv")
)
```

### Forest plot
```{r Forest plot, echo=TRUE, results='show', message=FALSE, warning=FALSE}
# =========================================================
# Forest plot: ONLY acetate, propionate, butyrate
# taxa on y-axis, SCFAs on top
# =========================================================

FDR_COL <- "p_fdr_global"   # or "p_fdr_outcome"

scfa_labels_basic <- c(
  acetate    = "Acetate",
  propionate = "Propionate",
  butyrate   = "Butyrate"
)

scfa_basic <- names(scfa_labels_basic)

# Keep only the 3 SCFAs
df_scfa_forest_basic_sp <- res_lmm_scfa_sp %>%
  dplyr::filter(
    Outcome %in% scfa_basic,
    !is.na(beta_SD),
    !is.na(conf.low),
    !is.na(conf.high)
  ) %>%
  dplyr::mutate(
    Outcome_label = dplyr::recode(Outcome, !!!as.list(scfa_labels_basic)),
    Outcome_label = factor(
      Outcome_label,
      levels = c("Acetate", "Propionate", "Butyrate")
    ),
    sig = !is.na(.data[[FDR_COL]]) & .data[[FDR_COL]] < 0.05
  )

# Optional: order taxa by their best FDR across the 3 SCFAs
taxa_order_forest_sp <- df_scfa_forest_basic_sp %>%
  dplyr::group_by(FeatureID) %>%
  dplyr::summarise(best_fdr = min(.data[[FDR_COL]], na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(best_fdr, FeatureID) %>%
  dplyr::pull(FeatureID)

df_scfa_forest_basic_sp <- df_scfa_forest_basic_sp %>%
  dplyr::mutate(
    FeatureID = factor(FeatureID, levels = rev(taxa_order_forest_sp))
  )

p_taxa_scfa_basic_sp <- ggplot(df_scfa_forest_basic_sp, aes(x = beta_SD, y = FeatureID)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    width = 0.2,
    linewidth = 0.8,
    orientation = "y"
  ) +
  geom_point(aes(fill = sig), shape = 21, size = 3, color = "black") +
  facet_grid(. ~ Outcome_label) +
  scale_fill_manual(
    name = "BH-FDR (q < 0.05)",
    values = c(`TRUE` = "black", `FALSE` = "white"),
    breaks = c(TRUE, FALSE),
    labels = c("Significant", "Not significant")
  ) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE)) +
  labs(
    title = "Association between lung taxa and sputum SCFAs (standardized)",
    subtitle = "Only acetate, propionate and butyrate; models adjusted for Time",
    x = "Standardized effect (SD outcome per 1 SD taxon)",
    y = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "grey85", linewidth = 0.6),
    panel.grid.minor = element_line(color = "grey92", linewidth = 0.4),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 9),
    strip.text  = element_text(face = "plain"),
    legend.position = "bottom"
  )

p_taxa_scfa_basic_sp

ggsave(
  filename = here::here("results", "forest_SCFAs_basic_taxa_on_y_sp.svg"),
  plot = p_taxa_scfa_basic_sp,
  device = "svg",
  width = 10,
  height = max(6, 0.28 * length(unique(df_scfa_forest_basic_sp$FeatureID))),
  units = "in"
)
```

#---------------------------------
# PANELS HEATMAPS AND FOREST PLOTS
#---------------------------------
### Heatmaps (feces + sputum)
```{r Panel heatmaps, echo=TRUE, results='show', message=FALSE, warning=FALSE}
# Vertical
panel_heatmap_taxa_scfa_fc_sp_vert <- (p_heat_taxa_scfa_fc) /
                (p_heat_taxa_scfa_sp) +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

panel_heatmap_taxa_scfa_fc_sp_vert <- panel_heatmap_taxa_scfa_fc_sp_vert + plot_annotation(tag_levels = "A")

ggsave(here("results","panel_heatmap_taxa_scfa_fc_sp_vert.svg"), panel_heatmap_taxa_scfa_fc_sp_vert, width = 6, height = 10, bg = "white")

# Horizontal
panel_heatmap_taxa_scfa_fc_sp_horiz <- (p_heat_taxa_scfa_fc | p_heat_taxa_scfa_sp) +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

panel_heatmap_taxa_scfa_fc_sp_horiz <- panel_heatmap_taxa_scfa_fc_sp_horiz + plot_annotation(tag_levels = "A")

ggsave(here("results","panel_heatmap_taxa_scfa_fc_sp_horiz.svg"), panel_heatmap_taxa_scfa_fc_sp_horiz, width = 12, height = 6, bg = "white")
```

### Forest (feces + sputum)
```{r Panel forests, echo=TRUE, results='show', message=FALSE, warning=FALSE}
panel_forest_taxa_scfa_fc_sp <- (p_taxa_scfa_basic_fc) /
                (p_taxa_scfa_basic_sp) +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

panel_forest_taxa_scfa_fc_sp <- panel_forest_taxa_scfa_fc_sp + plot_annotation(tag_levels = "A")

ggsave(here("results","panel_forest_taxa_scfa_fc_sp.svg"), panel_forest_taxa_scfa_fc_sp, width = 10, height = 11, bg = "white")
```





